### 2022-03-09 README.Rmd
First commit: add data and creat r script

---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
# Please put your title here to include it in the file below.
Title <- "Title of your paper goes here"
```

# tcsakoreanpaleolithic

[![Binder](https://mybinder.org/badge_logo.svg)](https://mybinder.org/v2/gh/parkgayoung/tcsakoreanpaleolithic/master?urlpath=rstudio)

This repository contains the data and code for our paper:

> Authors, (YYYY). _`r Title`_. Name of journal/book <https://doi.org/xxx/xxx>

Our pre-print is online here:

> Authors, (YYYY). _`r Title`_. Name of journal/book, Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

### How to cite

Please cite this compendium as:

> Authors, (`r format(Sys.Date(), "%Y")`). _Compendium of R code and data for `r Title`_. Accessed `r format(Sys.Date(), "%d %b %Y")`. Online at <https://doi.org/xxx/xxx>

## Contents

The **analysis** directory contains:

  - [:file\_folder: paper](/analysis/paper): R Markdown source document
    for manuscript. Includes code to reproduce the figures and tables
    generated by the analysis. It also has a rendered version,
    `paper.docx`, suitable for reading (the code is replaced by figures
    and tables in this file)
  - [:file\_folder: data](/analysis/data): Data used in the analysis.
  - [:file\_folder: figures](/analysis/figures): Plots and other
    illustrations
  - [:file\_folder:
    supplementary-materials](/analysis/supplementary-materials):
    Supplementary materials including notes and other documents
    prepared and collected during the analysis.

## How to run in your browser or download and run locally

This research compendium has been developed using the statistical programming
language R. To work with the compendium, you will need
installed on your computer the [R software](https://cloud.r-project.org/)
itself and optionally [RStudio Desktop](https://rstudio.com/products/rstudio/download/).

You can download the compendium as a zip from from this URL:
[master.zip](/archive/master.zip). After unzipping:
- open the `.Rproj` file in RStudio
- run `devtools::install()` to ensure you have the packages this analysis depends on (also listed in the
[DESCRIPTION](/DESCRIPTION) file).
- finally, open `analysis/paper/paper.Rmd` and knit to produce the `paper.docx`, or run `rmarkdown::render("analysis/paper/paper.Rmd")` in the R console

### Licenses

**Text and figures :**  [CC-BY-4.0](http://creativecommons.org/licenses/by/4.0/)

**Code :** See the [DESCRIPTION](DESCRIPTION) file

**Data :** [CC-0](http://creativecommons.org/publicdomain/zero/1.0/) attribution requested in reuse

### Contributions

We welcome contributions from everyone. Before you get started, please see our [contributor guidelines](CONTRIBUTING.md). Please note that this project is released with a [Contributor Code of Conduct](CONDUCT.md). By participating in this project you agree to abide by its terms.

### 2022-03-09 analysis/paper/000-tcsa-basic.R

@@ -1,0 +1,28 @@
+library(readxl)
+library(here)
+library(tidyr)
+library(dplyr)
+
+mydata <- read_excel(here("analysis/data/raw_data/TCSA_raw_data.xlsx"))
+
+tcsa <- mydata %>%
+  mutate(Tcsa = 0.5 *Width * Thickness) %>%
+  mutate(id = row_number())
+
+cv <- function(x, ... ) sd(x, ...) / mean(x, ...) *100
+
+mean(tcsa$Tcsa)
+sd(tcsa$Tcsa)
+cv(tcsa$Tcsa)
+
+## SP made out of blade
+tcsa_blade <- tcsa %>%
+  filter (Blank == "blade")
+
+mean(tcsa_blade$Tcsa)
+sd(tcsa_blade$Tcsa)
+cv(tcsa_blade$Tcsa)
+
+library(ggplot2)
+ggplot(tcsa, aes(Tcsa, id)) +
+  geom_point()

### 2022-03-10 analysis/paper/000-tcsa-basic.R
Basic analysis

@@ -2,27 +2,51 @@ library(readxl)
 library(here)
 library(tidyr)
 library(dplyr)
+library(ggplot2)
 
 mydata <- read_excel(here("analysis/data/raw_data/TCSA_raw_data.xlsx"))
 
 tcsa <- mydata %>%
-  mutate(Tcsa = 0.5 *Width * Thickness) %>%
+  mutate(min_tcsa = 0.5 *Width * Thickness) %>%
+  mutate(max_tcsa = 0.5 *Length * Thickness) %>%
   mutate(id = row_number())
 
 cv <- function(x, ... ) sd(x, ...) / mean(x, ...) *100
 
-mean(tcsa$Tcsa)
-sd(tcsa$Tcsa)
-cv(tcsa$Tcsa)
+mean(tcsa$min_tcsa)
+sd(tcsa$min_tcsa)
+cv(tcsa$min_tcsa)
 
-## SP made out of blade
-tcsa_blade <- tcsa %>%
-  filter (Blank == "blade")
+#### histogram for min tcsa
+all <- ggplot(tcsa, aes(min_tcsa)) +
+  geom_histogram(color="black", fill="green") +
+  theme_bw()
 
-mean(tcsa_blade$Tcsa)
-sd(tcsa_blade$Tcsa)
-cv(tcsa_blade$Tcsa)
+#Facet by raw material
+all +  facet_wrap(~Blank, ncol = 3)
 
-library(ggplot2)
-ggplot(tcsa, aes(Tcsa, id)) +
+
+#Facet by raw site
+all +  facet_wrap(~Site_name, ncol = 4) +
+      coord_cartesian(xlim = c(0,1000), ylim = c(0,20))
+
+#Facet by size class
+
+
+
+
+
+
+########figure 5#############3
+# use two columns for x-axis like Lombard's figure 5
+library(reshape2)
+two_tcsa <- melt(tcsa, id.vars = c("min_tcsa", "max_tcsa"))
+
+## combine min and max tcsas to one x-axis, and then make min again for y-axis
+two_tcsa <- tcsa %>%
+  pivot_longer(c(min_tcsa, max_tcsa), names_to = "var", values_to = "value") %>%
+  mutate(MIN_tcsa = 0.5 *Width * Thickness)
+
+ggplot(two_tcsa, aes(x = value, y = MIN_tcsa, shape = var)) +
   geom_point()
+

### 2022-03-10 analysis/paper/000-tcsa-basic.R
Create a plot similar to Lombard 2020 Figure 5

@@ -34,10 +34,7 @@ all +  facet_wrap(~Site_name, ncol = 4) +
 
 
 
-
-
-
-########figure 5#############3
+########figure 5#############
 # use two columns for x-axis like Lombard's figure 5
 library(reshape2)
 two_tcsa <- melt(tcsa, id.vars = c("min_tcsa", "max_tcsa"))
@@ -47,6 +44,8 @@ two_tcsa <- tcsa %>%
   pivot_longer(c(min_tcsa, max_tcsa), names_to = "var", values_to = "value") %>%
   mutate(MIN_tcsa = 0.5 *Width * Thickness)
 
-ggplot(two_tcsa, aes(x = value, y = MIN_tcsa, shape = var)) +
-  geom_point()
+ggplot(two_tcsa, aes(x = value, y = MIN_tcsa, color = var)) +
+  geom_point(size = 0.5) +
+  coord_cartesian(xlim = c(0,1000), ylim = c(0,1000)) +
+  theme_linedraw()
 

### 2022-03-11 analysis/paper/000-tcsa-basic.R
created a boxplot inspired by Lombard's 2021 qsr paper

@@ -17,6 +17,13 @@ mean(tcsa$min_tcsa)
 sd(tcsa$min_tcsa)
 cv(tcsa$min_tcsa)
 
+blank_tcsa <-tcsa %>%
+  filter(Blank == "blade")
+
+ggplot (blank_tcsa, aes(min_tcsa)) +
+  geom_boxplot()
+
+
 #### histogram for min tcsa
 all <- ggplot(tcsa, aes(min_tcsa)) +
   geom_histogram(color="black", fill="green") +

### 2022-03-11 analysis/paper/000-tcsa-basic.R
Explore the dataset

@@ -29,30 +29,34 @@ all <- ggplot(tcsa, aes(min_tcsa)) +
   geom_histogram(color="black", fill="green") +
   theme_bw()
 
-#Facet by raw material
-all +  facet_wrap(~Blank, ncol = 3)
+#Facet by Blank
+all +  facet_wrap(~Blank, ncol = 3) +
+  coord_cartesian(xlim = c(0,500))
 
+#Facet by raw materials
+all +  facet_wrap(~Raw_material, ncol = 3) +
+  coord_cartesian(xlim = c(0,500))
 
-#Facet by raw site
+#Facet by site
 all +  facet_wrap(~Site_name, ncol = 4) +
-      coord_cartesian(xlim = c(0,1000), ylim = c(0,20))
+      coord_cartesian(xlim = c(0,250), ylim = c(0,20))
 
 #Facet by size class
 
 
 
 ########figure 5#############
-# use two columns for x-axis like Lombard's figure 5
-library(reshape2)
-two_tcsa <- melt(tcsa, id.vars = c("min_tcsa", "max_tcsa"))
-
-## combine min and max tcsas to one x-axis, and then make min again for y-axis
-two_tcsa <- tcsa %>%
-  pivot_longer(c(min_tcsa, max_tcsa), names_to = "var", values_to = "value") %>%
-  mutate(MIN_tcsa = 0.5 *Width * Thickness)
-
-ggplot(two_tcsa, aes(x = value, y = MIN_tcsa, color = var)) +
-  geom_point(size = 0.5) +
-  coord_cartesian(xlim = c(0,1000), ylim = c(0,1000)) +
-  theme_linedraw()
+# use two columns for x-axis like Lombard's figure 5 : not applicable for our data
+# library(reshape2)
+# two_tcsa <- melt(tcsa, id.vars = c("min_tcsa", "max_tcsa"))
+#
+# ## combine min and max tcsas to one x-axis, and then make min again for y-axis
+# two_tcsa <- tcsa %>%
+#   pivot_longer(c(min_tcsa, max_tcsa), names_to = "var", values_to = "value") %>%
+#   mutate(MIN_tcsa = 0.5 *Width * Thickness)
+#
+# ggplot(two_tcsa, aes(x = value, y = MIN_tcsa, color = var)) +
+#   geom_point(size = 0.5) +
+#   coord_cartesian(xlim = c(0,1000), ylim = c(0,1000)) +
+#   theme_linedraw()
 

### 2022-03-14 analysis/paper/000-tcsa-basic.R
stash

@@ -23,6 +23,9 @@ all <- ggplot(tcsa, aes(min_tcsa)) +
   theme_bw()
 
 #Facet by raw material
+all +  facet_wrap(~Raw_material, ncol = 3)
+
+
 all +  facet_wrap(~Blank, ncol = 3)
 
 

### 2022-03-14 analysis/paper/000-tcsa-basic.R
add data points on boxplot

@@ -20,26 +20,21 @@ cv(tcsa$min_tcsa)
 blank_tcsa <-tcsa %>%
   filter(Blank == "blade")
 
-ggplot (blank_tcsa, aes(min_tcsa)) +
-  geom_boxplot()
-
+library(ggbeeswarm)
+ggplot (blank_tcsa,
+        aes(1, min_tcsa)) +
+  geom_boxplot() +
+  geom_quasirandom(alpha = 0.4,
+                   size = 2)
 
 #### histogram for min tcsa
 all <- ggplot(tcsa, aes(min_tcsa)) +
   geom_histogram(color="black", fill="green") +
   theme_bw()
 
-<<<<<<< HEAD
-#Facet by raw material
-all +  facet_wrap(~Raw_material, ncol = 3)
-
-
-all +  facet_wrap(~Blank, ncol = 3)
-=======
 #Facet by Blank
 all +  facet_wrap(~Blank, ncol = 3) +
   coord_cartesian(xlim = c(0,500))
->>>>>>> a0477033876dadc9e944cbd8c562fb31a6122a9f
 
 #Facet by raw materials
 all +  facet_wrap(~Raw_material, ncol = 3) +

### 2022-03-14 analysis/paper/paper.Rmd
Some basic exploration of the data, move code to Rmd

---
title: "Title Goes Here"
author:
  - Myfirstname Mylastname:
      email: fl@oneplace.edu
      institute: [UofO]
      correspondence: true
  - Yourfirstname Yourlastname:
      email: fl@another.edu
      institute: [UofA]
      correspondence: false
institute:
  - UofO: University of One Place
  - UofA: University of Another Place
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->

<!-- With the following code you can access and display values from the yml header above. -->

Keywords: `r rmarkdown::metadata$keywords`

Highlights: `r rmarkdown::metadata$highlights`

<!-- The following code chunk defines some general settings how code chunks should behave. -->

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/",
  dpi = 300
)
```

<!-- The actual document text starts here: -->

# Introduction

Here is a citation [@Marwick2017]

# Background

# Methods

# Results

```{r}
library(readxl)
library(here)
library(tidyverse)
library(ggbeeswarm)

mydata <- read_excel(here("analysis/data/raw_data/TCSA_raw_data.xlsx"))

tcsa <- mydata %>%
  mutate(min_tcsa = 0.5 *Width * Thickness) %>%
  mutate(max_tcsa = 0.5 *Length * Thickness) %>%
  #omit very large TCSA, could be measurement error
  filter(min_tcsa <= 250) %>%
  # group rare raw materials
  mutate(raw_materials_top = fct_lump_min(Raw_material, 10)) %>%
  mutate(site_name_top = fct_lump_min(Site_name, 5))
```

```{r}
# all artefacts overall
ggplot (tcsa,
        aes(1, min_tcsa)) +
  # show regions for tools from Table 3 of Lomabrd 2021
  # https://www-sciencedirect-com.offcampus.lib.washington.edu/science/article/pii/S0277379121002286
    annotate("rect", 
           xmin = -Inf,
           xmax = Inf,
           ymin = 4,
           ymax = 18,
           fill = "grey80") +
      annotate("text", 
               label = "Poisoned arrowheads",
           x = 1.3,
           y = (18-4) /2 + 4) +
      annotate("rect", 
           xmin = -Inf,
           xmax = Inf,
           ymin = 18,
           ymax = 53,
           fill = "grey50") +
      annotate("text", 
               label = "Arrowheads",
           x = 1.3,
           y = (53-13) /2 + 13) +
      annotate("rect", 
           xmin = -Inf,
           xmax = Inf,
           ymin = 53,
           ymax = 76,
           fill = "grey90") +
      annotate("text", 
               label = "Dart tips",
           x = 1.3,
           y = (76-53) /2 + 53 ) +
      annotate("rect", 
           xmin = -Inf,
           xmax = Inf,
           ymin = 76,
           ymax = 98,
           fill = "grey70") +
      annotate("text", 
               label = "Javelin",
           x = 1.3,
           y = (98-76) /2 + 76 ) +
     annotate("rect", 
           xmin = -Inf,
           xmax = Inf,
           ymin = 98,
           ymax = 230,
           fill = "grey95") +
      annotate("text", 
               label = "Thrusting spear",
           x = 1.3,
           y = (230-98) /2 + 98 ) +
  geom_boxplot(alpha = 0.1) +
  geom_quasirandom(alpha = 0.4,
                   size = 3) +

  theme_bw()
```

```{r}
#### histogram for min tcsa to prepare for exploring the data
all <- ggplot(tcsa,
              aes(min_tcsa)) +
  geom_histogram() +
  theme_bw()
```

## Raw materials

```{r}
# Facet by raw materials
all +  facet_wrap(~raw_materials_top,
                  scales = "free_y",
                  ncol = 1)
```

```{r}
# remove NAs
tcsa_mass_no_na <-
  tcsa %>%
 drop_na(`Weight(g)`)

# how does the size-tcsa relationship vary by raw material?
ggplot(tcsa_mass_no_na,
       aes(`Weight(g)`,
           min_tcsa,
           colour = raw_materials_top)) +
  geom_point() +
  theme_bw()
```

## Variation by site

```{r}
# Facet by site
all +  facet_wrap(~site_name_top,
                 scales = "free_y",
                  ncol = 1)
```

## Variation by artefact size

```{r}
# Facet by size class

# https://cran.r-project.org/web/packages/Ckmeans.1d.dp/vignettes/Ckmeans.1d.dp.html
library(Ckmeans.1d.dp)

# find clusters
result <- Ckmeans.1d.dp(tcsa_mass_no_na$`Weight(g)`)

tcsa_mass_no_na <-
tcsa_mass_no_na %>%
  mutate(mass_cluster = factor(result$cluster))

ggplot (tcsa_mass_no_na,
        aes(1, `Weight(g)`)) +
  geom_boxplot() +
  geom_quasirandom(aes(colour = mass_cluster),
                   alpha = 0.4,
                   size = 2) +
  theme_bw()

ggplot(tcsa_mass_no_na,
       aes(min_tcsa)) +
  geom_histogram() +
  theme_bw() +
  facet_wrap(~mass_cluster,
             ncol = 1,
             scales = "free_y")
```

# Discussion

# Conclusion

# Acknowledgements

<!-- The following line inserts a page break  -->

\newpage

# References

<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->

<div id="refs"></div>

\newpage

### Colophon

This report was generated on `r Sys.time()` using the following computational environment and dependencies:

```{r colophon, cache = FALSE}
# which R packages and versions?
if ("devtools" %in% installed.packages()) devtools::session_info()
```

The current Git commit details are:

```{r}
# what commit is this file at? 
if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
```

### 2022-03-14 analysis/paper/000-tcsa-basic.R

@@ -1,65 +1,14 @@
-library(readxl)
-library(here)
-library(tidyr)
-library(dplyr)
-library(ggplot2)
-
-mydata <- read_excel(here("analysis/data/raw_data/TCSA_raw_data.xlsx"))
-
-tcsa <- mydata %>%
-  mutate(min_tcsa = 0.5 *Width * Thickness) %>%
-  mutate(max_tcsa = 0.5 *Length * Thickness) %>%
-  mutate(id = row_number())
-
-cv <- function(x, ... ) sd(x, ...) / mean(x, ...) *100
-
-mean(tcsa$min_tcsa)
-sd(tcsa$min_tcsa)
-cv(tcsa$min_tcsa)
-
-blank_tcsa <-tcsa %>%
-  filter(Blank == "blade")
-
-library(ggbeeswarm)
-ggplot (blank_tcsa,
-        aes(1, min_tcsa)) +
-  geom_boxplot() +
-  geom_quasirandom(alpha = 0.4,
-                   size = 2)
-
-#### histogram for min tcsa
-all <- ggplot(tcsa, aes(min_tcsa)) +
-  geom_histogram(color="black", fill="green") +
-  theme_bw()
-
-#Facet by Blank
-all +  facet_wrap(~Blank, ncol = 3) +
-  coord_cartesian(xlim = c(0,500))
-
-#Facet by raw materials
-all +  facet_wrap(~Raw_material, ncol = 3) +
-  coord_cartesian(xlim = c(0,500))
-
-#Facet by site
-all +  facet_wrap(~Site_name, ncol = 4) +
-      coord_cartesian(xlim = c(0,250), ylim = c(0,20))
-
-#Facet by size class
 
 
 
-########figure 5#############
-# use two columns for x-axis like Lombard's figure 5 : not applicable for our data
-# library(reshape2)
-# two_tcsa <- melt(tcsa, id.vars = c("min_tcsa", "max_tcsa"))
-#
-# ## combine min and max tcsas to one x-axis, and then make min again for y-axis
-# two_tcsa <- tcsa %>%
-#   pivot_longer(c(min_tcsa, max_tcsa), names_to = "var", values_to = "value") %>%
-#   mutate(MIN_tcsa = 0.5 *Width * Thickness)
-#
-# ggplot(two_tcsa, aes(x = value, y = MIN_tcsa, color = var)) +
-#   geom_point(size = 0.5) +
[…]

### 2022-03-16 analysis/paper/paper.Rmd
editing rmd file (plots and text)

@@
 ---
-title: "Title Goes Here"
+title: "Why stemmed points? Understanding the usage of the Korea Late Paleolithic stemmed points using tip cross-sectional area"
 author:
-  - Myfirstname Mylastname:
-      email: fl@oneplace.edu
-      institute: [UofO]
+  - Gayoung Park: 
+      email: gayoungp@uw.edu
+      institute: [Uof W]
       correspondence: true
-  - Yourfirstname Yourlastname:
-      email: fl@another.edu
-      institute: [UofA]
+  - Marlize Lombard:
+      email: mlombard@uj.ac.za
+      institute: [UofJ]
+      correspondence: false
+  - Ben Marwick:
+      email: bmarwick@uw.edu
+      institute: [UofW]
       correspondence: false
 institute:
-  - UofO: University of One Place
-  - UofA: University of Another Place
+  - UofW: University of Washington
+  - UofJ: Palaeo-Research Institute, University of Johannesburg
+
 date: "`r format(Sys.time(), '%d %B, %Y')`"
@@
   mutate(site_name_top = fct_lump_min(Site_name, 5))
+
 ```
 
-```{r}
+```{r, fig.width= 4.5, fig.height=6}
 # all artefacts overall
@@
            ymax = 18,
-           fill = "grey80") +
+           fill = "green",
+           alpha = 0.2) +
       annotate("text", 
+               size = 3,
                label = "Poisoned arrowheads",
@@
            xmax = Inf,
-           ymin = 18,
+           ymin = 13,
            ymax = 53,
-           fill = "grey50") +
+           fill = "blue",
+           alpha = 0.2) +
       annotate("text", 
+               size = 3,
                label = "Arrowheads",
@@
            xmax = Inf,
-           ymin = 53,
+           ymin = 40,
            ymax = 76,
-           fill = "grey90") +
+           fill = "orange",
+           alpha = 0.2) +
       annotate("text", 
+               size = 3,
                label = "Dart tips",
            x = 1.3,
-           y = (76-53) /2 + 53 ) +
+           y = (76-40) /2 + 40 ) +
       annotate("rect", 
@@
            xmax = Inf,
-           ymin = 76,
+           ymin = 44,
            ymax = 98,
-           fill = "grey70") +
+           fill = "purple",
+           alpha = 0.2) +
       annotate("text", 
+               size = 3,
                label = "Javelin",
            x = 1.3,
-           y = (98-76) /2 + 76 ) +
+           y = (98-44) /2 + 44 ) +
      annotate("rect", 
@@
            xmax = Inf,
-           ymin = 98,
+           ymin = 88,
            ymax = 230,
-           fill = "grey95") +
+           fill = "yellow",
+           alpha = 0.2) +
       annotate("text", 
                label = "Thrusting spear",
+               size = 3,
            x = 1.3,
-           y = (230-98) /2 + 98 ) +
+           y = (230-88) /2 + 88 ) +
   geom_boxplot(alpha = 0.1) +
@@
                    size = 3) +
+  theme_bw() +
+  labs(title = "TCSA of Korean stemmed points", 
+       y = "TCSA")
+```
+1. The mean TCSA of the Korean stemmed point is `mean(tcsa$min_tcsa)`, and SD TCSA is `sd(tcsa$min_tcsa)`. 
+2. According to the Lombard's TCSA range, Korean stemmed points are included mostly in the categories of javelins and thrusting spear tips but overall the boxplot represents wide TCSA variation. 
+3. The individual weapon categories of TCSA range are overlapped a lot. We chose the specific colors for the annotation boxes to match with Figure 5 of Lombard's 2021 QSR paper.
 
-  theme_bw()
+```{r}
+
+### create column for "n = xx "
+tcsa_raw <- tcsa %>% 
+  group_by(raw_materials_top) %>% 
+  add_tally() %>% 
+  filter(n>1) %>% 
+  mutate(label_raw = as.factor(paste0(raw_materials_top,' (N = ', n, ")"))) %>%   
+  select(-n) %>%
+  nest(-raw_materials_top, -label_raw)
+
+tcsa_site <- tcsa %>% 
+  group_by(site_name_top) %>% 
+  add_tally() %>% 
+  filter(n>1) %>% 
+  mutate(label_site = as.factor(paste0(site_name_top,' (N = ', n, ")"))) %>%   
+  select(-n) %>%
+  nest(-site_name_top, -label_site) %>% 
+  select(-data)
+
+tcsa_label <- tcsa %>% 
+  left_join(tcsa_raw) %>% 
+  left_join(tcsa_site)
 ```
@@
 #### histogram for min tcsa to prepare for exploring the data
-all <- ggplot(tcsa,
+all <- ggplot(tcsa_label,
               aes(min_tcsa)) +
-  geom_histogram() +
-  theme_bw()
+  geom_histogram(color="black", fill="grey90") +
+  theme_bw() +
+  labs(x = "TCSA", 
+       y = "Count")
 ```
@@
 
-```{r}
+```{r, fig.height=5}
 # Facet by raw materials
-all +  facet_wrap(~raw_materials_top,
+all +  facet_wrap(~label_raw,
                   scales = "free_y",
                   ncol = 1)
+
 ```
 
+1) About half of the stemmed points made out of shale. 
+2) Acidic volcanic rocks tend to have more screwed and lower TCSA value. 
+3)
+
+## Size (Weight)
 ```{r}
+# how does the size-tcsa relationship vary by raw material?
+
 # remove NAs
@@
 
-# how does the size-tcsa relationship vary by raw material?
 ggplot(tcsa_mass_no_na,
@@
 # Facet by site
-all +  facet_wrap(~site_name_top,
+all +  facet_wrap(~label_site,
                  scales = "free_y",

### 2022-03-21 analysis/paper/paper.Rmd

@@
 
-Here is a citation [@Marwick2017]
-
 # Background
@@
                label = "Poisoned arrowheads",
-           x = 1.3,
+           x = 1.27,
            y = (18-4) /2 + 4) +
@@
   geom_quasirandom(alpha = 0.4,
-                   size = 3) +
+                   size = 2.3) +
   theme_bw() +
-  labs(title = "TCSA of Korean stemmed points", 
-       y = "TCSA")
+  labs(y = "TCSA")
 ```
-1. The mean TCSA of the Korean stemmed point is `mean(tcsa$min_tcsa)`, and SD TCSA is `sd(tcsa$min_tcsa)`. 
-2. According to the Lombard's TCSA range, Korean stemmed points are included mostly in the categories of javelins and thrusting spear tips but overall the boxplot represents wide TCSA variation. 
-3. The individual weapon categories of TCSA range are overlapped a lot. We chose the specific colors for the annotation boxes to match with Figure 5 of Lombard's 2021 QSR paper.
+1) The mean TCSA of the Korean stemmed point is `r paste0(round(mean(tcsa$min_tcsa), digits=1))`, and SD TCSA is `r paste0(round(sd(tcsa$min_tcsa), digits=1))`.  
+2) According to the Lombard's TCSA range, Korean stemmed points are included mostly in the categories of javelins and thrusting spear tips but overall the boxplot represents wide TCSA variation.  
+3) The individual weapon categories of TCSA range are overlapped a lot. We chose the specific colors for the annotation boxes to match with Figure 5 of Lombard's 2021 QSR paper.
 
@@
 2) Acidic volcanic rocks tend to have more screwed and lower TCSA value. 
-3)
-
-## Size (Weight)
-```{r}
-# how does the size-tcsa relationship vary by raw material?
-
-# remove NAs
-tcsa_mass_no_na <-
-  tcsa %>%
- drop_na(`Weight(g)`)
-
-ggplot(tcsa_mass_no_na,
-       aes(`Weight(g)`,
-           min_tcsa,
-           colour = raw_materials_top)) +
-  geom_point() +
-  theme_bw()
-```
+3) But overall, there is no clear pattern.
 
@@
 
-```{r}
+```{r, fig.height=5}
 # Facet by site
@@
 
+1) We computed TCSA for sites that include >= 5 stemmed points.
+2) Suyanggage has the most stemmed points and the range of TCSA is wide. We can assume that people used stemmed points with different purposes. 
+3) Nosanri and Sibuk have skewed and smaller TCSA values.
+4) The bigger ones are mainly in the category of Other.
+
 ## Variation by artefact size
+We wanted to check the relationship between TCSA and size. Weight can represent overall tool size.
 
-```{r}
+```{r, fig.width= 4.5, fig.height=4.6}
+# how does the size-tcsa relationship vary by raw material?
+
+# remove NAs
+tcsa_mass_no_na <-
+  tcsa_label %>%
+ drop_na(`Weight(g)`)
+
+ggplot(tcsa_mass_no_na,
+       aes(`Weight(g)`,
+           min_tcsa,
+           colour = label_raw)) +
+  geom_point() +
+    labs(x = "Weight", 
+       y = "TCSA",
+       colour='Raw material type') +
+  geom_smooth(method=lm, se=FALSE, size = 0.4)+
+  theme_bw() +
+  theme(
+  legend.title = element_text(size = 8),
+  legend.text = element_text(size = 7)
+  )
+```
+
+1) TCSA and Weight have positive relationship. 
+2) Shale stemmed points tend to be heavier than the other raw materials, while acidic volcanic rocks are lighter.
+
+```{r, fig.width= 4}
 # Facet by size class
@@
                    alpha = 0.4,
-                   size = 2) +
+                   size = 2.5) +
+  labs(colour='Cluster') +
   theme_bw()
+```
 
+1) Determined by Bayesian information criterion, our weight data is divided into three cluster.
+2) Most artifacts are under 20.
+
+```{r, fig.height=4}
 ggplot(tcsa_mass_no_na,
        aes(min_tcsa)) +
-  geom_histogram() +
+  geom_histogram(color="black", fill="grey90") +
   theme_bw() +
@@
              ncol = 1,
-             scales = "free_y")
+             scales = "free_y")+
+ labs(x = "TCSA", 
+       y = "Count")
 ```
+1) TCSA of the three cluster shows clear pattern.  
+2) Cluster 1, the smallest (lightest) artifacts, is the lower TCSA, compared Cluster 2. TCSA of Cluster 3 is the highest, except for one artifact, which is lower than 50.  
+3) We can assumed that artifacts in different clusters might be made for different purposes.
 

### 2022-03-21 analysis/paper/paper.Rmd
edits to GPs drafting and notes

--- 2022-03-14 analysis/paper/paper.Rmd
@@
 ---
-title: "Title Goes Here"
+title: "Variation in use of East Asian Late Paleolithic weapons: A study of tip cross-sectional area of stemmed points from Korea"
 author:
-  - Myfirstname Mylastname:
-      email: fl@oneplace.edu
-      institute: [UofO]
+  - Gayoung Park: 
+      email: gayoungp@uw.edu
+      institute: [UofW]
       correspondence: true
-  - Yourfirstname Yourlastname:
-      email: fl@another.edu
-      institute: [UofA]
+  - Marlize Lombard:
+      email: mlombard@uj.ac.za
+      institute: [UofJ]
+      correspondence: false
+  - Ben Marwick:
+      email: bmarwick@uw.edu
+      institute: [UofW]
       correspondence: false
 institute:
-  - UofO: University of One Place
-  - UofA: University of Another Place
+  - UofW: Department of Anthropology, University of Washington, Seattle. USA
+  - UofJ: Palaeo-Research Institute, University of Johannesburg, South Africa
+
 date: "`r format(Sys.time(), '%d %B, %Y')`"
@@
       fig_caption: yes
+      number_sections: false
+      global_numbering: true
       reference_docx: "../templates/template.docx" # Insert path for the DOCX file
@@
 
+<!-- hide these heading for now:
+
 # Introduction
-
-Here is a citation [@Marwick2017]
 
@@
 # Results
+
+-->
 
@@
   mutate(site_name_top = fct_lump_min(Site_name, 5))
+
 ```
 
-```{r}
+```{r, all-stemmed-points, fig.width= 4.5, fig.height=4, fig.cap="Distribution of TCSA values for all Korean stemmed points in the current dataset.
+Shaded boxes show TSCA ranges for various weapon types from Lombard (2021)."}
 # all artefacts overall
+
+point_function_group_label_y <- 0.75
+point_function_group_label_size <- 2
+
 ggplot (tcsa,
-        aes(1, min_tcsa)) +
+        aes(0.3, 
+            min_tcsa)) +
   # show regions for tools from Table 3 of Lomabrd 2021
@@
     annotate("rect", 
-           xmin = -Inf,
-           xmax = Inf,
+           xmin = 0.1,
+           xmax = 0.2,
            ymin = 4,
            ymax = 18,
-           fill = "grey80") +
+           fill = "green",
+           alpha = 0.2) +
       annotate("text", 
-               label = "Poisoned arrowheads",
-           x = 1.3,
+               size = point_function_group_label_size,
+               label = "Poisoned\narrowheads",
+               hjust = 0,
+           x = point_function_group_label_y,
            y = (18-4) /2 + 4) +
       annotate("rect", 
-           xmin = -Inf,
-           xmax = Inf,
-           ymin = 18,
+           xmin = 0.2,
+           xmax = 0.3,
+           ymin = 13,
            ymax = 53,
-           fill = "grey50") +
+           fill = "blue",
+           alpha = 0.2) +
       annotate("text", 
+               size = point_function_group_label_size,
                label = "Arrowheads",
-           x = 1.3,
+               hjust = 0,
+           x = point_function_group_label_y,
            y = (53-13) /2 + 13) +
       annotate("rect", 
-           xmin = -Inf,
-           xmax = Inf,
-           ymin = 53,
+           xmin = 0.3,
+           xmax = 0.4,
+           ymin = 40,
            ymax = 76,
-           fill = "grey90") +
+           fill = "orange",
+           alpha = 0.2) +
       annotate("text", 
+               size = point_function_group_label_size,
                label = "Dart tips",
-           x = 1.3,
-           y = (76-53) /2 + 53 ) +
+               hjust = 0,
+           x = point_function_group_label_y,
+           y = (76-40) /2 + 40 ) +
       annotate("rect", 
-           xmin = -Inf,
-           xmax = Inf,
-           ymin = 76,
+           xmin = 0.4,
+           xmax = 0.5,
+           ymin = 44,
            ymax = 98,
-           fill = "grey70") +
+           fill = "purple",
+           alpha = 0.2) +
       annotate("text", 
+               size = point_function_group_label_size,
                label = "Javelin",
-           x = 1.3,
-           y = (98-76) /2 + 76 ) +
+               hjust = 0,
+           x = point_function_group_label_y,
+           y = (98-44) /2 + 44 ) +
      annotate("rect", 
-           xmin = -Inf,
-           xmax = Inf,
-           ymin = 98,
+           xmin = 0.5,
+           xmax = 0.6,
+           ymin = 88,
            ymax = 230,
-           fill = "grey95") +
+           fill = "yellow",
+           alpha = 0.2) +
       annotate("text", 
-               label = "Thrusting spear",
-           x = 1.3,
-           y = (230-98) /2 + 98 ) +
+               label = "Thrusting\nspear",
+               size = point_function_group_label_size,
+               hjust = 0,
+           x = point_function_group_label_y,
+           y = (230-88) /2 + 88 ) +
   geom_boxplot(alpha = 0.1) +
   geom_quasirandom(alpha = 0.4,
-                   size = 3) +
-
-  theme_bw()
+                   size = 2.3) +
+  coord_cartesian(xlim = c(-0.1,1)) +
+  theme_bw(base_size = 6) +
+  theme(axis.ticks.x = element_blank(),
+        axis.text.x = element_blank(),
+        axis.title.x = element_blank(),
+        panel.grid.major = element_blank(), 
+        panel.grid.minor = element_blank()
+        ) +
+  labs(y = "TCSA")
 ```
 
+Notes on Figure \@ref(fig:all-stemmed-points):
+
+1) The mean TCSA of the Korean stemmed point is `r round(mean(tcsa$min_tcsa), digits=1)`, and standard deviation of TCSA is `r round(sd(tcsa$min_tcsa), digits=1)`.
+Overall we see wide variation of TCSA values in this boxplot.
+2) According to Lombard's (2021) TCSA ranges, Korean stemmed points are mostly in the categories of javelins and thrusting spear tips, with smaller numbers as dart tips and arrowheads.
+3) Poison arrowheads seem to be absent from these Korean assemblages
+
 ```{r}
-#### histogram for min tcsa to prepare for exploring the data
-all <- ggplot(tcsa,
-              aes(min_tcsa)) +
-  geom_histogram() +
-  theme_bw()
+
+### create column for "n = xx "
+tcsa_raw <- tcsa %>% 
+  group_by(raw_materials_top) %>% 
+  add_tally() %>% 
+  filter(n>1) %>% 
+  mutate(label_raw = as.factor(paste0(raw_materials_top,' (n = ', n, ")"))) %>%   
+  select(min_tcsa,
+         raw_materials_top, 
+       label_raw,
+       n)   %>% 
+  distinct_all()
+
+tcsa_site <- tcsa %>% 
+  group_by(site_name_top) %>% 
+  add_tally() %>% 
+  filter(n>1) %>% 
+  mutate(label_site = as.factor(paste0(site_name_top,' (n = ', n, ")"))) %>%   
+  select(min_tcsa,
+         site_name_top, 
+         label_site,
+         n) %>% 
+  distinct_all()
 ```
 
-## Raw materials
+## Variation by raw material
 
-```{r}
+```{r, raw-materials, fig.width= 4.5, fig.height=4, fig.cap="Histograms of TCSA values by lithic raw material."}
 # Facet by raw materials
-all +  facet_wrap(~raw_materials_top,
+ ggplot(tcsa_raw,
+              aes(min_tcsa, 
+                  fill = n)) +
+  geom_histogram() +
+  theme_bw(base_size = 6) +
+  labs(x = "TCSA", 
+       y = "Count of artefacts") +
+  facet_wrap(~label_raw,
                   scales = "free_y",
-                  ncol = 1)
+                  ncol = 1)  +
+  scale_fill_viridis_c(name = "Number\nof artefacts")
+
 ```
 
-```{r}
+Notes on Figure \@ref(fig:raw-materials):
+
+1) About half of all the stemmed points made out of shale. 
+2) Acidic volcanic rocks tend to have skewed lower TCSA values. 
+3) But overall, there is no clear pattern of different TCSA values by raw material.
+
+## Variation by site
+
+```{r, sites, fig.width= 4.5, fig.height=4, fig.cap="Histograms of TCSA values by archaeological site."}
+# Facet by site
+ ggplot(tcsa_site,
+              aes(min_tcsa, 
+                  fill = n)) +
+  geom_histogram() +
+  theme_bw(base_size = 6) +
+  labs(x = "TCSA", 
+       y = "Count of artefacts") +
+  facet_wrap(~label_site,
+                 scales = "free_y",
+                  ncol = 1) +
+    scale_fill_viridis_c(name = "Number\nof artefacts")
+```
+
+Notes on Figure \@ref(fig:sites):
+
+1) We computed TCSA for sites that include >= 5 stemmed points.
+2) Suyanggage has the most stemmed points and the range of TCSA is wide. We can assume that people used stemmed points for a variety of different purposes. 
+3) Nosanri and Sibuk have similarly small TCSA values, perhaps these sites had similar functions?
+Wolpyeng and Yongsujaeul have similar amounts of stemmed points to Nosanri and Sibuk, but different distribution of TCSA values, so perhaps Wolpyeng and Yongsujaeul had a different role in the foragers' land use system compared to Nosanri and Sibuk.
+
+## Variation by artefact size
+
+We wanted to check the relationship between TCSA and size. We use weight to represent overall tool size.
+
+```{r, size-tsca, fig.width= 4.5, fig.height=4, fig.cap="Artefact size and TSCA values by lithic raw material"}
+# how does the size-tcsa relationship vary by raw material?
+
 # remove NAs
@@
   geom_point() +
-  theme_bw()
+    labs(x = "Weight", 
+       y = "TCSA",
+       colour='Raw material') +
+  geom_smooth(method=lm, se=FALSE, size = 0.4)+
+  theme_bw(base_size = 6) +
+  theme(
+  legend.title = element_text(size = 8),
+  legend.text = element_text(size = 7),
+  legend.position = c(.8,.3)
+  )
 ```
 
-## Variation by site
+Notes on Figure \@ref(fig:size-tsca):
 
-```{r}
-# Facet by site
-all +  facet_wrap(~site_name_top,
-                 scales = "free_y",
-                  ncol = 1)
-```
+1) TCSA and artefact weight have a positive relationship.
+2) The relationship between TCSA and artefact weight varies by raw material, perhaps due to variation in the density of different lithic raw materials
+3) Shale stemmed points tend to have TCSA values that are strongly correlated with artefact size, while acidic volcanic points have TCSA values that vary over a very limited size range, suggesting size is less influential.
 
-## Variation by artefact size
-
-```{r}
+```{r, size-class, fig.width= 4.5, fig.height=4, fig.cap=""}
 # Facet by size class
@@
 
+plot_mass_cluster <- 
 ggplot (tcsa_mass_no_na,
-        aes(1, `Weight(g)`)) +
-  geom_boxplot() +
-  geom_quasirandom(aes(colour = mass_cluster),
+        aes(1, 
+            `Weight(g)`)) +
+  geom_boxplot(outlier.shape = NA) +
+  geom_quasirandom(aes(shape = mass_cluster),
                    alpha = 0.4,
-                   size = 2) +
-  theme_bw()
+                   size = 2.5) +
+  labs(shape='Cluster',
+       y = "Weight (g)") +
+  theme_bw(base_size = 6) +
+  theme(axis.ticks.x = element_blank(),
+        axis.text.x = element_blank(),
+        axis.title.x = element_blank(),
+        panel.grid.major = element_blank(), 
+        panel.grid.minor = element_blank()
+        )
+```
 
-ggplot(tcsa_mass_no_na,
-       aes(min_tcsa)) +
-  geom_histogram() +
-  theme_bw() +
-  facet_wrap(~mass_cluster,
-             ncol = 1,
-             scales = "free_y")
+```{r, artefact-size-cluster,fig.width= 4.5, fig.height=4, fig.cap="A.
+Distribution of artefact weight showing three clusters.
+B.
+TCSA values for all artefacts with size classes indicated by data point shape."}
+plot_tcsa_mass_cluster <- 
+ggplot (tcsa_mass_no_na,
+        aes(0.3, min_tcsa)) +
+  geom_boxplot(outlier.shape = NA) +
+  geom_quasirandom(aes(shape = mass_cluster),
+                   alpha = 0.4,
+                   size = 2.5) +
+  labs(shape='Cluster') +
+  theme_bw(base_size = 6) +
+  # show regions for tools from Table 3 of Lomabrd 2021
+  # https://www-sciencedirect-com.offcampus.lib.washington.edu/science/article/pii/S0277379121002286
+    annotate("rect", 
+           xmin = 0.1,
+           xmax = 0.2,
+           ymin = 4,
+           ymax = 18,
+           fill = "green",
+           alpha = 0.2) +
+      annotate("text", 
+               size = point_function_group_label_size,
+               label = "Poisoned\narrowheads",
+               hjust = 0,
+           x = point_function_group_label_y,
+           y = (18-4) /2 + 4) +
+      annotate("rect", 
+           xmin = 0.2,
+           xmax = 0.3,
+           ymin = 13,
+           ymax = 53,
+           fill = "blue",
+           alpha = 0.2) +
+      annotate("text", 
+               size = point_function_group_label_size,
+               label = "Arrowheads",
+               hjust = 0,
+           x = point_function_group_label_y,
+           y = (53-13) /2 + 13) +
+      annotate("rect", 
+           xmin = 0.3,
+           xmax = 0.4,
+           ymin = 40,
+           ymax = 76,
+           fill = "orange",
+           alpha = 0.2) +
+      annotate("text", 
+               size = point_function_group_label_size,
+               label = "Dart tips",
+               hjust = 0,
+           x = point_function_group_label_y,
+           y = (76-40) /2 + 40 ) +
+      annotate("rect", 
+           xmin = 0.4,
+           xmax = 0.5,
+           ymin = 44,
+           ymax = 98,
+           fill = "purple",
+           alpha = 0.2) +
+      annotate("text", 
+               size = point_function_group_label_size,
+               label = "Javelin",
+               hjust = 0,
+           x = point_function_group_label_y,
+           y = (98-44) /2 + 44 ) +
+     annotate("rect", 
+           xmin = 0.5,
+           xmax = 0.6,
+           ymin = 88,
+           ymax = 230,
+           fill = "yellow",
+           alpha = 0.2) +
+      annotate("text", 
+               label = "Thrusting\nspear",
+               size = point_function_group_label_size,
+               hjust = 0,
+           x = point_function_group_label_y,
+           y = (230-88) /2 + 88 ) +
+  coord_cartesian(xlim = c(-0.1,1.2)) +
+  theme_bw(base_size = 6) +
+  theme(axis.ticks.x = element_blank(),
+        axis.text.x = element_blank(),
+        axis.title.x = element_blank(),
+        panel.grid.major = element_blank(), 
+        panel.grid.minor = element_blank()
+        ) +
+  labs(y = "TCSA")
+
+library(cowplot)
+plot_grid(plot_mass_cluster + guides(shape = "none"),
+          plot_tcsa_mass_cluster,
+          labels = "AUTO",
+          rel_widths = c(0.5, 1))
 ```
+
+Notes on Figure \@ref(fig:artefact-size-cluster):
+
+1) Univariate cluster analysis of artefacts by weight reveals three groups or clusters 
+2) Cluster 1, the smallest (lightest) artifacts, is the lower TCSA, compared Cluster 2. TCSA of Cluster 3 is the highest, except for one artifact, which is lower than 50.  
+3) We can assumed that artifacts in different clusters might be made for different purposes.
+
+<!-- hide these for now
 
@@
 # Acknowledgements
+
+-->
 

### 2022-03-22 analysis/paper/paper.Rmd
updated the credit authorship

@@
 
+# CRediT authorship contribution statement
+Gayoung Park: Software, Validation, Formal analysis, Resources, Data curation,  Writing - original draft, Writing - Review & Editing, Visualization, Project administration.
+
+Marlize Lombard: Conceptualization, Methodology, Formal analysis, Writing - Review & Editing, Supervision
+
+Ben Marwick: Software, Validation, Formal analysis, Investigation, Writing - Review & Editing, Visualization, Supervision
+
 \newpage

### 2022-05-13 analysis/paper/paper.Rmd
Added citations and 4th author

@@
       correspondence: false
+  - Donghee Chong:
+      email: okidok2@khu.ac.kr
+      institute: [UofK]
+      correspondence: false
+
 institute:
@@
   - UofJ: Palaeo-Research Institute, University of Johannesburg, South Africa
+  - UofK: Kyung Hee University, South Korea
 
@@
 
-Marlize Lombard: Conceptualization, Methodology, Formal analysis, Writing - Review & Editing, Supervision
+Marlize Lombard: Conceptualization, Methodology, Writing - Review & Editing, Supervision
 
 Ben Marwick: Software, Validation, Formal analysis, Investigation, Writing - Review & Editing, Visualization, Supervision
+
+Donghee Chong: Data curation, Review & Editing
 

### 2022-05-26 analysis/paper/paper.Rmd
add site info and create the site map

@@
 library(ggbeeswarm)
+library(ggplot2)
 
@@
 
+```{r, map, fig.cap="Korean Paleolithic sites mentioned in this study."}
+# Facet by raw materials
+korean_archaeological_site_locations <-
+  read_excel(here::here("analysis/data/raw_data/TCSA_site_locations.xlsx"))
+
+korean_archaeological_site_locations <-
+  tibble::rowid_to_column(korean_archaeological_site_locations, "ID")
+
+library(ggmap)
+library(ggrepel)
+library(legendMap)
+
+map <-
+  get_stamenmap(bbox = c(left = 124,
+                         bottom = 33,
+                         right = 	130,
+                         top = 39),
+                zoom = 10)
+
+# map with site names next to the points
+map_of_sites <-
+  ggmap(map)  +
+  geom_point(data = korean_archaeological_site_locations,
+             aes(long_dd ,
+                 lat_dd),
+             colour = "red",
+             size = 2) +
+  # label points with site name
+  geom_text_repel(data = korean_archaeological_site_locations,
+                  aes(long_dd ,
+                      lat_dd,
+                      label = Site_name),
+                  size = 2,
+                  bg.color = "white",
+                  bg.r = 0.15) +
+  theme_minimal(base_size = 6) +
+  labs(x = "Longitute",
+       y = "Latitude") +
+  legendMap::scale_bar(
+    # edit these numbers to select a suitable location
+    # for the scale bar where it does not cover
+    # important details on the map
+    lon = 124.5,
+    lat = 33.5,
+    legend_size = 2,
+    # distance of one section of scale bar, in km
+    distance_lon = 50,
+    # height of the scale bar, in km
+    distance_lat = 5,
+    # distance between scale bar and units, in km
+    distance_legend = 20,
+    # units of scale bar
+    dist_unit = "km",
+    # add the north arrow
+    orientation = TRUE,
+    # length of N arrow, in km
+    arrow_length = 60,
+    # distance between scale bar & base of N arrow, in km
+    arrow_distance = 50,
+    # size of letter 'N' on N arrow, in km
+    arrow_north_size = 10) +
+  theme(axis.title.x = element_blank(),
+        axis.title.y = element_blank())
+
+# inset world map, from
+# https://stackoverflow.com/a/62104396/1036500
+library(maps)
+library(magrittr)
+library(maptools)
+library(raster)
+library(ggplot2)
+
+#Defining a general CRS
+mycrs <- "+proj=longlat +datum=WGS84 +no_defs"
+
+#Using the original maps package, then converting map into SpatialPolygons object
+world <- maps::map("world", fill=TRUE,  plot = FALSE) %$%
+  maptools::map2SpatialPolygons(., IDs=names,proj4string=CRS(mycrs))
+
+#The resulting map has self intersection problems so any further operation reports errors; using buffers of width 0 is a fast fix
+while(rgeos::gIsValid(world)==FALSE){
+  world <- rgeos::gBuffer(world, byid = TRUE, width = 0, quadsegs = 5, capStyle = "ROUND")
+}
+
+#Dissolving polygon's limits
+world <- raster::aggregate(world)
+
+worldplot <-
+  ggplot() +
+  geom_polygon(data = world,
+               aes(x=long, y=lat,
+                   group=group),
+               fill='NA',
+               color='black',
+               size=0.1) +
+  annotate("rect",
+           xmin = 118,
+           xmax = 137,
+           ymin = 28,
+           ymax = 47,
+           fill = NA,
+           colour = "red",
+           size = 0.5
+  ) +
+  theme_void() +
+  coord_fixed(1) +
+  labs(x = "", y = "") +
+  labs(x = NULL, y = NULL, title = NULL) +
+  scale_x_continuous(expand = c(0, 0)) +
+  scale_y_continuous(expand = c(0, 0)) +
+  theme(panel.background = element_rect(fill = "white"),
+        panel.border = element_blank(),
+        axis.title.x=element_blank(),
+        axis.text.x=element_blank(),
+        axis.ticks.x=element_blank(),
+        axis.title.y=element_blank(),
+        axis.text.y=element_blank(),
+        axis.ticks.y=element_blank())
+
+# combine site map and inset map
+library(cowplot)
+ggdraw() +
+  draw_plot(map_of_sites) +
+  draw_plot(plot = worldplot,
+            x = 0.6, # x location of inset placement
+            y = -0.09, # y location of inset placement
+            width = .4, # Inset width
+            height = .35, # Inset height
+            scale = 0.5 # Inset scale
+  ) +
+  theme(panel.background = element_rect(fill='white', colour="white"),
+        plot.background = element_rect(fill='white', colour="white"))
+
+```
+
 ```{r, all-stemmed-points, fig.width= 4.5, fig.height=4, fig.cap="Distribution of TCSA values for all Korean stemmed points in the current dataset.
@@
 
-ggplot (tcsa,
+ggplot(tcsa,
         aes(0.3, 
@@
   mutate(label_raw = as.factor(paste0(raw_materials_top,' (n = ', n, ")"))) %>%   
-  select(min_tcsa,
+  dplyr::select(min_tcsa,
          raw_materials_top, 

### 2022-05-27 analysis/paper/paper.Rmd
add more citations

@@
 library(tidyverse)
+library(dplyr)
 library(ggbeeswarm)
@@
   mutate(min_tcsa = 0.5 *Width * Thickness) %>%
-  mutate(max_tcsa = 0.5 *Length * Thickness) %>%
+ # mutate(max_tcsa = 0.5 *Length * Thickness) %>%
   #omit very large TCSA, could be measurement error
-  filter(min_tcsa <= 250) %>%
+  filter(min_tcsa <= 257) %>%
   # group rare raw materials
@@
   mutate(label_raw = as.factor(paste0(raw_materials_top,' (n = ', n, ")"))) %>%   
-  dplyr::select(min_tcsa,
+  select(min_tcsa,
          raw_materials_top, 

### 2022-06-09 analysis/paper/paper.Rmd
edit the site location
Added raw data (Dates) and created a new plot for temporal pattern

@@
   draw_plot(plot = worldplot,
-            x = 0.6, # x location of inset placement
-            y = -0.09, # y location of inset placement
+            x = 0.43, # x location of inset placement
+            y = -0.03, # y location of inset placement
             width = .4, # Inset width
@@
   mutate(label_site = as.factor(paste0(site_name_top,' (n = ', n, ")"))) %>%   
-  select(min_tcsa,
+  dplyr::select(min_tcsa,
          site_name_top, 
@@
                   ncol = 1)  +
-  scale_fill_viridis_c(name = "Number\nof artefacts")
+  scale_fill_viridis_c(name = "Number\nof artefacts") +
+  annotate("rect", 
+           xmin = 4,
+           xmax = 18,
+           ymin = 0,
+           ymax = 2.5,
+           fill = "green",
+           alpha = 0.2) +
+      annotate("text", 
+               size = point_function_group_label_size,
+               label = "Poisoned\narrowheads",
+               hjust = 0,
+           y = (0+2.5)/2,
+           x = 5.5) +
+      annotate("rect", 
+           xmin = 13,
+           xmax = 55,
+           ymin = 2.5,
+           ymax = 5,
+           fill = "blue",
+           alpha = 0.2) +
+      annotate("text", 
+               size = point_function_group_label_size,
+               label = "Arrowheads",
+               hjust = 0,
+           y = (2.5+5)/2,
+           x = 28) +
+      annotate("rect", 
+           xmin = 40,
+           xmax = 76,
+           ymin = 5,
+           ymax = 7.5,
+           fill = "orange",
+           alpha = 0.2) +
+      annotate("text", 
+               size = point_function_group_label_size,
+               label = "Dart tips",
+               hjust = 0,
+           y = (5+7.5)/2,
+           x = 53) +
+      annotate("rect", 
+           xmin = 44,
+           xmax = 98,
+           ymin = 7.5,
+           ymax = 10,
+           fill = "purple",
+           alpha = 0.2) +
+      annotate("text", 
+               size = point_function_group_label_size,
+               label = "Javelin",
+               hjust = 0,
+           y = (7.5+10)/2,
+           x = 70) +
+     annotate("rect", 
+           xmin = 88,
+           xmax = 230,
+           ymin = 10,
+           ymax = 12.5,
+           fill = "yellow",
+           alpha = 0.2) +
+      annotate("text", 
+               label = "Thrusting spear",
+               size = point_function_group_label_size,
+               hjust = 0,
+           y = (10+12.5)/2,
+           x = 160)
 
@@
 
+## Adding Radiocarbon dates info to the SP variation
+```{r}
+# radiocarbon data per assemblage
+sitedate <- read_excel(here("analysis/data/raw_data/TCSA_site_info.xlsx"))
+
+mydata_ages <-
+  sitedate %>%
+  separate(Period, into = c('age', 'error'),
+           sep = "±") %>%
+  mutate(age_ka = parse_number(age) / 1000,
+         error = parse_number(error))
+
+# combine the assemblage with age data
+
+tcsa_site_ass <- tcsa %>% 
+  left_join(mydata_ages) %>% 
+  group_by(Assemblage) %>% 
+ drop_na(age_ka) %>% 
+  add_tally() %>% 
+  mutate(label_site = as.factor(paste0(Assemblage,' (n = ', n, '), (', {round(age_ka,1)}, 'ka)' ))) %>%   
+  dplyr::select(age_ka,
+         min_tcsa,
+         Assemblage, 
+         label_site,
+         n) %>% 
+    distinct_all()
+
+ ggplot(tcsa_site_ass,
+              # aes(reorder(min_tcsa, -age_ka), 
+               aes(min_tcsa,
+                  fill = n)) +
+  geom_histogram() +
+  theme_bw(base_size = 6) +
+  labs(x = "TCSA", 
+       y = "Count of artefacts") +
+  facet_wrap(~label_site,
+                 scales = "free_y",
+                  ncol = 1) +
+    scale_fill_viridis_c(name = "Number\nof artefacts")
+
+```
+
 <!-- hide these for now

### 2022-06-09 analysis/paper/paper.Rmd
added environmental data
added some codes for climate model

@@
 
+## Paleo climate data
+```{r}
+library(ncdf4)
+library(lattice)
+
+# you'll need to get this file from
+# https://figshare.com/articles/LateQuaternary_Environment_nc/12293345/3
+# and move to our raw_data folder
+file <- here::here("analysis/data/raw_data/LateQuaternary_Environment.nc")
+
+env_nc      <- ncdf4::nc_open(file)
+longitude   <- ncdf4::ncvar_get(env_nc, "longitude")
+latitude    <- ncdf4::ncvar_get(env_nc, "latitude")
+years       <- ncdf4::ncvar_get(env_nc, "time")
+months      <- ncdf4::ncvar_get(env_nc, "month")
+temperature <- ncdf4::ncvar_get(env_nc, "temperature")
+biome       <- ncdf4::ncvar_get(env_nc, "biome")
+ncdf4::nc_close(env_nc)
+
+# this takes a few minutes
+mean_annual_temperature <- apply(temperature, c(1, 2, 4), mean)
+
+site_locations_tbl_temps <-
+      korean_archaeological_site_locations %>% 
+      dplyr::select(Site_name,
+         lat_dd,
+         long_dd) %>%
+  rowwise()  %>%
+  mutate(lonID = which.min(abs(longitude - long_dd)),
+         latID = which.min(abs(latitude - lat_dd))) %>%
+  mutate(mean_annual_temperature = list(tibble(year = years,
+                                               mean_annual_temperature = mean_annual_temperature[lonID, latID, ]))) %>%
+  unnest(mean_annual_temperature) %>%
+  filter(between(year, -50000, -10000))
+```
+
 <!-- hide these for now

### 2022-06-09 analysis/paper/paper.Rmd
added scatter plot for site&dates

@@
 
- ggplot(tcsa_site_ass,
-              # aes(reorder(min_tcsa, -age_ka), 
-               aes(min_tcsa,
-                  fill = n)) +
-  geom_histogram() +
-  theme_bw(base_size = 6) +
+ # ggplot(tcsa_site_ass,
+ #               aes(min_tcsa,
+ #                  fill = n)) +
+ #  geom_histogram() +
+ #  theme_bw(base_size = 6) +
+ #  labs(x = "TCSA", 
+ #       y = "Count of artefacts") +
+ #  facet_wrap(~label_site,
+ #                 scales = "free_y",
+ #                  ncol = 1) +
+ #    scale_fill_viridis_c(name = "Number\nof artefacts")
+
+ ggplot(tcsa_site_ass, aes(x = min_tcsa, y = reorder(label_site, age_ka), color = label_site)) +
+   geom_point() +
+   theme(legend.position="none") +
    labs(x = "TCSA", 
-       y = "Count of artefacts") +
-  facet_wrap(~label_site,
-                 scales = "free_y",
-                  ncol = 1) +
-    scale_fill_viridis_c(name = "Number\nof artefacts")
+       y = "Assemblage")
 

### 2022-08-31 analysis/paper/paper.qmd
Created qmd file

---
title: "Variation in use of East Asian Late Paleolithic weapons: A study of tip cross-sectional area of stemmed points from Korea"
author:
  - Gayoung Park:
      institute: uw
      correspondence: "yes"
      email: gayoungp@uw.edu
  - Marlize Lombard:
      institute: uj
  - Ben Marwick:
      institute: uw
  - Donghee Chong:
      institute: uk

institute:
  - uw:
      name: Department of Anthropology, University of Washington
      address: Seattle, WA, USA
 - uj:
      name: Palaeo-Research Institute, University of Johannesburg
      address: Johannesburg, South Africa
  - uk:
      name: Department of History, Kyung Hee University
      address: Seoul, South Korea

format: 
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: false
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
editor: visual
date: now
date-format: long
title-block-published: "Last updated" 
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style

abstract: |

keywords: |
  Stemmed points; Korean Paleolithic; Tip Cross-Sectional Area
---

## Quarto

Quarto enables you to weave together content and executable code into a finished document. To learn more about Quarto see <https://quarto.org>.

## Running Code

When you click the **Render** button a document will be generated that includes both content and the output of embedded code. You can embed code like this:

```{r}
1 + 1
```

You can add options to executable code like this

```{r}
#| echo: false
2 * 2
```

The `echo: false` option disables the printing of code (only output is displayed).

### 2022-09-07 analysis/paper/paper.qmd
Add abstract and try rendering

@@
 title-block-published: "Last updated"
-filters:
-  - ../templates/scholarly-metadata.lua
-  - ../templates/author-info-blocks.lua
-  - ../templates/pagebreak.lua
+
 bibliography: references.bib
@@
 abstract: |
-
+ The introduction of blade technology, stemmed points, end scrapers, burins, denticulates, and finer grained materials led to the transition from the Early to Late Paleolithic in Korea.
+Stemmed points have been considered a representative tool that led this whole set of changes.
+In this research, we examine the function of stemmed points to understand the role that they played during the technological transition as well as throughout the Late Paleolithic.
+Our main questions are: What were stemmed points used for?
+How diverse were their functions?
+What are the temporal patterns in stemmed point functions?
+We measured tip cross-sectional areas (TCSA) to discriminate different functional classes of projectile points, for example, poisoned arrowheads or thrusting spear.
+We analyze TCSA with other variables including raw materials, weight, site and radiocarbon dates.
+Our results show that the stemmed points mostly functioned as javelins and thrusting spear tips, with smaller numbers as dart tips and arrowheads.
+TCSA values are depending on size and raw material types.
+We found different usage of stemmed points in different sites, which could indicate people used stemmed points in different ways depending on the environment.
+However, some sites show a wide range of TCSA values that represent multi-purpose usage of stemmed points.
+The temporal pattern of TCSA values is one of little change throughout the Late Paleolithic period.
+We conclude that stemmed points were mainly used as Javelin but they were multi-functional tools.
 keywords: |
@@
 ---
-
-## Quarto
-
-Quarto enables you to weave together content and executable code into a finished document. To learn more about Quarto see <https://quarto.org>.
-
-## Running Code
-
-When you click the **Render** button a document will be generated that includes both content and the output of embedded code. You can embed code like this:
-
-```{r}
-1 + 1
-```
-
-You can add options to executable code like this
-
-```{r}
-#| echo: false
-2 * 2
-```
-
-The `echo: false` option disables the printing of code (only output is displayed).

### 2022-09-29 analysis/paper/paper.qmd
Add citations

@@
 ---
+
+# Introduction {.unnumbered}
+
+With the introduction of new technologies, the Korean Paleolithic transitioned from the Early to the Late Paleolithic periods.
+The transition includes blade technology, stemmed points, end scrapers, burins, denticulates, etc [@bae2017origin; @nakazawa2018quaternary; @bae2017late; @lee2017korean; @seong2016eastern].
+Another notable change is the selective use of raw materials along with the emergence of new tools.
+Previously quartzite and vein quartz were the most commonly used for core and flake tools but finer grained materials such as silicified tuff (shale), chert, hornfels, and obsidian became more important to the lithic technology during the Late Paleolithic.
+While people still used coarse materials with existing tools, they selectively chose finer materials for newly introduced tools.
+[@Seong_2003; @seong2004quartzite].
+Stemmed points are considered the first evidence of the new technology, as well as the Late Paleolithic, since they led to the whole set of changes [@seong2016eastern; @seong2008tanged].
+Due to the fact that stemmed points originated in Korea in Northeast Asia and are closely associated with mobility, site formation, and occupation diversity [@o2018origins; @Chong_2021; @park2022did].
+Despite the importance of stemmed points, only a few studies have examined their use, whereas researchers often discuss stemmed points relating to their origin and chronology of the Korean Late Paleolithic and their relationship with Japan [@chang2013human; @Park_2013; @lee2019were; @Chong_2021].[@chang2013human; @Park_2013; @lee2019were; @Chong_2021].

### 2022-10-02 analysis/paper/paper.qmd
edited and added introduction

@@
 Stemmed points are considered the first evidence of the new technology, as well as the Late Paleolithic, since they led to the whole set of changes [@seong2016eastern; @seong2008tanged].
-Due to the fact that stemmed points originated in Korea in Northeast Asia and are closely associated with mobility, site formation, and occupation diversity [@o2018origins; @Chong_2021; @park2022did].
+This is related to the fact that stemmed points originate from Korea in Northeast Asia and have a close association with mobility, site formation, and occupation diversity [@o2018origins; @Chong_2021; @park2022did].
 Despite the importance of stemmed points, only a few studies have examined their use, whereas researchers often discuss stemmed points relating to their origin and chronology of the Korean Late Paleolithic and their relationship with Japan [@chang2013human; @Park_2013; @lee2019were; @Chong_2021].[@chang2013human; @Park_2013; @lee2019were; @Chong_2021].
+
+To understand the role stemmed points played in the technological transition, this study examines the functions of stemmed points.
+We use the tip cross-sectional area (TCSA) metric to calculate artifact function and compare the results to other archaeological and ethnographic assemblages based on the idea that different shapes correspond to different functions [@lombard2021variation].
+We then explore the relationship between and function raw materials, artifact size and discard location, and how these change over time.
+Our main questions are: What were stemmed points used for?
+How diverse were their functions?
+What are the temporal patterns in stemmed point functions?
+As part of our discussion, we examine possible connections between the roles of stemmed points and population dynamics or environmental change during the Late Paleolithic period.
+In order to understand how a certain weapon-tip type was chosen, we premise that human groups had the cognitive capacity to choose proper weapons depending on their hunting strategies and socio-environmental circumstances.

### 2022-10-03 analysis/paper/paper.qmd
add figure 1&2 and method text

@@
       institute: uk
-
 institute:
@@
     address: Seoul, South Korea
-
 format: 
@@
 title-block-published: "Last updated"
-
 bibliography: references.bib
@@
 As part of our discussion, we examine possible connections between the roles of stemmed points and population dynamics or environmental change during the Late Paleolithic period.
-In order to understand how a certain weapon-tip type was chosen, we premise that human groups had the cognitive capacity to choose proper weapons depending on their hunting strategies and socio-environmental circumstances.
+In order to understand how a certain weapon-tip type was chosen, we purposely borrow evolutionary perspective and premise that human groups had the cognitive capacity to choose proper weapons depending on their hunting strategies and socio-environmental circumstances.
+
+## Why Stemmed Points?
+
+Stemmed points (Sumbejjirugae) are projectile points made on an elongated blade-like flake or blade with two parallel facets and a single or two arises that converge to form an inverted “Y” [@pratt2020circum] (@fig-sp-hand-made).
+Slight retouch is located on the proximal end to shape an acute tip and on the distal end to make a stem, which connects to a wooden shaft.
+These types of artifacts are often called tanged points, but we prefer ‘stemmed point’ here to distinguish from Bronze Age stone projectile points that have long been called ‘tanged points’ in Korea [@park2022did].
+After the first discovery at the Suyanggae site, more than 450 stemmed points have been found in over 30 sites [@Chong_2021].
+The emergence of stemmed points is relevant to questions about the origins of projectile technology, the emergence of the Late Paleoltihic, and relationships to adjacent regions in East Asia, such as Japan.
+
+The stemmed point is the first composite tool in the Korean Paleolithic that requires two different parts, consisting of a stone point and shaft, presumably made out of wood [@seong2008tanged].
+Using blades enables mass production for this composite tool and its shape becomes more standardized [@lee2015characteristics; @park2022did].
+Therefore, -@o2018origins claims that understanding the emergence of projectile technology gives insights to greater cultural, evolutionary, and behavioral cognitive flexibility.
+
+Since the stemmed points represent the beginning of the Korean Late Paleolithic, investigating their origin(s) becomes a key for understanding the technological transition from the Early Paleoltihic, modern human dispersals, and existence of the ‘Middle’ Paleolithic in Korea [@bae2010origin; @bae2017late; @norton2009evolution; @seong2016eastern].
+There has been a northworthy debate about explaining the origin(s) of stemmed points and it can be summarized into two competing models: ‘heterogenic’ migration [@bae2010origin], in situ evolution[@seong2009emergence].
+The migration model claims that the new blade industry including stemmed points, and the earlier coarse flake tradition including large cores, polyhedrons, choppers, and even handaxes, came from different origins (e.g.
+1: North route: Siberia, Mongolia, or other regions of northeast China, and 2: South route: southern China) as the result of continuing influx of modern human migration from two routes.
+The ‘in situ’ model argues that stemmed points and other Late Paleolithic technologies, including blade industries, autonomously emerged in the south of the Korean peninsula, with no apparent external influence.
+The difference between the two models comes from how to understand a few early sites that contain stemmed points but no blades and blade coares.
+The ‘in situ’ model claims that the stemmed points appeared before blades by acknowledging those early sites including Bonggok, Songamri, Yonghodong, and Hwadaeri sites while migration model supports the traditional definition of the Late or Upper Palaeolithic in Europe and Africa that started with emergence and establishment of blade technology [@seong2016eastern].
+
+Stemmed points are important proxies to understand human behaviors during the Late Paleolithic.
+-@park2022did examine mobility and site occupation patterns through lithic assemblages and concepts of human behavioral ecology and find that forager groups using stemmed points may have been associated with occupation of marginal or extreme environments compared to the groups with no stemmed points.
+Also, they were more associated with expedient technologies, indicating residential and less mobile behaviors.
+-@Chong_2021 claims that the morphological variation of the stemmed points along with tool types in lithic assemblages, assemblage size, use of raw materials, and types of blanks could represent the characteristics of occupations such as “limited activity station” and “residential base camp.” For example, stemmed points with high morphological variations in a tool size, shape of edge, degrees of damage, and types of edge retouching from Yongsandong site indicate that the site was used for specific or limited activities such as hunting  [@kim2004yongsan; @seong2015diversity; @Chong_2021].
+
+The presence of stemmed points in both Korean and Japan has been studied since the late 1980s as a part of evidence for long-distance/maritime cultural interchanges or social networks [@lee2015characteristics].
+Stemmed points from the Bonggok site in Korea are currently accepted as the oldest ones among Northeast Asia, dated to ca.
+41.5 ka, and made on elongated flakes [@seong2009emergence; @seong2015diversity; @bae2017origin].
+After their first appearance in Korea, stemmed points (Hakuhensenntouki in Japanese) appear in Kyushu, Japan during late MIS 3.
+In addition to the stemmed points, there are similar artifacts found in both regions such as microblade cores, Moppule-seokgi (Kakusuijyosekki in Japanese), backed knife, bilateral points, bifacial points, and transport of obsidian [@lee2012characteristics; @chang2013human; @lee2015characteristics; @kim2021evidence].
+In addition, stemmed points also help understand the connections and technological transitions between islands in Japan [@chang2013human].
+
+```{r}
+
+#| label: fig-sp-hand-made
+#| fig-cap: Stemmed points from Yongsandong site. 1-6: plain stemmed points, 7-8: one side denticulate stemmed points.
+
+knitr::include_graphics(here::here("analysis/figures/hand-made-figures/TCSA_sp.png"))
+```
+
+## Previous studies about the stemmed points function
+
+Though the stemmed points are generally assumed as hunting armatures [@lee2019were; @lee2011study; @seong2008tanged; @lee2002new; @chang2013human], it is difficult to understand the function of the tools without knowing their complete shape attached to other components.
+Preserved wooden components of projectile tools are too few and rare to generalize the overall scale and variability of the tools used in prehistory [@shea2006origins].
+-@Lee_Kong_2002 claimed to name the stemmed points as ‘stemmed tools’ because of the diverse shape of stemmed points, which could represent different functions as well as the uncertainty of complete tool shape.
+Other researchers agree that some stemmed points that are not symmetrical and have retouch on one side of the artifacts or have denticulate blades on one side should not be called as stemmed points but stemmed knives, stemmed side-scrapers, stemmed end-scrapers, or stemmed burins [@seong2008tanged; @Lee_Jang_2011; @Kim] (@fig-sp-hand-made: 7-8).
+
+Most stemmed points are symmetrical from the tip to the tang, having the central axis function as a line of symmetry [@Lee_Jang_2011].
+The studies of the stemmed points manufacturing process and the patterns of broken pieces show that stemmed points might be used mainly as spear tips [@Lee_1985; @Chang_2002].
+There are a high proportion of broken tips and stems and the reused tools were repaired based on keeping the symmetry [@Park_2013; @Kim].
+In addition to the morphological aspect of stemmed points, investigating the whole site and tool composition of an assemblage show that stemmed points or stemmed tools are strongly associated with hunting activities including peeling the animal skin after slaughtering or separating the bones from the flesh [@seong2008tanged; @Chong_2021].
+
+A recent study conducts use-wear analysis on 95 stemmed points from the Jingeuneul site along with fracture patterns, TCSA, tip cross-sectional perimeter (TCSP), and neck width to understand the use of the tools [@lee2019were].
+Their results show that a considerable number of the stemmed points show diagnostic impact fractures (DIFs) on the surface, which could occur due to longitudinal force from the shaft when a stone tip hits a target.
+In other words, their research indicates that stemmed points could be used as spear-throwers or bows [@sano2015backed].
+
+## Tip Cross-Sectional Area
+
+The tip cross-sectional area (TCSA) has been used as a ballistically significant standard to discriminate different functional classes of projectile armatures such as spearthrower (a.k.a.
+atlatl) dart tips, arrowheads and large experimental stabbing/thrusting spears [@sisk2011african; @lombard2021variation; @lombard2020tip; @lombard2021did; @o2018origins].
+TCSA metric indicates the part of the tool that cuts the target’s hide and relates to weapon flight and penetration dynamics [@hughes1998getting; @sitton2020tip; @lombard2021variation].
+This method was first proposed by -@hughes1998getting and and tested by -@shea2006origins through comparing archeological examples to ethnographically collected samples of known use.
+-@shea2009impact also applied this approach to compare projectile points from Africa, the Levant, and Europe and claimed that projectile weapons first appeared in Africa.
+The main advantage of the TCSA is, regardless of point type, only maximum width and thickness measurements are needed for calculating the TCSA value (0.5 × maximum width × maximum thickness) [@sisk2011african; @lombard2020tip].
+Later, -@sisk2011african proposed an alternate measure, tip cross-sectional perimeter (TCSP), for more accurate measure for the force needed to penetrate a target to a lethal depth while TCSA metric is more associated with cutting aspect.
+However, TCSP has a few disadvantages to directly apply to our case of stemmed points.
+The force and penetration depth are not only affected by the stone tip but also by the mass of the shaft, which cannot be known for archaeological stone-tipped weapons [@lombard2020tip].
+-@sisk2011 also mentioned that TCSP cannot be applied to backed pieces that are hafted as projectile armatures.
+
+-@lee2019were first applied TCSA to stemmed points from Korea along with their use-wear analysis.
+The stemmed points for their research were discovered from Jingeuneul site located in southwest of Korea, where the largest number of stemmed points (n = 99) were found in a single site.
+For TCSA, they were only able to use ten stemmed points because they needed to select the stemmed points that retained the widest and thickest part of the specimens for TCSA metric and displayed diagnostic impact fractures for the use-wear analysis.
+The purpose of using TCSA in their research is to compare the values to North American dart tips and arrowheads.
+Their results show that the TCSA range of Jingeuneul stemmed points is relatively wide overlapped with both North American dart tips and arrowheads.
+Inspired by -@lee2019were in addition to other prior research that developed and applied TCSA into archaeological materials from other parts of the world, we explore stemmed points from all over South Korea to better understand their function to understand the role that stemmed points represent during the Late Paleolithic.
+
+# Methods
+
+## Archaeological sites
+After the first evidence of the Korean Late Paleolithic discovered at Seokjangri site including blade and a stemmed base, around 300 stemmed points have been found in Korea [@Sohn_1967; @lee2019were].
+Most sites contain only a few stemmed points and only a few sites have many more, such as Jingeuneul, Suyanggae (n = 55), and Yongsandong (n = 38) [@Kim_2017].
+Among these stemmed points, we gathered those that retained the widest and thickest part of specimens.
+Our dataset contains stemmed points that were found during field surveys as well as those were from sites that were never dated but accompanied with other Late Paleoltihic artifacts.
+Those criteria resulted in a sample of 173 stemmed points from 36 assemblages unearthed from 29 sites spanning the period 44-10 ka (Figure 2).
+The dimensions of the 173 stemmed points were obtained from published excavation reports and by direct measurements during our visits to the collections of Korean local museums and archaeological institutions.
+
+We distinguished multiple assemblages in one site where numerous excavations have been conducted in different spots and by different institutions under the same site name.
+For example, Suyanggae site, one of the Korean National Heritage sites, has been excavated more than ten times since 1980 by the local university museum and later by archaeology research institutions.
+There are six different excavation spots apart from each other from a few meters to a few kilometers.
+The archaeologists identified multiple assemblages in each excavation or even a trench by culturally sterile deposits, or where distinct artifact-bearing stratigraphic units could be identified by major differences in the texture, color, and composition of the sedimentary deposits.
+In this research, we used four assemblages from the Suyanggae site and we separated one assemblage among the four by using a different site name, Hajinri.
+Because Hajinri was from the sixth excavation area, which is 3.5 kilometers apart from the other spots and dated to much earlier (around 42-30 ka) than the other assemblages (around 31-15 ka).
+According to the excavation reports for Hajinri, stemmed points first appeared in Hajinri around 42 ka with the earliest blade and blade cores [@Lee_Woo_Lee_An_Yun_Park_Otani_Kim_Kim_Han_et al._2018].
+We consider this assemblage as an unusual outlier because the stemmed points and blades from Hajinri are highly standardized and refined, which are only found at other sites much later in time.
+We consider Bonggok site, which has the second-earliest dates (41.5 ka) in our collection, is the actual earliest site.
+Bonggok site includes blades or elongate flakes but it is hard to define without accompanying blade core [@park2022did].
+
+```{r}
+#| label: import-data-and-compute-tcsa
+
+library(readxl)
+library(here)
+library(tidyverse)
+library(dplyr)
+library(ggbeeswarm)
+library(ggplot2)
+
+# import raw data
+mydata <- read_excel(here("analysis/data/raw_data/TCSA_raw_data.xlsx"))
+
+# compute TCSA
+tcsa <- mydata %>%
+  mutate(min_tcsa = 0.5 *Width * Thickness) %>%
+ # mutate(max_tcsa = 0.5 *Length * Thickness) %>%
+  #omit very large TCSA, could be measurement error
+  dplyr::filter(min_tcsa <= 257) %>%
+  # group rare raw materials
+  mutate(raw_materials_top = fct_lump_min(Raw_material, 10)) %>%
+  mutate(site_name_top = fct_lump_min(Site_name, 5))
+
+```
+
+```{r}
+
+#| label: fig-map
+#| fig-cap: Korean Paleolithic sites mentioned in this study
+#| fig-width:   4.45
+#| fig-height:  4.5
+
+# Facet by raw materials
+korean_archaeological_site_locations <-
+  read_excel(here::here("analysis/data/raw_data/TCSA_site_locations.xlsx"))
+
+korean_archaeological_site_locations <-
+  tibble::rowid_to_column(korean_archaeological_site_locations, "ID")
+
+library(ggmap)
+library(ggrepel)
+# devtools::install_github('3wen/legendMap')
+library(legendMap)
+
+map <-
+  get_stamenmap(bbox = c(left = 124,
+                         bottom = 33,
+                         right = 	130,
+                         top = 39),
+                zoom = 10)
+
+# map with site names next to the points
+map_of_sites <-
+  ggmap(map)  +
+  geom_point(data = korean_archaeological_site_locations,
+             aes(long_dd ,
+                 lat_dd),
+             colour = "red",
+             size = 2) +
+  # label points with site name
+  geom_text_repel(data = korean_archaeological_site_locations,
+                  aes(long_dd ,
+                      lat_dd,
+                      label = Site_name),
+                  size = 2,
+                  bg.color = "white",
+                  bg.r = 0.15) +
+  theme_minimal(base_size = 6) +
+  labs(x = "Longitute",
+       y = "Latitude") +
+  legendMap::scale_bar(
+    # edit these numbers to select a suitable location
+    # for the scale bar where it does not cover
+    # important details on the map
+    lon = 124.5,
+    lat = 33.5,
+    legend_size = 2,
+    # distance of one section of scale bar, in km
+    distance_lon = 50,
+    # height of the scale bar, in km
+    distance_lat = 5,
+    # distance between scale bar and units, in km
+    distance_legend = 20,
+    # units of scale bar
+    dist_unit = "km",
+    # add the north arrow
+    orientation = TRUE,
+    # length of N arrow, in km
+    arrow_length = 60,
+    # distance between scale bar & base of N arrow, in km
+    arrow_distance = 50,
+    # size of letter 'N' on N arrow, in km
+    arrow_north_size = 10) +
+  theme(axis.title.x = element_blank(),
+        axis.title.y = element_blank())
+
+# inset world map, from
+# https://stackoverflow.com/a/62104396/1036500
+library(maps)
+library(magrittr)
+library(maptools)
+library(raster)
+library(ggplot2)
+
+#Defining a general CRS
+mycrs <- "+proj=longlat +datum=WGS84 +no_defs"
+
+#Using the original maps package, then converting map into SpatialPolygons object
+world <- maps::map("world", fill=TRUE,  plot = FALSE) %$%
+  maptools::map2SpatialPolygons(., IDs=names,proj4string=CRS(mycrs))
+
+#The resulting map has self intersection problems so any further operation reports errors; using buffers of width 0 is a fast fix
+while(rgeos::gIsValid(world)==FALSE){
+  world <- rgeos::gBuffer(world, byid = TRUE, width = 0, quadsegs = 5, capStyle = "ROUND")
+}
+
+#Dissolving polygon's limits
+world <- raster::aggregate(world)
+
+worldplot <-
+  ggplot() +
+  geom_polygon(data = world,
+               aes(x=long, y=lat,
+                   group=group),
+               fill='NA',
+               color='black',
+               size=0.1) +
+  annotate("rect",
+           xmin = 118,
+           xmax = 137,
+           ymin = 28,
+           ymax = 47,
+           fill = NA,
+           colour = "red",
+           size = 0.5
+  ) +
+  theme_void() +
+  coord_fixed(1) +
+  labs(x = "", y = "") +
+  labs(x = NULL, y = NULL, title = NULL) +
+  scale_x_continuous(expand = c(0, 0)) +
+  scale_y_continuous(expand = c(0, 0)) +
+  theme(panel.background = element_rect(fill = "white"),
+        panel.border = element_blank(),
+        axis.title.x=element_blank(),
+        axis.text.x=element_blank(),
+        axis.ticks.x=element_blank(),
+        axis.title.y=element_blank(),
+        axis.text.y=element_blank(),
+        axis.ticks.y=element_blank())
+
+# combine site map and inset map
+library(cowplot)
+ggdraw() +
+  draw_plot(map_of_sites) +
+  draw_plot(plot = worldplot,
+            x = 0.43, # x location of inset placement
+            y = -0.03, # y location of inset placement
+            width = .4, # Inset width
+            height = .35, # Inset height
+            scale = 0.5 # Inset scale
+  ) +
+  theme(panel.background = element_rect(fill='white', colour="white"),
+        plot.background = element_rect(fill='white', colour="white"))
+
+```
+
+# Acknowledgements
+
+<!-- The following line inserts a page break  -->
+
+# CRediT authorship contribution statement
+
+Gayoung Park: Software, Validation, Formal analysis, Resources, Data curation,  Writing - original draft, Writing - Review & Editing, Visualization, Project administration.
+
+Marlize Lombard: Conceptualization, Methodology, Writing - Review & Editing, Supervision
+
+Ben Marwick: Software, Validation, Formal analysis, Investigation, Writing - Review & Editing, Visualization, Supervision
+
+Donghee Chong: Data curation, Review & Editing
+
+\newpage
+
+# References {.unnumbered}
+
+<!-- The following line ensures the references appear here for the MS Word or HTML output files, rather than right at the end of the document (this will not work for PDF files):  -->
+
+::: {#refs}
+:::
+
+\newpage
+
+### Colophon
+
+This report was generated on `r Sys.time()` using the following computational environment and dependencies:
+
+```{r colophon, cache = FALSE}
+# which R packages and versions?
+if ("devtools" %in% installed.packages()) devtools::session_info()
+```
+
+The current Git commit details are:
+
+```{r}
+# what commit is this file at? 
+if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
+```

### 2022-10-03 analysis/paper/paper.Rmd
Move all figures to qmd

@@
 ```{r}
-
-### create column for "n = xx "
+# create column for "n = xx "
 tcsa_raw <- tcsa %>% 

### 2022-10-03 analysis/paper/paper.qmd
Move all figures to qmd

@@
 
-Stemmed points (Sumbejjirugae) are projectile points made on an elongated blade-like flake or blade with two parallel facets and a single or two arises that converge to form an inverted “Y” [@pratt2020circum] (@fig-sp-hand-made).
+Stemmed points (Sumbejjirugae) are projectile points made on an elongated blade-like flake or blade with two parallel facets and a single or two arises that converge to form an inverted "Y" [@pratt2020circum] (@fig-sp-hand-made).
 Slight retouch is located on the proximal end to shape an acute tip and on the distal end to make a stem, which connects to a wooden shaft.
-These types of artifacts are often called tanged points, but we prefer ‘stemmed point’ here to distinguish from Bronze Age stone projectile points that have long been called ‘tanged points’ in Korea [@park2022did].
+These types of artifacts are often called tanged points, but we prefer 'stemmed point' here to distinguish from Bronze Age stone projectile points that have long been called 'tanged points' in Korea [@park2022did].
 After the first discovery at the Suyanggae site, more than 450 stemmed points have been found in over 30 sites [@Chong_2021].
@@
 
-Since the stemmed points represent the beginning of the Korean Late Paleolithic, investigating their origin(s) becomes a key for understanding the technological transition from the Early Paleoltihic, modern human dispersals, and existence of the ‘Middle’ Paleolithic in Korea [@bae2010origin; @bae2017late; @norton2009evolution; @seong2016eastern].
-There has been a northworthy debate about explaining the origin(s) of stemmed points and it can be summarized into two competing models: ‘heterogenic’ migration [@bae2010origin], in situ evolution[@seong2009emergence].
+Since the stemmed points represent the beginning of the Korean Late Paleolithic, investigating their origin(s) becomes a key for understanding the technological transition from the Early Paleoltihic, modern human dispersals, and existence of the 'Middle' Paleolithic in Korea [@bae2010origin; @bae2017late; @norton2009evolution; @seong2016eastern].
+There has been a northworthy debate about explaining the origin(s) of stemmed points and it can be summarized into two competing models: 'heterogenic' migration [@bae2010origin], in situ evolution[@seong2009emergence].
 The migration model claims that the new blade industry including stemmed points, and the earlier coarse flake tradition including large cores, polyhedrons, choppers, and even handaxes, came from different origins (e.g.
 1: North route: Siberia, Mongolia, or other regions of northeast China, and 2: South route: southern China) as the result of continuing influx of modern human migration from two routes.
-The ‘in situ’ model argues that stemmed points and other Late Paleolithic technologies, including blade industries, autonomously emerged in the south of the Korean peninsula, with no apparent external influence.
+The 'in situ' model argues that stemmed points and other Late Paleolithic technologies, including blade industries, autonomously emerged in the south of the Korean peninsula, with no apparent external influence.
 The difference between the two models comes from how to understand a few early sites that contain stemmed points but no blades and blade coares.
-The ‘in situ’ model claims that the stemmed points appeared before blades by acknowledging those early sites including Bonggok, Songamri, Yonghodong, and Hwadaeri sites while migration model supports the traditional definition of the Late or Upper Palaeolithic in Europe and Africa that started with emergence and establishment of blade technology [@seong2016eastern].
+The 'in situ' model claims that the stemmed points appeared before blades by acknowledging those early sites including Bonggok, Songamri, Yonghodong, and Hwadaeri sites while migration model supports the traditional definition of the Late or Upper Palaeolithic in Europe and Africa that started with emergence and establishment of blade technology [@seong2016eastern].
 
@@
 Also, they were more associated with expedient technologies, indicating residential and less mobile behaviors.
--@Chong_2021 claims that the morphological variation of the stemmed points along with tool types in lithic assemblages, assemblage size, use of raw materials, and types of blanks could represent the characteristics of occupations such as “limited activity station” and “residential base camp.” For example, stemmed points with high morphological variations in a tool size, shape of edge, degrees of damage, and types of edge retouching from Yongsandong site indicate that the site was used for specific or limited activities such as hunting  [@kim2004yongsan; @seong2015diversity; @Chong_2021].
+-@Chong_2021 claims that the morphological variation of the stemmed points along with tool types in lithic assemblages, assemblage size, use of raw materials, and types of blanks could represent the characteristics of occupations such as "limited activity station" and "residential base camp." For example, stemmed points with high morphological variations in a tool size, shape of edge, degrees of damage, and types of edge retouching from Yongsandong site indicate that the site was used for specific or limited activities such as hunting [@kim2004yongsan; @seong2015diversity; @Chong_2021].
 
@@
 Preserved wooden components of projectile tools are too few and rare to generalize the overall scale and variability of the tools used in prehistory [@shea2006origins].
--@Lee_Kong_2002 claimed to name the stemmed points as ‘stemmed tools’ because of the diverse shape of stemmed points, which could represent different functions as well as the uncertainty of complete tool shape.
+-@Lee_Kong_2002 claimed to name the stemmed points as 'stemmed tools' because of the diverse shape of stemmed points, which could represent different functions as well as the uncertainty of complete tool shape.
 Other researchers agree that some stemmed points that are not symmetrical and have retouch on one side of the artifacts or have denticulate blades on one side should not be called as stemmed points but stemmed knives, stemmed side-scrapers, stemmed end-scrapers, or stemmed burins [@seong2008tanged; @Lee_Jang_2011; @Kim] (@fig-sp-hand-made: 7-8).
@@
 atlatl) dart tips, arrowheads and large experimental stabbing/thrusting spears [@sisk2011african; @lombard2021variation; @lombard2020tip; @lombard2021did; @o2018origins].
-TCSA metric indicates the part of the tool that cuts the target’s hide and relates to weapon flight and penetration dynamics [@hughes1998getting; @sitton2020tip; @lombard2021variation].
+TCSA metric indicates the part of the tool that cuts the target's hide and relates to weapon flight and penetration dynamics [@hughes1998getting; @sitton2020tip; @lombard2021variation].
 This method was first proposed by -@hughes1998getting and and tested by -@shea2006origins through comparing archeological examples to ethnographically collected samples of known use.
@@
 
--@lee2019were first applied TCSA to stemmed points from Korea along with their use-wear analysis.
+\-@lee2019were first applied TCSA to stemmed points from Korea along with their use-wear analysis.
 The stemmed points for their research were discovered from Jingeuneul site located in southwest of Korea, where the largest number of stemmed points (n = 99) were found in a single site.
@@
 ## Archaeological sites
+
 After the first evidence of the Korean Late Paleolithic discovered at Seokjangri site including blade and a stemmed base, around 300 stemmed points have been found in Korea [@Sohn_1967; @lee2019were].
@@
 Because Hajinri was from the sixth excavation area, which is 3.5 kilometers apart from the other spots and dated to much earlier (around 42-30 ka) than the other assemblages (around 31-15 ka).
-According to the excavation reports for Hajinri, stemmed points first appeared in Hajinri around 42 ka with the earliest blade and blade cores [@Lee_Woo_Lee_An_Yun_Park_Otani_Kim_Kim_Han_et al._2018].
+According to the excavation reports for Hajinri, stemmed points first appeared in Hajinri around 42 ka with the earliest blade and blade cores [@Lee_Woo_Lee_An_Yun_Park_Otani_Kim_Kim_Han_et_al._2018].
 We consider this assemblage as an unusual outlier because the stemmed points and blades from Hajinri are highly standardized and refined, which are only found at other sites much later in time.
@@
 
+## Applying TCSA
+
+To answer our research questions, What were stemmed points used for?
+How diverse were their functions?
+What are the temporal patterns in stemmed point functions?
+Are these temporal patterns associated with population dynamics or environmental processes?
+We calculated TCSA values for stemmed points and explored the usage for the stemmed points depending on raw materials, location, and size.
+We applied radiocarbon dates to the results to examine the temporal patterns.
+In the discussion, we compared those temporal patterns with demographic and environmental proxies.
+We used the results of summed probability distributions (SPD) and annual temperature changes from -@park2022did which covers our target period and area.
+Following the TCSA ranges for different weapon-delivery systems that -@lombard2021variation created by summarizing -@wadley2008segment' study of backed microliths (Table 1?), we excluded 12 artifacts from our dataset.
+Because the TCSA values of those 12 artifacts from Bupyeongri, Hwadaeri, Haga, Sinhwari, Gokcheon, and Sachang sites are greater than 257.
+Their TCSA range is from 263 to 2622.
+Therefore, a total of 161 stemmed points from 32 assemblages unearthed from 26 sites are used for the following analysis.
+
+| Weapon type               | N of tools | Mean TCSA | SD  | TCSA Range |
+|---------------------------|------------|-----------|-----|------------|
+| Poisoned arrowheads       | 434        | 11        | 7   | 4-18       |
+| Arroheads                 | 118        | 33        | 20  | 13-53      |
+| Dart tips                 | 40         | 58        | 18  | 40-76      |
+| All javelins of known use | 137        | 71        | 27  | 44-98      |
+| Thrusting spear tips      | 75         | 159       | 71  | 88-230     |
+
+: TCSA range from Lombard [-@lombard2021variation]
+
+### Raw materials
+
+Selective use of raw material is one of the main technological transitions to the Korea Late Paleolithic.
+Previously used quartzite and vein quartz were continuously selected for the existing core and flake tools and finer grained materials were newly added to assemblages for producing the newly introduced tools [@seong2004quartzite].
+We analyze TCSA values per each raw material and examine why types of raw materials were selected for certain types of tool.
+
+### Regional Pattern
+
+We present TCSA values per site that contains more than 4 stemmed points (more than and equal to 5?) based on the evolutionary concept that people choose their tools according to the surrounding environment and available resources.
+Through the TCSA distribution, we explore which function(s) the stemmed points had performed at the sites.
+
+### Weight
+
+Different sizes of stone artifacts can represent different functions.
+Overall size of stone points has been used as a potential proxy to separate different armature types [@thomas1978arrowheads; @sahle2019assessment].
+We chose weight as a size indicator for the stemmed points.
+The weight is a reliable discriminator between tools and a descriptive attribute which can be determined objectively and which serves as a criterion of likeness or dissimilarity in projectile point description [@fenenga1953weights; @shea2006origins].
+Excluding 9 points that we were not able to measure or get the records of weight, we calculate TCSA for 152 artifacts.
+
+### Temporal pattern
+
+We add radiocarbon dates to explore the temporal pattern of the function of stemmed points.
+We excluded the assemblages that have no radiocarbon dates.
+We arrange 24 assemblages dated from 45ka to 14.8ka periodically, and compute TCSA.
+
+## Choice of weapon-tip types
+
+The introduction of new types reflect major implications for social transformation.
+For example, the bow and arrow enabled individuals to work without hunting teams and provided opportunities for non-elite hunters to produce their own subsistence or pursue individual wealth [@bettinger2013effects; @angelbeck2014faustian; @rorabaugh2015timing].
+In evolutionary terms, we assume that human groups carefully selected a specific stone tool based on its functional advantage over other alternatives once they acquired the necessary physiological and cognitive abilities to innovate [@lombard2022tip].
+The choice of weapon-tip types, therefore, reflects the socio-environmental circumstances that they had to encounter and overcome.
+-@eren2022plains claim that Clovis points were more variable in shape than Folsom points because Clovis foragers were exposed to largely unfamiliar landscape and employed their points as multifunctional tools that perform a wider range of tasks including cutting and sawing.
+Folsom points show a narrow range of variation, indicating a specific task.
+They used TCSA to assess the morphological variance of the points as well as understand the potential functional costs or benefits of one artifact variant relative to another.
+
+Similar to -@eren2022plains's approach, we hypothesize two scenarios that explain the regional or/and temporal pattern of TCSA range.
+If stemmed points represent a narrow TCSA range, people produced tools that performed certain tasks.
+There could be less surprise in the surrounding environment.
+On the other hand, a wide TCSA range indicates that stemmed points could be multifunctional tools and people might need to be prepared for unexpected situations such as moving into an unfamiliar landscape or their patch becoming less productive.
+We expect to observe various TCSA ranges, which reflect morphologically and functionally different stemmed points, across different regions or environments.
+However, we acknowledge that lithic morphology is highly dependent on the quality of raw materials and their availability.
+
+The entire R code [@core2021team] and data files used for all the analysis and visualizations contained in this paper is openly available at https://doi-org/xxxx to enable re-use of materials and improve reproducibility and transparency [@marwick2017computational].
+All of the figures, tables, statistical test results presented here in addition to the raw data can be independently reproduced with the code and data in this repository.
+The code is released under the MIT license, the data as CC-0, and figures as CC-BY, to enable maximum re-use.
+
+# Results
+
+## TCSA Variation
+
+```{r}
+#| label: tcsa-all-sp
+#| fig-cap: Distribution of TCSA values for all Korean stemmed points in the current dataset. Shaded boxes show TSCA ranges for various weapon types.
+#| fig-width:   4.5
+#| fig-height:  4.4
+
+# all artefacts overall
+
+point_function_group_label_y <- 0.75
+point_function_group_label_size <- 2
+
+ggplot(tcsa,
+        aes(0.3, 
+            min_tcsa)) +
+  # show regions for tools from Table 3 of Lomabrd 2021
+  # https://www-sciencedirect-com.offcampus.lib.washington.edu/science/article/pii/S0277379121002286
+    annotate("rect", 
+           xmin = 0.1,
+           xmax = 0.2,
+           ymin = 4,
+           ymax = 18,
+           fill = "green",
+           alpha = 0.2) +
+      annotate("text", 
+               size = point_function_group_label_size,
+               label = "Poisoned\narrowheads",
+               hjust = 0,
+           x = point_function_group_label_y,
+           y = (18-4) /2 + 4) +
+      annotate("rect", 
+           xmin = 0.2,
+           xmax = 0.3,
+           ymin = 13,
+           ymax = 53,
+           fill = "blue",
+           alpha = 0.2) +
+      annotate("text", 
+               size = point_function_group_label_size,
+               label = "Arrowheads",
+               hjust = 0,
+           x = point_function_group_label_y,
+           y = (53-13) /2 + 13) +
+      annotate("rect", 
+           xmin = 0.3,
+           xmax = 0.4,
+           ymin = 40,
+           ymax = 76,
+           fill = "orange",
+           alpha = 0.2) +
+      annotate("text", 
+               size = point_function_group_label_size,
+               label = "Dart tips",
+               hjust = 0,
+           x = point_function_group_label_y,
+           y = (76-40) /2 + 40 ) +
+      annotate("rect", 
+           xmin = 0.4,
+           xmax = 0.5,
+           ymin = 44,
+           ymax = 98,
+           fill = "purple",
+           alpha = 0.2) +
+      annotate("text", 
+               size = point_function_group_label_size,
+               label = "Javelin",
+               hjust = 0,
+           x = point_function_group_label_y,
+           y = (98-44) /2 + 44 ) +
+     annotate("rect", 
+           xmin = 0.5,
+           xmax = 0.6,
+           ymin = 88,
+           ymax = 230,
+           fill = "yellow",
+           alpha = 0.2) +
+      annotate("text", 
+               label = "Thrusting\nspear",
+               size = point_function_group_label_size,
+               hjust = 0,
+           x = point_function_group_label_y,
+           y = (230-88) /2 + 88 ) +
+  geom_boxplot(alpha = 0.1) +
+  geom_quasirandom(alpha = 0.4,
+                   size = 2.3) +
+  coord_cartesian(xlim = c(-0.1,1)) +
+  theme_bw(base_size = 6) +
+  theme(axis.ticks.x = element_blank(),
+        axis.text.x = element_blank(),
+        axis.title.x = element_blank(),
+        panel.grid.major = element_blank(), 
+        panel.grid.minor = element_blank()
+        ) +
+  labs(y = "TCSA")
+```
+1) The mean TCSA of the Korean stemmed point is `r round(mean(tcsa$min_tcsa), digits=1)`, and standard deviation of TCSA is `r round(sd(tcsa$min_tcsa), digits=1)`.
+Overall we see wide variation of TCSA values in this boxplot.
+2) According to Lombard's (2021) TCSA ranges, Korean stemmed points are mostly in the categories of javelins and thrusting spear tips, with smaller numbers as dart tips and arrowheads.
+3) Poison arrowheads seem to be absent from these Korean assemblages
+
+## Variation by raw material
+
+```{r}
+
+#| label: tcsa-raw-materials
+#| fig-cap: Histograms of TCSA values by lithic raw material.
+#| fig-width:   4.5
+#| fig-height:  4
+
+# create column for "n = xx "
+tcsa_raw <- tcsa %>% 
+  group_by(raw_materials_top) %>% 
+  add_tally() %>% 
+  filter(n>1) %>% 
+  mutate(label_raw = as.factor(paste0(raw_materials_top,' (n = ', n, ")"))) %>%   
+  dplyr::select(min_tcsa,
+         raw_materials_top, 
+       label_raw,
+       n) %>% 
+  distinct_all()
+
+tcsa_site <- tcsa %>% 
+  group_by(site_name_top) %>% 
+  add_tally() %>% 
+  filter(n>1) %>% 
+  mutate(label_site = as.factor(paste0(site_name_top,' (n = ', n, ")"))) %>%   
+  dplyr::select(min_tcsa,
+         site_name_top, 
+         label_site,
+         n) %>% 
+  distinct_all()
+
+# Facet by raw materials
+ ggplot(tcsa_raw,
+              aes(min_tcsa, 
+                  fill = n)) +
+  geom_histogram() +
+  theme_bw(base_size = 6) +
+  labs(x = "TCSA", 
+       y = "Count of artefacts") +
+  facet_wrap(~label_raw,
+                  scales = "free_y",
+                  ncol = 1)  +
+  scale_fill_viridis_c(name = "Number\nof artefacts") +
+  annotate("rect", 
+           xmin = 4,
+           xmax = 18,
+           ymin = 0,
+           ymax = 2.5,
+           fill = "green",
+           alpha = 0.2) +
+      annotate("text", 
+               size = point_function_group_label_size,
+               label = "Poisoned\narrowheads",
+               hjust = 0,
+           y = (0+2.5)/2,
+           x = 5.5) +
+      annotate("rect", 
+           xmin = 13,
+           xmax = 55,
+           ymin = 2.5,
+           ymax = 5,
+           fill = "blue",
+           alpha = 0.2) +
+      annotate("text", 
+               size = point_function_group_label_size,
+               label = "Arrowheads",
+               hjust = 0,
+           y = (2.5+5)/2,
+           x = 28) +
+      annotate("rect", 
+           xmin = 40,
+           xmax = 76,
+           ymin = 5,
+           ymax = 7.5,
+           fill = "orange",
+           alpha = 0.2) +
+      annotate("text", 
+               size = point_function_group_label_size,
+               label = "Dart tips",
+               hjust = 0,
+           y = (5+7.5)/2,
+           x = 53) +
+      annotate("rect", 
+           xmin = 44,
+           xmax = 98,
+           ymin = 7.5,
+           ymax = 10,
+           fill = "purple",
+           alpha = 0.2) +
+      annotate("text", 
+               size = point_function_group_label_size,
+               label = "Javelin",
+               hjust = 0,
+           y = (7.5+10)/2,
+           x = 70) +
+     annotate("rect", 
+           xmin = 88,
+           xmax = 230,
+           ymin = 10,
+           ymax = 12.5,
+           fill = "yellow",
+           alpha = 0.2) +
+      annotate("text", 
+               label = "Thrusting spear",
+               size = point_function_group_label_size,
+               hjust = 0,
+           y = (10+12.5)/2,
+           x = 160)
+
+```
+
+1) About half of all the stemmed points made out of shale. 
+2) Acidic volcanic rocks tend to have skewed lower TCSA values. 
+3) But overall, there is no clear pattern of different TCSA values by raw material.
+
+## Variation by site (Regional pattern)
+
+```{r}
+
+#| label: tcsa-sites
+#| fig-cap: Histograms of TCSA values by archaeological site.
+#| fig-width:   4.5
+#| fig-height:  4
+
+ ggplot(tcsa_site,
+              aes(min_tcsa, 
+                  fill = n)) +
+  geom_histogram() +
+  theme_bw(base_size = 6) +
+  labs(x = "TCSA", 
+       y = "Count of artefacts") +
+  facet_wrap(~label_site,
+                 scales = "free_y",
+                  ncol = 1) +
+    scale_fill_viridis_c(name = "Number\nof artefacts")
+```
+1) We computed TCSA for sites that include >= 5 stemmed points.
+2) Suyanggage has the most stemmed points and the range of TCSA is wide. We can assume that people used stemmed points for a variety of different purposes. 
+3) Nosanri and Sibuk have similarly small TCSA values, perhaps these sites had similar functions?
+Wolpyeng and Yongsujaeul have similar amounts of stemmed points to Nosanri and Sibuk, but different distribution of TCSA values, so perhaps Wolpyeng and Yongsujaeul had a different role in the foragers' land use system compared to Nosanri and Sibuk.
+
+## Variation by size (weight)
+
+```{r}
+
+#| label: tcsa-size
+#| fig-cap: Artifact size and TSCA values by lithic raw material
+#| fig-width:   4.5
+#| fig-height:  4
+
+# remove NAs
+tcsa_mass_no_na <-
+  tcsa %>%
+ drop_na(`Weight(g)`)
+
+ggplot(tcsa_mass_no_na,
+       aes(`Weight(g)`,
+           min_tcsa,
+           colour = raw_materials_top)) +
+  geom_point() +
+    labs(x = "Weight", 
+       y = "TCSA",
+       colour='Raw material') +
+  geom_smooth(method=lm, se=FALSE, size = 0.4)+
+  theme_bw(base_size = 6) +
+  theme(
+  legend.title = element_text(size = 8),
+  legend.text = element_text(size = 7),
+  legend.position = c(.8,.3)
+  )
+```
+1) TCSA and artefact weight have a positive relationship.
+2) The relationship between TCSA and artefact weight varies by raw material, perhaps due to variation in the density of different lithic raw materials
+3) Shale stemmed points tend to have TCSA values that are strongly correlated with artefact size, while acidic volcanic points have TCSA values that vary over a very limited size range, suggesting size is less influential.
+
+```{r}
+
+#| label: artifact-size-cluster
+#| fig-cap: A. Distribution of artefact weight showing three clusters. B. TCSA values for all artefacts with size classes indicated by data point shape.
+#| fig-width:   4.5
+#| fig-height:  4
+
+# Facet by size class
+# https://cran.r-project.org/web/packages/Ckmeans.1d.dp/vignettes/Ckmeans.1d.dp.html
+library(Ckmeans.1d.dp)
+
+# find clusters
+result <- Ckmeans.1d.dp(tcsa_mass_no_na$`Weight(g)`)
+
+tcsa_mass_no_na <-
+tcsa_mass_no_na %>%
+  mutate(mass_cluster = factor(result$cluster))
+
+plot_mass_cluster <- 
+ggplot (tcsa_mass_no_na,
+        aes(1, 
+            `Weight(g)`)) +
+  geom_boxplot(outlier.shape = NA) +
+  geom_quasirandom(aes(shape = mass_cluster),
+                   alpha = 0.4,
+                   size = 2.5) +
+  labs(shape='Cluster',
+       y = "Weight (g)") +
+  theme_bw(base_size = 6) +
+  theme(axis.ticks.x = element_blank(),
+        axis.text.x = element_blank(),
+        axis.title.x = element_blank(),
+        panel.grid.major = element_blank(), 
+        panel.grid.minor = element_blank()
+        )
+
+plot_tcsa_mass_cluster <- 
+ggplot (tcsa_mass_no_na,
+        aes(0.3, min_tcsa)) +
+  geom_boxplot(outlier.shape = NA) +
+  geom_quasirandom(aes(shape = mass_cluster),
+                   alpha = 0.4,
+                   size = 2.5) +
+  labs(shape='Cluster') +
+  theme_bw(base_size = 6) +
+  # show regions for tools from Table 3 of Lomabrd 2021
+  # https://www-sciencedirect-com.offcampus.lib.washington.edu/science/article/pii/S0277379121002286
+    annotate("rect", 
+           xmin = 0.1,
+           xmax = 0.2,
+           ymin = 4,
+           ymax = 18,
+           fill = "green",
+           alpha = 0.2) +
+      annotate("text", 
+               size = point_function_group_label_size,
+               label = "Poisoned\narrowheads",
+               hjust = 0,
+           x = point_function_group_label_y,
+           y = (18-4) /2 + 4) +
+      annotate("rect", 
+           xmin = 0.2,
+           xmax = 0.3,
+           ymin = 13,
+           ymax = 53,
+           fill = "blue",
+           alpha = 0.2) +
+      annotate("text", 
+               size = point_function_group_label_size,
+               label = "Arrowheads",
+               hjust = 0,
+           x = point_function_group_label_y,
+           y = (53-13) /2 + 13) +
+      annotate("rect", 
+           xmin = 0.3,
+           xmax = 0.4,
+           ymin = 40,
+           ymax = 76,
+           fill = "orange",
+           alpha = 0.2) +
+      annotate("text", 
+               size = point_function_group_label_size,
+               label = "Dart tips",
+               hjust = 0,
+           x = point_function_group_label_y,
+           y = (76-40) /2 + 40 ) +
+      annotate("rect", 
+           xmin = 0.4,
+           xmax = 0.5,
+           ymin = 44,
+           ymax = 98,
+           fill = "purple",
+           alpha = 0.2) +
+      annotate("text", 
+               size = point_function_group_label_size,
+               label = "Javelin",
+               hjust = 0,
+           x = point_function_group_label_y,
+           y = (98-44) /2 + 44 ) +
+     annotate("rect", 
+           xmin = 0.5,
+           xmax = 0.6,
+           ymin = 88,
+           ymax = 230,
+           fill = "yellow",
+           alpha = 0.2) +
+      annotate("text", 
+               label = "Thrusting\nspear",
+               size = point_function_group_label_size,
+               hjust = 0,
+           x = point_function_group_label_y,
+           y = (230-88) /2 + 88 ) +
+  coord_cartesian(xlim = c(-0.1,1.2)) +
+  theme_bw(base_size = 6) +
+  theme(axis.ticks.x = element_blank(),
+        axis.text.x = element_blank(),
+        axis.title.x = element_blank(),
+        panel.grid.major = element_blank(), 
+        panel.grid.minor = element_blank()
+        ) +
+  labs(y = "TCSA")
+
+library(cowplot)
+plot_grid(plot_mass_cluster + guides(shape = "none"),
+          plot_tcsa_mass_cluster,
+          labels = "AUTO",
+          rel_widths = c(0.5, 1))
+
+```
+1) Univariate cluster analysis of artefacts by weight reveals three groups or clusters 
+2) Cluster 1, the smallest (lightest) artifacts, is the lower TCSA, compared Cluster 2. TCSA of Cluster 3 is the highest, except for one artifact, which is lower than 50.  
+3) We can assumed that artifacts in different clusters might be made for different purposes.
+
+## Adding Radiocarbon dates info to the SP variation (Temoporal pattern)
+
+```{r}
+
+#| label: tcsa-radio-carbon
+#| fig-cap: Distribution of TCSA values by time
+#| fig-width:   4.5
+#| fig-height:  4
+
+# radiocarbon data per assemblage
+sitedate <- read_excel(here("analysis/data/raw_data/TCSA_site_info.xlsx"))
+
+mydata_ages <-
+  sitedate %>%
+  separate(Period, into = c('age', 'error'),
+           sep = "±") %>%
+  mutate(age_ka = parse_number(age) / 1000,
+         error = parse_number(error))
+
+# combine the assemblage with age data
+
+tcsa_site_ass <- tcsa %>% 
+  left_join(mydata_ages) %>% 
+  group_by(Assemblage) %>% 
+ drop_na(age_ka) %>% 
+  add_tally() %>% 
+  mutate(label_site = as.factor(paste0(Assemblage,' (n = ', n, '), (', {round(age_ka,1)}, 'ka)' ))) %>%   
+  dplyr::select(age_ka,
+         min_tcsa,
+         Assemblage, 
+         label_site,
+         n) %>% 
+    distinct_all()
+
+ ggplot(tcsa_site_ass, aes(x = min_tcsa, y = reorder(label_site, age_ka), color = label_site)) +
+   geom_point() +
+   theme(legend.position="none") +
+   labs(x = "TCSA", 
+       y = "Assemblage")
+
+```
+
+# Discussion
+
+# Conclusion
+
 # Acknowledgements

### 2022-10-03 analysis/paper/paper.qmd
Added text for the first results

@@
 
-## TCSA Variation
+## TCSA range of Korean stemmed points
 
@@
 ```
-1) The mean TCSA of the Korean stemmed point is `r round(mean(tcsa$min_tcsa), digits=1)`, and standard deviation of TCSA is `r round(sd(tcsa$min_tcsa), digits=1)`.
-Overall we see wide variation of TCSA values in this boxplot.
-2) According to Lombard's (2021) TCSA ranges, Korean stemmed points are mostly in the categories of javelins and thrusting spear tips, with smaller numbers as dart tips and arrowheads.
-3) Poison arrowheads seem to be absent from these Korean assemblages
+
+To gain a general idea of the role of stemmed points, we calculated TCSA for all stemmed points.
+@tcsa-all-sp shows the mean TCSA is `r round(mean(tcsa$min_tcsa), digits=1)`, and standard deviation of TCSA is `r round(sd(tcsa$min_tcsa), digits=1)`.
+Overall we see a wide variation of TCSA values in this boxplot.
+According to -@lombard2021variation TCSA ranges (Table 1), Korean stemmed points are mostly in the categories of javelins and thrusting spear tips, with smaller numbers as dart tips and arrowheads.
+Among the weapon-tip types, only poison arrowheads appear to be absent from these Korean assemblages.
 

### 2022-10-03 analysis/paper/paper.qmd
update scholarly-metadata filter and we can render again

@@
 csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
+filters:
+  - ../templates/scholarly-metadata.lua
+  - ../templates/author-info-blocks.lua
+  - ../templates/pagebreak.lua
 
@@
 ```
-1) We computed TCSA for sites that include >= 5 stemmed points.
+
+1)  We computed TCSA for sites that include \>= 5 stemmed points.
 2)  Suyanggage has the most stemmed points and the range of TCSA is wide. We can assume that people used stemmed points for a variety of different purposes.
@@
 ```
+
 1)  TCSA and artefact weight have a positive relationship.
@@
 ```
+
 1)  Univariate cluster analysis of artefacts by weight reveals three groups or clusters
-2) Cluster 1, the smallest (lightest) artifacts, is the lower TCSA, compared Cluster 2. TCSA of Cluster 3 is the highest, except for one artifact, which is lower than 50.  
+2)  Cluster 1, the smallest (lightest) artifacts, is the lower TCSA, compared Cluster 2. TCSA of Cluster 3 is the highest, except for one artifact, which is lower than 50.\
 3)  We can assumed that artifacts in different clusters might be made for different purposes.

### 2022-10-03 analysis/paper/paper.qmd
the chunk label needs fig-xxx for captions to work

@@
 ```{r}
-#| label: tcsa-all-sp
+#| label: fig-tcsa-all-sp
 #| fig-cap: Distribution of TCSA values for all Korean stemmed points in the current dataset. Shaded boxes show TSCA ranges for various weapon types.
@@
 To gain a general idea of the role of stemmed points, we calculated TCSA for all stemmed points.
-@tcsa-all-sp shows the mean TCSA is `r round(mean(tcsa$min_tcsa), digits=1)`, and standard deviation of TCSA is `r round(sd(tcsa$min_tcsa), digits=1)`.
+@fig-tcsa-all-sp shows the mean TCSA is `r round(mean(tcsa$min_tcsa), digits=1)`, and standard deviation of TCSA is `r round(sd(tcsa$min_tcsa), digits=1)`.
 Overall we see a wide variation of TCSA values in this boxplot.

### 2022-10-03 analysis/paper/paper.qmd
"fig-" option doesn't work for other figure. The figure that Ben tried was already working except for showing "Figure. 1".

@@
 
-#| label: tcsa-raw-materials
+#| label: fig-tcsa-raw-materials
 #| fig-cap: Histograms of TCSA values by lithic raw material.
@@
 
-1)  About half of all the stemmed points made out of shale.
+1)  About half of all the stemmed points made out of shale (@fig-tcsa-raw-materials).
 2)  Acidic volcanic rocks tend to have skewed lower TCSA values.

### 2022-10-04 analysis/paper/paper.qmd
Deleted rmd file. Tried cache = false, then deleted related files

@@
 
-```{r}
+```{r, include = FALSE}
 
@@
 
+## Paleo climate data
+```{r, include = FALSE}
+library(ncdf4)
+library(lattice)
+
+# you'll need to get this file from
+# https://figshare.com/articles/LateQuaternary_Environment_nc/12293345/3
+# and move to our raw_data folder
+file <- here::here("analysis/data/raw_data/LateQuaternary_Environment.nc")
+
+env_nc      <- ncdf4::nc_open(file)
+longitude   <- ncdf4::ncvar_get(env_nc, "longitude")
+latitude    <- ncdf4::ncvar_get(env_nc, "latitude")
+years       <- ncdf4::ncvar_get(env_nc, "time")
+months      <- ncdf4::ncvar_get(env_nc, "month")
+temperature <- ncdf4::ncvar_get(env_nc, "temperature")
+biome       <- ncdf4::ncvar_get(env_nc, "biome")
+ncdf4::nc_close(env_nc)
+
+# this takes a few minutes
+mean_annual_temperature <- apply(temperature, c(1, 2, 4), mean)
+
+site_locations_tbl_temps <-
+      korean_archaeological_site_locations %>% 
+      dplyr::select(Site_name,
+         lat_dd,
+         long_dd) %>%
+  rowwise()  %>%
+  mutate(lonID = which.min(abs(longitude - long_dd)),
+         latID = which.min(abs(latitude - lat_dd))) %>%
+  mutate(mean_annual_temperature = list(tibble(year = years,
+                                               mean_annual_temperature = mean_annual_temperature[lonID, latID, ]))) %>%
+  unnest(mean_annual_temperature) %>%
+  filter(between(year, -50000, -10000))
+```
+
 # Conclusion

### 2022-10-04 analysis/paper/paper.qmd
Deleted unused code chunk and fix the authors' institutions. <-- the way of putting institution for our CT paper didn't work well. I simplified the code.

--- 2022-10-03 analysis/paper/paper.qmd
@@
   - Gayoung Park:
-      institute: uw
+      institute: University of Washington
       correspondence: "yes"
@@
   - Marlize Lombard:
-      institute: uj
+      institute: University of Johannesburg
   - Ben Marwick:
-      institute: uw
+      institute: University of Washington
   - Donghee Chong:
-      institute: uk
-institute:
-  - uw:
-    name: Department of Anthropology, University of Washington
-    address: Seattle, WA, USA
-  - uj:
-    name: Palaeo-Research Institute, University of Johannesburg
-    address: Johannesburg, South Africa
-  - uk:
-    name: Department of History, Kyung Hee University
-    address: Seoul, South Korea
+      institute: Kyung Hee University
 format: 
@@
 
+## Paleo climate data?
+
 # Conclusion

### 2022-10-06 analysis/paper/paper.qmd
adjusting the annotation (size) in raw material plot

@@
 #| fig-width:   4.5
-#| fig-height:  4
+#| fig-height:  6
 
@@
       annotate("text", 
-               size = point_function_group_label_size,
+               size = 1.5,
                label = "Poisoned\narrowheads",
                hjust = 0,
+               lineheight = 0.75,
            y = (0+2.5)/2,

### 2022-10-06 analysis/paper/paper.qmd
adjust the plot for site (regional pattern)

@@
   dplyr::filter(min_tcsa <= 257) %>%
-  # group rare raw materials
+  # group rare raw materials > 10
   mutate(raw_materials_top = fct_lump_min(Raw_material, 10)) %>%
@@
 # Facet by raw materials
+
 korean_archaeological_site_locations <-
@@
 
-1)  About half of all the stemmed points made out of shale (@fig-tcsa-raw-materials).
-2)  Acidic volcanic rocks tend to have skewed lower TCSA values.
-3)  But overall, there is no clear pattern of different TCSA values by raw material.
+In order to explore the relationship between raw material and the function of the stemmed points, we calculated TCSA per raw material type.
+@fig-tcsa-raw-materials shows that about half of the stemmed points were made from shale and its TCSA range is widely distributed.
+Acidic volcanic rocks tend to have skewed lower TCSA values.
+The other raw materials including hornfels, rhyolite, and tuff show a wide distribution, similar to shale.
+The category of ‘other’ raw materials include porphyry, trachyte, felsite, chert, quartz, quartzite, granite, mudstone, and unidentified rocks.
+Overall, there is no clear pattern of TCSA among raw materials.
 
@@
                   ncol = 1) +
-    scale_fill_viridis_c(name = "Number\nof artefacts")
+    scale_fill_viridis_c(name = "Number\nof artefacts") +
+  annotate("rect", 
+           xmin = 4,
+           xmax = 18,
+           ymin = 0,
+           ymax = 2.5,
+           fill = "green",
+           alpha = 0.2) +
+      annotate("text", 
+               size = 1.5,
+               label = "Poisoned\narrowheads",
+               hjust = 0,
+               lineheight = 0.75,
+           y = (0+2.5)/2,
+           x = 5.5) +
+      annotate("rect", 
+           xmin = 13,
+           xmax = 55,
+           ymin = 2.5,
+           ymax = 5,
+           fill = "blue",
+           alpha = 0.2) +
+      annotate("text", 
+               size = point_function_group_label_size,
+               label = "Arrowheads",
+               hjust = 0,
+           y = (2.5+5)/2,
+           x = 28) +
+      annotate("rect", 
+           xmin = 40,
+           xmax = 76,
+           ymin = 5,
+           ymax = 7.5,
+           fill = "orange",
+           alpha = 0.2) +
+      annotate("text", 
+               size = point_function_group_label_size,
+               label = "Dart tips",
+               hjust = 0,
+           y = (5+7.5)/2,
+           x = 53) +
+      annotate("rect", 
+           xmin = 44,
+           xmax = 98,
+           ymin = 7.5,
+           ymax = 10,
+           fill = "purple",
+           alpha = 0.2) +
+      annotate("text", 
+               size = point_function_group_label_size,
+               label = "Javelin",
+               hjust = 0,
+           y = (7.5+10)/2,
+           x = 70) +
+     annotate("rect", 
+           xmin = 88,
+           xmax = 230,
+           ymin = 10,
+           ymax = 12.5,
+           fill = "yellow",
+           alpha = 0.2) +
+      annotate("text", 
+               label = "Thrusting spear",
+               size = point_function_group_label_size,
+               hjust = 0,
+           y = (10+12.5)/2,
+           x = 160)
 ```

### 2022-10-06 analysis/paper/paper.qmd
added text to method section and tried to add 'fig' for caption. It didn't work...

@@
 
-1)  We computed TCSA for sites that include \>= 5 stemmed points.
-2)  Suyanggage has the most stemmed points and the range of TCSA is wide. We can assume that people used stemmed points for a variety of different purposes.
-3)  Nosanri and Sibuk have similarly small TCSA values, perhaps these sites had similar functions?
-Wolpyeng and Yongsujaeul have similar amounts of stemmed points to Nosanri and Sibuk, but different distribution of TCSA values, so perhaps Wolpyeng and Yongsujaeul had a different role in the foragers' land use system compared to Nosanri and Sibuk.
+We computed TCSA for sites that contain more than 5 stemmed points to observe the pattern.
+As shown in @tcsa-sites, the sites with fewer stemmed points are grouped under the category of 'other'.
+Among the selected sites, Suyanggae has the most stemmed points and the widest range of TCSA.
+We can assume that people made stemmed points for a variety of different purposes in Suyanggae.
+The TCSA values for Nosanri and Sibuk are skewed, which indicates less diverse functions for stemmed points, and lower.
+Wolpyeng and Yongsujaeul have similar amounts of stemmed points to Nosanri and Sibuk, but exhibit different distribution of TCSA values.
+Yongsandong has the second most stemmed points and TCSA shows a similar distribution to Yongsakul.
 
@@
 
-#| label: tcsa-size
+#| label: fig-tcsa-size
 #| fig-cap: Artifact size and TSCA values by lithic raw material
@@
 
-1)  TCSA and artefact weight have a positive relationship.
+1)  TCSA and artefact weight have a positive relationship. @fig-tcsa-size shows~
 2)  The relationship between TCSA and artefact weight varies by raw material, perhaps due to variation in the density of different lithic raw materials

### 2022-10-06 analysis/paper/paper.qmd
add text to the result section

@@
 
-1)  TCSA and artefact weight have a positive relationship. @fig-tcsa-size shows~
-2)  The relationship between TCSA and artefact weight varies by raw material, perhaps due to variation in the density of different lithic raw materials
-3)  Shale stemmed points tend to have TCSA values that are strongly correlated with artefact size, while acidic volcanic points have TCSA values that vary over a very limited size range, suggesting size is less influential.
+We chose weight as a size proxy and examined the relationship between size and function of the stemmed points.
+We combined the raw material information as another variable for the analysis.
+@fig-tcsa-size shows a positive relationship between TCSA and artifact weight.
+The relationship varies by raw material types, perhaps due to variation in the density of different lithic raw materials.
+There is a stronger correlation between TCSA and artifact weight for acidic volcanic points (i.e.
+closer to the regression line), whereas the correlation for shale stemmed points is less pronounced.
+The other raw materials show different distribution patterns around the regression line, reconfirming that raw material is less influential (as @fig-tcsa-raw-materials).
 
@@
 
-1)  Univariate cluster analysis of artefacts by weight reveals three groups or clusters
-2)  Cluster 1, the smallest (lightest) artifacts, is the lower TCSA, compared Cluster 2. TCSA of Cluster 3 is the highest, except for one artifact, which is lower than 50.\
-3)  We can assumed that artifacts in different clusters might be made for different purposes.
+We then conducted univariate cluster analysis of stemmed points by weight and it revealed three groups of clusters (A of @artifact-size-cluster, mean = `r round(mean(tcsa_mass_no_na$'Weight(g)'), digits=1)`, SD = `r round(sd(tcsa_mass_no_na$'Weight(g)'), digits=1)`).
+Cluster 1, the smallest (lightest) artifacts, is the lower TCSA, compared to Cluster 2.
+TCSA of Cluster 3 is the highest, except for one artifact, which is lower than 50 (A of @artifact-size-cluster, mean = `r round(mean(tcsa_mass_no_na$'min_tcsa'), digits=1)`, SD = `r round(sd(tcsa_mass_no_na$'min_tcsa'), digits=1)`).
+There are a few overlaps, but we can still assume that artifacts in different clusters might be made for different purposes.
 
-## Adding Radiocarbon dates info to the SP variation (Temoporal pattern)
+## Temporal pattern of TCSA
 
@@
 ```
+Using radiocarbon dates, we arranged 24 assemblages in chronological order (@tcsa-radio-carbon).
+Our analysis shows that early assemblages (i.e.
+45-24 ka) tend to possess numerous stemmed points such as two from the Suyanggae site and they have a wide range of TCSA, indicating multiple usages of stemmed points.
+Overall, TCSA from assemblages that have more than 3 stemmed points vary considerably, which makes it difficult to define a single function of the artifacts.
+There is also no clear temporal pattern in assemblages with a fewer number of artifacts.
 

### 2022-10-07 analysis/paper/paper.qmd
Update paper.qmd

@@
 ```{r}
-
 #| label: fig-sp-hand-made
-#| fig-cap: Stemmed points from Yongsandong site. 1-6: plain stemmed points, 7-8: one side denticulate stemmed points.
+#| fig-cap: "Stemmed points from Yongsandong site. 1-6: plain stemmed points, 7-8: one side denticulate stemmed points."
 
@@
 -@Lee_Kong_2002 claimed to name the stemmed points as 'stemmed tools' because of the diverse shape of stemmed points, which could represent different functions as well as the uncertainty of complete tool shape.
-Other researchers agree that some stemmed points that are not symmetrical and have retouch on one side of the artifacts or have denticulate blades on one side should not be called as stemmed points but stemmed knives, stemmed side-scrapers, stemmed end-scrapers, or stemmed burins [@seong2008tanged; @Lee_Jang_2011; @Kim] (@fig-sp-hand-made: 7-8).
+Other researchers agree that some stemmed points that are not symmetrical and have retouch on one side of the artifacts or have denticulate blades on one side should not be called as stemmed points but stemmed knives, stemmed side-scrapers, stemmed end-scrapers, or stemmed burins [@seong2008tanged; @Lee_Jang_2011; @Kim] @fig-sp-hand-made: 7-8.
 
@@
 Our dataset contains stemmed points that were found during field surveys as well as those were from sites that were never dated but accompanied with other Late Paleoltihic artifacts.
-Those criteria resulted in a sample of 173 stemmed points from 36 assemblages unearthed from 29 sites spanning the period 44-10 ka (Figure 2).
+Those criteria resulted in a sample of 173 stemmed points from 36 assemblages unearthed from 29 sites spanning the period 44-10 ka (@fig-map).
 The dimensions of the 173 stemmed points were obtained from published excavation reports and by direct measurements during our visits to the collections of Korean local museums and archaeological institutions.
@@
 ```{r}
-
 #| label: fig-map
-#| fig-cap: Korean Paleolithic sites mentioned in this study
+#| fig-cap: "Korean Paleolithic sites mentioned in this study"
 #| fig-width:   4.45
@@
 
+# if we have already downloaded the map tiles, skip doing it again
+if (file.exists(here::here("analysis/data/raw_data/map.RData"))) {
+
+  # load the the map tiles have already been downloaded
+  load(file = here::here("analysis/data/raw_data/map.RData"))
+
+  } else {
+
+    # download map tiles for our study area if we don't 
+    # have them already
 map <-
@@
                 zoom = 10)
+# and save the tiles so we don't have to do this every time we render
+save(map, file = here::here("analysis/data/raw_data/map.RData"))
+
+}
 
@@
 #| label: fig-tcsa-all-sp
-#| fig-cap: Distribution of TCSA values for all Korean stemmed points in the current dataset. Shaded boxes show TSCA ranges for various weapon types.
+#| fig-cap: "Distribution of TCSA values for all Korean stemmed points in the current dataset. Shaded boxes show TSCA ranges for various weapon types."
 #| fig-width:   4.5
@@
 ```{r}
-
 #| label: fig-tcsa-raw-materials
-#| fig-cap: Histograms of TCSA values by lithic raw material.
+#| fig-cap: "Histograms of TCSA values by lithic raw material."
 #| fig-width:   4.5
@@
 ```{r}
-
-#| label: tcsa-sites
-#| fig-cap: Histograms of TCSA values by archaeological site.
+#| label: fig-tcsa-sites
+#| fig-cap: "Histograms of TCSA values by archaeological site."
 #| fig-width:   4.5
@@
 We computed TCSA for sites that contain more than 5 stemmed points to observe the pattern.
-As shown in @tcsa-sites, the sites with fewer stemmed points are grouped under the category of 'other'.
+As shown in @fig-tcsa-sites, the sites with fewer stemmed points are grouped under the category of 'other'.
 Among the selected sites, Suyanggae has the most stemmed points and the widest range of TCSA.
@@
 ```{r}
-
 #| label: fig-tcsa-size
-#| fig-cap: Artifact size and TSCA values by lithic raw material
+#| fig-cap: "Artifact size and TSCA values by lithic raw material"
 #| fig-width:   4.5
@@
 ```{r}
-
-#| label: artifact-size-cluster
-#| fig-cap: A. Distribution of artefact weight showing three clusters. B. TCSA values for all artefacts with size classes indicated by data point shape.
+#| label: fig-artifact-size-cluster
+#| fig-cap: "A. Distribution of artefact weight showing three clusters. B. TCSA values for all artefacts with size classes indicated by data point shape."
 #| fig-width:   4.5
@@
 
-We then conducted univariate cluster analysis of stemmed points by weight and it revealed three groups of clusters (A of @artifact-size-cluster, mean = `r round(mean(tcsa_mass_no_na$'Weight(g)'), digits=1)`, SD = `r round(sd(tcsa_mass_no_na$'Weight(g)'), digits=1)`).
+We then conducted univariate cluster analysis of stemmed points by weight and it revealed three groups of clusters (A of @fig-artifact-size-cluster, mean = `r round(mean(tcsa_mass_no_na$'Weight(g)'), digits=1)`, SD = `r round(sd(tcsa_mass_no_na$'Weight(g)'), digits=1)`).
 Cluster 1, the smallest (lightest) artifacts, is the lower TCSA, compared to Cluster 2.
-TCSA of Cluster 3 is the highest, except for one artifact, which is lower than 50 (A of @artifact-size-cluster, mean = `r round(mean(tcsa_mass_no_na$'min_tcsa'), digits=1)`, SD = `r round(sd(tcsa_mass_no_na$'min_tcsa'), digits=1)`).
+TCSA of Cluster 3 is the highest, except for one artifact, which is lower than 50 (A of @fig-artifact-size-cluster, mean = `r round(mean(tcsa_mass_no_na$'min_tcsa'), digits=1)`, SD = `r round(sd(tcsa_mass_no_na$'min_tcsa'), digits=1)`).
 There are a few overlaps, but we can still assume that artifacts in different clusters might be made for different purposes.
@@
 ```{r}
-
-#| label: tcsa-radio-carbon
-#| fig-cap: Distribution of TCSA values by time
+#| label: fig-tcsa-radio-carbon
+#| fig-cap: "Distribution of TCSA values by time"
 #| fig-width:   4.5
@@
    geom_point() +
+   theme_minimal() +
    theme(legend.position="none") +
@@
 ```
-Using radiocarbon dates, we arranged 24 assemblages in chronological order (@tcsa-radio-carbon).
+
+Using radiocarbon dates, we arranged 24 assemblages in chronological order (@fig-tcsa-radio-carbon).
 Our analysis shows that early assemblages (i.e.

### 2022-10-08 analysis/paper/paper.qmd
add map data
fix some bad in text citations

@@
 Using blades enables mass production for this composite tool and its shape becomes more standardized [@lee2015characteristics; @park2022did].
-Therefore, -@o2018origins claims that understanding the emergence of projectile technology gives insights to greater cultural, evolutionary, and behavioral cognitive flexibility.
+Therefore, @o2018origins claims that understanding the emergence of projectile technology gives insights to greater cultural, evolutionary, and behavioral cognitive flexibility.
 
 Since the stemmed points represent the beginning of the Korean Late Paleolithic, investigating their origin(s) becomes a key for understanding the technological transition from the Early Paleoltihic, modern human dispersals, and existence of the 'Middle' Paleolithic in Korea [@bae2010origin; @bae2017late; @norton2009evolution; @seong2016eastern].
-There has been a northworthy debate about explaining the origin(s) of stemmed points and it can be summarized into two competing models: 'heterogenic' migration [@bae2010origin], in situ evolution[@seong2009emergence].
+There has been a northworthy debate about explaining the origin(s) of stemmed points and it can be summarized into two competing models: 'heterogenic' migration [@bae2010origin], in situ evolution [@seong2009emergence].
 The migration model claims that the new blade industry including stemmed points, and the earlier coarse flake tradition including large cores, polyhedrons, choppers, and even handaxes, came from different origins (e.g.
@@
 Stemmed points are important proxies to understand human behaviors during the Late Paleolithic.
--@park2022did examine mobility and site occupation patterns through lithic assemblages and concepts of human behavioral ecology and find that forager groups using stemmed points may have been associated with occupation of marginal or extreme environments compared to the groups with no stemmed points.
+@park2022did examine mobility and site occupation patterns through lithic assemblages and concepts of human behavioral ecology and find that forager groups using stemmed points may have been associated with occupation of marginal or extreme environments compared to the groups with no stemmed points.
 Also, they were more associated with expedient technologies, indicating residential and less mobile behaviors.
--@Chong_2021 claims that the morphological variation of the stemmed points along with tool types in lithic assemblages, assemblage size, use of raw materials, and types of blanks could represent the characteristics of occupations such as "limited activity station" and "residential base camp." For example, stemmed points with high morphological variations in a tool size, shape of edge, degrees of damage, and types of edge retouching from Yongsandong site indicate that the site was used for specific or limited activities such as hunting [@kim2004yongsan; @seong2015diversity; @Chong_2021].
+@Chong_2021 claims that the morphological variation of the stemmed points along with tool types in lithic assemblages, assemblage size, use of raw materials, and types of blanks could represent the characteristics of occupations such as "limited activity station" and "residential base camp." For example, stemmed points with high morphological variations in a tool size, shape of edge, degrees of damage, and types of edge retouching from Yongsandong site indicate that the site was used for specific or limited activities such as hunting [@kim2004yongsan; @seong2015diversity; @Chong_2021].
 
@@
 Preserved wooden components of projectile tools are too few and rare to generalize the overall scale and variability of the tools used in prehistory [@shea2006origins].
--@Lee_Kong_2002 claimed to name the stemmed points as 'stemmed tools' because of the diverse shape of stemmed points, which could represent different functions as well as the uncertainty of complete tool shape.
+@Lee_Kong_2002 claimed to name the stemmed points as 'stemmed tools' because of the diverse shape of stemmed points, which could represent different functions as well as the uncertainty of complete tool shape.
 Other researchers agree that some stemmed points that are not symmetrical and have retouch on one side of the artifacts or have denticulate blades on one side should not be called as stemmed points but stemmed knives, stemmed side-scrapers, stemmed end-scrapers, or stemmed burins [@seong2008tanged; @Lee_Jang_2011; @Kim] @fig-sp-hand-made: 7-8.
@@
 TCSA metric indicates the part of the tool that cuts the target's hide and relates to weapon flight and penetration dynamics [@hughes1998getting; @sitton2020tip; @lombard2021variation].
-This method was first proposed by -@hughes1998getting and and tested by -@shea2006origins through comparing archeological examples to ethnographically collected samples of known use.
--@shea2009impact also applied this approach to compare projectile points from Africa, the Levant, and Europe and claimed that projectile weapons first appeared in Africa.
+This method was first proposed by @hughes1998getting and and tested by @shea2006origins through comparing archeological examples to ethnographically collected samples of known use.
+@shea2009impact also applied this approach to compare projectile points from Africa, the Levant, and Europe and claimed that projectile weapons first appeared in Africa.
 The main advantage of the TCSA is, regardless of point type, only maximum width and thickness measurements are needed for calculating the TCSA value (0.5 × maximum width × maximum thickness) [@sisk2011african; @lombard2020tip].
-Later, -@sisk2011african proposed an alternate measure, tip cross-sectional perimeter (TCSP), for more accurate measure for the force needed to penetrate a target to a lethal depth while TCSA metric is more associated with cutting aspect.
+Later, @sisk2011african proposed an alternate measure, tip cross-sectional perimeter (TCSP), for more accurate measure for the force needed to penetrate a target to a lethal depth while TCSA metric is more associated with cutting aspect.
 However, TCSP has a few disadvantages to directly apply to our case of stemmed points.
 The force and penetration depth are not only affected by the stone tip but also by the mass of the shaft, which cannot be known for archaeological stone-tipped weapons [@lombard2020tip].
--@sisk2011 also mentioned that TCSP cannot be applied to backed pieces that are hafted as projectile armatures.
-
-\-@lee2019were first applied TCSA to stemmed points from Korea along with their use-wear analysis.
+@sisk2011 also mentioned that TCSP cannot be applied to backed pieces that are hafted as projectile armatures.
+
+@lee2019were first applied TCSA to stemmed points from Korea along with their use-wear analysis.
 The stemmed points for their research were discovered from Jingeuneul site located in southwest of Korea, where the largest number of stemmed points (n = 99) were found in a single site.
@@
 Their results show that the TCSA range of Jingeuneul stemmed points is relatively wide overlapped with both North American dart tips and arrowheads.
-Inspired by -@lee2019were in addition to other prior research that developed and applied TCSA into archaeological materials from other parts of the world, we explore stemmed points from all over South Korea to better understand their function to understand the role that stemmed points represent during the Late Paleolithic.
+Inspired by @lee2019were in addition to other prior research that developed and applied TCSA into archaeological materials from other parts of the world, we explore stemmed points from all over South Korea to better understand their function to understand the role that stemmed points represent during the Late Paleolithic.
 
@@
 In the discussion, we compared those temporal patterns with demographic and environmental proxies.
-We used the results of summed probability distributions (SPD) and annual temperature changes from -@park2022did which covers our target period and area.
-Following the TCSA ranges for different weapon-delivery systems that -@lombard2021variation created by summarizing -@wadley2008segment' study of backed microliths (Table 1?), we excluded 12 artifacts from our dataset.
+We used the results of summed probability distributions (SPD) and annual temperature changes from @park2022did which covers our target period and area.
+Following the TCSA ranges for different weapon-delivery systems that @lombard2021variation created by summarizing @wadley2008segment' study of backed microliths (Table 1?), we excluded 12 artifacts from our dataset.
 Because the TCSA values of those 12 artifacts from Bupyeongri, Hwadaeri, Haga, Sinhwari, Gokcheon, and Sachang sites are greater than 257.
@@
 The choice of weapon-tip types, therefore, reflects the socio-environmental circumstances that they had to encounter and overcome.
--@eren2022plains claim that Clovis points were more variable in shape than Folsom points because Clovis foragers were exposed to largely unfamiliar landscape and employed their points as multifunctional tools that perform a wider range of tasks including cutting and sawing.
+@eren2022plains claim that Clovis points were more variable in shape than Folsom points because Clovis foragers were exposed to largely unfamiliar landscape and employed their points as multifunctional tools that perform a wider range of tasks including cutting and sawing.
 Folsom points show a narrow range of variation, indicating a specific task.
@@
 
-Similar to -@eren2022plains's approach, we hypothesize two scenarios that explain the regional or/and temporal pattern of TCSA range.
+Similar to @eren2022plains's approach, we hypothesize two scenarios that explain the regional or/and temporal pattern of TCSA range.
 If stemmed points represent a narrow TCSA range, people produced tools that performed certain tasks.
@@
 Overall we see a wide variation of TCSA values in this boxplot.
-According to -@lombard2021variation TCSA ranges (Table 1), Korean stemmed points are mostly in the categories of javelins and thrusting spear tips, with smaller numbers as dart tips and arrowheads.
+According to @lombard2021variation TCSA ranges (Table 1), Korean stemmed points are mostly in the categories of javelins and thrusting spear tips, with smaller numbers as dart tips and arrowheads.
 Among the weapon-tip types, only poison arrowheads appear to be absent from these Korean assemblages.
@@
 The other raw materials including hornfels, rhyolite, and tuff show a wide distribution, similar to shale.
-The category of ‘other’ raw materials include porphyry, trachyte, felsite, chert, quartz, quartzite, granite, mudstone, and unidentified rocks.
+The category of 'other' raw materials include porphyry, trachyte, felsite, chert, quartz, quartzite, granite, mudstone, and unidentified rocks.
 Overall, there is no clear pattern of TCSA among raw materials.

### 2022-10-09 analysis/paper/paper.qmd
Add text to the discussion section

@@
 Our results show that the stemmed points mostly functioned as javelins and thrusting spear tips, with smaller numbers as dart tips and arrowheads.
-TCSA values are depending on size and raw material types.
+TCSA values are depending mostly on size rather than raw material types.
 We found different usage of stemmed points in different sites, which could indicate people used stemmed points in different ways depending on the environment.
@@
 The temporal pattern of TCSA values is one of little change throughout the Late Paleolithic period.
-We conclude that stemmed points were mainly used as Javelin but they were multi-functional tools.
+We conclude that stemmed points were mainly designed for Javelin and thrusting spear tips but they were multi-functional tools.
 keywords: |
@@
 
-## Choice of weapon-tip types
+## Two models for weapon-tip types choice
 
@@
 
+To investigate the role that stemmed points played in the technological transition during the Korean Late Paleolithic, we applied TCSA metric to estimate artifact function by comparing the results to other archaeological and ethnographic cases.
+We asked three questions to achieve our research goal: What were stemmed points used for?
+How diverse were their functions?
+What are the temporal patterns in stemmed point functions?
+In our assumptions, human groups were capable of intentionally choosing and designing tools according to their socio-environmental circumstances.
+We came up with two models: (1) people created stemmed points and used them as a specific type of tool; or (2) they created stemmed points for multifunctional tools.
+
+Our results indicate that javelins and thrusting spear tips are the best matches for stemmed points.
+There are a few stemmed points that can be used as dart tips and arrowheads, but none of them are suitable for poison arrowheads (@fig-tcsa-all-sp).
+We found stemmed points played a diverse role, by showing a wide range of TCSA values.
+Some sites, such as Nosanri and Sinbuk, used stemmed points as limited-function tools.
+Meanwhile, stemmed points were used in multiple ways at other sites, especially Suyanggae.
+In general, stemmed points played multiple roles in more sites (@fig-tcsa-sites).
+Through the Late Paleolithic period, TCSA values hardly changed (@fig-tcsa-radio-carbon).
+The chronological sequence of TCSA varies from region to region with no clear pattern.
+Therefore, we conclude that stemmed points were mainly designed for Javelin and thrusting spear tips but they were multi-functional tools.
+
+We are aware of the concern that discriminating the function of small projectile points could be arbitrary work [@erlandson2014darts].
+Because TCSA covers the critical elements of projectiles, flight and penetration dynamics (i.e.
+increased or decrease by shape of tip and cross section), we consider it a useful metric for discriminating between weapon tip functions [@hughes1998getting; @sitton2020tip; @lombard2021variation].
+Stemmed points have been considered so far as long-distance hunting amateurs pursuing large bodies of prey such as a deer [@lee2019were; @lee2011study; @seong2008tanged; @lee2002new; @chang2013human].
+TCSA enabled us to gain a more detailed understanding of the function of stemmed points.
+
+TCSA can, however, be impacted by other factors such as raw materials and portability.
+-@eren2022plains explain that smaller TCSA could be the result of pursuing production economy and transport efficiency.
+We explore the relationship between TCSA and raw material types and weight in order to examine these factors.
+Weight plays an important role in portability and penetration of projectiles [@hughes1998getting].
+According to our results, raw materials do not have a clear pattern, but weight is strongly correlated with raw materials (@fig-tcsa-raw-materials and @fig-tcsa-size).
+Different clusters of weight are matched with different types of weapon tips (@fig-artifact-size-cluster).
+Based on these analyses, we assume that people were able to produce the ideal size tools regardless of raw materials.
+
 ## Paleo climate data?

### 2022-10-10 analysis/paper/paper.qmd
Discussion text done.

@@
 
-## Paleo climate data?
+Our models premise that the choice of weapon-tip types reflects the socio-environmental circumstances.
+To examine the possible connections between the roles of stemmed points and population dynamics or environmental change during the Late Paleolithic period, we compared the temporal pattern of TCSA to results of Summed probability distribution (SPD) and simulated Mean Annual Temperature (MAT) from @park2022did.
+In those two analyses, -@park2022did covered assemblages with and without stemmed points, which is useful for us to explore the overall socio-environmental dynamics.
+As we discussed, there is no clear pattern in the temporal or regional pattern of TCSA (@fig-tcsa-radio-carbon or @fig-socio-climate: A).
+We consider that TCSA, or the function of the stemmed points, are strongly dependent on the surrounding environment or their situation.
+@fig-socio-climate: B shows that the temperature slowly decreased until around 22 ka and then went up again with a little range of fluctuations.
+Temperatures differ significantly between sites, with a maximum difference of 5℃.
+Results of SPD show several peaks, as indicated by the positive deviations from the three null models (shaded in red), and downs (shaded in blue), but no drastic growth or decline (@fig-socio-climate: C).
+From the climate and population study, we could observe that there is no dominating change but minor and often fluctuations throughout the Late Paleolithic and variations per site.
+Overall this pattern is well matched with the TCSA results, indicating the people produced stemmed points as multifunctional tools suitable for unexpected or various situations in their habitats similar to the case of Clovis points (@eren2022plains).
+
+```{r}
+#| label: fig-socio-climate
+#| fig-cap: "A.Distribution of TCSA values by time.
+B.Mean Annual Temperature (MAT) of the Korean Paleolithic assemblages from 50 ka to 10 ka.
+The gray lines indicate the MAT for each site and the black line is the mean temperature of all sites.
+The light gray area in the middle of the plot indicates the duration of MIS 2, and the dark gray area represents the duration of the LGM (from @park2022did: Figure 9).
+C.
+Summed probability distribution (SPD) of 100 radiocarbon dates.
+The black solid line represents actual radiocarbon ages.
+The gray shaded region shows the Monte Carlo envelope that encompasses the 95% confidence interval for the null models.
+The red and blue vertical bands highlight the portions of the SPD where positive and negative deviations are detected.
+The best fit model has the lowest AIC score, which is the linear model in this case (from @park2022did: Figure 10)."
+#| fig-width:   4.5
+#| fig-height:  12
+
+tem_pattern <- ggplot(tcsa_site_ass, aes(x = min_tcsa, y = reorder(label_site, age_ka))) +
+   geom_point(color="gray", size =1.5) +
+   theme_minimal() +
+   theme(legend.position="none") +
+   labs(x = "TCSA", 
+       y = "Assemblage")
+
+# imput images from the Park&Marwick 2022 paper
+# https://github.com/parkgayoung/koreapaleolithicmobilityoccupation
+
+climate_pattern <- knitr::include_graphics(here::here("analysis/figures/figures-from-other-paper/007-tcsa-climate-model-sites-panel-plot.png"))
+
+population_pattern <- knitr::include_graphics(here::here("analysis/figures/figures-from-other-paper/008-summed-probability-distribution-models.png"))
+
+#convert images to png
+library(png)
+climate_pattern_img<-readPNG(climate_pattern)
+population_pattern_img<-readPNG(population_pattern)
+
+library(cowplot)
+plot_grid(tem_pattern, climate_pattern_img, population_pattern_img,
+          nrow= 1,
+          labels = "AUTO")
+```
 

### 2022-10-10 analysis/paper/paper.qmd
convert table to imported CSV

--- 2022-10-08 analysis/paper/paper.qmd
@@
 We used the results of summed probability distributions (SPD) and annual temperature changes from @park2022did which covers our target period and area.
-Following the TCSA ranges for different weapon-delivery systems that @lombard2021variation created by summarizing @wadley2008segment' study of backed microliths (Table 1?), we excluded 12 artifacts from our dataset.
+Following the TCSA ranges for different weapon-delivery systems that @lombard2021variation created by summarizing @wadley2008segment' study of backed microliths (@tbl-tcsa-ranges), we excluded 12 artifacts from our dataset.
 Because the TCSA values of those 12 artifacts from Bupyeongri, Hwadaeri, Haga, Sinhwari, Gokcheon, and Sachang sites are greater than 257.
@@
 
-| Weapon type               | N of tools | Mean TCSA | SD  | TCSA Range |
-|---------------------------|------------|-----------|-----|------------|
-| Poisoned arrowheads       | 434        | 11        | 7   | 4-18       |
-| Arroheads                 | 118        | 33        | 20  | 13-53      |
-| Dart tips                 | 40         | 58        | 18  | 40-76      |
-| All javelins of known use | 137        | 71        | 27  | 44-98      |
-| Thrusting spear tips      | 75         | 159       | 71  | 88-230     |
-
-: TCSA range from Lombard [-@lombard2021variation]
+```{r}
+#| label: tbl-tcsa-ranges
+#| tbl-cap: "TCSA range from Lombard (2021)"
+
+library(gt)
+read_csv(here::here("analysis/data/raw_data/tcsa-range-from-lombard.csv")) %>% 
+  gt() %>% 
+  tab_options(row.striping.include_table_body = FALSE) %>%
+  tab_style(
+    style = cell_borders(
+      sides = c("top"),
+      color = "black",
+      weight = px(3),
+      style = "solid"
+    ),
+    locations = cells_body(
+      columns = everything(),
+      rows = c(1)
+    )
+  ) %>% 
+  tab_options(
+    table.border.top.style = "hidden",
+    table.border.bottom.style = "hidden"
+  )
+```
 

### 2022-10-15 analysis/paper/paper.qmd
editing the abstract and intro

@@
 abstract: |
- The introduction of blade technology, stemmed points, end scrapers, burins, denticulates, and finer grained materials led to the transition from the Early to Late Paleolithic in Korea.
-Stemmed points have been considered a representative tool that led this whole set of changes.
-In this research, we examine the function of stemmed points to understand the role that they played during the technological transition as well as throughout the Late Paleolithic.
+ The introduction of blade technology, stemmed points, end scrapers, burins, denticulates, and finer grained materials characterises the transition from the Early to Late Paleolithic in Korea.
+Stemmed points have been considered a representative tool that led this set of changes.
+In this research, we examine the possible role that stemmed points played during the technological transition, as well as throughout the Late Paleolithic.
 Our main questions are: What were stemmed points used for?
-How diverse were their functions?
-What are the temporal patterns in stemmed point functions?
-We measured tip cross-sectional areas (TCSA) to discriminate different functional classes of projectile points, for example, poisoned arrowheads or thrusting spear.
-We analyze TCSA with other variables including raw materials, weight, site and radiocarbon dates.
-Our results show that the stemmed points mostly functioned as javelins and thrusting spear tips, with smaller numbers as dart tips and arrowheads.
-TCSA values are depending mostly on size rather than raw material types.
-We found different usage of stemmed points in different sites, which could indicate people used stemmed points in different ways depending on the environment.
-However, some sites show a wide range of TCSA values that represent multi-purpose usage of stemmed points.
+How diverse were their likely uses?
+What are the temporal patterns in stemmed point use?
+We measured tip cross-sectional area (TCSA) to discriminate different likely use classes of projectile points, for example, as poisoned arrowheads or as thrusting spears.
+We analyzed TCSA with other variables including raw materials, weight, site and radiocarbon dates.
+Our results show that the stemmed points were likely used  as javelins and thrusting spear tips, with smaller numbers as dart tips and arrowheads.
+TCSA values are controlled mostly by size rather than raw material types.
+We found different usage of stemmed points at different sites, which could indicate people used stemmed points in different ways depending on the local environment.
+Some sites show a wide range of TCSA values that represent multi-purpose usage of stemmed points.
 The temporal pattern of TCSA values is one of little change throughout the Late Paleolithic period.
-We conclude that stemmed points were mainly designed for Javelin and thrusting spear tips but they were multi-functional tools.
+We conclude that stemmed points were  multi-functional tools, with many likely designed for javelin and thrusting spear tips.
 keywords: |
@@
 
-With the introduction of new technologies, the Korean Paleolithic transitioned from the Early to the Late Paleolithic periods.
-The transition includes blade technology, stemmed points, end scrapers, burins, denticulates, etc [@bae2017origin; @nakazawa2018quaternary; @bae2017late; @lee2017korean; @seong2016eastern].
-Another notable change is the selective use of raw materials along with the emergence of new tools.
-Previously quartzite and vein quartz were the most commonly used for core and flake tools but finer grained materials such as silicified tuff (shale), chert, hornfels, and obsidian became more important to the lithic technology during the Late Paleolithic.
-While people still used coarse materials with existing tools, they selectively chose finer materials for newly introduced tools.
-[@Seong_2003; @seong2004quartzite].
-Stemmed points are considered the first evidence of the new technology, as well as the Late Paleolithic, since they led to the whole set of changes [@seong2016eastern; @seong2008tanged].
-This is related to the fact that stemmed points originate from Korea in Northeast Asia and have a close association with mobility, site formation, and occupation diversity [@o2018origins; @Chong_2021; @park2022did].
-Despite the importance of stemmed points, only a few studies have examined their use, whereas researchers often discuss stemmed points relating to their origin and chronology of the Korean Late Paleolithic and their relationship with Japan [@chang2013human; @Park_2013; @lee2019were; @Chong_2021].[@chang2013human; @Park_2013; @lee2019were; @Chong_2021].
+The introduction of new stone artefact technologies marks a major transition in the Korean Paleolithic from the Early to the Late Paleolithic periods.
+The transition includes the appearance blade technology, stemmed points, end scrapers, burins, denticulates, etc [@bae2017origin; @nakazawa2018quaternary; @bae2017late; @lee2017korean; @seong2016eastern].
+Stemmed points are considered the first evidence of a suite of new technologies defining the Late Paleolithic in this region [@seong2016eastern; @seong2008tanged].
+This is related to the fact that stemmed points originate from Korea and spread throughout Northeast Asia, and they have a close association with mobility, site formation, and occupation diversity [@o2018origins; @Chong_2021; @park2022did].
+Despite the importance of stemmed points, only a few studies to date have examined their likely uses.
+Previous work has mostly debated their origin, the chronology of the Korean Late Paleolithic, and their relationship with Japan [@chang2013human; @Park_2013; @lee2019were; @Chong_2021].[@chang2013human; @Park_2013; @lee2019were; @Chong_2021].
 

### 2022-10-15 analysis/paper/paper.qmd
finish editing introduction

@@
 
-To understand the role stemmed points played in the technological transition, this study examines the functions of stemmed points.
-We use the tip cross-sectional area (TCSA) metric to calculate artifact function and compare the results to other archaeological and ethnographic assemblages based on the idea that different shapes correspond to different functions [@lombard2021variation].
-We then explore the relationship between and function raw materials, artifact size and discard location, and how these change over time.
-Our main questions are: What were stemmed points used for?
-How diverse were their functions?
-What are the temporal patterns in stemmed point functions?
+To understand the likely roles that stemmed points played in the technological transition, this study examines the possible uses of stemmed points.
+We use the tip cross-sectional area (TCSA) metric to infer likely artifact use, and compare the results to other archaeological and ethnographic assemblages based on the idea that different shapes correspond to different weapon types [@lombard2021variation].
+We then explore the relationship between likely use and raw materials, artifact size and discard location, and how these change over time.
+Our main questions are: What was the best-fit ballistic probability for the stemmed points if they were hafted as weapon tips?
+How diverse were their likely uses?
+What are the temporal patterns in stemmed point uses?
 As part of our discussion, we examine possible connections between the roles of stemmed points and population dynamics or environmental change during the Late Paleolithic period.
-In order to understand how a certain weapon-tip type was chosen, we purposely borrow evolutionary perspective and premise that human groups had the cognitive capacity to choose proper weapons depending on their hunting strategies and socio-environmental circumstances.
+In order to understand how a certain weapon-tip type was chosen, we apply an evolutionary perspective with the assumption that people chose their weapon tip types depending on their hunting strategies and socio-environmental circumstances.
 

### 2022-10-15 analysis/paper/paper.qmd
finish editing background sections

@@
 
-## Why Stemmed Points?
+## Stemmed Points in Korea and East Asia
 
 Stemmed points (Sumbejjirugae) are projectile points made on an elongated blade-like flake or blade with two parallel facets and a single or two arises that converge to form an inverted "Y" [@pratt2020circum] (@fig-sp-hand-made).
-Slight retouch is located on the proximal end to shape an acute tip and on the distal end to make a stem, which connects to a wooden shaft.
-These types of artifacts are often called tanged points, but we prefer 'stemmed point' here to distinguish from Bronze Age stone projectile points that have long been called 'tanged points' in Korea [@park2022did].
-After the first discovery at the Suyanggae site, more than 450 stemmed points have been found in over 30 sites [@Chong_2021].
-The emergence of stemmed points is relevant to questions about the origins of projectile technology, the emergence of the Late Paleoltihic, and relationships to adjacent regions in East Asia, such as Japan.
+Slight retouch is typically located on the proximal end to shape an acute tip and on the distal end to make a stem, which connects to a wooden shaft.
+Elsewhere in the world these types of artifacts are often called tanged points, but we prefer 'stemmed point' here to distinguish from Bronze Age stone projectile points that have long been called 'tanged points' in Korea [@park2022did].
+After the first discovery at the Suyanggae site, more than 450 stemmed points have been found in over 30 sites throughout South Korea [@Chong_2021].
+Understanding the emergence of stemmed points is relevant to general questions about the origins of projectile technology, the emergence of the Late Palaeolithic, and relationships to adjacent regions in East Asia, such as Japan.
 
 The stemmed point is the first composite tool in the Korean Paleolithic that requires two different parts, consisting of a stone point and shaft, presumably made out of wood [@seong2008tanged].
-Using blades enables mass production for this composite tool and its shape becomes more standardized [@lee2015characteristics; @park2022did].
+Using blades as the blank for the point enables mass production for this composite tool and its shape can become more standardized [@lee2015characteristics; @park2022did].
 Therefore, @o2018origins claims that understanding the emergence of projectile technology gives insights to greater cultural, evolutionary, and behavioral cognitive flexibility.
 
-Since the stemmed points represent the beginning of the Korean Late Paleolithic, investigating their origin(s) becomes a key for understanding the technological transition from the Early Paleoltihic, modern human dispersals, and existence of the 'Middle' Paleolithic in Korea [@bae2010origin; @bae2017late; @norton2009evolution; @seong2016eastern].
-There has been a northworthy debate about explaining the origin(s) of stemmed points and it can be summarized into two competing models: 'heterogenic' migration [@bae2010origin], in situ evolution [@seong2009emergence].
+Since stemmed points represent the beginning of the Korean Late Paleolithic, investigating their origin(s) becomes a key for understanding the technological transition from the Early Palaeolithic, modern human dispersals, and claims for the existence of the 'Middle' Paleolithic in Korea [@bae2010origin; @bae2017late; @norton2009evolution; @seong2016eastern].
+There has been a noteworthy debate about explaining the origin(s) of stemmed points which can be summarized into two competing models: a 'heterogenic' migration [@bae2010origin], and an *in situ* evolution model [@seong2009emergence].
 The migration model claims that the new blade industry including stemmed points, and the earlier coarse flake tradition including large cores, polyhedrons, choppers, and even handaxes, came from different origins (e.g.
-1: North route: Siberia, Mongolia, or other regions of northeast China, and 2: South route: southern China) as the result of continuing influx of modern human migration from two routes.
-The 'in situ' model argues that stemmed points and other Late Paleolithic technologies, including blade industries, autonomously emerged in the south of the Korean peninsula, with no apparent external influence.
-The difference between the two models comes from how to understand a few early sites that contain stemmed points but no blades and blade coares.
-The 'in situ' model claims that the stemmed points appeared before blades by acknowledging those early sites including Bonggok, Songamri, Yonghodong, and Hwadaeri sites while migration model supports the traditional definition of the Late or Upper Palaeolithic in Europe and Africa that started with emergence and establishment of blade technology [@seong2016eastern].
-
-Stemmed points are important proxies to understand human behaviors during the Late Paleolithic.
-@park2022did examine mobility and site occupation patterns through lithic assemblages and concepts of human behavioral ecology and find that forager groups using stemmed points may have been associated with occupation of marginal or extreme environments compared to the groups with no stemmed points.
-Also, they were more associated with expedient technologies, indicating residential and less mobile behaviors.
-@Chong_2021 claims that the morphological variation of the stemmed points along with tool types in lithic assemblages, assemblage size, use of raw materials, and types of blanks could represent the characteristics of occupations such as "limited activity station" and "residential base camp." For example, stemmed points with high morphological variations in a tool size, shape of edge, degrees of damage, and types of edge retouching from Yongsandong site indicate that the site was used for specific or limited activities such as hunting [@kim2004yongsan; @seong2015diversity; @Chong_2021].
-
-The presence of stemmed points in both Korean and Japan has been studied since the late 1980s as a part of evidence for long-distance/maritime cultural interchanges or social networks [@lee2015characteristics].
+1: North route: Siberia, Mongolia, or other regions of northeast China, and 2: South route: southern China) as the result of a continuing influx of modern human migration from two routes.
+The *in situ* model argues that stemmed points and other Late Paleolithic technologies, including blade industries, autonomously emerged in the south of the Korean peninsula, with no apparent external influence.
+The difference between the two models comes from how to understand a few early sites that contain stemmed points but no blades and blade cores.
+The *in situ* model claims that stemmed points appeared before blades by acknowledging those early sites, including Bonggok, Songamri, Yonghodong, and Hwadaeri, while the migration model supports the traditional definition of the Late or Upper Palaeolithic in Europe and Africa that started with emergence and establishment of blade technology [@seong2016eastern].
+
+Stemmed points are important proxies to understand human behaviors during the Late Palaeolithic.
+@park2022did examined mobility and site occupation patterns by applying concepts of human behavioral ecology to lithic assemblages and found that forager groups using stemmed points may have been associated with the occupation of marginal or extreme environments, compared to groups with no stemmed points.
+Also, stemmed points were more frequently associated with expedient technologies, indicating residential and less mobile behaviors.
+@Chong_2021 claims that the morphological variation of the stemmed points along with tool types in lithic assemblages, assemblage size, use of raw materials, and types of blanks could represent specific characteristics of occupation such as a "limited activity station" and a "residential base camp." For example, stemmed points with high morphological variations in a tool size, shape of edge, degrees of damage, and types of edge retouching from Yongsandong site may indicate that the site was used for specific or limited activities such as hunting [@kim2004yongsan; @seong2015diversity; @Chong_2021].
+
+Stemmed points in both Korea and Japan have been studied since the late 1980s as a part of evidence for long-distance/maritime cultural interchanges or social networks [@lee2015characteristics].
 Stemmed points from the Bonggok site in Korea are currently accepted as the oldest ones among Northeast Asia, dated to ca.
@@
 In addition to the stemmed points, there are similar artifacts found in both regions such as microblade cores, Moppule-seokgi (Kakusuijyosekki in Japanese), backed knife, bilateral points, bifacial points, and transport of obsidian [@lee2012characteristics; @chang2013human; @lee2015characteristics; @kim2021evidence].
-In addition, stemmed points also help understand the connections and technological transitions between islands in Japan [@chang2013human].
+Stemmed points are thus important for understanding the connections and technological transitions between islands in Japan [@chang2013human].
 
@@
 
-## Previous studies about the stemmed points function
-
-Though the stemmed points are generally assumed as hunting armatures [@lee2019were; @lee2011study; @seong2008tanged; @lee2002new; @chang2013human], it is difficult to understand the function of the tools without knowing their complete shape attached to other components.
+## Uses of stemmed points
+
+Though stemmed points are generally assumed to have been hunting armatures [@lee2019were; @lee2011study; @seong2008tanged; @lee2002new; @chang2013human], it is difficult to understand the likely uses of the tools without knowing their complete shape when attached to other components.
 Preserved wooden components of projectile tools are too few and rare to generalize the overall scale and variability of the tools used in prehistory [@shea2006origins].
-@Lee_Kong_2002 claimed to name the stemmed points as 'stemmed tools' because of the diverse shape of stemmed points, which could represent different functions as well as the uncertainty of complete tool shape.
-Other researchers agree that some stemmed points that are not symmetrical and have retouch on one side of the artifacts or have denticulate blades on one side should not be called as stemmed points but stemmed knives, stemmed side-scrapers, stemmed end-scrapers, or stemmed burins [@seong2008tanged; @Lee_Jang_2011; @Kim] @fig-sp-hand-made: 7-8.
+@Lee_Kong_2002 claimed stemmed points were 'stemmed tools' because of their diverse shapes, which could represent different functions, as well as the uncertainty of complete tool shape.
+Other researchers agree that some stemmed points that are not symmetrical and have retouch on one side of the artifacts or have denticulate blades on one side should not be called stemmed points but instead stemmed knives, stemmed side-scrapers, stemmed end-scrapers, or stemmed burins [@seong2008tanged; @Lee_Jang_2011; @Kim] @fig-sp-hand-made: 7-8.
 
 Most stemmed points are symmetrical from the tip to the tang, having the central axis function as a line of symmetry [@Lee_Jang_2011].
-The studies of the stemmed points manufacturing process and the patterns of broken pieces show that stemmed points might be used mainly as spear tips [@Lee_1985; @Chang_2002].
+Studies of stemmed point manufacturing processes and the patterns of broken pieces show that stemmed points may have been used mainly as spear tips [@Lee_1985; @Chang_2002].
 There are a high proportion of broken tips and stems and the reused tools were repaired based on keeping the symmetry [@Park_2013; @Kim].
-In addition to the morphological aspect of stemmed points, investigating the whole site and tool composition of an assemblage show that stemmed points or stemmed tools are strongly associated with hunting activities including peeling the animal skin after slaughtering or separating the bones from the flesh [@seong2008tanged; @Chong_2021].
-
-A recent study conducts use-wear analysis on 95 stemmed points from the Jingeuneul site along with fracture patterns, TCSA, tip cross-sectional perimeter (TCSP), and neck width to understand the use of the tools [@lee2019were].
-Their results show that a considerable number of the stemmed points show diagnostic impact fractures (DIFs) on the surface, which could occur due to longitudinal force from the shaft when a stone tip hits a target.
-In other words, their research indicates that stemmed points could be used as spear-throwers or bows [@sano2015backed].
+In addition to the morphological aspect of stemmed points, investigation of the whole site and tool composition of an assemblage shows that stemmed points or stemmed tools are strongly associated with hunting activities including peeling the animal skin after slaughtering or separating the bones from the flesh [@seong2008tanged; @Chong_2021].
+
+A recent study conducted use-wear analysis on 95 stemmed points from the Jingeuneul site, along with examining fracture patterns, TCSA, tip cross-sectional perimeter (TCSP), and neck width to understand the use of the tools [@lee2019were].
+Their results show that a considerable number of the stemmed points show diagnostic impact fractures (DIFs) on the surface, which likely occurred due to longitudinal forces from the shaft when a stone tip hits a target.
+In other words, their research indicates that stemmed points may have been used as spear-throwers or bows [@sano2015backed].
 
@@
 
-The tip cross-sectional area (TCSA) has been used as a ballistically significant standard to discriminate different functional classes of projectile armatures such as spearthrower (a.k.a.
-atlatl) dart tips, arrowheads and large experimental stabbing/thrusting spears [@sisk2011african; @lombard2021variation; @lombard2020tip; @lombard2021did; @o2018origins].
-TCSA metric indicates the part of the tool that cuts the target's hide and relates to weapon flight and penetration dynamics [@hughes1998getting; @sitton2020tip; @lombard2021variation].
-This method was first proposed by @hughes1998getting and and tested by @shea2006origins through comparing archeological examples to ethnographically collected samples of known use.
-@shea2009impact also applied this approach to compare projectile points from Africa, the Levant, and Europe and claimed that projectile weapons first appeared in Africa.
-The main advantage of the TCSA is, regardless of point type, only maximum width and thickness measurements are needed for calculating the TCSA value (0.5 × maximum width × maximum thickness) [@sisk2011african; @lombard2020tip].
+Tip cross-sectional area (TCSA) of stone artefacts has been used as a ballistically significant standard to discriminate different functional classes of projectile armatures such as spearthrower (a.k.a.
+atlatl) dart tips, arrowheads and large stabbing/thrusting spears [@sisk2011african; @lombard2021variation; @lombard2020tip; @lombard2021did; @o2018origins].
+The TCSA metric indicates the part of the tool that cuts the target's hide and relates to weapon flight and penetration dynamics [@hughes1998getting; @sitton2020tip; @lombard2021variation].
+This method was first proposed by @hughes1998getting and and tested by @shea2006origins by comparing archaeological examples to ethnographically collected samples of known use.
+@shea2009impact also applied this approach to compare projectile points from Africa, the Levant, and Europe, and claimed that projectile weapons first appeared in Africa.
+A key advantage of the TCSA metric is its convenience: regardless of point type, only maximum width and thickness measurements are needed for calculating the TCSA value (0.5 × maximum width × maximum thickness) [@sisk2011african; @lombard2020tip].
 Later, @sisk2011african proposed an alternate measure, tip cross-sectional perimeter (TCSP), for more accurate measure for the force needed to penetrate a target to a lethal depth while TCSA metric is more associated with cutting aspect.
-However, TCSP has a few disadvantages to directly apply to our case of stemmed points.
-The force and penetration depth are not only affected by the stone tip but also by the mass of the shaft, which cannot be known for archaeological stone-tipped weapons [@lombard2020tip].
+However, TCSP has a few disadvantages that limit its applicability to our case study of Korean stemmed points.
+The force and penetration depth are not only affected by the stone tip, but also by the mass of the shaft, which cannot be known for most archaeological stone-tipped weapons [@lombard2020tip].
 @sisk2011 also mentioned that TCSP cannot be applied to backed pieces that are hafted as projectile armatures.
 
-@lee2019were first applied TCSA to stemmed points from Korea along with their use-wear analysis.
-The stemmed points for their research were discovered from Jingeuneul site located in southwest of Korea, where the largest number of stemmed points (n = 99) were found in a single site.
-For TCSA, they were only able to use ten stemmed points because they needed to select the stemmed points that retained the widest and thickest part of the specimens for TCSA metric and displayed diagnostic impact fractures for the use-wear analysis.
-The purpose of using TCSA in their research is to compare the values to North American dart tips and arrowheads.
-Their results show that the TCSA range of Jingeuneul stemmed points is relatively wide overlapped with both North American dart tips and arrowheads.
-Inspired by @lee2019were in addition to other prior research that developed and applied TCSA into archaeological materials from other parts of the world, we explore stemmed points from all over South Korea to better understand their function to understand the role that stemmed points represent during the Late Paleolithic.
+@lee2019were first applied TCSA to stemmed points from Korea along with use-wear analysis.
+In that study, stemmed points were analysed from Jingeuneul, located in southwest of Korea, which has the largest number of stemmed points (n = 99) ever found in a single site.
+For TCSA, they were only able to use ten stemmed points because they needed to select the stemmed points that retained the widest and thickest part of the specimens for TCSA metric and also displayed diagnostic impact fractures for the use-wear analysis.
+The purpose of using TCSA in their research was to compare the values to North American dart tips and arrowheads.
+Their results show that the TCSA range of Jingeuneul stemmed points is relatively wide, overlapping with both North American dart tips and arrowheads.
+Inspired by @lee2019were and TCSA research from lithic assemblages in other parts of the world, we aimed to investigate TCSA values for a large number of stemmed points from all over South Korea to better understand their likely uses during the Late Paleolithic.
 

### 2022-10-15 analysis/paper/paper.qmd
finish editing methods - archaeological sites

@@
 
-After the first evidence of the Korean Late Paleolithic discovered at Seokjangri site including blade and a stemmed base, around 300 stemmed points have been found in Korea [@Sohn_1967; @lee2019were].
+Since the first evidence of the Korean Late Paleolithic was discovered at Seokjangri site, including blades and a stemmed base, around 300 stemmed points have been found in Korea [@Sohn_1967; @lee2019were].
 Most sites contain only a few stemmed points and only a few sites have many more, such as Jingeuneul, Suyanggae (n = 55), and Yongsandong (n = 38) [@Kim_2017].
-Among these stemmed points, we gathered those that retained the widest and thickest part of specimens.
-Our dataset contains stemmed points that were found during field surveys as well as those were from sites that were never dated but accompanied with other Late Paleoltihic artifacts.
-Those criteria resulted in a sample of 173 stemmed points from 36 assemblages unearthed from 29 sites spanning the period 44-10 ka (@fig-map).
-The dimensions of the 173 stemmed points were obtained from published excavation reports and by direct measurements during our visits to the collections of Korean local museums and archaeological institutions.
-
-We distinguished multiple assemblages in one site where numerous excavations have been conducted in different spots and by different institutions under the same site name.
-For example, Suyanggae site, one of the Korean National Heritage sites, has been excavated more than ten times since 1980 by the local university museum and later by archaeology research institutions.
-There are six different excavation spots apart from each other from a few meters to a few kilometers.
-The archaeologists identified multiple assemblages in each excavation or even a trench by culturally sterile deposits, or where distinct artifact-bearing stratigraphic units could be identified by major differences in the texture, color, and composition of the sedimentary deposits.
-In this research, we used four assemblages from the Suyanggae site and we separated one assemblage among the four by using a different site name, Hajinri.
+Among these stemmed points, we selected those that retained their widest and thickest parts.
+Our dataset contains stemmed points that were found during field surveys as well as those were from sites that were never dated but were associtaed with other Late Palaeoltihic artifacts.
+Applying these sampling criteria resulted in a sample of 173 stemmed points from 36 assemblages unearthed from 29 sites spanning the period 44-10 ka (@fig-map).
+The dimensions of the 173 stemmed points were obtained from published excavation reports and by direct measurements during our visits to the collections of local museums and archaeological institutions in Korea.
+
+We distinguished between multiple assemblages at a single site where numerous excavations have been conducted in different locations at the site, and by different institutions under the same site name.
+For example, Suyanggae site, a registered Korean National Heritage site, has been excavated more than ten times since 1980 by the local university museum and later by archaeological research institutions.
+There are six different excavation locations that range from a few meters apart to a few kilometers.
+Similarly, we identified multiple assemblages in a single excavation or even a trench where archaeological deposits were separated by culturally sterile deposits, or where distinct artifact-bearing stratigraphic units could be identified by major differences in the texture, color, and composition of the sedimentary deposits.
+As a result, in this research we used four assemblages from the Suyanggae site.
+We separated one assemblage among the four by using a different site name, Hajinri, following the convention established the excavators of that location.
 Because Hajinri was from the sixth excavation area, which is 3.5 kilometers apart from the other spots and dated to much earlier (around 42-30 ka) than the other assemblages (around 31-15 ka).
 According to the excavation reports for Hajinri, stemmed points first appeared in Hajinri around 42 ka with the earliest blade and blade cores [@Lee_Woo_Lee_An_Yun_Park_Otani_Kim_Kim_Han_et_al._2018].
-We consider this assemblage as an unusual outlier because the stemmed points and blades from Hajinri are highly standardized and refined, which are only found at other sites much later in time.
-We consider Bonggok site, which has the second-earliest dates (41.5 ka) in our collection, is the actual earliest site.
-Bonggok site includes blades or elongate flakes but it is hard to define without accompanying blade core [@park2022did].
+We consider this assemblage as an unusual outlier because the stemmed points and blades from Hajinri are so early and also so highly standardized and refined, which are only found at other sites much later in time.
+We have more confidence in finds from Bonggok, which has the second-earliest dates (41.5 ka) in our collection, as the first appearance of stemmed points.
+Bonggok includes blades or elongate flakes, but without any accompanying blade cores [@park2022did].
 

### 2022-10-15 analysis/paper/paper.qmd
adjust position of inset map

@@
   draw_plot(plot = worldplot,
-            x = 0.43, # x location of inset placement
-            y = -0.03, # y location of inset placement
-            width = .4, # Inset width
-            height = .35, # Inset height
+            x = 0.37, # x location of inset placement
+            y = -0.23, # y location of inset placement
+            width = .7, # Inset width
+            height = .7, # Inset height
             scale = 0.5 # Inset scale

### 2022-10-16 analysis/paper/paper.qmd
add osf address

--- 2022-10-15 analysis/paper/paper.qmd
@@
 
-The entire R code [@core2021team] and data files used for all the analysis and visualizations contained in this paper is openly available at https://doi-org/xxxx to enable re-use of materials and improve reproducibility and transparency [@marwick2017computational].
+The entire R code [@core2021team] and data files used for all the analysis and visualizations contained in this paper is openly available at https://osf.io/dqna8/ to enable re-use of materials and improve reproducibility and transparency [@marwick2017computational].
 All of the figures, tables, statistical test results presented here in addition to the raw data can be independently reproduced with the code and data in this repository.

### 2022-10-16 analysis/paper/paper.qmd
finish editing methods section

--- 2022-10-15 analysis/paper/paper.qmd
@@
 
-Tip cross-sectional area (TCSA) of stone artefacts has been used as a ballistically significant standard to discriminate different functional classes of projectile armatures such as spearthrower (a.k.a.
+Tip cross-sectional area (TCSA) of stone artefacts has been used as a ballistically relevant standard to probabalistically discriminate different between functional classes of projectile armatures such as spearthrower (a.k.a.
 atlatl) dart tips, arrowheads and large stabbing/thrusting spears [@sisk2011african; @lombard2021variation; @lombard2020tip; @lombard2021did; @o2018origins].
+It is important to note that the TCSA metric alone cannot unambiguously determine artefact function, it only suggests a best-fit ballistic probability for the points if they were hafted as weapon tips.
 The TCSA metric indicates the part of the tool that cuts the target's hide and relates to weapon flight and penetration dynamics [@hughes1998getting; @sitton2020tip; @lombard2021variation].
@@
 @shea2009impact also applied this approach to compare projectile points from Africa, the Levant, and Europe, and claimed that projectile weapons first appeared in Africa.
-A key advantage of the TCSA metric is its convenience: regardless of point type, only maximum width and thickness measurements are needed for calculating the TCSA value (0.5 × maximum width × maximum thickness) [@sisk2011african; @lombard2020tip].
+A key advantage of the TCSA metric is its convenience: regardless of point type, only maximum width and thickness measurements are needed for calculating the TCSA value (0.5 x maximum width x maximum thickness) [@sisk2011african; @lombard2020tip].
 Later, @sisk2011african proposed an alternate measure, tip cross-sectional perimeter (TCSP), for more accurate measure for the force needed to penetrate a target to a lethal depth while TCSA metric is more associated with cutting aspect.
@@
 
-## Applying TCSA
-
-To answer our research questions, What were stemmed points used for?
-How diverse were their functions?
-What are the temporal patterns in stemmed point functions?
-Are these temporal patterns associated with population dynamics or environmental processes?
-We calculated TCSA values for stemmed points and explored the usage for the stemmed points depending on raw materials, location, and size.
-We applied radiocarbon dates to the results to examine the temporal patterns.
-In the discussion, we compared those temporal patterns with demographic and environmental proxies.
-We used the results of summed probability distributions (SPD) and annual temperature changes from @park2022did which covers our target period and area.
-Following the TCSA ranges for different weapon-delivery systems that @lombard2021variation created by summarizing @wadley2008segment' study of backed microliths (@tbl-tcsa-ranges), we excluded 12 artifacts from our dataset.
-Because the TCSA values of those 12 artifacts from Bupyeongri, Hwadaeri, Haga, Sinhwari, Gokcheon, and Sachang sites are greater than 257.
-Their TCSA range is from 263 to 2622.
-Therefore, a total of 161 stemmed points from 32 assemblages unearthed from 26 sites are used for the following analysis.
+## Investigating patterns in TCSA values
+
+To answer our research questions about the likely uses of stemmed points, we calculated TCSA values for the stemmed points in our sample, and explored the interaction of TCSA values with raw materials, site location, and artefact size.
+To examine the temporal patterns in TCSA values we investigated radiocarbon ages associated with the points.
+We then compared those temporal patterns with demographic and environmental proxies based on summed probability distributions (SPD) and annual temperature changes from @park2022did.
+We followed the TCSA ranges for different weapon-delivery systems that @lombard2021variation created by summarizing @wadley2008segment' study of backed microliths (@tbl-tcsa-ranges).
+We excluded 12 artifacts from our dataset with TCSA greater than 257 (263-2622) and thus outside of the range of the comparative data.
+For this study we include TCSA values from a total of 161 stemmed points from 32 assemblages unearthed from 26 sites.
 
@@
 
-Selective use of raw material is one of the main technological transitions to the Korea Late Paleolithic.
-Previously used quartzite and vein quartz were continuously selected for the existing core and flake tools and finer grained materials were newly added to assemblages for producing the newly introduced tools [@seong2004quartzite].
-We analyze TCSA values per each raw material and examine why types of raw materials were selected for certain types of tool.
-
-### Regional Pattern
-
-We present TCSA values per site that contains more than 4 stemmed points (more than and equal to 5?) based on the evolutionary concept that people choose their tools according to the surrounding environment and available resources.
-Through the TCSA distribution, we explore which function(s) the stemmed points had performed at the sites.
+Selective use of raw material is a key characteristic of the technological transition of the Korean Late Paleolithic.
+Prior to the Late Paleolithic people used quartzite and vein quartz for core and flake tools.
+Then, finer grained materials were newly added to assemblages for producing the newly introduced tools [@seong2004quartzite].
+We analyzed TCSA values for each raw material type to examine the interaction of raw material types and TCSA values.
+
+### Geographical Patterns
+
+To explore geographical patterns in TCSA values, we retained only sites that contains more than four stemmed points.
+This resulted in a sample of XXX sites and XXX points.
+We summarised the distribution of TCSA values for each site to identify variation that might be influenced b environmental differences relating to the different locations of the sites.
 
@@
 
-Different sizes of stone artifacts can represent different functions.
+Different sizes of stone artifacts can constrain or enable different functions.
 Overall size of stone points has been used as a potential proxy to separate different armature types [@thomas1978arrowheads; @sahle2019assessment].
-We chose weight as a size indicator for the stemmed points.
-The weight is a reliable discriminator between tools and a descriptive attribute which can be determined objectively and which serves as a criterion of likeness or dissimilarity in projectile point description [@fenenga1953weights; @shea2006origins].
-Excluding 9 points that we were not able to measure or get the records of weight, we calculate TCSA for 152 artifacts.
+We chose weight as a proxy for overall size of the stemmed points.
+Weight is a reliable discriminator between tools of different sizes and a descriptive attribute which can be determined rapidly and objectively and serve as a criterion of likeness or dissimilarity in projectile point description [@fenenga1953weights; @shea2006origins].
+Excluding nine points that we were not able to measure or obtain the records of their weight, we explored the relationship between TCSA and weight for 152 artifacts.
 
@@
 
-We add radiocarbon dates to explore the temporal pattern of the function of stemmed points.
-We excluded the assemblages that have no radiocarbon dates.
-We arrange 24 assemblages dated from 45ka to 14.8ka periodically, and compute TCSA.
-
-## Two models for weapon-tip types choice
-
-The introduction of new types reflect major implications for social transformation.
-For example, the bow and arrow enabled individuals to work without hunting teams and provided opportunities for non-elite hunters to produce their own subsistence or pursue individual wealth [@bettinger2013effects; @angelbeck2014faustian; @rorabaugh2015timing].
-In evolutionary terms, we assume that human groups carefully selected a specific stone tool based on its functional advantage over other alternatives once they acquired the necessary physiological and cognitive abilities to innovate [@lombard2022tip].
-The choice of weapon-tip types, therefore, reflects the socio-environmental circumstances that they had to encounter and overcome.
-@eren2022plains claim that Clovis points were more variable in shape than Folsom points because Clovis foragers were exposed to largely unfamiliar landscape and employed their points as multifunctional tools that perform a wider range of tasks including cutting and sawing.
-Folsom points show a narrow range of variation, indicating a specific task.
-They used TCSA to assess the morphological variance of the points as well as understand the potential functional costs or benefits of one artifact variant relative to another.
-
-Similar to @eren2022plains's approach, we hypothesize two scenarios that explain the regional or/and temporal pattern of TCSA range.
-If stemmed points represent a narrow TCSA range, people produced tools that performed certain tasks.
-There could be less surprise in the surrounding environment.
-On the other hand, a wide TCSA range indicates that stemmed points could be multifunctional tools and people might need to be prepared for unexpected situations such as moving into an unfamiliar landscape or their patch becoming less productive.
-We expect to observe various TCSA ranges, which reflect morphologically and functionally different stemmed points, across different regions or environments.
-However, we acknowledge that lithic morphology is highly dependent on the quality of raw materials and their availability.
+We used XXX radiocarbon ages to explore the temporal pattern of the function of stemmed points.
+We excluded assemblages that have no radiocarbon dates.
+We arranged 24 assemblages dated from 45ka to 14.8ka to explore variation in TCSA over time.
+
+## Modelling weapon-tip type selection
+
+The process of the introduction of new technologies can vary depending on the environmental and social contexts of the transmission of the techniques for making the new tools, and the new tools themselves can alter the typical contexts of tool use.
+For example, the introduction of the bow and arrow in American enabled individuals to work without hunting in teams and provided opportunities for non-elite hunters to produce their own subsistence or pursue individual wealth [@bettinger2013effects; @angelbeck2014faustian; @rorabaugh2015timing].
+Taking an evolutionary approach, we assume that, given an opportunity to explore alternative technologies, human groups carefully selected a specific stone tool technology based on its advantages over other alternatives, according to their performance in a variety of domains, such as physical and social functions [@lombard2022tip].
+Thus, the selection of weapon-tip types is likely to reflect the socio-environmental circumstances that people encountered and managed.
+For example, @eren2022plains used TCSA to compare the morphological variance of Clovis and Folsom points and claimed that Clovis points were more variable in shape than Folsom points because Clovis foragers were exposed to largely unfamiliar landscape.
+Clovis points were used as multifunctional tools that performed a wider range of tasks, including cutting and sawing.
+Folsom points show a narrower range of variation, indicating a they were more likely used for specific tasks.
+
+Inspired by @eren2022plains's approach, we hypothesized two basic scenarios that might explain the regional or/and temporal pattern of TCSA range.
+If stemmed points represent a narrow TCSA range, then people likely produced tools that performed a small set tasks.
+This may be due to low levels of uncertainty in the forager's social and physical environments.
+On the other hand, a wide TCSA range may indicate that stemmed points were multifunctional tools, suggesting that people were responding to unfamiliar situations, such as moving into an unfamiliar landscape or unpredictable changes in patch productivity and travel times \[CITE PAPERS ON ARCHAEOLOGY AND PATCH CHOICE MODEL\].
+We expect to observe different TCSA ranges, reflecting morphologically and functionally different stemmed points, across different regions or environments in Korea, and over time.
 
@@
 
-## Variation by site (Regional pattern)
+## Geographical Patterns
 

### 2022-10-16 analysis/paper/paper.qmd
add DOI

--- 2022-10-16 analysis/paper/paper.qmd
@@
 
-The entire R code [@core2021team] and data files used for all the analysis and visualizations contained in this paper is openly available at https://osf.io/dqna8/ to enable re-use of materials and improve reproducibility and transparency [@marwick2017computational].
+The entire R code [@core2021team] and data files used for all the analysis and visualizations contained in this paper is openly available at DOI: 10.17605/OSF.IO/DQNA8 to enable re-use of materials and improve reproducibility and transparency [@marwick2017computational].
 All of the figures, tables, statistical test results presented here in addition to the raw data can be independently reproduced with the code and data in this repository.

### 2022-10-16 analysis/paper/paper.qmd

--- 2022-10-16 analysis/paper/paper.qmd
@@
 
-The entire R code [@core2021team] and data files used for all the analysis and visualizations contained in this paper is openly available at https://doi-org/xxxx to enable re-use of materials and improve reproducibility and transparency [@marwick2017computational].
+The entire R code [@core2021team] and data files used for all the analysis and visualizations contained in this paper is openly available at DOI: 10.17605/OSF.IO/DQNA8 to enable re-use of materials and improve reproducibility and transparency [@marwick2017computational].
 All of the figures, tables, statistical test results presented here in addition to the raw data can be independently reproduced with the code and data in this repository.

### 2022-10-16 analysis/paper/paper.qmd
finish editing results section

--- 2022-10-16 analysis/paper/paper.qmd
@@
 
+```{r}
+# ADD CODE TO REPLACE NULL WITH ACTUAL VALUES
+n_sites <- NULL # replace this with the number
+min_n_points <- NULL # replace this with the number
+n_radiocarbon_ages <- NULL # replace this with the number
+n_sites_with_radiocarbon_ages <- NULL # replace this with the number
+
+```
+
 To explore geographical patterns in TCSA values, we retained only sites that contains more than four stemmed points.
-This resulted in a sample of XXX sites and XXX points.
+This resulted in a sample of `r n_sites` sites and `r min_n_points` points.
 We summarised the distribution of TCSA values for each site to identify variation that might be influenced b environmental differences relating to the different locations of the sites.
@@
 
-We used XXX radiocarbon ages to explore the temporal pattern of the function of stemmed points.
+We used `r n_radiocarbon_ages` radiocarbon ages to explore the temporal pattern of the function of stemmed points.
 We excluded assemblages that have no radiocarbon dates.
-We arranged 24 assemblages dated from 45ka to 14.8ka to explore variation in TCSA over time.
+We arranged `r n_sites_with_radiocarbon_ages` assemblages dated from 45ka to 14.8ka to explore variation in TCSA over time.
 
@@
 #| fig-cap: "Distribution of TCSA values for all Korean stemmed points in the current dataset. Shaded boxes show TSCA ranges for various weapon types."
-#| fig-width:   4.5
-#| fig-height:  4.4
+#| fig-width:   3.5
+#| fig-height:  2.5
 
@@
 
-To gain a general idea of the role of stemmed points, we calculated TCSA for all stemmed points.
-@fig-tcsa-all-sp shows the mean TCSA is `r round(mean(tcsa$min_tcsa), digits=1)`, and standard deviation of TCSA is `r round(sd(tcsa$min_tcsa), digits=1)`.
+@fig-tcsa-all-sp shows TCSA values for all stemmed points in our sample, with shaded rectangles to assist in the interpretation of their likely function.
+The ranges indicated by the shaded rectangles were taken from @lombard2021variation.
 Overall we see a wide variation of TCSA values in this boxplot.
-According to @lombard2021variation TCSA ranges (Table 1), Korean stemmed points are mostly in the categories of javelins and thrusting spear tips, with smaller numbers as dart tips and arrowheads.
+The sample mean TCSA is `r round(mean(tcsa$min_tcsa), digits=1)`, and standard deviation of TCSA is `r round(sd(tcsa$min_tcsa), digits=1)`.
+According to the TCSA ranges presented in Table 1 of @lombard2021variation, Korean stemmed points are overall mostly in the categories of javelins and thrusting spear tips, with smaller numbers as dart tips and arrowheads.
 Among the weapon-tip types, only poison arrowheads appear to be absent from these Korean assemblages.
@@
 #| fig-width:   4.5
-#| fig-height:  6
+#| fig-height:  4
 
@@
 
-In order to explore the relationship between raw material and the function of the stemmed points, we calculated TCSA per raw material type.
 @fig-tcsa-raw-materials shows that about half of the stemmed points were made from shale and its TCSA range is widely distributed.
@@
 
-## Geographical Patterns
+## Geographical patterns
 
@@
 #| fig-width:   4.5
-#| fig-height:  6
+#| fig-height:  4.5
 
@@
 
-We computed TCSA for sites that contain more than 5 stemmed points to observe the pattern.
-As shown in @fig-tcsa-sites, the sites with fewer stemmed points are grouped under the category of 'other'.
-Among the selected sites, Suyanggae has the most stemmed points and the widest range of TCSA.
-We can assume that people made stemmed points for a variety of different purposes in Suyanggae.
-The TCSA values for Nosanri and Sibuk are skewed, which indicates less diverse functions for stemmed points, and lower.
-Wolpyeng and Yongsujaeul have similar amounts of stemmed points to Nosanri and Sibuk, but exhibit different distribution of TCSA values.
+We computed artefact TCSA values for `r n_sites` sites that contain more than `r min_n_points` stemmed points to observe variation between sites (@fig-tcsa-sites) Sites with fewer than `r min_n_points` stemmed points were grouped under the category of 'other'.
+Among the sites in our sample, Suyanggae has the most stemmed points and the widest range of TCSA values.
+This suggests that people made stemmed points for a variety of different purposes in Suyanggae.
+The TCSA values for Nosanri and Sibuk are have a narrower range of lower values, which indicates less diverse functions for stemmed points.
+Wolpyeng and Yongsujaeul have similar amounts of stemmed points to Nosanri and Sibuk, but exhibit wider distributions of TCSA values.
 Yongsandong has the second most stemmed points and TCSA shows a similar distribution to Yongsakul.
 
-## Variation by size (weight)
-
-```{r}
-#| label: fig-tcsa-size
-#| fig-cap: "Artifact size and TSCA values by lithic raw material"
-#| fig-width:   4.5
-#| fig-height:  4
+## Variation by artefact size and raw material
+
+```{r}
+
+# Facet by size class
+# https://cran.r-project.org/web/packages/Ckmeans.1d.dp/vignettes/Ckmeans.1d.dp.html
+library(Ckmeans.1d.dp)
 
@@
 
-ggplot(tcsa_mass_no_na,
-       aes(`Weight(g)`,
-           min_tcsa,
-           colour = raw_materials_top)) +
-  geom_point() +
-    labs(x = "Weight", 
-       y = "TCSA",
-       colour='Raw material') +
-  geom_smooth(method=lm, se=FALSE, size = 0.4)+
-  theme_bw(base_size = 6) +
-  theme(
-  legend.title = element_text(size = 8),
-  legend.text = element_text(size = 7),
-  legend.position = c(.8,.3)
-  )
-```
-
-We chose weight as a size proxy and examined the relationship between size and function of the stemmed points.
-We combined the raw material information as another variable for the analysis.
-@fig-tcsa-size shows a positive relationship between TCSA and artifact weight.
-The relationship varies by raw material types, perhaps due to variation in the density of different lithic raw materials.
-There is a stronger correlation between TCSA and artifact weight for acidic volcanic points (i.e.
-closer to the regression line), whereas the correlation for shale stemmed points is less pronounced.
-The other raw materials show different distribution patterns around the regression line, reconfirming that raw material is less influential (as @fig-tcsa-raw-materials).
+# find clusters
+result <- Ckmeans.1d.dp(tcsa_mass_no_na$`Weight(g)`)
+
+tcsa_mass_no_na <-
+tcsa_mass_no_na %>%
+  mutate(mass_cluster = factor(result$cluster))
+
+```
+
+Using weight as a size proxy we examined the relationship between size and likely use of the stemmed points inferred from TCSA values.
+We conducted a univariate cluster analysis INSERT CITATION TO PAPER ON THIS METHOD of stemmed points by weight and it revealed three clusters of artefact sizes (A of @fig-artifact-size-cluster, mean = `r round(mean(tcsa_mass_no_na$'Weight(g)'), digits=1)`, SD = `r round(sd(tcsa_mass_no_na$'Weight(g)'), digits=1)`).
+Cluster 1, the smallest (lightest) artifacts, have a lower range of TCSA values, compared to Cluster 2 (B of @fig-artifact-size-cluster).
+TCSA values for Cluster 3 are the highest, except for one artifact, which is lower than 50 (A of @fig-artifact-size-cluster, mean = `r round(mean(tcsa_mass_no_na$'min_tcsa'), digits=1)`, SD = `r round(sd(tcsa_mass_no_na$'min_tcsa'), digits=1)`).
+There are a few overlaps, but we can still assume that artifacts in different size clusters might be made for different purposes.
 
@@
 #| fig-width:   4.5
-#| fig-height:  4
-
-# Facet by size class
-# https://cran.r-project.org/web/packages/Ckmeans.1d.dp/vignettes/Ckmeans.1d.dp.html
-library(Ckmeans.1d.dp)
-
-# find clusters
-result <- Ckmeans.1d.dp(tcsa_mass_no_na$`Weight(g)`)
-
-tcsa_mass_no_na <-
-tcsa_mass_no_na %>%
-  mutate(mass_cluster = factor(result$cluster))
+#| fig-height:  3
 
@@
 
-We then conducted univariate cluster analysis of stemmed points by weight and it revealed three groups of clusters (A of @fig-artifact-size-cluster, mean = `r round(mean(tcsa_mass_no_na$'Weight(g)'), digits=1)`, SD = `r round(sd(tcsa_mass_no_na$'Weight(g)'), digits=1)`).
-Cluster 1, the smallest (lightest) artifacts, is the lower TCSA, compared to Cluster 2.
-TCSA of Cluster 3 is the highest, except for one artifact, which is lower than 50 (A of @fig-artifact-size-cluster, mean = `r round(mean(tcsa_mass_no_na$'min_tcsa'), digits=1)`, SD = `r round(sd(tcsa_mass_no_na$'min_tcsa'), digits=1)`).
-There are a few overlaps, but we can still assume that artifacts in different clusters might be made for different purposes.
-
-## Temporal pattern of TCSA
+We combined the raw material information as another variable for the analysis.
+@fig-tcsa-size shows a positive relationship between TCSA and artifact weight.
+The relationship varies by raw material types, perhaps due to variation in the density of different lithic raw materials.
+There is a stronger correlation between TCSA and artifact weight for acidic volcanic points (i.e.
+points are closer to the regression line), whereas the correlation for shale stemmed points is weaker.
+The other raw materials show varied distribution patterns around the regression line, reconfirming that raw material appears not to have strongly influenced TCSA (see also @fig-tcsa-raw-materials).
+
+```{r}
+#| label: fig-tcsa-size
+#| fig-cap: "Artifact size and TSCA values by lithic raw material"
+#| fig-width:   3.5
+#| fig-height:  3
+
+ggplot(tcsa_mass_no_na,
+       aes(`Weight(g)`,
+           min_tcsa,
+           colour = raw_materials_top)) +
+  geom_point() +
+    labs(x = "Weight (g)", 
+       y = "TCSA",
+       colour='Raw material') +
+  geom_smooth(method=lm, se=FALSE, size = 0.4)+
+  theme_bw(base_size = 6) +
+  theme(
+  legend.title = element_text(size = 5),
+  legend.text = element_text(size = 4),
+  legend.position = c(.8,.3)
+  )
+```
+
+## Temporal pattern of TCSA values
 
@@
 #| label: fig-tcsa-radio-carbon
-#| fig-cap: "Distribution of TCSA values by time"
+#| fig-cap: "Distribution of TCSA values over time"
 #| fig-width:   4.5
-#| fig-height:  4
+#| fig-height:  2.5
 
@@
 
- ggplot(tcsa_site_ass, aes(x = min_tcsa, y = reorder(label_site, age_ka), color = label_site)) +
-   geom_point() +
+ggplot(tcsa_site_ass, 
+       aes(y = min_tcsa,
+           x = reorder(label_site, -age_ka))) +
+   geom_boxplot() +
    theme_minimal() +
    theme(legend.position="none") +
-   labs(x = "TCSA", 
-       y = "Assemblage")
-
-```
-
-Using radiocarbon dates, we arranged 24 assemblages in chronological order (@fig-tcsa-radio-carbon).
+   labs(y = "TCSA", 
+        x = "") +
+  theme(axis.text.x = element_text(size = 4)) +
+  scale_x_discrete(guide = guide_axis(angle = 45))
+
+```
+
+We arranged `r n_sites_with_radiocarbon_ages` assemblages in chronological order (@fig-tcsa-radio-carbon).
 Our analysis shows that early assemblages (i.e.
-45-24 ka) tend to possess numerous stemmed points such as two from the Suyanggae site and they have a wide range of TCSA, indicating multiple usages of stemmed points.
-Overall, TCSA from assemblages that have more than 3 stemmed points vary considerably, which makes it difficult to define a single function of the artifacts.
-There is also no clear temporal pattern in assemblages with a fewer number of artifacts.
+45-24 ka) tend to possess more stemmed points, such as Suyanggae, and these have a wide range of TCSA values, indicating multiple likely usages of stemmed points.
+After about 38 ka stemmed points become less common in archaeological assemblages.
+Overall, TCSA values from assemblages that have more than `r min_n_points` stemmed points vary considerably, which makes it difficult to discern a clear temporal pattern in these assemblages.
 
@@
 
-To investigate the role that stemmed points played in the technological transition during the Korean Late Paleolithic, we applied TCSA metric to estimate artifact function by comparing the results to other archaeological and ethnographic cases.
-We asked three questions to achieve our research goal: What were stemmed points used for?
+To investigate the role that stemmed points played in the technological transition during the Korean Late Paleolithic, we computed the TCSA metric to infer likely artifact uses by comparing the results to other archaeological and ethnographic studies We asked three questions to achieve our research goal: What were stemmed points used for?
 How diverse were their functions?
@@
 #| fig-width:   4.5
-#| fig-height:  12
+#| fig-height:  4
 

### 2022-10-16 analysis/paper/paper.qmd
replace tcsa range csv to updated range

--- 2022-10-16 analysis/paper/paper.qmd
@@
 We then compared those temporal patterns with demographic and environmental proxies based on summed probability distributions (SPD) and annual temperature changes from @park2022did.
-We followed the TCSA ranges for different weapon-delivery systems that @lombard2021variation created by summarizing @wadley2008segment' study of backed microliths (@tbl-tcsa-ranges).
+We followed the TCSA ranges for different weapon-delivery systems that @lombard2022tip and @lombard2021variation created by summarizing @wadley2008segment' study of backed microliths (@tbl-tcsa-ranges).
 We excluded 12 artifacts from our dataset with TCSA greater than 257 (263-2622) and thus outside of the range of the comparative data.
@@
 #| label: tbl-tcsa-ranges
-#| tbl-cap: "TCSA range from Lombard (2021)"
+#| tbl-cap: TCSA range from Lombard (2021; 2022)
 
 library(gt)
-read_csv(here::here("analysis/data/raw_data/tcsa-range-from-lombard.csv")) %>% 
+read_csv(here::here("analysis/data/raw_data/tcsa-range-from-lombard-2022.csv")) %>% 
   gt() %>% 

### 2022-10-16 analysis/paper/paper.qmd
udpated tcsa range

@@
 We then compared those temporal patterns with demographic and environmental proxies based on summed probability distributions (SPD) and annual temperature changes from @park2022did.
-We followed the TCSA ranges for different weapon-delivery systems that @lombard2022tip and @lombard2021variation created by summarizing @wadley2008segment' study of backed microliths (@tbl-tcsa-ranges).
+We followed the TCSA ranges for different weapon-delivery systems that @lombard2022tip and @lombard2021variation created by summarizing @wadley2008segment' study of backed microliths and calibrating existing ranges (@tbl-tcsa-ranges).
 We excluded 12 artifacts from our dataset with TCSA greater than 257 (263-2622) and thus outside of the range of the comparative data.
@@
 #| label: tbl-tcsa-ranges
-#| tbl-cap: TCSA range from Lombard (2021; 2022)
+#| tbl-cap: TCSA range from Lombard (2021 & 2022)
 

### 2022-10-16 analysis/paper/paper.qmd
updated doi address

@@
 
-The entire R code [@core2021team] and data files used for all the analysis and visualizations contained in this paper is openly available at DOI: 10.17605/OSF.IO/DQNA8 to enable re-use of materials and improve reproducibility and transparency [@marwick2017computational].
+The entire R code [@core2021team] and data files used for all the analysis and visualizations contained in this paper is openly available at https://doi.org/10.17605/osf.io/dqna8 to enable re-use of materials and improve reproducibility and transparency [@marwick2017computational].
 All of the figures, tables, statistical test results presented here in addition to the raw data can be independently reproduced with the code and data in this repository.

### 2022-10-16 analysis/paper/paper.qmd
updated insert map size

@@
   draw_plot(plot = worldplot,
-            x = 0.37, # x location of inset placement
-            y = -0.23, # y location of inset placement
-            width = .7, # Inset width
-            height = .7, # Inset height
+            x = 0.45, # x location of inset placement
+            y = -0.27, # y location of inset placement
+            width = .6, # Inset width
+            height = .75, # Inset height
             scale = 0.5 # Inset scale

### 2022-10-16 analysis/paper/paper.qmd
finish edits on discussion and conclusion, reorder methods and results sections

--- 2022-10-16 analysis/paper/paper.qmd
@@
       institute: University of Johannesburg
+  - Donghee Chong:
+      institute: Kyung Hee University
   - Ben Marwick:
       institute: University of Washington
-  - Donghee Chong:
-      institute: Kyung Hee University
 format: 
@@
 We then explore the relationship between likely use and raw materials, artifact size and discard location, and how these change over time.
-Our main questions are: What was the best-fit ballistic probability for the stemmed points if they were hafted as weapon tips?
+Our main questions are: What were the best-fit ballistic probabilities for the stemmed points if they were hafted as weapon tips?
 How diverse were their likely uses?
@@
 We then compared those temporal patterns with demographic and environmental proxies based on summed probability distributions (SPD) and annual temperature changes from @park2022did.
-We followed the TCSA ranges for different weapon-delivery systems that @lombard2022tip and @lombard2021variation created by summarizing @wadley2008segment' study of backed microliths and calibrating existing ranges (@tbl-tcsa-ranges).
+We followed the TCSA ranges for different weapon-delivery systems that @lombard2021variation created by summarizing @wadley2008segment' study of backed microliths (@tbl-tcsa-ranges).
 We excluded 12 artifacts from our dataset with TCSA greater than 257 (263-2622) and thus outside of the range of the comparative data.
@@
 #| label: tbl-tcsa-ranges
-#| tbl-cap: TCSA range from Lombard (2021 & 2022)
+#| tbl-cap: "TCSA range from Lombard (2021)"
 
 library(gt)
-read_csv(here::here("analysis/data/raw_data/tcsa-range-from-lombard-2022.csv")) %>% 
+read_csv(here::here("analysis/data/raw_data/tcsa-range-from-lombard.csv")) %>% 
   gt() %>% 
@@
 
+### Weight
+
+Different sizes of stone artifacts can constrain or enable different functions.
+Overall size of stone points has been used as a potential proxy to separate different armature types [@thomas1978arrowheads; @sahle2019assessment].
+We chose weight as a proxy for overall size of the stemmed points.
+Weight is a reliable discriminator between tools of different sizes and a descriptive attribute which can be determined rapidly and objectively and serve as a criterion of likeness or dissimilarity in projectile point description [@fenenga1953weights; @shea2006origins].
+Excluding nine points that we were not able to measure or obtain the records of their weight, we explored the relationship between TCSA and weight for 152 artifacts.
+
 ### Geographical Patterns
@@
 We summarised the distribution of TCSA values for each site to identify variation that might be influenced b environmental differences relating to the different locations of the sites.
-
-### Weight
-
-Different sizes of stone artifacts can constrain or enable different functions.
-Overall size of stone points has been used as a potential proxy to separate different armature types [@thomas1978arrowheads; @sahle2019assessment].
-We chose weight as a proxy for overall size of the stemmed points.
-Weight is a reliable discriminator between tools of different sizes and a descriptive attribute which can be determined rapidly and objectively and serve as a criterion of likeness or dissimilarity in projectile point description [@fenenga1953weights; @shea2006origins].
-Excluding nine points that we were not able to measure or obtain the records of their weight, we explored the relationship between TCSA and weight for 152 artifacts.
+I THINK WE NEED SOMETHING MORE IN HERE.
+LIKE NEAREST NEIGHBOUR DISTANCE OR DISTANCE TO THE COAST OR ELEVATION OR SOME GEOGRAPHICAL VARIABLE THAT IS RELEVANT TO OUR MODEL OF SINGLE-FUNCTION VS MULTI-FUNCTION.
 
@@
 
-## Variation by raw material
+## Variation in TCSA values by raw material
 
@@
 
-## Geographical patterns
-
-```{r}
-#| label: fig-tcsa-sites
-#| fig-cap: "Histograms of TCSA values by archaeological site."
-#| fig-width:   4.5
-#| fig-height:  4.5
-
- ggplot(tcsa_site,
-              aes(min_tcsa, 
-                  fill = n)) +
-  geom_histogram() +
-  theme_bw(base_size = 6) +
-  labs(x = "TCSA", 
-       y = "Count of artefacts") +
-  facet_wrap(~label_site,
-                 scales = "free_y",
-                  ncol = 1) +
-    scale_fill_viridis_c(name = "Number\nof artefacts") +
-  annotate("rect", 
-           xmin = 4,
-           xmax = 18,
-           ymin = 0,
-           ymax = 2.5,
-           fill = "green",
-           alpha = 0.2) +
-      annotate("text", 
-               size = 1.5,
-               label = "Poisoned\narrowheads",
-               hjust = 0,
-               lineheight = 0.75,
-           y = (0+2.5)/2,
-           x = 5.5) +
-      annotate("rect", 
-           xmin = 13,
-           xmax = 55,
-           ymin = 2.5,
-           ymax = 5,
-           fill = "blue",
-           alpha = 0.2) +
-      annotate("text", 
-               size = point_function_group_label_size,
-               label = "Arrowheads",
-               hjust = 0,
-           y = (2.5+5)/2,
-           x = 28) +
-      annotate("rect", 
-           xmin = 40,
-           xmax = 76,
-           ymin = 5,
-           ymax = 7.5,
-           fill = "orange",
-           alpha = 0.2) +
-      annotate("text", 
-               size = point_function_group_label_size,
-               label = "Dart tips",
-               hjust = 0,
-           y = (5+7.5)/2,
-           x = 53) +
-      annotate("rect", 
-           xmin = 44,
-           xmax = 98,
-           ymin = 7.5,
-           ymax = 10,
-           fill = "purple",
-           alpha = 0.2) +
-      annotate("text", 
-               size = point_function_group_label_size,
-               label = "Javelin",
-               hjust = 0,
-           y = (7.5+10)/2,
-           x = 70) +
-     annotate("rect", 
-           xmin = 88,
-           xmax = 230,
-           ymin = 10,
-           ymax = 12.5,
-           fill = "yellow",
-           alpha = 0.2) +
-      annotate("text", 
-               label = "Thrusting spear",
-               size = point_function_group_label_size,
-               hjust = 0,
-           y = (10+12.5)/2,
-           x = 160)
-```
-
-We computed artefact TCSA values for `r n_sites` sites that contain more than `r min_n_points` stemmed points to observe variation between sites (@fig-tcsa-sites) Sites with fewer than `r min_n_points` stemmed points were grouped under the category of 'other'.
-Among the sites in our sample, Suyanggae has the most stemmed points and the widest range of TCSA values.
-This suggests that people made stemmed points for a variety of different purposes in Suyanggae.
-The TCSA values for Nosanri and Sibuk are have a narrower range of lower values, which indicates less diverse functions for stemmed points.
-Wolpyeng and Yongsujaeul have similar amounts of stemmed points to Nosanri and Sibuk, but exhibit wider distributions of TCSA values.
-Yongsandong has the second most stemmed points and TCSA shows a similar distribution to Yongsakul.
-
-## Variation by artefact size and raw material
+## Variation in TCSA values by artefact size
 
@@
 
-We combined the raw material information as another variable for the analysis.
-@fig-tcsa-size shows a positive relationship between TCSA and artifact weight.
+@fig-tcsa-size shows a positive relationship between TCSA, artifact weight and raw material.
 The relationship varies by raw material types, perhaps due to variation in the density of different lithic raw materials.
@@
 The other raw materials show varied distribution patterns around the regression line, reconfirming that raw material appears not to have strongly influenced TCSA (see also @fig-tcsa-raw-materials).
+
+## Geographical patterns of TCSA values
+
+```{r}
+#| label: fig-tcsa-sites
+#| fig-cap: "Histograms of TCSA values by archaeological site."
+#| fig-width:   4.5
+#| fig-height:  4.5
+
+ ggplot(tcsa_site,
+              aes(min_tcsa, 
+                  fill = n)) +
+  geom_histogram() +
+  theme_bw(base_size = 6) +
+  labs(x = "TCSA", 
+       y = "Count of artefacts") +
+  facet_wrap(~label_site,
+                 scales = "free_y",
+                  ncol = 1) +
+    scale_fill_viridis_c(name = "Number\nof artefacts") +
+  annotate("rect", 
+           xmin = 4,
+           xmax = 18,
+           ymin = 0,
+           ymax = 2.5,
+           fill = "green",
+           alpha = 0.2) +
+      annotate("text", 
+               size = 1.5,
+               label = "Poisoned\narrowheads",
+               hjust = 0,
+               lineheight = 0.75,
+           y = (0+2.5)/2,
+           x = 5.5) +
+      annotate("rect", 
+           xmin = 13,
+           xmax = 55,
+           ymin = 2.5,
+           ymax = 5,
+           fill = "blue",
+           alpha = 0.2) +
+      annotate("text", 
+               size = point_function_group_label_size,
+               label = "Arrowheads",
+               hjust = 0,
+           y = (2.5+5)/2,
+           x = 28) +
+      annotate("rect", 
+           xmin = 40,
+           xmax = 76,
+           ymin = 5,
+           ymax = 7.5,
+           fill = "orange",
+           alpha = 0.2) +
+      annotate("text", 
+               size = point_function_group_label_size,
+               label = "Dart tips",
+               hjust = 0,
+           y = (5+7.5)/2,
+           x = 53) +
+      annotate("rect", 
+           xmin = 44,
+           xmax = 98,
+           ymin = 7.5,
+           ymax = 10,
+           fill = "purple",
+           alpha = 0.2) +
+      annotate("text", 
+               size = point_function_group_label_size,
+               label = "Javelin",
+               hjust = 0,
+           y = (7.5+10)/2,
+           x = 70) +
+     annotate("rect", 
+           xmin = 88,
+           xmax = 230,
+           ymin = 10,
+           ymax = 12.5,
+           fill = "yellow",
+           alpha = 0.2) +
+      annotate("text", 
+               label = "Thrusting spear",
+               size = point_function_group_label_size,
+               hjust = 0,
+           y = (10+12.5)/2,
+           x = 160)
+```
+
+We computed artefact TCSA values for `r n_sites` sites that contain more than `r min_n_points` stemmed points to observe variation between sites (@fig-tcsa-sites) Sites with fewer than `r min_n_points` stemmed points were grouped under the category of 'other'.
+Among the sites in our sample, Suyanggae has the most stemmed points and the widest range of TCSA values.
+This suggests that people made stemmed points for a variety of different purposes in Suyanggae.
+The TCSA values for Nosanri and Sibuk are have a narrower range of lower values, which indicates less diverse functions for stemmed points.
+Wolpyeng and Yongsujaeul have similar amounts of stemmed points to Nosanri and Sibuk, but exhibit wider distributions of TCSA values.
+Yongsandong has the second most stemmed points and TCSA shows a similar distribution to Yongsakul.
 
@@
 
-## Temporal pattern of TCSA values
+## Temporal patterns of TCSA values
 
@@
 
-# Discussion
-
-To investigate the role that stemmed points played in the technological transition during the Korean Late Paleolithic, we computed the TCSA metric to infer likely artifact uses by comparing the results to other archaeological and ethnographic studies We asked three questions to achieve our research goal: What were stemmed points used for?
-How diverse were their functions?
-What are the temporal patterns in stemmed point functions?
-In our assumptions, human groups were capable of intentionally choosing and designing tools according to their socio-environmental circumstances.
-We came up with two models: (1) people created stemmed points and used them as a specific type of tool; or (2) they created stemmed points for multifunctional tools.
-
-Our results indicate that javelins and thrusting spear tips are the best matches for stemmed points.
-There are a few stemmed points that can be used as dart tips and arrowheads, but none of them are suitable for poison arrowheads (@fig-tcsa-all-sp).
-We found stemmed points played a diverse role, by showing a wide range of TCSA values.
-Some sites, such as Nosanri and Sinbuk, used stemmed points as limited-function tools.
-Meanwhile, stemmed points were used in multiple ways at other sites, especially Suyanggae.
+# Discussion ORDER OF TOPICS HERE MUST FOLLOW EXACTLY THE ORDER OF THE METHODS AND RESULTS SECTION
+
+To investigate the role that stemmed points played in the technological transition during the Korean Late Paleolithic, we computed the TCSA metric to infer likely artifact uses by comparing the results to other archaeological and ethnographic studies Our main questions were: What were the best-fit ballistic probabilities for the stemmed points if they were hafted as weapon tips?
+How diverse were their likely uses?
+What are the temporal patterns in stemmed point uses?
+
+Our results indicate that javelins and thrusting spear tips are the most probable ballistic uses for stemmed points.
+There are a few stemmed points in our sample that may have been used as dart tips and arrowheads, but none of them have TCSA values in the range of poison arrowheads (@fig-tcsa-all-sp).
+We found a wide range of TCSA values in our sample, indicating that stemmed points played a diverse role in forager toolkits.
+Some sites, such as Nosanri and Sinbuk, likely used stemmed points as restricted-function tools.
+Meanwhile, stemmed points were used for a wider set of tasks at other sites, especially Suyanggae.
 In general, stemmed points played multiple roles in more sites (@fig-tcsa-sites).
-Through the Late Paleolithic period, TCSA values hardly changed (@fig-tcsa-radio-carbon).
+Through the Late Paleolithic period, TCSA values showed little directional change (@fig-tcsa-radio-carbon).
 The chronological sequence of TCSA varies from region to region with no clear pattern.
-Therefore, we conclude that stemmed points were mainly designed for Javelin and thrusting spear tips but they were multi-functional tools.
-
-We are aware of the concern that discriminating the function of small projectile points could be arbitrary work [@erlandson2014darts].
+Most stemmed points appear to be have designed for use as javelin and thrusting spear tips, but there are also large numbers of tools that likely could fulfil multiple roles as armatures.
+
+TCSA can, however, be impacted by other factors such as raw materials and portability.
+-@eren2022plains explain that smaller TCSA could be the result of pursuing production economy and transport efficiency.
+We explore the relationship between TCSA and raw material types and weight in order to examine these factors.
+Weight plays an important role in portability and penetration of projectiles [@hughes1998getting].
+According to our results, raw materials do not have a clear pattern, but weight is correlated with raw materials (@fig-tcsa-raw-materials and @fig-tcsa-size).
+Different clusters of weight are matched with different types of weapon tips (@fig-artifact-size-cluster).
+Based on these analyses, we assume that people were able to produce the ideal size tools regardless of raw materials.
+
+Our models premise that the choice of weapon-tip types reflects the socio-environmental circumstances.
+To examine the possible connections between the roles of stemmed points and population dynamics or environmental change during the Late Paleolithic period, we compared the temporal pattern of TCSA to results of Summed probability distribution (SPD) and simulated Mean Annual Temperature (MAT) from @park2022did.
+In those two analyses, -@park2022did covered assemblages with and without stemmed points, which is useful for us to explore the overall socio-environmental dynamics here.
+As we previously noted, there is no clear pattern in the temporal or regional pattern of TCSA (@fig-tcsa-radio-carbon).
+THIS DISCUSSION IS INCOMPLETE, WE NEED TO EXPLORE THESE RELATIONSHIPS ABOUT GEOGRAPHICAL VARIATION AND TEMPORAL VARIATION FURTHER, AND IN A NEW WAY THAT IS NOT THE SAME AS THE QSR PAPER, E.G.
+FOCUS ON THE LGM AS I MENTIONED IN OUR LAST MEETING.
+
+# Conclusion
+
+RESTATE QUESTIONS AND ANSWERS VERY BRIELFY
+
+SUGGEST SOME POSSIBLE DIRECTIONS FOR FUTURE WORK, E.G. EXPERIMENTAL ARCHAEOLOGY, USEWEAR ANALYSIS
+
+We are aware of the arbitrary that discriminating the function of small projectile points could be arbitrary work [@erlandson2014darts].
 Because TCSA covers the critical elements of projectiles, flight and penetration dynamics (i.e.
@@
 
-TCSA can, however, be impacted by other factors such as raw materials and portability.
--@eren2022plains explain that smaller TCSA could be the result of pursuing production economy and transport efficiency.
-We explore the relationship between TCSA and raw material types and weight in order to examine these factors.
-Weight plays an important role in portability and penetration of projectiles [@hughes1998getting].
-According to our results, raw materials do not have a clear pattern, but weight is strongly correlated with raw materials (@fig-tcsa-raw-materials and @fig-tcsa-size).
-Different clusters of weight are matched with different types of weapon tips (@fig-artifact-size-cluster).
-Based on these analyses, we assume that people were able to produce the ideal size tools regardless of raw materials.
-
-Our models premise that the choice of weapon-tip types reflects the socio-environmental circumstances.
-To examine the possible connections between the roles of stemmed points and population dynamics or environmental change during the Late Paleolithic period, we compared the temporal pattern of TCSA to results of Summed probability distribution (SPD) and simulated Mean Annual Temperature (MAT) from @park2022did.
-In those two analyses, -@park2022did covered assemblages with and without stemmed points, which is useful for us to explore the overall socio-environmental dynamics.
-As we discussed, there is no clear pattern in the temporal or regional pattern of TCSA (@fig-tcsa-radio-carbon or @fig-socio-climate: A).
-We consider that TCSA, or the function of the stemmed points, are strongly dependent on the surrounding environment or their situation.
-@fig-socio-climate: B shows that the temperature slowly decreased until around 22 ka and then went up again with a little range of fluctuations.
-Temperatures differ significantly between sites, with a maximum difference of 5℃.
-Results of SPD show several peaks, as indicated by the positive deviations from the three null models (shaded in red), and downs (shaded in blue), but no drastic growth or decline (@fig-socio-climate: C).
-From the climate and population study, we could observe that there is no dominating change but minor and often fluctuations throughout the Late Paleolithic and variations per site.
-Overall this pattern is well matched with the TCSA results, indicating the people produced stemmed points as multifunctional tools suitable for unexpected or various situations in their habitats similar to the case of Clovis points (@eren2022plains).
-
-```{r}
-#| label: fig-socio-climate
-#| fig-cap: "A.Distribution of TCSA values by time.
-B.Mean Annual Temperature (MAT) of the Korean Paleolithic assemblages from 50 ka to 10 ka.
-The gray lines indicate the MAT for each site and the black line is the mean temperature of all sites.
-The light gray area in the middle of the plot indicates the duration of MIS 2, and the dark gray area represents the duration of the LGM (from @park2022did: Figure 9).
-C.
-Summed probability distribution (SPD) of 100 radiocarbon dates.
-The black solid line represents actual radiocarbon ages.
-The gray shaded region shows the Monte Carlo envelope that encompasses the 95% confidence interval for the null models.
-The red and blue vertical bands highlight the portions of the SPD where positive and negative deviations are detected.
-The best fit model has the lowest AIC score, which is the linear model in this case (from @park2022did: Figure 10)."
-#| fig-width:   4.5
-#| fig-height:  4
-
-tem_pattern <- ggplot(tcsa_site_ass, aes(x = min_tcsa, y = reorder(label_site, age_ka))) +
-   geom_point(color="gray", size =1.5) +
-   theme_minimal() +
-   theme(legend.position="none") +
-   labs(x = "TCSA", 
-       y = "Assemblage")
-
-# imput images from the Park&Marwick 2022 paper
-# https://github.com/parkgayoung/koreapaleolithicmobilityoccupation
-
-climate_pattern <- knitr::include_graphics(here::here("analysis/figures/figures-from-other-paper/007-tcsa-climate-model-sites-panel-plot.png"))
-
-population_pattern <- knitr::include_graphics(here::here("analysis/figures/figures-from-other-paper/008-summed-probability-distribution-models.png"))
-
-#convert images to png
-library(png)
-climate_pattern_img<-readPNG(climate_pattern)
-population_pattern_img<-readPNG(population_pattern)
-
-library(cowplot)
-plot_grid(tem_pattern, climate_pattern_img, population_pattern_img,
-          nrow= 1,
-          labels = "AUTO")
-```
-
-# Conclusion
-
 # Acknowledgements
@@
 
+Donghee Chong: Data curation, Review & Editing
+
 Ben Marwick: Software, Validation, Formal analysis, Investigation, Writing - Review & Editing, Visualization, Supervision
-
-Donghee Chong: Data curation, Review & Editing
 

### 2022-10-16 analysis/paper/paper.qmd

@@
 We then compared those temporal patterns with demographic and environmental proxies based on summed probability distributions (SPD) and annual temperature changes from @park2022did.
-We followed the TCSA ranges for different weapon-delivery systems that @lombard2021variation created by summarizing @wadley2008segment' study of backed microliths (@tbl-tcsa-ranges).
+We followed the TCSA ranges for different weapon-delivery systems that @lombard2022tip and @lombard2021variation created by summarizing @wadley2008segment' study of backed microliths and calibrating existing ranges (@tbl-tcsa-ranges).
 We excluded 12 artifacts from our dataset with TCSA greater than 257 (263-2622) and thus outside of the range of the comparative data.
@@
 #| label: tbl-tcsa-ranges
-#| tbl-cap: "TCSA range from Lombard (2021)"
+#| tbl-cap: TCSA range from Lombard (2021 & 2022)
 
 library(gt)
-read_csv(here::here("analysis/data/raw_data/tcsa-range-from-lombard.csv")) %>% 
+read_csv(here::here("analysis/data/raw_data/tcsa-range-from-lombard-2022.csv")) %>% 
   gt() %>% 

### 2022-10-16 analysis/paper/paper.qmd
got rid of the plot in discussion and related files

--- 2022-10-16 analysis/paper/paper.qmd
@@
             min_tcsa)) +
-  # show regions for tools from Table 3 of Lomabrd 2021
+  # show regions for tools from Lomabrd 2021 & 2022
   # https://www-sciencedirect-com.offcampus.lib.washington.edu/science/article/pii/S0277379121002286
+  # https://assets.researchsquare.com/files/rs-2016964/v1/00d381a3-3ecf-4d5b-b2ec-89a41227bbd5.pdf?c=1662475267
     annotate("rect", 
@@
                size = point_function_group_label_size,
-               label = "Poisoned\narrowheads",
-               hjust = 0,
+               label = "Poisoned\narrow tips",
+               hjust = 0,
+               lineheight = 0.75,
            x = point_function_group_label_y,
@@
            xmax = 0.3,
-           ymin = 13,
-           ymax = 53,
+           ymin = 17,
+           ymax = 47,
            fill = "blue",
@@
                size = point_function_group_label_size,
-               label = "Arrowheads",
-               hjust = 0,
+               label = "Un-poisoned\narrow tips",
+               hjust = 0,
+               lineheight = 0.75,
            x = point_function_group_label_y,
-           y = (53-13) /2 + 13) +
+           y = (47-17) /2 + 17) +
       annotate("rect", 
@@
            xmax = 0.5,
-           ymin = 44,
-           ymax = 98,
+           ymin = 42,
+           ymax = 90,
            fill = "purple",
@@
                size = point_function_group_label_size,
-               label = "Javelin",
+               label = "Javelin tips",
                hjust = 0,
            x = point_function_group_label_y,
-           y = (98-44) /2 + 44 ) +
+           y = (90-42) /2 + 42 ) +
      annotate("rect", 
@@
            xmax = 0.6,
-           ymin = 88,
-           ymax = 230,
+           ymin = 80,
+           ymax = 200,
            fill = "yellow",
@@
       annotate("text", 
-               label = "Thrusting\nspear",
-               size = point_function_group_label_size,
-               hjust = 0,
+               label = "Stabbing\nspear tips",
+               size = point_function_group_label_size,
+               hjust = 0,
+               lineheight = 0.75,
            x = point_function_group_label_y,
-           y = (230-88) /2 + 88 ) +
+           y = (200-80) /2 + 80 ) +
   geom_boxplot(alpha = 0.1) +
@@
                size = 1.5,
-               label = "Poisoned\narrowheads",
+               label = "Poisoned\narrow tips",
                hjust = 0,
@@
       annotate("rect", 
-           xmin = 13,
-           xmax = 55,
+           xmin = 17,
+           xmax = 47,
            ymin = 2.5,
@@
                size = point_function_group_label_size,
-               label = "Arrowheads",
-               hjust = 0,
+               label = "Un-poisoned\narrow tips",
+               hjust = 0,
+               lineheight = 0.75,
            y = (2.5+5)/2,
@@
       annotate("rect", 
-           xmin = 44,
-           xmax = 98,
+           xmin = 42,
+           xmax = 90,
            ymin = 7.5,
@@
                size = point_function_group_label_size,
-               label = "Javelin",
+               label = "Javelin tips",
                hjust = 0,
@@
      annotate("rect", 
-           xmin = 88,
-           xmax = 230,
+           xmin = 80,
+           xmax = 200,
            ymin = 10,
@@
       annotate("text", 
-               label = "Thrusting spear",
+               label = "Stabbing spear tips",
                size = point_function_group_label_size,
@@
            x = point_function_group_label_y,
-           y = (76-40) /2 + 40 ) +
+           y = (76-40) /2 + 40) +
       annotate("rect", 
@@
                size = point_function_group_label_size,
-               label = "Javelin",
+               label = "Javelin tips",
                hjust = 0,
            x = point_function_group_label_y,
-           y = (98-44) /2 + 44 ) +
+           y = (90-42) /2 + 42) +
      annotate("rect", 
@@
       annotate("text", 
-               label = "Thrusting\nspear",
-               size = point_function_group_label_size,
-               hjust = 0,
+               label = "Stabbing\nspear tips",
+               size = point_function_group_label_size,
+               hjust = 0,
+               lineheight = 0.75,
            x = point_function_group_label_y,
-           y = (230-88) /2 + 88 ) +
+           y = (200-80) /2 + 80 ) +
   coord_cartesian(xlim = c(-0.1,1.2)) +
@@
 In those two analyses, -@park2022did covered assemblages with and without stemmed points, which is useful for us to explore the overall socio-environmental dynamics.
-As we discussed, there is no clear pattern in the temporal or regional pattern of TCSA (@fig-tcsa-radio-carbon or @fig-socio-climate: A).
+As we discussed, there is no clear pattern in the temporal or regional pattern of TCSA (@fig-tcsa-radio-carbon or @no-figure-no-more: A).
 We consider that TCSA, or the function of the stemmed points, are strongly dependent on the surrounding environment or their situation.
-@fig-socio-climate: B shows that the temperature slowly decreased until around 22 ka and then went up again with a little range of fluctuations.
+@no-figure-no-more: B shows that the temperature slowly decreased until around 22 ka and then went up again with a little range of fluctuations.
 Temperatures differ significantly between sites, with a maximum difference of 5℃.
-Results of SPD show several peaks, as indicated by the positive deviations from the three null models (shaded in red), and downs (shaded in blue), but no drastic growth or decline (@fig-socio-climate: C).
+Results of SPD show several peaks, as indicated by the positive deviations from the three null models (shaded in red), and downs (shaded in blue), but no drastic growth or decline (@no-figure-no-more: C).
 From the climate and population study, we could observe that there is no dominating change but minor and often fluctuations throughout the Late Paleolithic and variations per site.
 Overall this pattern is well matched with the TCSA results, indicating the people produced stemmed points as multifunctional tools suitable for unexpected or various situations in their habitats similar to the case of Clovis points (@eren2022plains).
-
-```{r}
-#| label: fig-socio-climate
-#| fig-cap: "A.Distribution of TCSA values by time.
-B.Mean Annual Temperature (MAT) of the Korean Paleolithic assemblages from 50 ka to 10 ka.
-The gray lines indicate the MAT for each site and the black line is the mean temperature of all sites.
-The light gray area in the middle of the plot indicates the duration of MIS 2, and the dark gray area represents the duration of the LGM (from @park2022did: Figure 9).
-C.
-Summed probability distribution (SPD) of 100 radiocarbon dates.
-The black solid line represents actual radiocarbon ages.
-The gray shaded region shows the Monte Carlo envelope that encompasses the 95% confidence interval for the null models.
-The red and blue vertical bands highlight the portions of the SPD where positive and negative deviations are detected.
-The best fit model has the lowest AIC score, which is the linear model in this case (from @park2022did: Figure 10)."
-#| fig-width:   4.5
-#| fig-height:  4
-
-tem_pattern <- ggplot(tcsa_site_ass, aes(x = min_tcsa, y = reorder(label_site, age_ka))) +
-   geom_point(color="gray", size =1.5) +
-   theme_minimal() +
-   theme(legend.position="none") +
-   labs(x = "TCSA", 
-       y = "Assemblage")
-
-# imput images from the Park&Marwick 2022 paper
-# https://github.com/parkgayoung/koreapaleolithicmobilityoccupation
-
-climate_pattern <- knitr::include_graphics(here::here("analysis/figures/figures-from-other-paper/007-tcsa-climate-model-sites-panel-plot.png"))
-
-population_pattern <- knitr::include_graphics(here::here("analysis/figures/figures-from-other-paper/008-summed-probability-distribution-models.png"))
-
-#convert images to png
-library(png)
-climate_pattern_img<-readPNG(climate_pattern)
-population_pattern_img<-readPNG(population_pattern)
-
-library(cowplot)
-plot_grid(tem_pattern, climate_pattern_img, population_pattern_img,
-          nrow= 1,
-          labels = "AUTO")
-```
 

### 2022-10-17 analysis/paper/paper.qmd
refine plot sizes for the plots we are settled on, please don't change anything before the discussion section!

--- 2022-10-16 analysis/paper/paper.qmd
@@
 
-Stemmed points (Sumbejjirugae) are projectile points made on an elongated blade-like flake or blade with two parallel facets and a single or two arises that converge to form an inverted "Y" [@pratt2020circum] (@fig-sp-hand-made).
+Stemmed points (Sumbejjirugae in Korean) are projectile points made on an elongated blade-like flake or blade with two parallel facets and a single or two arises that converge to form an inverted "Y" [@pratt2020circum] (@fig-sp-hand-made).
 Slight retouch is typically located on the proximal end to shape an acute tip and on the distal end to make a stem, which connects to a wooden shaft.
@@
 
-The stemmed point is the first composite tool in the Korean Paleolithic that requires two different parts, consisting of a stone point and shaft, presumably made out of wood [@seong2008tanged].
+Stemmed points are the first composite tools in the Korean Paleolithic that required two different parts, consisting of a stone point and shaft, presumably made out of wood [@seong2008tanged].
 Using blades as the blank for the point enables mass production for this composite tool and its shape can become more standardized [@lee2015characteristics; @park2022did].
@@
 
-Since stemmed points represent the beginning of the Korean Late Paleolithic, investigating their origin(s) becomes a key for understanding the technological transition from the Early Palaeolithic, modern human dispersals, and claims for the existence of the 'Middle' Paleolithic in Korea [@bae2010origin; @bae2017late; @norton2009evolution; @seong2016eastern].
+Since stemmed points represent the beginning of the Korean Late Paleolithic, investigating their origins is important for understanding the technological transition from the Early Palaeolithic, modern human dispersals into the region, and claims for the existence of the 'Middle' Paleolithic in Korea [@bae2010origin; @bae2017late; @norton2009evolution; @seong2016eastern].
 There has been a noteworthy debate about explaining the origin(s) of stemmed points which can be summarized into two competing models: a 'heterogenic' migration [@bae2010origin], and an *in situ* evolution model [@seong2009emergence].
@@
 #| fig-cap: "Distribution of TCSA values for all Korean stemmed points in the current dataset. Shaded boxes show TSCA ranges for various weapon types."
-#| fig-width:   3.5
-#| fig-height:  2.5
+#| fig-width:   3
+#| fig-height:  2
 
@@
 point_function_group_label_y <- 0.75
-point_function_group_label_size <- 2
+point_function_group_label_size <- 1
 
@@
       annotate("text", 
-               label = "Thrusting\nspear",
-               size = point_function_group_label_size,
-               hjust = 0,
+               label = "Stabbing\nspear tips",
+               size = point_function_group_label_size,
+               hjust = 0,
+               lineheight = 0.75,
            x = point_function_group_label_y,
-           y = (230-88) /2 + 88 ) +
+           y = (200-80) /2 + 80 ) +
   geom_boxplot(alpha = 0.1) +
   geom_quasirandom(alpha = 0.4,
-                   size = 2.3) +
+                   size = 1.5) +
   coord_cartesian(xlim = c(-0.1,1)) +
@@
       annotate("text", 
-               size = 1.5,
-               label = "Poisoned\narrowheads",
+               size = point_function_group_label_size,
+               label = "Poisoned\narrow tips",
                hjust = 0,
@@
 
-## Variation in TCSA values by artefact size
-
-```{r}
-
-# Facet by size class
-# https://cran.r-project.org/web/packages/Ckmeans.1d.dp/vignettes/Ckmeans.1d.dp.html
-library(Ckmeans.1d.dp)
-
-# remove NAs
-tcsa_mass_no_na <-
-  tcsa %>%
- drop_na(`Weight(g)`)
-
-# find clusters
-result <- Ckmeans.1d.dp(tcsa_mass_no_na$`Weight(g)`)
-
-tcsa_mass_no_na <-
-tcsa_mass_no_na %>%
-  mutate(mass_cluster = factor(result$cluster))
-
-```
-
-Using weight as a size proxy we examined the relationship between size and likely use of the stemmed points inferred from TCSA values.
-We conducted a univariate cluster analysis INSERT CITATION TO PAPER ON THIS METHOD of stemmed points by weight and it revealed three clusters of artefact sizes (A of @fig-artifact-size-cluster, mean = `r round(mean(tcsa_mass_no_na$'Weight(g)'), digits=1)`, SD = `r round(sd(tcsa_mass_no_na$'Weight(g)'), digits=1)`).
-Cluster 1, the smallest (lightest) artifacts, have a lower range of TCSA values, compared to Cluster 2 (B of @fig-artifact-size-cluster).
-TCSA values for Cluster 3 are the highest, except for one artifact, which is lower than 50 (A of @fig-artifact-size-cluster, mean = `r round(mean(tcsa_mass_no_na$'min_tcsa'), digits=1)`, SD = `r round(sd(tcsa_mass_no_na$'min_tcsa'), digits=1)`).
-There are a few overlaps, but we can still assume that artifacts in different size clusters might be made for different purposes.
-
-```{r}
-#| label: fig-artifact-size-cluster
-#| fig-cap: "A. Distribution of artefact weight showing three clusters. B. TCSA values for all artefacts with size classes indicated by data point shape."
-#| fig-width:   4.5
-#| fig-height:  3
-
-plot_mass_cluster <- 
-ggplot (tcsa_mass_no_na,
-        aes(1, 
-            `Weight(g)`)) +
-  geom_boxplot(outlier.shape = NA) +
-  geom_quasirandom(aes(shape = mass_cluster),
-                   alpha = 0.4,
-                   size = 2.5) +
-  labs(shape='Cluster',
-       y = "Weight (g)") +
-  theme_bw(base_size = 6) +
-  theme(axis.ticks.x = element_blank(),
-        axis.text.x = element_blank(),
-        axis.title.x = element_blank(),
-        panel.grid.major = element_blank(), 
-        panel.grid.minor = element_blank()
-        )
-
-plot_tcsa_mass_cluster <- 
-ggplot (tcsa_mass_no_na,
-        aes(0.3, min_tcsa)) +
-  geom_boxplot(outlier.shape = NA) +
-  geom_quasirandom(aes(shape = mass_cluster),
-                   alpha = 0.4,
-                   size = 2.5) +
-  labs(shape='Cluster') +
-  theme_bw(base_size = 6) +
-  # show regions for tools from Table 3 of Lomabrd 2021
-  # https://www-sciencedirect-com.offcampus.lib.washington.edu/science/article/pii/S0277379121002286
-    annotate("rect", 
-           xmin = 0.1,
-           xmax = 0.2,
-           ymin = 4,
-           ymax = 18,
-           fill = "green",
-           alpha = 0.2) +
-      annotate("text", 
-               size = point_function_group_label_size,
-               label = "Poisoned\narrowheads",
-               hjust = 0,
-           x = point_function_group_label_y,
-           y = (18-4) /2 + 4) +
-      annotate("rect", 
-           xmin = 0.2,
-           xmax = 0.3,
-           ymin = 13,
-           ymax = 53,
-           fill = "blue",
-           alpha = 0.2) +
-      annotate("text", 
-               size = point_function_group_label_size,
-               label = "Arrowheads",
-               hjust = 0,
-           x = point_function_group_label_y,
-           y = (53-13) /2 + 13) +
-      annotate("rect", 
-           xmin = 0.3,
-           xmax = 0.4,
-           ymin = 40,
-           ymax = 76,
-           fill = "orange",
-           alpha = 0.2) +
-      annotate("text", 
-               size = point_function_group_label_size,
-               label = "Dart tips",
-               hjust = 0,
-           x = point_function_group_label_y,
-           y = (76-40) /2 + 40 ) +
-      annotate("rect", 
-           xmin = 0.4,
-           xmax = 0.5,
-           ymin = 44,
-           ymax = 98,
-           fill = "purple",
-           alpha = 0.2) +
-      annotate("text", 
-               size = point_function_group_label_size,
-               label = "Javelin",
-               hjust = 0,
-           x = point_function_group_label_y,
-           y = (98-44) /2 + 44 ) +
-     annotate("rect", 
-           xmin = 0.5,
-           xmax = 0.6,
-           ymin = 88,
-           ymax = 230,
-           fill = "yellow",
-           alpha = 0.2) +
-      annotate("text", 
-               label = "Thrusting\nspear",
-               size = point_function_group_label_size,
-               hjust = 0,
-           x = point_function_group_label_y,
-           y = (230-88) /2 + 88 ) +
-  coord_cartesian(xlim = c(-0.1,1.2)) +
-  theme_bw(base_size = 6) +
-  theme(axis.ticks.x = element_blank(),
-        axis.text.x = element_blank(),
-        axis.title.x = element_blank(),
-        panel.grid.major = element_blank(), 
-        panel.grid.minor = element_blank()
-        ) +
-  labs(y = "TCSA")
-
-library(cowplot)
-plot_grid(plot_mass_cluster + guides(shape = "none"),
-          plot_tcsa_mass_cluster,
-          labels = "AUTO",
-          rel_widths = c(0.5, 1))
-
-```
-
-@fig-tcsa-size shows a positive relationship between TCSA, artifact weight and raw material.
-The relationship varies by raw material types, perhaps due to variation in the density of different lithic raw materials.
-There is a stronger correlation between TCSA and artifact weight for acidic volcanic points (i.e.
-points are closer to the regression line), whereas the correlation for shale stemmed points is weaker.
-The other raw materials show varied distribution patterns around the regression line, reconfirming that raw material appears not to have strongly influenced TCSA (see also @fig-tcsa-raw-materials).
-
-## Geographical patterns of TCSA values
+## Geographical patterns
 
@@
 #| fig-width:   4.5
-#| fig-height:  4.5
+#| fig-height:  4
 
@@
       annotate("text", 
+               size = point_function_group_label_size,
+               label = "Poisoned\narrow tips",
+               hjust = 0,
+               lineheight = 0.75,
+           y = (0+2.5)/2,
+           x = 5.5) +
+      annotate("rect", 
+           xmin = 17,
+           xmax = 47,
+           ymin = 2.5,
+           ymax = 5,
+           fill = "blue",
+           alpha = 0.2) +
+      annotate("text", 
+               size = point_function_group_label_size,
+               label = "Un-poisoned\narrow tips",
+               hjust = 0,
+               lineheight = 0.75,
+           y = (2.5+5)/2,
+           x = 28) +
+      annotate("rect", 
+           xmin = 40,
+           xmax = 76,
+           ymin = 5,
+           ymax = 7.5,
+           fill = "orange",
+           alpha = 0.2) +
+      annotate("text", 
+               size = point_function_group_label_size,
+               label = "Dart tips",
+               hjust = 0,
+           y = (5+7.5)/2,
+           x = 53) +
+      annotate("rect", 
+           xmin = 42,
+           xmax = 90,
+           ymin = 7.5,
+           ymax = 10,
+           fill = "purple",
+           alpha = 0.2) +
+      annotate("text", 
+               size = point_function_group_label_size,
+               label = "Javelin tips",
+               hjust = 0,
+           y = (7.5+10)/2,
+           x = 70) +
+     annotate("rect", 
+           xmin = 80,
+           xmax = 200,
+           ymin = 10,
+           ymax = 12.5,
+           fill = "yellow",
+           alpha = 0.2) +
+      annotate("text", 
+               label = "Stabbing spear tips",
+               size = point_function_group_label_size,
+               hjust = 0,
+           y = (10+12.5)/2,
+           x = 160)
+```
+
+We computed artefact TCSA values for `r n_sites` sites that contain more than `r min_n_points` stemmed points to observe variation between sites (@fig-tcsa-sites) Sites with fewer than `r min_n_points` stemmed points were grouped under the category of 'other'.
+Among the sites in our sample, Suyanggae has the most stemmed points and the widest range of TCSA values.
+This suggests that people made stemmed points for a variety of different purposes in Suyanggae.
+The TCSA values for Nosanri and Sibuk are have a narrower range of lower values, which indicates less diverse functions for stemmed points.
+Wolpyeng and Yongsujaeul have similar amounts of stemmed points to Nosanri and Sibuk, but exhibit wider distributions of TCSA values.
+Yongsandong has the second most stemmed points and TCSA shows a similar distribution to Yongsakul.
+
+## Variation in TCSA values by artefact size
+
+```{r}
+
+# Facet by size class
+# https://cran.r-project.org/web/packages/Ckmeans.1d.dp/vignettes/Ckmeans.1d.dp.html
+library(Ckmeans.1d.dp)
+
+# remove NAs
+tcsa_mass_no_na <-
+  tcsa %>%
+ drop_na(`Weight(g)`)
+
+# find clusters
+result <- Ckmeans.1d.dp(tcsa_mass_no_na$`Weight(g)`)
+
+tcsa_mass_no_na <-
+tcsa_mass_no_na %>%
+  mutate(mass_cluster = factor(result$cluster))
+
+```
+
+Using weight as a size proxy we examined the relationship between size and likely use of the stemmed points inferred from TCSA values.
+We conducted a univariate cluster analysis INSERT CITATION TO PAPER ON THIS METHOD of stemmed points by weight and it revealed three clusters of artefact sizes (A of @fig-artifact-size-cluster, mean = `r round(mean(tcsa_mass_no_na$'Weight(g)'), digits=1)`, SD = `r round(sd(tcsa_mass_no_na$'Weight(g)'), digits=1)`).
+Cluster 1, the smallest (lightest) artifacts, have a lower range of TCSA values, compared to Cluster 2 (B of @fig-artifact-size-cluster).
+TCSA values for Cluster 3 are the highest, except for one artifact, which is lower than 50 (A of @fig-artifact-size-cluster, mean = `r round(mean(tcsa_mass_no_na$'min_tcsa'), digits=1)`, SD = `r round(sd(tcsa_mass_no_na$'min_tcsa'), digits=1)`).
+There are a few overlaps, but we can still assume that artifacts in different size clusters might be made for different purposes.
+
+```{r}
+#| label: fig-artifact-size-cluster
+#| fig-cap: "A. Distribution of artefact weight showing three clusters. B. TCSA values for all artefacts with size classes indicated by data point shape."
+#| fig-width:   4
+#| fig-height:  2
+
+plot_mass_cluster <- 
+ggplot (tcsa_mass_no_na,
+        aes(1, 
+            `Weight(g)`)) +
+  geom_boxplot(outlier.shape = NA) +
+  geom_quasirandom(aes(shape = mass_cluster),
+                   alpha = 0.4,
+                   size = 1.5) +
+  labs(shape='Cluster',
+       y = "Weight (g)") +
+  theme_bw(base_size = 6) +
+  theme(axis.ticks.x = element_blank(),
+        axis.text.x = element_blank(),
+        axis.title.x = element_blank(),
+        panel.grid.major = element_blank(), 
+        panel.grid.minor = element_blank()
+        )
+
+plot_tcsa_mass_cluster <- 
+ggplot (tcsa_mass_no_na,
+        aes(0.3, min_tcsa)) +
+  geom_boxplot(outlier.shape = NA) +
+  geom_quasirandom(aes(shape = mass_cluster),
+                   alpha = 0.4,
+                   size = 1.5) +
+  labs(shape='Cluster') +
+  theme_bw(base_size = 6) +
+  # show regions for tools from Table 3 of Lomabrd 2021
+  # https://www-sciencedirect-com.offcampus.lib.washington.edu/science/article/pii/S0277379121002286
+    annotate("rect", 
+           xmin = 0.1,
+           xmax = 0.2,
+           ymin = 4,
+           ymax = 18,
+           fill = "green",
+           alpha = 0.2) +
+      annotate("text", 
+               size = point_function_group_label_size,
+               label = "Poisoned\narrow tips",
+               hjust = 0,
+               lineheight = 0.75,
+           x = point_function_group_label_y,
+           y = (18-4) /2 + 4) +
+      annotate("rect", 
+           xmin = 0.2,
+           xmax = 0.3,
+           ymin = 17,
+           ymax = 47,
+           fill = "blue",
+           alpha = 0.2) +
+      annotate("text", 
+               size = point_function_group_label_size,
+               label = "Un-poisoned\narrow tips",
+               hjust = 0,
+               lineheight = 0.75,
+           x = point_function_group_label_y,
+           y = (47-17) /2 + 17) +
+      annotate("rect", 
+           xmin = 0.3,
+           xmax = 0.4,
+           ymin = 40,
+           ymax = 76,
+           fill = "orange",
+           alpha = 0.2) +
+      annotate("text", 
+               size = point_function_group_label_size,
+               label = "Dart tips",
+               hjust = 0,
+           x = point_function_group_label_y,
+           y = (76-40) /2 + 40) +
+      annotate("rect", 
+           xmin = 0.4,
+           xmax = 0.5,
+           ymin = 42,
+           ymax = 90,
+           fill = "purple",
+           alpha = 0.2) +
+      annotate("text", 
+               size = point_function_group_label_size,
+               label = "Javelin tips",
+               hjust = 0,
+           x = point_function_group_label_y,
+           y = (90-42) /2 + 42) +
+     annotate("rect", 
+           xmin = 0.5,
+           xmax = 0.6,
+           ymin = 80,
+           ymax = 200,
+           fill = "yellow",
+           alpha = 0.2) +
+      annotate("text", 
+               label = "Stabbing\nspear tips",
+               size = point_function_group_label_size,
+               hjust = 0,
+               lineheight = 0.75,
+           x = point_function_group_label_y,
+           y = (200-80) /2 + 80 ) +
+  coord_cartesian(xlim = c(-0.1,1.2)) +
+  theme_bw(base_size = 6) +
+  theme(axis.ticks.x = element_blank(),
+        axis.text.x = element_blank(),
+        axis.title.x = element_blank(),
+        panel.grid.major = element_blank(), 
+        panel.grid.minor = element_blank()
+        ) +
+  labs(y = "TCSA")
+
+library(cowplot)
+plot_grid(plot_mass_cluster + guides(shape = "none"),
+          plot_tcsa_mass_cluster,
+          labels = "AUTO",
+          rel_widths = c(0.5, 1))
+
+```
+
+@fig-tcsa-size shows a positive relationship between TCSA, artifact weight and raw material.
+The relationship varies by raw material types, perhaps due to variation in the density of different lithic raw materials.
+There is a stronger correlation between TCSA and artifact weight for acidic volcanic points (i.e.
+points are closer to the regression line), whereas the correlation for shale stemmed points is weaker.
+The other raw materials show varied distribution patterns around the regression line, reconfirming that raw material appears not to have strongly influenced TCSA (see also @fig-tcsa-raw-materials).
+
+## Geographical patterns of TCSA values
+
+```{r}
+#| label: fig-tcsa-per-sites
+#| fig-cap: "Histograms of TCSA values by archaeological site."
+#| fig-width:   4.5
+#| fig-height:  4
+
+ ggplot(tcsa_site,
+              aes(min_tcsa, 
+                  fill = n)) +
+  geom_histogram() +
+  theme_bw(base_size = 6) +
+  labs(x = "TCSA", 
+       y = "Count of artefacts") +
+  facet_wrap(~label_site,
+                 scales = "free_y",
+                  ncol = 1) +
+    scale_fill_viridis_c(name = "Number\nof artefacts") +
+  annotate("rect", 
+           xmin = 4,
+           xmax = 18,
+           ymin = 0,
+           ymax = 2.5,
+           fill = "green",
+           alpha = 0.2) +
+      annotate("text", 
                size = 1.5,
@@
 
-We computed artefact TCSA values for `r n_sites` sites that contain more than `r min_n_points` stemmed points to observe variation between sites (@fig-tcsa-sites) Sites with fewer than `r min_n_points` stemmed points were grouped under the category of 'other'.
+We computed artefact TCSA values for `r n_sites` sites that contain more than `r min_n_points` stemmed points to observe variation between sites (@fig-tcsa-per-sites) Sites with fewer than `r min_n_points` stemmed points were grouped under the category of 'other'.
 Among the sites in our sample, Suyanggae has the most stemmed points and the widest range of TCSA values.
@@
 #| fig-width:   4.5
-#| fig-height:  2.5
+#| fig-height:  4
 
@@
 
-ggplot(tcsa_site_ass, 
+temporal_tcsa <- ggplot(tcsa_site_ass, 
        aes(y = min_tcsa,
@@
 
+# add shady box for MIS stage
+
+temporal_tcsa +
+ annotate("rect", 
+           xmin = 0,
+           xmax = 12.2,
+           ymin = 0,
+           ymax = 300,
+           fill = "green",
+           alpha = 0.2) +
+  annotate("text", 
+               size = 3.5,
+               label = "MIS 3",
+               hjust = 0,
+           x = 5,
+           y = 260) +
+   annotate("rect", 
+           xmin = 12.2,
+           xmax = 24.5,
+           ymin = 0,
+           ymax = 300,
+           fill = "yellow",
+           alpha = 0.2) +
+  annotate("text", 
+               size = 3.5,
+               label = "MIS 2",
+               hjust = 0,
+           x = 17,
+           y = 260) +
+   annotate("rect", 
+           xmin = 13.5,
+           xmax = 21,
+           ymin = 0,
+           ymax = 230,
+           fill = "yellow",
+           alpha = 0.2) +
+  annotate("text", 
+               size = 2.3,
+               label = "LGM",
+               hjust = 0,
+           x = 17.3,
+           y = 210)
+
+```
+
+```{r}
+
+# aggregate all the MIS 3 sites, MIS 2 sites
+tcsa_site_ass_mis <- tcsa_site_ass %>% 
+                      mutate(mis_stage = case_when (
+                        age_ka > 29 ~ '3', age_ka < 29 ~ '2'))
+
+# box plot for each MIS stage
+  ggplot(tcsa_site_ass_mis, 
+       aes(y = min_tcsa,
+           x = mis_stage)) +
+   geom_boxplot() +
+   theme_minimal() +
+   theme(legend.position="none") +
+   labs(y = "TCSA", 
+        x = "MIS") +
+  geom_quasirandom(alpha = 0.4,
+                   size = 2.3) +
+    scale_x_discrete(limits=rev) +
+     annotate("rect", 
+           xmin = 2.1,
+           xmax = 2.2,
+           ymin = 4,
+           ymax = 18,
+           fill = "green",
+           alpha = 0.2) +
+      annotate("text", 
+               size = point_function_group_label_size,
+               label = "Poisoned\narrow tips",
+               hjust = 0,
+               lineheight = 0.75,
+           x = 2.1,
+           y = (18-4) /2 + 4) +
+      annotate("rect", 
+           xmin = 2.2,
+           xmax = 2.3,
+           ymin = 17,
+           ymax = 47,
+           fill = "blue",
+           alpha = 0.2) +
+      annotate("text", 
+               size = point_function_group_label_size,
+               label = "Un-poisoned\narrow tips",
+               hjust = 0,
+               lineheight = 0.75,
+           x = 2.2,
+           y = (47-17) /2 + 17) +
+      annotate("rect", 
+           xmin = 2.3,
+           xmax = 2.4,
+           ymin = 40,
+           ymax = 76,
+           fill = "orange",
+           alpha = 0.2) +
+      annotate("text", 
+               size = point_function_group_label_size,
+               label = "Dart tips",
+               hjust = 0,
+           x = 2.3,
+           y = (76-40) /2 + 40) +
+      annotate("rect", 
+           xmin = 2.4,
+           xmax = 2.5,
+           ymin = 42,
+           ymax = 90,
+           fill = "purple",
+           alpha = 0.2) +
+      annotate("text", 
+               size = point_function_group_label_size,
+               label = "Javelin tips",
+               hjust = 0,
+           x = 2.4,
+           y = (90-42) /2 + 42) +
+     annotate("rect", 
+           xmin = 2.5,
+           xmax = 2.6,
+           ymin = 80,
+           ymax = 200,
+           fill = "yellow",
+           alpha = 0.2) +
+      annotate("text", 
+               label = "Stabbing\nspear tips",
+               size = point_function_group_label_size,
+               hjust = 0,
+               lineheight = 0.75,
+           x = 2.5,
+           y = (200-80) /2 + 80 )
+
 ```
@@
 The chronological sequence of TCSA varies from region to region with no clear pattern.
-Most stemmed points appear to be have designed for use as javelin and thrusting spear tips, but there are also large numbers of tools that likely could fulfil multiple roles as armatures.
+Most stemmed points appear to be have designed for use as javelin and thrusting spear tips, but there are also large numbers of tools that likely could fulfill multiple roles as armatures.
 
@@
 To examine the possible connections between the roles of stemmed points and population dynamics or environmental change during the Late Paleolithic period, we compared the temporal pattern of TCSA to results of Summed probability distribution (SPD) and simulated Mean Annual Temperature (MAT) from @park2022did.
+In those two analyses, -@park2022did covered assemblages with and without stemmed points, which is useful for us to explore the overall socio-environmental dynamics.
+As we discussed, there is no clear pattern in the temporal or regional pattern of TCSA (@fig-tcsa-radio-carbon or @no-figure-no-more: A).
+We consider that TCSA, or the function of the stemmed points, are strongly dependent on the surrounding environment or their situation.
+@no-figure-no-more: B shows that the temperature slowly decreased until around 22 ka and then went up again with a little range of fluctuations.
+Temperatures differ significantly between sites, with a maximum difference of 5℃.
+Results of SPD show several peaks, as indicated by the positive deviations from the three null models (shaded in red), and downs (shaded in blue), but no drastic growth or decline (@no-figure-no-more: C).
+From the climate and population study, we could observe that there is no dominating change but minor and often fluctuations throughout the Late Paleolithic and variations per site.
+Overall this pattern is well matched with the TCSA results, indicating the people produced stemmed points as multifunctional tools suitable for unexpected or various situations in their habitats similar to the case of Clovis points (@eren2022plains).
+
+```{r}
+
+korean_archaeological_site_locations_tcsa <- 
+  tcsa_site_ass %>% # maybe some sites are missing TSCA values in this table?
+  mutate(Site_name = str_remove(Assemblage, "_.*")) %>%  
+  full_join(korean_archaeological_site_locations) %>% 
+  drop_na(min_tcsa)
+
+ggmap(map)  +
+  geom_jitter(data = korean_archaeological_site_locations_tcsa,
+             aes(long_dd ,
+                 lat_dd,
+                 colour = min_tcsa),
+             size = 3,
+             width = 0.25, 
+             height = 0.25) +
+  scale_color_viridis_c(option = "A") +
+  theme_minimal(base_size = 6) +
+  labs(x = "Longitute",
+       y = "Latitude") +
+  theme(axis.title.x = element_blank(),
+        axis.title.y = element_blank())
+
+```
+
+Our models premise that the choice of weapon-tip types reflects the socio-environmental circumstances.
+To examine the possible connections between the roles of stemmed points and population dynamics or environmental change during the Late Paleolithic period, we compared the temporal pattern of TCSA to results of Summed probability distribution (SPD) and simulated Mean Annual Temperature (MAT) from @park2022did.
 In those two analyses, -@park2022did covered assemblages with and without stemmed points, which is useful for us to explore the overall socio-environmental dynamics here.

### 2022-10-18 analysis/paper/paper.qmd
Edited tip type names in text. (GP did it for the plots previously)

@@
 What are the temporal patterns in stemmed point use?
-We measured tip cross-sectional area (TCSA) to discriminate different likely use classes of projectile points, for example, as poisoned arrowheads or as thrusting spears.
+We measured tip cross-sectional area (TCSA) to discriminate different likely use classes of projectile points, for example, as poisoned arrow tips or as stabbing spears.
 We analyzed TCSA with other variables including raw materials, weight, site and radiocarbon dates.
-Our results show that the stemmed points were likely used  as javelins and thrusting spear tips, with smaller numbers as dart tips and arrowheads.
+Our results show that the stemmed points were likely used  as javelin tips and stabbing spear tips, with smaller numbers as dart tips and un-poisoned arrow tips.
 TCSA values are controlled mostly by size rather than raw material types.
@@
 The temporal pattern of TCSA values is one of little change throughout the Late Paleolithic period.
-We conclude that stemmed points were  multi-functional tools, with many likely designed for javelin and thrusting spear tips.
+We conclude that stemmed points were  multi-functional tools, with many likely designed for javelin and stabbing spear tips.
 keywords: |
@@
 Tip cross-sectional area (TCSA) of stone artefacts has been used as a ballistically relevant standard to probabalistically discriminate different between functional classes of projectile armatures such as spearthrower (a.k.a.
-atlatl) dart tips, arrowheads and large stabbing/thrusting spears [@sisk2011african; @lombard2021variation; @lombard2020tip; @lombard2021did; @o2018origins].
+atlatl) dart tips, un-poisoned arrow tips and large stabbing/thrusting spears [@sisk2011african; @lombard2021variation; @lombard2020tip; @lombard2021did; @o2018origins].
 It is important to note that the TCSA metric alone cannot unambiguously determine artefact function, it only suggests a best-fit ballistic probability for the points if they were hafted as weapon tips.
@@
 The sample mean TCSA is `r round(mean(tcsa$min_tcsa), digits=1)`, and standard deviation of TCSA is `r round(sd(tcsa$min_tcsa), digits=1)`.
-According to the TCSA ranges presented in Table 1 of @lombard2021variation, Korean stemmed points are overall mostly in the categories of javelins and thrusting spear tips, with smaller numbers as dart tips and arrowheads.
-Among the weapon-tip types, only poison arrowheads appear to be absent from these Korean assemblages.
+According to the TCSA ranges presented in Table 1 of @lombard2021variation, Korean stemmed points are overall mostly in the categories of javelin tips and stabbing spear tips, with smaller numbers as dart tips and un-poisoned arrow tips.
+Among the weapon-tip types, only poison arrow tips appear to be absent from these Korean assemblages.
 
@@
                size = point_function_group_label_size,
-               label = "Arrowheads",
-               hjust = 0,
+               label = "Un-poisoned\narrow tips",
+               hjust = 0,
+              lineheight = 0.75,
            y = (2.5+5)/2,
@@
       annotate("text", 
-               label = "Thrusting spear",
+               label = "STabbing spear tips",
                size = point_function_group_label_size,
@@
 
-Our results indicate that javelins and thrusting spear tips are the most probable ballistic uses for stemmed points.
-There are a few stemmed points in our sample that may have been used as dart tips and arrowheads, but none of them have TCSA values in the range of poison arrowheads (@fig-tcsa-all-sp).
+Our results indicate that javelin tips and stabbing spear tips are the most probable ballistic uses for stemmed points.
+There are a few stemmed points in our sample that may have been used as dart tips and un-poisoned arrow tips, but none of them have TCSA values in the range of poison arrow tips (@fig-tcsa-all-sp).
 We found a wide range of TCSA values in our sample, indicating that stemmed points played a diverse role in forager toolkits.
@@
 The chronological sequence of TCSA varies from region to region with no clear pattern.
-Most stemmed points appear to be have designed for use as javelin and thrusting spear tips, but there are also large numbers of tools that likely could fulfill multiple roles as armatures.
+Most stemmed points appear to be have designed for use as javelin and stabbing spear tips, but there are also large numbers of tools that likely could fulfill multiple roles as armatures.
 

### 2022-10-19 analysis/paper/paper.qmd
Edit two boxplots for temporal pattern

@@
            ymax = 300,
-           fill = "green",
+           fill = "lightgray",
            alpha = 0.2) +
@@
            ymax = 300,
-           fill = "yellow",
            alpha = 0.2) +
@@
            ymax = 230,
-           fill = "yellow",
            alpha = 0.2) +
@@
                       mutate(mis_stage = case_when (
-                        age_ka > 29 ~ '3', age_ka < 29 ~ '2'))
+                        age_ka > 26.5 ~ 'Before' , age_ka > 19 ~ 'LGM', age_ka < 19 ~ 'After'))
 
@@
        aes(y = min_tcsa,
-           x = mis_stage)) +
+           x = factor(mis_stage, level = c('After', 'LGM', 'Before')))) +
    geom_boxplot() +
@@
      annotate("rect", 
-           xmin = 2.1,
-           xmax = 2.2,
+           xmin = 3.4,
+           xmax = 3.5,
            ymin = 4,
@@
                lineheight = 0.75,
-           x = 2.1,
+           x = 3.4,
            y = (18-4) /2 + 4) +
       annotate("rect", 
-           xmin = 2.2,
-           xmax = 2.3,
+           xmin = 3.5,
+           xmax = 3.6,
            ymin = 17,
@@
                lineheight = 0.75,
-           x = 2.2,
+           x = 3.5,
            y = (47-17) /2 + 17) +
       annotate("rect", 
-           xmin = 2.3,
-           xmax = 2.4,
+           xmin = 3.6,
+           xmax = 3.7,
            ymin = 40,
@@
                hjust = 0,
-           x = 2.3,
+           x = 3.6,
            y = (76-40) /2 + 40) +
       annotate("rect", 
-           xmin = 2.4,
-           xmax = 2.5,
+           xmin = 3.7,
+           xmax = 3.8,
            ymin = 42,
@@
                hjust = 0,
-           x = 2.4,
+           x = 3.7,
            y = (90-42) /2 + 42) +
      annotate("rect", 
-           xmin = 2.5,
-           xmax = 2.6,
+           xmin = 3.8,
+           xmax = 3.9,
            ymin = 80,
@@
                lineheight = 0.75,
-           x = 2.5,
+           x = 3.8,
            y = (200-80) /2 + 80 )

### 2022-10-19 analysis/paper/paper.qmd
put code for null ##

@@
 ```{r}
-# ADD CODE TO REPLACE NULL WITH ACTUAL VALUES
-n_sites <- NULL # replace this with the number
-min_n_points <- NULL # replace this with the number
-n_radiocarbon_ages <- NULL # replace this with the number
-n_sites_with_radiocarbon_ages <- NULL # replace this with the number
-
+
+tcsa_site <- tcsa %>% 
+  group_by(site_name_top) %>% 
+  add_tally() %>% 
+  filter(n>1) %>% 
+  mutate(label_site = as.factor(paste0(site_name_top,' (n = ', n, ")"))) %>%   
+  dplyr::select(min_tcsa,
+         site_name_top, 
+         label_site,
+         n) %>% 
+  distinct_all()
+
+# get rid of 'Other'
+tcsa_site_num <- tcsa_site %>% 
+  filter(site_name_top != 'Other')
+
+###########
+# radiocarbon data per assemblage
+sitedate <- read_excel(here("analysis/data/raw_data/TCSA_site_info.xlsx"))
+
+mydata_ages <-
+  sitedate %>%
+  separate(Period, into = c('age', 'error'),
+           sep = "±") %>%
+  mutate(age_ka = parse_number(age) / 1000,
+         error = parse_number(error))
+
+# combine the assemblage with age data
+
+tcsa_site_ass <- tcsa %>% 
+  left_join(mydata_ages) %>% 
+  group_by(Assemblage) %>% 
+ drop_na(age_ka) %>% 
+  add_tally() %>% 
+  mutate(label_site = as.factor(paste0(Assemblage,' (n = ', n, '), (', {round(age_ka,1)}, 'ka)' ))) %>%   
+  dplyr::select(age_ka,
+         min_tcsa,
+         Assemblage, 
+         label_site,
+         n) %>% 
+    distinct_all()
+
+########
+mydata_ages_na <- mydata_ages %>%
+  drop_na(age)
+
+# count radiocarbon ages
+
+n_sites <- length(unique(tcsa$Site_name))
+min_n_points <- nrow(tcsa_site) # should I get rid of "Other" in the figure?, if then, it's nrow(tcsa_site_num)
+# n_radiocarbon_ages <- NULL # we only have one age per assemblage
+n_sites_with_radiocarbon_ages <- nrow(mydata_ages_na)
 ```
@@
 
-We used `r n_radiocarbon_ages` radiocarbon ages to explore the temporal pattern of the function of stemmed points.
+We used radiocarbon ages to explore the temporal pattern of the function of stemmed points.
 We excluded assemblages that have no radiocarbon dates.
@@
 
-tcsa_site <- tcsa %>% 
-  group_by(site_name_top) %>% 
-  add_tally() %>% 
-  filter(n>1) %>% 
-  mutate(label_site = as.factor(paste0(site_name_top,' (n = ', n, ")"))) %>%   
-  dplyr::select(min_tcsa,
-         site_name_top, 
-         label_site,
-         n) %>% 
-  distinct_all()
-
 # Facet by raw materials
@@
 
-We computed artefact TCSA values for `r n_sites` sites that contain more than `r min_n_points` stemmed points to observe variation between sites (@fig-tcsa-sites) Sites with fewer than `r min_n_points` stemmed points were grouped under the category of 'other'.
+We computed artefact TCSA values for `r n_sites` sites that contain more than `r min_n_points` stemmed points to observe variation between sites (@fig-tcsa-sites) Sites with fewer than 4 stemmed points were grouped under the category of 'other'.
 Among the sites in our sample, Suyanggae has the most stemmed points and the widest range of TCSA values.
@@
 
-# radiocarbon data per assemblage
-sitedate <- read_excel(here("analysis/data/raw_data/TCSA_site_info.xlsx"))
-
-mydata_ages <-
-  sitedate %>%
-  separate(Period, into = c('age', 'error'),
-           sep = "±") %>%
-  mutate(age_ka = parse_number(age) / 1000,
-         error = parse_number(error))
-
-# combine the assemblage with age data
-
-tcsa_site_ass <- tcsa %>% 
-  left_join(mydata_ages) %>% 
-  group_by(Assemblage) %>% 
- drop_na(age_ka) %>% 
-  add_tally() %>% 
-  mutate(label_site = as.factor(paste0(Assemblage,' (n = ', n, '), (', {round(age_ka,1)}, 'ka)' ))) %>%   
-  dplyr::select(age_ka,
-         min_tcsa,
-         Assemblage, 
-         label_site,
-         n) %>% 
-    distinct_all()
-
 temporal_tcsa <- ggplot(tcsa_site_ass, 
@@
 
-korean_archaeological_site_locations_tcsa <- 
-  tcsa_site_ass %>% # maybe some sites are missing TSCA values in this table?
-  mutate(Site_name = str_remove(Assemblage, "_.*")) %>%  
+# korean_archaeological_site_locations_tcsa <- 
+#   tcsa_site_ass %>% 
+#   mutate(Site_name = str_remove(Assemblage, "_.*")) %>%  
+#   full_join(korean_archaeological_site_locations) %>% 
+#   drop_na(min_tcsa)
+
+korean_archaeological_site_locations_tcsa <- tcsa %>% 
   full_join(korean_archaeological_site_locations) %>% 

### 2022-10-26 analysis/paper/paper.qmd
added shapefile
Added shapefile data
stash
exploring GIS files

@@
 
-# korean_archaeological_site_locations_tcsa <- 
-#   tcsa_site_ass %>% 
-#   mutate(Site_name = str_remove(Assemblage, "_.*")) %>%  
-#   full_join(korean_archaeological_site_locations) %>% 
-#   drop_na(min_tcsa)
-
-korean_archaeological_site_locations_tcsa <- tcsa %>% 
-  full_join(korean_archaeological_site_locations) %>% 
+korean_archaeological_site_locations_tcsa <- 
+  tcsa %>% 
+  mutate(Site_name_to_join = str_squish(Site_name)) %>% 
+  left_join(korean_archaeological_site_locations %>% 
+              mutate(Site_name_to_join  = str_squish(Site_name)),
+                     by = "Site_name_to_join") %>% 
   drop_na(min_tcsa)
@@
 
+library(sf)
+
+veg <- read_sf(here("analysis/data/gis_data/vegetation_polygons.shp"))
+
+# take a look
+ggplot() +
+  geom_sf(data = veg)
+
+# BM: basic plan is the do a spatial join with the veg zones and the TCSA values.
+Then table of artefacts will get a new column telling us what veg zone it is in.
+Then we can make a plot to compare the distribution of tcsa values across the different veg zones
+
+# Determine which points fall inside which polygons
+
+# Convert to class 'sf'
+pts <- st_as_sf(korean_archaeological_site_locations_tcsa,
+                coords = c( "long_dd", "lat_dd"),
+                crs = st_crs(4326))
+
+# take a look
+ggplot() +
+  geom_sf(data = pts)
+
+points_in_veg_zones <- 
+st_join(pts, 
+        veg, 
+        join = st_contains)
+
 ```

### 2022-10-26 analysis/paper/paper.qmd
Changed cords on shapefile
exploring GIS files and crs
made QGIS project

@@
 
-veg <- read_sf(here("analysis/data/gis_data/vegetation_polygons.shp"))
+veg <- read_sf(here("analysis/data/gis_data/Vegetation_polygon_2.shp"))
+
+st_crs(veg)
 
@@
 ggplot() +
-  geom_sf(data = veg)
+  geom_sf(data = veg) +
+  coord_sf()
 
@@
                 coords = c( "long_dd", "lat_dd"),
-                crs = st_crs(4326))
+                crs = st_crs(veg))
 

### 2022-10-26 analysis/paper/paper.qmd
sketch of code for spatial analysis with veg regions

@@
 
-veg <- read_sf(here("analysis/data/gis_data/Vegetation_polygon_2.shp"))
-
-st_crs(veg)
+veg <- read_sf(here("analysis/data/gis_data/vegetation_polygons.shp"))
+
+# st_crs(veg)
+
+veg <- st_transform(veg, crs = st_crs(4162))
 
@@
 points_in_veg_zones <- 
-st_join(pts, 
-        veg, 
+st_join(veg, 
+        pts, 
         join = st_contains)
 
+ggplot() +
+  geom_sf(data = veg) +
+  geom_sf(data = pts)
+
+ggplot(points_in_veg_zones) +
+  aes(Vegetation,
+      min_tcsa) +
+  geom_boxplot() +
+  geom_quasirandom()
+
+summary(aov(min_tcsa ~ as.factor(Vegetation), points_in_veg_zones))
 ```

### 2022-10-27 analysis/paper/paper.qmd
made plots for only LGM artifacts

@@
   tcsa %>% 
+  left_join(mydata_ages) %>% # add age data
+  group_by(Assemblage) %>% 
   mutate(Site_name_to_join = str_squish(Site_name)) %>% 
   left_join(korean_archaeological_site_locations %>% 
-              mutate(Site_name_to_join  = str_squish(Site_name)),
+              mutate(Site_name_to_join  = str_squish(Site_name)), # deleted invisible space in the data set probably created during data collecting
                      by = "Site_name_to_join") %>% 
-  drop_na(min_tcsa)
+  drop_na(min_tcsa) %>% 
+  drop_na(age)
+
+# data only for LGM period
+tcsa_lgm <- korean_archaeological_site_locations_tcsa %>% 
+  filter(age< 26.5 & age> 19)
 
 ggmap(map)  +
-  geom_jitter(data = korean_archaeological_site_locations_tcsa,
-             aes(long_dd ,
-                 lat_dd,
+  geom_jitter(data = tcsa_lgm,
+             aes(long_dd.y ,
+                 lat_dd.y,
                  colour = min_tcsa),
@@
 # Convert to class 'sf'
-pts <- st_as_sf(korean_archaeological_site_locations_tcsa,
-                coords = c(   "long_dd", "lat_dd"),
+pts <- st_as_sf(tcsa_lgm,
+                coords = c(   "long_dd.y", "lat_dd.y"),
                 crs = st_crs(veg))
@@
 ggplot() +
-  geom_sf(data = veg) +
+  geom_sf(data = veg, fill=c("#CCCCCC", "#FFFFFF", "#999999", "#666666")) + #add color for vegetation polygons
   geom_sf(data = pts)

### 2022-10-27 analysis/paper/paper.qmd
added colors to the gis plot
updated shapefiles for new eco-regions, and tried to use the previous codes to create plots

@@
 ggplot() +
-  geom_sf(data = veg, fill=c("#CCCCCC", "#FFFFFF", "#999999", "#666666")) + #add color for vegetation polygons
-  geom_sf(data = pts)
+  geom_sf(data = veg, aes(fill = veg$Vegetation)) + #add color for vegetation polygons
+  geom_sf(data = pts, aes(color = pts$min_tcsa)) +
+  theme_minimal()
+
+# fill=c("#CCCCCC", "#FFFFFF", "#999999", "#666666")
 
@@
 summary(aov(min_tcsa ~ as.factor(Vegetation), points_in_veg_zones))
+
+############ explore tcsa with MODERN eco_regions #####
+
+veg_all <- read_sf(here("analysis/data/gis_data/eco_region.shp"))
+
+veg_all <- st_transform(veg_all, crs = st_crs(4162))
+
+ggplot() +
+  geom_sf(data = veg_all) +
+  coord_sf()
+
+pts_all <- st_as_sf(korean_archaeological_site_locations_tcsa,
+                coords = c(   "long_dd.y", "lat_dd.y"),
+                crs = st_crs(veg_all))
+ggplot() +
+  geom_sf(data = pts_all)
+
+points_in_veg_zones_all <- 
+st_join(veg_all, 
+        pts_all, 
+        join = st_contains)
+
+ggplot() +
+  geom_sf(data = veg_all, aes(fill = veg_all$ecoregion)) + #add color for vegetation polygons
+  geom_sf(data = pts_all, aes(color = pts_all$min_tcsa)) +
+  theme_minimal()
+
+ggplot(points_in_veg_zones_all) +
+  aes(ecoregion,
+      min_tcsa) +
+  geom_boxplot() +
+  geom_quasirandom()
+
+summary(aov(min_tcsa ~ as.factor(ecoregion), points_in_veg_zones_all))
+
 ```

### 2022-10-28 analysis/paper/paper.qmd
Fixed error by trying this --> https://github.com/r-spatial/sf/issues/1762

@@
 
-veg_all <- read_sf(here("analysis/data/gis_data/eco_region.shp"))
-
-veg_all <- st_transform(veg_all, crs = st_crs(4162))
+veg_all <- read_sf(here("analysis/data/gis_data/eco_re_two.shp"))
+
+veg_all <- st_transform(veg_all, crs = st_crs(4326))
 
@@
 pts_all <- st_as_sf(korean_archaeological_site_locations_tcsa,
-                coords = c(   "long_dd.y", "lat_dd.y"),
+                coords = c("long_dd.y", "lat_dd.y"),
                 crs = st_crs(veg_all))
@@
   geom_sf(data = pts_all)
+
+#sf_use_s2(FALSE)
 

### 2022-10-30 analysis/paper/paper.qmd
Added codes for eco&forest zones

@@
 
-############ explore tcsa with MODERN eco_regions #####
+############ explore tcsa with MODERN eco_regions #######
 
@@
 summary(aov(min_tcsa ~ as.factor(ecoregion), points_in_veg_zones_all))
+
+############ explore tcsa with MODERN eco_regions & forest zones #######
+
+veg_forest <- read_sf(here("analysis/data/gis_data/eco_forest_two.shp"))
+
+veg_forest <- st_transform(veg_forest, crs = st_crs(4326))
+
+ggplot() +
+  geom_sf(data = veg_forest) +
+  coord_sf()
+
+pts_forest <- st_as_sf(korean_archaeological_site_locations_tcsa,
+                coords = c("long_dd.y", "lat_dd.y"),
+                crs = st_crs(veg_forest))
+ggplot() +
+  geom_sf(data = pts_forest)
+
+#sf_use_s2(FALSE)
+
+points_in_veg_zones_forest <- 
+st_join(veg_forest, 
+        pts_forest, 
+        join = st_contains)
+
+ggplot() +
+  geom_sf(data = veg_forest, aes(fill = veg_forest$Zone)) + #add color for vegetation polygons
+  geom_sf(data = pts_forest, aes(color = pts_forest$min_tcsa)) +
+  theme_minimal()
+
+ggplot(points_in_veg_zones_forest) +
+  aes(Zone,
+      min_tcsa) +
+  geom_boxplot() +
+  geom_quasirandom()
+
+summary(aov(min_tcsa ~ as.factor(Zone), points_in_veg_zones_forest))
 

### 2022-10-30 analysis/paper/paper.qmd
Moved the gis chunk to the results

@@
 
-# Discussion ORDER OF TOPICS HERE MUST FOLLOW EXACTLY THE ORDER OF THE METHODS AND RESULTS SECTION
-
-To investigate the role that stemmed points played in the technological transition during the Korean Late Paleolithic, we computed the TCSA metric to infer likely artifact uses by comparing the results to other archaeological and ethnographic studies Our main questions were: What were the best-fit ballistic probabilities for the stemmed points if they were hafted as weapon tips?
-How diverse were their likely uses?
-What are the temporal patterns in stemmed point uses?
-
-Our results indicate that javelin tips and stabbing spear tips are the most probable ballistic uses for stemmed points.
-There are a few stemmed points in our sample that may have been used as dart tips and un-poisoned arrow tips, but none of them have TCSA values in the range of poison arrow tips (@fig-tcsa-all-sp).
-We found a wide range of TCSA values in our sample, indicating that stemmed points played a diverse role in forager toolkits.
-Some sites, such as Nosanri and Sinbuk, likely used stemmed points as restricted-function tools.
-Meanwhile, stemmed points were used for a wider set of tasks at other sites, especially Suyanggae.
-In general, stemmed points played multiple roles in more sites (@fig-tcsa-sites).
-Through the Late Paleolithic period, TCSA values showed little directional change (@fig-tcsa-radio-carbon).
-The chronological sequence of TCSA varies from region to region with no clear pattern.
-Most stemmed points appear to be have designed for use as javelin and stabbing spear tips, but there are also large numbers of tools that likely could fulfill multiple roles as armatures.
-
-TCSA can, however, be impacted by other factors such as raw materials and portability.
--@eren2022plains explain that smaller TCSA could be the result of pursuing production economy and transport efficiency.
-We explore the relationship between TCSA and raw material types and weight in order to examine these factors.
-Weight plays an important role in portability and penetration of projectiles [@hughes1998getting].
-According to our results, raw materials do not have a clear pattern, but weight is correlated with raw materials (@fig-tcsa-raw-materials and @fig-tcsa-size).
-Different clusters of weight are matched with different types of weapon tips (@fig-artifact-size-cluster).
-Based on these analyses, we assume that people were able to produce the ideal size tools regardless of raw materials.
-
-Our models premise that the choice of weapon-tip types reflects the socio-environmental circumstances.
-To examine the possible connections between the roles of stemmed points and population dynamics or environmental change during the Late Paleolithic period, we compared the temporal pattern of TCSA to results of Summed probability distribution (SPD) and simulated Mean Annual Temperature (MAT) from @park2022did.
-In those two analyses, -@park2022did covered assemblages with and without stemmed points, which is useful for us to explore the overall socio-environmental dynamics.
-As we discussed, there is no clear pattern in the temporal or regional pattern of TCSA (@fig-tcsa-radio-carbon or @no-figure-no-more: A).
-We consider that TCSA, or the function of the stemmed points, are strongly dependent on the surrounding environment or their situation.
-@no-figure-no-more: B shows that the temperature slowly decreased until around 22 ka and then went up again with a little range of fluctuations.
-Temperatures differ significantly between sites, with a maximum difference of 5℃.
-Results of SPD show several peaks, as indicated by the positive deviations from the three null models (shaded in red), and downs (shaded in blue), but no drastic growth or decline (@no-figure-no-more: C).
-From the climate and population study, we could observe that there is no dominating change but minor and often fluctuations throughout the Late Paleolithic and variations per site.
-Overall this pattern is well matched with the TCSA results, indicating the people produced stemmed points as multifunctional tools suitable for unexpected or various situations in their habitats similar to the case of Clovis points (@eren2022plains).
-
 ```{r}
@@
 ggplot() +
-  geom_sf(data = veg_forest, aes(fill = veg_forest$Zone)) + #add color for vegetation polygons
-  geom_sf(data = pts_forest, aes(color = pts_forest$min_tcsa)) +
+  geom_sf(data = veg_forest, aes(fill = Zone)) + #add color for vegetation polygons
+  geom_sf(data = pts_forest,size = 3, aes(color = min_tcsa)) +
   theme_minimal()
 
-ggplot(points_in_veg_zones_forest) +
+# get rid of zones that has no artifact
+points_in_veg_zones_forest_box <- points_in_veg_zones_forest %>% 
+  drop_na(Region)
+
+ggplot(points_in_veg_zones_forest_box) +
   aes(Zone,
@@
 
-summary(aov(min_tcsa ~ as.factor(Zone), points_in_veg_zones_forest))
-
-```
+summary(aov(min_tcsa ~ as.factor(Zone), points_in_veg_zones_forest_box))
+
+```
+
+# Discussion ORDER OF TOPICS HERE MUST FOLLOW EXACTLY THE ORDER OF THE METHODS AND RESULTS SECTION
+
+To investigate the role that stemmed points played in the technological transition during the Korean Late Paleolithic, we computed the TCSA metric to infer likely artifact uses by comparing the results to other archaeological and ethnographic studies Our main questions were: What were the best-fit ballistic probabilities for the stemmed points if they were hafted as weapon tips?
+How diverse were their likely uses?
+What are the temporal patterns in stemmed point uses?
+
+Our results indicate that javelin tips and stabbing spear tips are the most probable ballistic uses for stemmed points.
+There are a few stemmed points in our sample that may have been used as dart tips and un-poisoned arrow tips, but none of them have TCSA values in the range of poison arrow tips (@fig-tcsa-all-sp).
+We found a wide range of TCSA values in our sample, indicating that stemmed points played a diverse role in forager toolkits.
+Some sites, such as Nosanri and Sinbuk, likely used stemmed points as restricted-function tools.
+Meanwhile, stemmed points were used for a wider set of tasks at other sites, especially Suyanggae.
+In general, stemmed points played multiple roles in more sites (@fig-tcsa-sites).
+Through the Late Paleolithic period, TCSA values showed little directional change (@fig-tcsa-radio-carbon).
+The chronological sequence of TCSA varies from region to region with no clear pattern.
+Most stemmed points appear to be have designed for use as javelin and stabbing spear tips, but there are also large numbers of tools that likely could fulfill multiple roles as armatures.
+
+TCSA can, however, be impacted by other factors such as raw materials and portability.
+-@eren2022plains explain that smaller TCSA could be the result of pursuing production economy and transport efficiency.
+We explore the relationship between TCSA and raw material types and weight in order to examine these factors.
+Weight plays an important role in portability and penetration of projectiles [@hughes1998getting].
+According to our results, raw materials do not have a clear pattern, but weight is correlated with raw materials (@fig-tcsa-raw-materials and @fig-tcsa-size).
+Different clusters of weight are matched with different types of weapon tips (@fig-artifact-size-cluster).
+Based on these analyses, we assume that people were able to produce the ideal size tools regardless of raw materials.
+
+Our models premise that the choice of weapon-tip types reflects the socio-environmental circumstances.
+To examine the possible connections between the roles of stemmed points and population dynamics or environmental change during the Late Paleolithic period, we compared the temporal pattern of TCSA to results of Summed probability distribution (SPD) and simulated Mean Annual Temperature (MAT) from @park2022did.
+In those two analyses, -@park2022did covered assemblages with and without stemmed points, which is useful for us to explore the overall socio-environmental dynamics.
+As we discussed, there is no clear pattern in the temporal or regional pattern of TCSA (@fig-tcsa-radio-carbon or @no-figure-no-more: A).
+We consider that TCSA, or the function of the stemmed points, are strongly dependent on the surrounding environment or their situation.
+@no-figure-no-more: B shows that the temperature slowly decreased until around 22 ka and then went up again with a little range of fluctuations.
+Temperatures differ significantly between sites, with a maximum difference of 5℃.
+Results of SPD show several peaks, as indicated by the positive deviations from the three null models (shaded in red), and downs (shaded in blue), but no drastic growth or decline (@no-figure-no-more: C).
+From the climate and population study, we could observe that there is no dominating change but minor and often fluctuations throughout the Late Paleolithic and variations per site.
+Overall this pattern is well matched with the TCSA results, indicating the people produced stemmed points as multifunctional tools suitable for unexpected or various situations in their habitats similar to the case of Clovis points (@eren2022plains).
 
@@
 
+SPACIAL ANAlysis
+@prates2022changes
+
 # Conclusion

### 2022-10-31 analysis/paper/paper.qmd
Fixed missing name of zones for the eco-forest figure

@@
 
-SPACIAL ANAlysis
-@prates2022changes
+SPACIAL ANAlysis @prates2022changes provde strong relationship between the temporal density and spatial distribution of megafauna and Fishtail projectile points, and proposed that this weapon was designed and used for megafauna hunting, contributing to their extinction by exploring changes in projectile point technology, as well as the body mass of potential megafaunal prey, and show that Fishtails were strongly related to the largest extinct megafaunal species.
 

### 2022-10-31 analysis/paper/paper.qmd
rearranged the sequence of analysis

--- 2022-10-30 analysis/paper/paper.qmd
@@
 How diverse were their likely uses?
-What are the temporal patterns in stemmed point use?
+What are the temporal and spatial patterns in stemmed point use?
 We measured tip cross-sectional area (TCSA) to discriminate different likely use classes of projectile points, for example, as poisoned arrow tips or as stabbing spears.
@@
 Some sites show a wide range of TCSA values that represent multi-purpose usage of stemmed points.
-The temporal pattern of TCSA values is one of little change throughout the Late Paleolithic period.
-We conclude that stemmed points were  multi-functional tools, with many likely designed for javelin and stabbing spear tips.
+The temporal pattern of TCSA values is one of little change throughout the Late Paleolithic period.<!-- SP were mainly used before Last Glacial Maximum (LGM).
+SP was mainly located in certain latitude.
+why?
+--> We conclude that stemmed points were  multi-functional tools, with many likely designed for javelin and stabbing spear tips.
 keywords: |
@@
 
-The introduction of new stone artefact technologies marks a major transition in the Korean Paleolithic from the Early to the Late Paleolithic periods.
+The introduction of new stone artifact technologies marks a major transition in the Korean Paleolithic from the Early to the Late Paleolithic periods.
 The transition includes the appearance blade technology, stemmed points, end scrapers, burins, denticulates, etc [@bae2017origin; @nakazawa2018quaternary; @bae2017late; @lee2017korean; @seong2016eastern].
@@
 How diverse were their likely uses?
-What are the temporal patterns in stemmed point uses?
+What are the temporal and spatial patterns in stemmed point uses?
 As part of our discussion, we examine possible connections between the roles of stemmed points and population dynamics or environmental change during the Late Paleolithic period.
@@
 Slight retouch is typically located on the proximal end to shape an acute tip and on the distal end to make a stem, which connects to a wooden shaft.
-Elsewhere in the world these types of artifacts are often called tanged points, but we prefer 'stemmed point' here to distinguish from Bronze Age stone projectile points that have long been called 'tanged points' in Korea [@park2022did].
+Elsewhere in the world these types of artifacts are often called 'tanged points', but we prefer 'stemmed point' here to distinguish from Bronze Age stone projectile points that have long been called 'tanged points' in Korea [@park2022did].
 After the first discovery at the Suyanggae site, more than 450 stemmed points have been found in over 30 sites throughout South Korea [@Chong_2021].
@@
 atlatl) dart tips, un-poisoned arrow tips and large stabbing/thrusting spears [@sisk2011african; @lombard2021variation; @lombard2020tip; @lombard2021did; @o2018origins].
-It is important to note that the TCSA metric alone cannot unambiguously determine artefact function, it only suggests a best-fit ballistic probability for the points if they were hafted as weapon tips.
+It is important to note that the TCSA metric alone cannot unambiguously determine artifact function, it only suggests a best-fit ballistic probability for the points if they were hafted as weapon tips.
 The TCSA metric indicates the part of the tool that cuts the target's hide and relates to weapon flight and penetration dynamics [@hughes1998getting; @sitton2020tip; @lombard2021variation].
@@
 
-Since the first evidence of the Korean Late Paleolithic was discovered at Seokjangri site, including blades and a stemmed base, around 300 stemmed points have been found in Korea [@Sohn_1967; @lee2019were].
-Most sites contain only a few stemmed points and only a few sites have many more, such as Jingeuneul, Suyanggae (n = 55), and Yongsandong (n = 38) [@Kim_2017].
-Among these stemmed points, we selected those that retained their widest and thickest parts.
-Our dataset contains stemmed points that were found during field surveys as well as those were from sites that were never dated but were associtaed with other Late Palaeoltihic artifacts.
-Applying these sampling criteria resulted in a sample of 173 stemmed points from 36 assemblages unearthed from 29 sites spanning the period 44-10 ka (@fig-map).
-The dimensions of the 173 stemmed points were obtained from published excavation reports and by direct measurements during our visits to the collections of local museums and archaeological institutions in Korea.
-
-We distinguished between multiple assemblages at a single site where numerous excavations have been conducted in different locations at the site, and by different institutions under the same site name.
-For example, Suyanggae site, a registered Korean National Heritage site, has been excavated more than ten times since 1980 by the local university museum and later by archaeological research institutions.
-There are six different excavation locations that range from a few meters apart to a few kilometers.
-Similarly, we identified multiple assemblages in a single excavation or even a trench where archaeological deposits were separated by culturally sterile deposits, or where distinct artifact-bearing stratigraphic units could be identified by major differences in the texture, color, and composition of the sedimentary deposits.
-As a result, in this research we used four assemblages from the Suyanggae site.
-We separated one assemblage among the four by using a different site name, Hajinri, following the convention established the excavators of that location.
-Because Hajinri was from the sixth excavation area, which is 3.5 kilometers apart from the other spots and dated to much earlier (around 42-30 ka) than the other assemblages (around 31-15 ka).
-According to the excavation reports for Hajinri, stemmed points first appeared in Hajinri around 42 ka with the earliest blade and blade cores [@Lee_Woo_Lee_An_Yun_Park_Otani_Kim_Kim_Han_et_al._2018].
-We consider this assemblage as an unusual outlier because the stemmed points and blades from Hajinri are so early and also so highly standardized and refined, which are only found at other sites much later in time.
-We have more confidence in finds from Bonggok, which has the second-earliest dates (41.5 ka) in our collection, as the first appearance of stemmed points.
-Bonggok includes blades or elongate flakes, but without any accompanying blade cores [@park2022did].
-
 ```{r}
@@
  # mutate(max_tcsa = 0.5 *Length * Thickness) %>%
-  #omit very large TCSA, could be measurement error
-  dplyr::filter(min_tcsa <= 257) %>%
+  dplyr::filter(min_tcsa <= 250) %>% #omit very large TCSA, could be measurement error
   # group rare raw materials > 10
@@
 
+```{r}
+# move dataframes from the Results section to create codes for numbers in the text
+
+tcsa_site <- tcsa %>% 
+  group_by(site_name_top) %>% 
+  add_tally() %>% 
+  filter(n>1) %>% 
+  mutate(label_site = as.factor(paste0(site_name_top,' (n = ', n, ")"))) %>%   
+  dplyr::select(min_tcsa,
+         site_name_top, 
+         label_site,
+         n) %>% 
+  distinct_all()
+
+# get rid of 'Other'
+tcsa_site_num <- tcsa_site %>% 
+  filter(site_name_top != 'Other')
+
+###########
+# radiocarbon data per assemblage
+sitedate <- read_excel(here("analysis/data/raw_data/TCSA_site_info.xlsx"))
+
+mydata_ages <-
+  sitedate %>%
+  separate(Period, into = c('age', 'error'),
+           sep = "±") %>%
+  mutate(age_ka = parse_number(age) / 1000,
+         error = parse_number(error))
+
+# combine the assemblage with age data
+
+tcsa_site_ass <- tcsa %>% 
+  left_join(mydata_ages) %>% 
+  group_by(Assemblage) %>% 
+ drop_na(age_ka) %>% 
+  add_tally() %>% 
+  mutate(label_site = as.factor(paste0(Assemblage,' (n = ', n, '), (', {round(age_ka,1)}, 'ka)' ))) %>%   
+  dplyr::select(age_ka,
+         min_tcsa,
+         Assemblage, 
+         label_site,
+         n) %>% 
+    distinct_all()
+
+# create column for "n = xx "
+tcsa_raw <- tcsa %>% 
+  group_by(raw_materials_top) %>% 
+  add_tally() %>% 
+  filter(n>1) %>% 
+  mutate(label_raw = as.factor(paste0(raw_materials_top,' (n = ', n, ")"))) %>%   
+  dplyr::select(min_tcsa,
+         raw_materials_top, 
+       label_raw,
+       n) %>% 
+  distinct_all()
+
+#
+mydata_ages_na <- mydata_ages %>%
+  drop_na(age)
+
+# excluded artifact #
+tcsa_omit <- mydata %>%
+  mutate(min_tcsa = 0.5 *Width * Thickness) %>%
+  dplyr::filter (min_tcsa > 250)
+
+# code for numbers in the text
+total_site <- nrow(korean_archaeological_site_locations)
+total_sp <- nrow(mydata)
+total_ass <- nrow(mydata_ages)
+n_sites <- length(unique(tcsa$Site_name))
+min_n_points <- nrow(tcsa_site) 
+# should I get rid of "Other" in the figure?, if then, it's nrow(tcsa_site_num)
+# n_radiocarbon_ages <- NULL # we only have one age per assemblage
+n_sites_with_radiocarbon_ages <- nrow(mydata_ages_na)
+n_omit <- nrow(tcsa_omit)
+n_ass <- length(unique(tcsa$Assemblage))
+n_raw <- nrow(tcsa_raw)
+```
+
+Since the first evidence of the Korean Late Paleolithic was discovered at Seokjangri site, including blades and a stemmed base, around 300 stemmed points have been found in Korea [@Sohn_1967; @lee2019were].
+Most sites contain only a few stemmed points and only a few sites have many more, such as Jingeuneul, Suyanggae (n = 55), and Yongsandong (n = 38) [@Kim_2017].
+Among these stemmed points, we selected those that retained their widest and thickest parts.
+Our dataset contains stemmed points that were found during field surveys as well as those were from sites that were never dated but were associtaed with other Late Palaeoltihic artifacts.
+Applying these sampling criteria resulted in a sample of `r total_sp` stemmed points from `r total_ass` assemblages unearthed from `r total_site` sites spanning the period 44-10 ka (@fig-map).
+The dimensions of the `r total_sp` stemmed points were obtained from published excavation reports and by direct measurements during our visits to the collections of local museums and archaeological institutions in Korea.
+
+We distinguished between multiple assemblages at a single site where numerous excavations have been conducted in different locations at the site, and by different institutions under the same site name.
+For example, Suyanggae site, a registered Korean National Heritage site, has been excavated more than ten times since 1980 by the local university museum and later by archaeological research institutions.
+There are six different excavation locations that range from a few meters apart to a few kilometers.
+Similarly, we identified multiple assemblages in a single excavation or even a trench where archaeological deposits were separated by culturally sterile deposits, or where distinct artifact-bearing stratigraphic units could be identified by major differences in the texture, color, and composition of the sedimentary deposits.
+As a result, in this research we used four assemblages from the Suyanggae site.
+We separated one assemblage among the four by using a different site name, Hajinri, following the convention established the excavators of that location.
+Because Hajinri was from the sixth excavation area, which is 3.5 kilometers apart from the other spots and dated to much earlier (around 42-30 ka) than the other assemblages (around 31-15 ka).
+According to the excavation reports for Hajinri, stemmed points first appeared in Hajinri around 42 ka with the earliest blade and blade cores [@Lee_Woo_Lee_An_Yun_Park_Otani_Kim_Kim_Han_et_al._2018].
+We consider this assemblage as an unusual outlier because the stemmed points and blades from Hajinri are so early and also so highly standardized and refined, which are only found at other sites much later in time.
+We have more confidence in finds from Bonggok, which has the second-earliest dates (41.5 ka) in our collection, as the first appearance of stemmed points.
+Bonggok includes blades or elongate flakes, but without any accompanying blade cores [@park2022did].
+
 ## Investigating patterns in TCSA values
 
-To answer our research questions about the likely uses of stemmed points, we calculated TCSA values for the stemmed points in our sample, and explored the interaction of TCSA values with raw materials, site location, and artefact size.
+To answer our research questions about the likely uses of stemmed points, we calculated TCSA values for the stemmed points in our sample, and explored the interaction of TCSA values with raw materials, and artefact size.
 To examine the temporal patterns in TCSA values we investigated radiocarbon ages associated with the points.
-We then compared those temporal patterns with demographic and environmental proxies based on summed probability distributions (SPD) and annual temperature changes from @park2022did.
+We distinguish the research time period into three based on the major climate event during the Late Pleistocene, Last Glacial Maximum (LGM), and compare with the temporal pattern.
+We explore the relationship between TCSA values and the location of assemblages to explore spatial pattern.
 We followed the TCSA ranges for different weapon-delivery systems that @lombard2022tip and @lombard2021variation created by summarizing @wadley2008segment' study of backed microliths and calibrating existing ranges (@tbl-tcsa-ranges).
-We excluded 12 artifacts from our dataset with TCSA greater than 257 (263-2622) and thus outside of the range of the comparative data.
-For this study we include TCSA values from a total of 161 stemmed points from 32 assemblages unearthed from 26 sites.
+We excluded `r n_omit` artifacts from our dataset with TCSA greater than 250 and thus outside of the range of the comparative data.
+For this study we include TCSA values from a total of `r min_n_points` stemmed points from `r n_ass` assemblages unearthed from `r n_sites` sites.
 
@@
 
+`r n_raw`
+
 ### Weight
@@
 Weight is a reliable discriminator between tools of different sizes and a descriptive attribute which can be determined rapidly and objectively and serve as a criterion of likeness or dissimilarity in projectile point description [@fenenga1953weights; @shea2006origins].
-Excluding nine points that we were not able to measure or obtain the records of their weight, we explored the relationship between TCSA and weight for 152 artifacts.
-
-### Geographical Patterns
-
-```{r}
-
-tcsa_site <- tcsa %>% 
-  group_by(site_name_top) %>% 
-  add_tally() %>% 
-  filter(n>1) %>% 
-  mutate(label_site = as.factor(paste0(site_name_top,' (n = ', n, ")"))) %>%   
-  dplyr::select(min_tcsa,
-         site_name_top, 
-         label_site,
-         n) %>% 
-  distinct_all()
-
-# get rid of 'Other'
-tcsa_site_num <- tcsa_site %>% 
-  filter(site_name_top != 'Other')
-
-###########
-# radiocarbon data per assemblage
-sitedate <- read_excel(here("analysis/data/raw_data/TCSA_site_info.xlsx"))
-
-mydata_ages <-
-  sitedate %>%
-  separate(Period, into = c('age', 'error'),
-           sep = "±") %>%
-  mutate(age_ka = parse_number(age) / 1000,
-         error = parse_number(error))
-
-# combine the assemblage with age data
-
-tcsa_site_ass <- tcsa %>% 
-  left_join(mydata_ages) %>% 
-  group_by(Assemblage) %>% 
- drop_na(age_ka) %>% 
-  add_tally() %>% 
-  mutate(label_site = as.factor(paste0(Assemblage,' (n = ', n, '), (', {round(age_ka,1)}, 'ka)' ))) %>%   
-  dplyr::select(age_ka,
-         min_tcsa,
-         Assemblage, 
-         label_site,
-         n) %>% 
-    distinct_all()
-
-########
-mydata_ages_na <- mydata_ages %>%
-  drop_na(age)
-
-# count radiocarbon ages
-
-n_sites <- length(unique(tcsa$Site_name))
-min_n_points <- nrow(tcsa_site) # should I get rid of "Other" in the figure?, if then, it's nrow(tcsa_site_num)
-# n_radiocarbon_ages <- NULL # we only have one age per assemblage
-n_sites_with_radiocarbon_ages <- nrow(mydata_ages_na)
-```
-
-To explore geographical patterns in TCSA values, we retained only sites that contains more than four stemmed points.
+Excluding points that we were not able to measure or obtain the records of their weight, we explored the relationship between TCSA and weight for 152 artifacts.
+
+### Temporal pattern
+
+We used radiocarbon ages to explore the temporal pattern of the function of stemmed points.
+We excluded assemblages that have no radiocarbon dates.
+We arranged `r n_sites_with_radiocarbon_ages` assemblages dated from 45ka to 14.8ka to explore variation in TCSA over time.
+
+### Spatial Patterns
+
+To explore spatial patterns in TCSA values, we retained only sites that contains more than four stemmed points.
 This resulted in a sample of `r n_sites` sites and `r min_n_points` points.
@@
 LIKE NEAREST NEIGHBOUR DISTANCE OR DISTANCE TO THE COAST OR ELEVATION OR SOME GEOGRAPHICAL VARIABLE THAT IS RELEVANT TO OUR MODEL OF SINGLE-FUNCTION VS MULTI-FUNCTION.
-
-### Temporal pattern
-
-We used radiocarbon ages to explore the temporal pattern of the function of stemmed points.
-We excluded assemblages that have no radiocarbon dates.
-We arranged `r n_sites_with_radiocarbon_ages` assemblages dated from 45ka to 14.8ka to explore variation in TCSA over time.
 
@@
 
-# create column for "n = xx "
-tcsa_raw <- tcsa %>% 
-  group_by(raw_materials_top) %>% 
-  add_tally() %>% 
-  filter(n>1) %>% 
-  mutate(label_raw = as.factor(paste0(raw_materials_top,' (n = ', n, ")"))) %>%   
-  dplyr::select(min_tcsa,
-         raw_materials_top, 
-       label_raw,
-       n) %>% 
-  distinct_all()
-
 # Facet by raw materials
@@
 Overall, there is no clear pattern of TCSA among raw materials.
-
-## Geographical patterns
-
-```{r}
-#| label: fig-tcsa-sites
-#| fig-cap: "Histograms of TCSA values by archaeological site."
-#| fig-width:   4.5
-#| fig-height:  4
-
- ggplot(tcsa_site,
-              aes(min_tcsa, 
-                  fill = n)) +
-  geom_histogram() +
-  theme_bw(base_size = 6) +
-  labs(x = "TCSA", 
-       y = "Count of artefacts") +
-  facet_wrap(~label_site,
-                 scales = "free_y",
-                  ncol = 1) +
-    scale_fill_viridis_c(name = "Number\nof artefacts") +
-  annotate("rect", 
-           xmin = 4,
-           xmax = 18,
-           ymin = 0,
-           ymax = 2.5,
-           fill = "green",
-           alpha = 0.2) +
-      annotate("text", 
-               size = point_function_group_label_size,
-               label = "Poisoned\narrow tips",
-               hjust = 0,
-               lineheight = 0.75,
-           y = (0+2.5)/2,
-           x = 5.5) +
-      annotate("rect", 
-           xmin = 17,
-           xmax = 47,
-           ymin = 2.5,
-           ymax = 5,
-           fill = "blue",
-           alpha = 0.2) +
-      annotate("text", 
-               size = point_function_group_label_size,
-               label = "Un-poisoned\narrow tips",
-               hjust = 0,
-               lineheight = 0.75,
-           y = (2.5+5)/2,
-           x = 28) +
-      annotate("rect", 
-           xmin = 40,
-           xmax = 76,
-           ymin = 5,
-           ymax = 7.5,
-           fill = "orange",
-           alpha = 0.2) +
-      annotate("text", 
-               size = point_function_group_label_size,
-               label = "Dart tips",
-               hjust = 0,
-           y = (5+7.5)/2,
-           x = 53) +
-      annotate("rect", 
-           xmin = 42,
-           xmax = 90,
-           ymin = 7.5,
-           ymax = 10,
-           fill = "purple",
-           alpha = 0.2) +
-      annotate("text", 
-               size = point_function_group_label_size,
-               label = "Javelin tips",
-               hjust = 0,
-           y = (7.5+10)/2,
-           x = 70) +
-     annotate("rect", 
-           xmin = 80,
-           xmax = 200,
-           ymin = 10,
-           ymax = 12.5,
-           fill = "yellow",
-           alpha = 0.2) +
-      annotate("text", 
-               label = "Stabbing spear tips",
-               size = point_function_group_label_size,
-               hjust = 0,
-           y = (10+12.5)/2,
-           x = 160)
-```
-
-We computed artefact TCSA values for `r n_sites` sites that contain more than `r min_n_points` stemmed points to observe variation between sites (@fig-tcsa-sites) Sites with fewer than 4 stemmed points were grouped under the category of 'other'.
-Among the sites in our sample, Suyanggae has the most stemmed points and the widest range of TCSA values.
-This suggests that people made stemmed points for a variety of different purposes in Suyanggae.
-The TCSA values for Nosanri and Sibuk are have a narrower range of lower values, which indicates less diverse functions for stemmed points.
-Wolpyeng and Yongsujaeul have similar amounts of stemmed points to Nosanri and Sibuk, but exhibit wider distributions of TCSA values.
-Yongsandong has the second most stemmed points and TCSA shows a similar distribution to Yongsakul.
 
@@
 
-## Geographical patterns of TCSA values
-
-```{r}
-#| label: fig-tcsa-per-sites
-#| fig-cap: "Histograms of TCSA values by archaeological site."
-#| fig-width:   4.5
-#| fig-height:  4
-
- ggplot(tcsa_site,
-              aes(min_tcsa, 
-                  fill = n)) +
-  geom_histogram() +
-  theme_bw(base_size = 6) +
-  labs(x = "TCSA", 
-       y = "Count of artefacts") +
-  facet_wrap(~label_site,
-                 scales = "free_y",
-                  ncol = 1) +
-    scale_fill_viridis_c(name = "Number\nof artefacts") +
-  annotate("rect", 
-           xmin = 4,
-           xmax = 18,
-           ymin = 0,
-           ymax = 2.5,
-           fill = "green",
-           alpha = 0.2) +
-      annotate("text", 
-               size = 1.5,
-               label = "Poisoned\narrow tips",
-               hjust = 0,
-               lineheight = 0.75,
-           y = (0+2.5)/2,
-           x = 5.5) +
-      annotate("rect", 
-           xmin = 17,
-           xmax = 47,
-           ymin = 2.5,
-           ymax = 5,
-           fill = "blue",
-           alpha = 0.2) +
-      annotate("text", 
-               size = point_function_group_label_size,
-               label = "Un-poisoned\narrow tips",
-               hjust = 0,
-              lineheight = 0.75,
-           y = (2.5+5)/2,
-           x = 28) +
-      annotate("rect", 
-           xmin = 40,
-           xmax = 76,
-           ymin = 5,
-           ymax = 7.5,
-           fill = "orange",
-           alpha = 0.2) +
-      annotate("text", 
-               size = point_function_group_label_size,
-               label = "Dart tips",
-               hjust = 0,
-           y = (5+7.5)/2,
-           x = 53) +
-      annotate("rect", 
-           xmin = 42,
-           xmax = 90,
-           ymin = 7.5,
-           ymax = 10,
-           fill = "purple",
-           alpha = 0.2) +
-      annotate("text", 
-               size = point_function_group_label_size,
-               label = "Javelin tips",
-               hjust = 0,
-           y = (7.5+10)/2,
-           x = 70) +
-     annotate("rect", 
-           xmin = 80,
-           xmax = 200,
-           ymin = 10,
-           ymax = 12.5,
-           fill = "yellow",
-           alpha = 0.2) +
-      annotate("text", 
-               label = "STabbing spear tips",
-               size = point_function_group_label_size,
-               hjust = 0,
-           y = (10+12.5)/2,
-           x = 160)
-```
-
-We computed artefact TCSA values for `r n_sites` sites that contain more than `r min_n_points` stemmed points to observe variation between sites (@fig-tcsa-per-sites) Sites with fewer than `r min_n_points` stemmed points were grouped under the category of 'other'.
-Among the sites in our sample, Suyanggae has the most stemmed points and the widest range of TCSA values.
-This suggests that people made stemmed points for a variety of different purposes in Suyanggae.
-The TCSA values for Nosanri and Sibuk are have a narrower range of lower values, which indicates less diverse functions for stemmed points.
-Wolpyeng and Yongsujaeul have similar amounts of stemmed points to Nosanri and Sibuk, but exhibit wider distributions of TCSA values.
-Yongsandong has the second most stemmed points and TCSA shows a similar distribution to Yongsakul.
-
 ```{r}
@@
    labs(y = "TCSA", 
-        x = "MIS") +
+        x = "Time Period") +
   geom_quasirandom(alpha = 0.4,
@@
 
+summary(aov(min_tcsa ~ as.factor(mis_stage), tcsa_site_ass_mis))
+
 ```
@@
 Overall, TCSA values from assemblages that have more than `r min_n_points` stemmed points vary considerably, which makes it difficult to discern a clear temporal pattern in these assemblages.
+
+- Unclear pattern, but mainly made before LGM
+- Function is not closely related to the period based on p-value but before LGM show wide variation. 
+ - some assemblages show wide variation while the other show narrow variation.
+
+## Spatial patterns
+
+```{r}
+#| label: fig-tcsa-sites
+#| fig-cap: "Histograms of TCSA values by archaeological site."
+#| fig-width:   4.5
+#| fig-height:  4
+
+ ggplot(tcsa_site,
+              aes(min_tcsa, 
+                  fill = n)) +
+  geom_histogram() +
+  theme_bw(base_size = 6) +
+  labs(x = "TCSA", 
+       y = "Count of artefacts") +
+  facet_wrap(~label_site,
+                 scales = "free_y",
+                  ncol = 1) +
+    scale_fill_viridis_c(name = "Number\nof artefacts") +
+  annotate("rect", 
+           xmin = 4,
+           xmax = 18,
+           ymin = 0,
+           ymax = 2.5,
+           fill = "green",
+           alpha = 0.2) +
+      annotate("text", 
+               size = point_function_group_label_size,
+               label = "Poisoned\narrow tips",
+               hjust = 0,
+               lineheight = 0.75,
+           y = (0+2.5)/2,
+           x = 5.5) +
+      annotate("rect", 
+           xmin = 17,
+           xmax = 47,
+           ymin = 2.5,
+           ymax = 5,
+           fill = "blue",
+           alpha = 0.2) +
+      annotate("text", 
+               size = point_function_group_label_size,
+               label = "Un-poisoned\narrow tips",
+               hjust = 0,
+               lineheight = 0.75,
+           y = (2.5+5)/2,
+           x = 28) +
+      annotate("rect", 
+           xmin = 40,
+           xmax = 76,
+           ymin = 5,
+           ymax = 7.5,
+           fill = "orange",
+           alpha = 0.2) +
+      annotate("text", 
+               size = point_function_group_label_size,
+               label = "Dart tips",
+               hjust = 0,
+           y = (5+7.5)/2,
+           x = 53) +
+      annotate("rect", 
+           xmin = 42,
+           xmax = 90,
+           ymin = 7.5,
+           ymax = 10,
+           fill = "purple",
+           alpha = 0.2) +
+      annotate("text", 
+               size = point_function_group_label_size,
+               label = "Javelin tips",
+               hjust = 0,
+           y = (7.5+10)/2,
+           x = 70) +
+     annotate("rect", 
+           xmin = 80,
+           xmax = 200,
+           ymin = 10,
+           ymax = 12.5,
+           fill = "yellow",
+           alpha = 0.2) +
+      annotate("text", 
+               label = "Stabbing spear tips",
+               size = point_function_group_label_size,
+               hjust = 0,
+           y = (10+12.5)/2,
+           x = 160)
+```
+
+We computed artefact TCSA values for `r n_sites` sites that contain more than `r min_n_points` stemmed points to observe variation between sites (@fig-tcsa-per-sites) Sites with fewer than `r min_n_points` stemmed points were grouped under the category of 'other'.
+Among the sites in our sample, Suyanggae has the most stemmed points and the widest range of TCSA values.
+This suggests that people made stemmed points for a variety of different purposes in Suyanggae.
+The TCSA values for Nosanri and Sibuk are have a narrower range of lower values, which indicates less diverse functions for stemmed points.
+Wolpyeng and Yongsujaeul have similar amounts of stemmed points to Nosanri and Sibuk, but exhibit wider distributions of TCSA values.
+Yongsandong has the second most stemmed points and TCSA shows a similar distribution to Yongsakul.
 
@@
 How diverse were their likely uses?
-What are the temporal patterns in stemmed point uses?
+What are the temporal and spatial patterns in stemmed point uses?
 
@@
 
+<--!
+In discussion we then compare those temporal patterns with demographic and environmental proxies based on summed probability distributions (SPD) and annual temperature changes from @park2022did.
+-->
+
 Our models premise that the choice of weapon-tip types reflects the socio-environmental circumstances.
@@
 
-SPACIAL ANAlysis
-@prates2022changes
+SPATIAL ANAlysis @prates2022changes provde strong relationship between the temporal density and spatial distribution of megafauna and Fishtail projectile points, and proposed that this weapon was designed and used for megafauna hunting, contributing to their extinction by exploring changes in projectile point technology, as well as the body mass of potential megafaunal prey, and show that Fishtails were strongly related to the largest extinct megafaunal species.
 

### 2022-10-31 analysis/paper/paper.qmd
Changed numbers to codes

@@
 
-# create column for "n = xx "
+# raw material : create column for "n = xx "
 tcsa_raw <- tcsa %>% 
@@
 
-#
+# remove NAs for weight
+tcsa_mass_no_na <-
+  tcsa %>%
+ drop_na(`Weight(g)`)
+
+# drop na for dates
 mydata_ages_na <- mydata_ages %>%
@@
 min_n_points <- nrow(tcsa_site) 
-# should I get rid of "Other" in the figure?, if then, it's nrow(tcsa_site_num)
 # n_radiocarbon_ages <- NULL # we only have one age per assemblage
@@
 n_raw <- nrow(tcsa_raw)
+n_size <- nrow(tcsa_mass_no_na)
+n_spatial <-nrow(tcsa_site_num) # got rid of "Other" in the dataframe
+n_spatial_ass <- length(unique(tcsa_site_num$Assemblage))
 ```
@@
 Then, finer grained materials were newly added to assemblages for producing the newly introduced tools [@seong2004quartzite].
-We analyzed TCSA values for each raw material type to examine the interaction of raw material types and TCSA values.
-
-`r n_raw`
+We analyzed TCSA values for each raw material type with `r n_raw` artifacts to examine the interaction of raw material types and TCSA values.
 
@@
 Weight is a reliable discriminator between tools of different sizes and a descriptive attribute which can be determined rapidly and objectively and serve as a criterion of likeness or dissimilarity in projectile point description [@fenenga1953weights; @shea2006origins].
-Excluding points that we were not able to measure or obtain the records of their weight, we explored the relationship between TCSA and weight for 152 artifacts.
+We then explored the relationship between weight and raw materials on TCSA values.
+Excluding points that we were not able to measure or obtain the records of their weight, we explored the relationship between TCSA and weight for `r n_size` artifacts.
 
@@
 We used radiocarbon ages to explore the temporal pattern of the function of stemmed points.
-We excluded assemblages that have no radiocarbon dates.
-We arranged `r n_sites_with_radiocarbon_ages` assemblages dated from 45ka to 14.8ka to explore variation in TCSA over time.
+After excluding assemblages that have no radiocarbon dates, we arranged `r n_sites_with_radiocarbon_ages` assemblages dated from 45ka to 14.8ka to explore variation in TCSA over time.
+The we divided the artifacts into three groups based on LGM: before LGM, LGM, and after LGM to examine the impact of this major climate event on the material culture.
 
@@
 
-To explore spatial patterns in TCSA values, we retained only sites that contains more than four stemmed points.
-This resulted in a sample of `r n_sites` sites and `r min_n_points` points.
-We summarised the distribution of TCSA values for each site to identify variation that might be influenced b environmental differences relating to the different locations of the sites.
-I THINK WE NEED SOMETHING MORE IN HERE.
-LIKE NEAREST NEIGHBOUR DISTANCE OR DISTANCE TO THE COAST OR ELEVATION OR SOME GEOGRAPHICAL VARIABLE THAT IS RELEVANT TO OUR MODEL OF SINGLE-FUNCTION VS MULTI-FUNCTION.
+We sat sites that contains fewer than 5 stemmed points aside and named them as "Other".
+This resulted in a sample of `r n_spatial_ass` assemblages and `r n_spatial_ass` points.
+We summarized the distribution of TCSA values for each site to examine spatial patterns in TCSA values.
+We then explore possible environmental impact by connecting stemmed points on the map with eco-regional zones.
+We used QGIS 3.28.0 'Firenze' (https://qgis.org/en/site/index.html) to draw a map based on the ecological study [CITE RESEARCH OF FINAL CHOICE] and used longitudes and latitudes of assemblages to locate the individual artifacts.
 
@@
 This may be due to low levels of uncertainty in the forager's social and physical environments.
-On the other hand, a wide TCSA range may indicate that stemmed points were multifunctional tools, suggesting that people were responding to unfamiliar situations, such as moving into an unfamiliar landscape or unpredictable changes in patch productivity and travel times \[CITE PAPERS ON ARCHAEOLOGY AND PATCH CHOICE MODEL\].
+On the other hand, a wide TCSA range may indicate that stemmed points were multifunctional tools, suggesting that people were responding to unfamiliar situations, such as moving into an unfamiliar landscape or unpredictable changes in patch productivity and travel times [@bird2006behavioral; @bettinger2016marginal; @kelly2007foraging].
 We expect to observe different TCSA ranges, reflecting morphologically and functionally different stemmed points, across different regions or environments in Korea, and over time.
@@
 library(Ckmeans.1d.dp)
-
-# remove NAs
-tcsa_mass_no_na <-
-  tcsa %>%
- drop_na(`Weight(g)`)
 

### 2022-10-31 analysis/paper/paper.qmd
edit texts & added citations

@@
   dplyr::filter(min_tcsa <= 250) %>% #omit very large TCSA, could be measurement error
-  # group rare raw materials > 10
+  # group rare raw materials < 10
   mutate(raw_materials_top = fct_lump_min(Raw_material, 10)) %>%
@@
 n_size <- nrow(tcsa_mass_no_na)
-n_spatial <-nrow(tcsa_site_num) # got rid of "Other" in the dataframe
-n_spatial_ass <- length(unique(tcsa_site_num$Assemblage))
+# n_spatial <-nrow(tcsa_site_num) # got rid of "Other" in the dataframe
+# n_spatial_ass <- length(unique(tcsa_site_num$Assemblage))
 ```
@@
 We analyzed TCSA values for each raw material type with `r n_raw` artifacts to examine the interaction of raw material types and TCSA values.
+We categorized rare raw materials, which have less than 10 artifacts, as "Other".
 
@@
 
-We sat sites that contains fewer than 5 stemmed points aside and named them as "Other".
-This resulted in a sample of `r n_spatial_ass` assemblages and `r n_spatial_ass` points.
 We summarized the distribution of TCSA values for each site to examine spatial patterns in TCSA values.
@@
 We used QGIS 3.28.0 'Firenze' (https://qgis.org/en/site/index.html) to draw a map based on the ecological study [CITE RESEARCH OF FINAL CHOICE] and used longitudes and latitudes of assemblages to locate the individual artifacts.
+Among `r n_sites` artifacts, we sat sites that contains fewer than 5 stemmed points aside and named them as "Other".
 
@@
 The other raw materials including hornfels, rhyolite, and tuff show a wide distribution, similar to shale.
-The category of 'other' raw materials include porphyry, trachyte, felsite, chert, quartz, quartzite, granite, mudstone, and unidentified rocks.
+The category of "Other" raw materials include porphyry, trachyte, felsite, chert, quartz, quartzite, granite, mudstone, and unidentified rocks.
 Overall, there is no clear pattern of TCSA among raw materials.
 
-## Variation in TCSA values by artefact size
+## Variation in TCSA values by artifact size
 
@@
 Using weight as a size proxy we examined the relationship between size and likely use of the stemmed points inferred from TCSA values.
-We conducted a univariate cluster analysis INSERT CITATION TO PAPER ON THIS METHOD of stemmed points by weight and it revealed three clusters of artefact sizes (A of @fig-artifact-size-cluster, mean = `r round(mean(tcsa_mass_no_na$'Weight(g)'), digits=1)`, SD = `r round(sd(tcsa_mass_no_na$'Weight(g)'), digits=1)`).
+We conducted a univariate cluster analysis [@song2020efficient; @wang2011ckmeans] of stemmed points by weight and it revealed three clusters of artefact sizes (A of @fig-artifact-size-cluster, mean = `r round(mean(tcsa_mass_no_na$'Weight(g)'), digits=1)`, SD = `r round(sd(tcsa_mass_no_na$'Weight(g)'), digits=1)`).
 Cluster 1, the smallest (lightest) artifacts, have a lower range of TCSA values, compared to Cluster 2 (B of @fig-artifact-size-cluster).

### 2022-10-31 analysis/paper/paper.qmd
Edit before spatial pattern

@@
 After excluding assemblages that have no radiocarbon dates, we arranged `r n_sites_with_radiocarbon_ages` assemblages dated from 45ka to 14.8ka to explore variation in TCSA over time.
-The we divided the artifacts into three groups based on LGM: before LGM, LGM, and after LGM to examine the impact of this major climate event on the material culture.
+Then we divided the artifacts into three groups based on LGM: before LGM, LGM, and after LGM to examine the impact of this major climate event on the material culture.
 
@@
 We then explore possible environmental impact by connecting stemmed points on the map with eco-regional zones.
-We used QGIS 3.28.0 'Firenze' (https://qgis.org/en/site/index.html) to draw a map based on the ecological study [CITE RESEARCH OF FINAL CHOICE] and used longitudes and latitudes of assemblages to locate the individual artifacts.
+We used QGIS 3.22.12 'Białowieża' (https://qgis.org/en/site/index.html) to draw a map based on the ecological study [CITE RESEARCH OF FINAL CHOICE] and used longitudes and latitudes of assemblages to locate the individual artifacts.
 Among `r n_sites` artifacts, we sat sites that contains fewer than 5 stemmed points aside and named them as "Other".
@@
 Using weight as a size proxy we examined the relationship between size and likely use of the stemmed points inferred from TCSA values.
-We conducted a univariate cluster analysis [@song2020efficient; @wang2011ckmeans] of stemmed points by weight and it revealed three clusters of artefact sizes (A of @fig-artifact-size-cluster, mean = `r round(mean(tcsa_mass_no_na$'Weight(g)'), digits=1)`, SD = `r round(sd(tcsa_mass_no_na$'Weight(g)'), digits=1)`).
+We conducted a univariate cluster analysis [@song2020efficient; @wang2011ckmeans] of stemmed points by weight and it revealed three clusters of artifact sizes (A of @fig-artifact-size-cluster, mean = `r round(mean(tcsa_mass_no_na$'Weight(g)'), digits=1)`, SD = `r round(sd(tcsa_mass_no_na$'Weight(g)'), digits=1)`).
 Cluster 1, the smallest (lightest) artifacts, have a lower range of TCSA values, compared to Cluster 2 (B of @fig-artifact-size-cluster).
@@
 
+We arranged `r n_sites_with_radiocarbon_ages` assemblages in chronological order (@fig-tcsa-radio-carbon).
+Our analysis shows that earlier assemblages before LGM  tend to possess more stemmed points, such as Suyanggae, and these have a wide range of TCSA values, indicating multiple likely usages of stemmed points.
+Overall, there is no clear pattern in these assemblages over time.
+Depending on the assemblage, there may be wide variation or narrow variation.
+
 ```{r}
 #| label: fig-tcsa-radio-carbon
-#| fig-cap: "Distribution of TCSA values over time"
+#| fig-cap: "Distribution of TCSA values over time. The light gray shade indicates Marine isotope stages (MIS) 3, the gray shade indicates MIS 2 and the darkest shade indicates LGM period."
 #| fig-width:   4.5
@@
         x = "") +
-  theme(axis.text.x = element_text(size = 4)) +
+  theme(axis.text.x = element_text(size = 6)) +
   scale_x_discrete(guide = guide_axis(angle = 45))
@@
 ```{r}
+
+#| label: fig-tcsa-radio-carbon-box
+#| fig-cap: "Distribution of TCSA values grouped by LGM event"
+#| fig-width:   4.5
+#| fig-height:  4
 
@@
 
-summary(aov(min_tcsa ~ as.factor(mis_stage), tcsa_site_ass_mis))
-
-```
-
-We arranged `r n_sites_with_radiocarbon_ages` assemblages in chronological order (@fig-tcsa-radio-carbon).
-Our analysis shows that early assemblages (i.e.
-45-24 ka) tend to possess more stemmed points, such as Suyanggae, and these have a wide range of TCSA values, indicating multiple likely usages of stemmed points.
-After about 38 ka stemmed points become less common in archaeological assemblages.
-Overall, TCSA values from assemblages that have more than `r min_n_points` stemmed points vary considerably, which makes it difficult to discern a clear temporal pattern in these assemblages.
-
-- Unclear pattern, but mainly made before LGM
-- Function is not closely related to the period based on p-value but before LGM show wide variation. 
- - some assemblages show wide variation while the other show narrow variation.
+p_lgm <- round(summary(aov(min_tcsa ~ as.factor(mis_stage), tcsa_site_ass_mis))[[1]][[5]][1], digits = 2)
+
+```
+We then grouped stemmed points into three categories based on LGM event: before, LGM, and after to examine possibility of environmental impact on the use of the artifacts (@fig-tcsa-radio-carbon-box).
+Our results shows that stemmed points were made predominantly before LGM and only a few left after LGM.
+Each category shows wide variation in LGM, which indicates diverse uses of stemmed points.
+There is no strong correlation between the TCSA values and the time period (p-value = `r p_lgm`).
 
@@
 
-ggmap(map)  +
-  geom_jitter(data = tcsa_lgm,
-             aes(long_dd.y ,
-                 lat_dd.y,
-                 colour = min_tcsa),
-             size = 3,
-             width = 0.25, 
-             height = 0.25) +
-  scale_color_viridis_c(option = "A") +
-  theme_minimal(base_size = 6) +
-  labs(x = "Longitute",
-       y = "Latitude") +
-  theme(axis.title.x = element_blank(),
-        axis.title.y = element_blank())
+# ggmap(map)  +
+#   geom_jitter(data = tcsa_lgm,
+#              aes(long_dd.y ,
+#                  lat_dd.y,
+#                  colour = min_tcsa),
+#              size = 3,
+#              width = 0.25, 
+#              height = 0.25) +
+#   scale_color_viridis_c(option = "A") +
+#   theme_minimal(base_size = 6) +
+#   labs(x = "Longitute",
+#        y = "Latitude") +
+#   theme(axis.title.x = element_blank(),
+#         axis.title.y = element_blank())
 
@@
 
-#sf_use_s2(FALSE)
+sf_use_s2(FALSE) # to prevent error. Check here for more info: https://github.com/r-spatial/sf/issues/1762
 

### 2022-10-31 analysis/paper/paper.qmd
Trimed gis map

@@
 ggplot() +
-  geom_sf(data = veg_all, aes(fill = veg_all$ecoregion)) + #add color for vegetation polygons
-  geom_sf(data = pts_all, aes(color = pts_all$min_tcsa)) +
-  theme_minimal()
+  geom_sf(data = veg_all, aes(fill = ecoregion)) + #add color for vegetation polygons
+  geom_sf(data = pts_all, aes(color = min_tcsa)) +
+  theme_minimal() +
+  scale_fill_brewer(palette="Greys") +
+  labs(fill="Ecoregion", color="TCSA")
 

### 2022-11-01 analysis/paper/paper.qmd
edit GIS maps

--- 2022-10-31 analysis/paper/paper.qmd
@@
 
-ggplot() +
-  geom_sf(data = veg_all, aes(fill = veg_all$ecoregion)) + #add color for vegetation polygons
-  geom_sf(data = pts_all, aes(color = pts_all$min_tcsa)) +
-  theme_minimal()
-
-ggplot(points_in_veg_zones_all) +
+eco_plot <- ggplot() +
+  geom_sf(data = veg_all, aes(fill = "grey")) +
+   geom_sf_text(data = veg_all, aes(label = ecoregion), size =3.5)+
+  geom_sf(data = pts_all, aes(color = min_tcsa)) +
+  theme_minimal() +
+  scale_fill_brewer(palette="Greys") +
+  labs(color="TCSA") +
+  guides(fill = FALSE) +
+  theme(axis.text.x = element_text(size = 6))
+
+eco_box <- ggplot(points_in_veg_zones_all) +
   aes(ecoregion,
@@
 
-ggplot() +
-  geom_sf(data = veg_forest, aes(fill = Zone)) + #add color for vegetation polygons
+for_plot <- ggplot() +
+  geom_sf(data = veg_forest, aes(fill = "grey")) +
+  geom_sf_text(data = veg_forest, aes(label = Zone), size =3.5)+
   geom_sf(data = pts_forest,size = 3, aes(color = min_tcsa)) +
-  theme_minimal()
+  theme_minimal()+
+  scale_fill_brewer(palette="Greys") +
+  labs(color="TCSA") +
+  guides(fill = FALSE)
 
@@
 
-ggplot(points_in_veg_zones_forest_box) +
+for_plot <- ggplot(points_in_veg_zones_forest_box) +
   aes(Zone,
@@
 
+plot_grid(eco_plot, for_plot, eco_box, for_plot, labels = c('A', 'C', 'B', 'D'))
 ```

### 2022-11-01 analysis/paper/paper.qmd
combined gis maps

--- 2022-10-31 analysis/paper/paper.qmd
@@
 
-# data only for LGM period
-tcsa_lgm <- korean_archaeological_site_locations_tcsa %>% 
-  filter(age< 26.5 & age> 19)
-
-# ggmap(map)  +
-#   geom_jitter(data = tcsa_lgm,
-#              aes(long_dd.y ,
-#                  lat_dd.y,
-#                  colour = min_tcsa),
-#              size = 3,
-#              width = 0.25, 
-#              height = 0.25) +
-#   scale_color_viridis_c(option = "A") +
-#   theme_minimal(base_size = 6) +
-#   labs(x = "Longitute",
-#        y = "Latitude") +
-#   theme(axis.title.x = element_blank(),
-#         axis.title.y = element_blank())
-
-library(sf)
-
-veg <- read_sf(here("analysis/data/gis_data/vegetation_polygons.shp"))
-
-# st_crs(veg)
-
-veg <- st_transform(veg, crs = st_crs(4162))
-
-# take a look
-ggplot() +
-  geom_sf(data = veg) +
-  coord_sf()
-
-# BM: basic plan is the do a spatial join with the veg zones and the TCSA values.
-Then table of artefacts will get a new column telling us what veg zone it is in.
-Then we can make a plot to compare the distribution of tcsa values across the different veg zones
-
-# Determine which points fall inside which polygons
-
-# Convert to class 'sf'
-pts <- st_as_sf(tcsa_lgm,
-                coords = c(   "long_dd.y", "lat_dd.y"),
-                crs = st_crs(veg))
-
-# take a look
-ggplot() +
-  geom_sf(data = pts)
-
-points_in_veg_zones <- 
-st_join(veg, 
-        pts, 
-        join = st_contains)
-
-ggplot() +
-  geom_sf(data = veg, aes(fill = veg$Vegetation)) + #add color for vegetation polygons
-  geom_sf(data = pts, aes(color = pts$min_tcsa)) +
-  theme_minimal()
-
-# fill=c("#CCCCCC", "#FFFFFF", "#999999", "#666666")
-
-ggplot(points_in_veg_zones) +
-  aes(Vegetation,
-      min_tcsa) +
-  geom_boxplot() +
-  geom_quasirandom()
-
-summary(aov(min_tcsa ~ as.factor(Vegetation), points_in_veg_zones))
-
 ############ explore tcsa with MODERN eco_regions #######
@@
 
-ggplot() +
-  geom_sf(data = veg_all) +
-  coord_sf()
+# ggplot() +
+#   geom_sf(data = veg_all) +
+#   coord_sf()
 
@@
                 crs = st_crs(veg_all))
-ggplot() +
-  geom_sf(data = pts_all)
+# ggplot() +
+#   geom_sf(data = pts_all)
 
@@
 
-ggplot() +
-  geom_sf(data = veg_all, aes(fill = veg_all$ecoregion)) + #add color for vegetation polygons
-  geom_sf(data = pts_all, aes(color = pts_all$min_tcsa)) +
-  theme_minimal()
-
-ggplot(points_in_veg_zones_all) +
+eco_plot <- ggplot() +
+  geom_sf(data = veg_all, aes(fill = "grey")) +
+   geom_sf_text(data = veg_all, aes(label = ecoregion), size =3.5)+
+  geom_sf(data = pts_all, aes(color = min_tcsa)) +
+  theme_minimal() +
+  scale_fill_brewer(palette="Greys") +
+  labs(color="TCSA") +
+  guides(fill = FALSE) +
+  theme(axis.text.x = element_text(size = 6))+
+  theme(axis.title.x = element_blank(),
+          axis.title.y = element_blank())
+
+eco_box <- ggplot(points_in_veg_zones_all) +
   aes(ecoregion,
@@
   geom_boxplot() +
-  geom_quasirandom()
+  geom_quasirandom() +
+  xlab("Vegetation") + 
+  ylab("TCSA")
 
@@
 
-ggplot() +
-  geom_sf(data = veg_forest) +
-  coord_sf()
+# ggplot() +
+#   geom_sf(data = veg_forest) +
+#   coord_sf()
 
@@
                 crs = st_crs(veg_forest))
-ggplot() +
-  geom_sf(data = pts_forest)
+# ggplot() +
+#   geom_sf(data = pts_forest)
 
@@
 
-ggplot() +
-  geom_sf(data = veg_forest, aes(fill = Zone)) + #add color for vegetation polygons
+for_plot <- ggplot() +
+  geom_sf(data = veg_forest, aes(fill = "grey")) +
+  geom_sf_text(data = veg_forest, aes(label = Zone), size =3.5)+
   geom_sf(data = pts_forest,size = 3, aes(color = min_tcsa)) +
-  theme_minimal()
+  theme_minimal()+
+  scale_fill_brewer(palette="Greys") +
+  labs(color="TCSA") +
+  guides(fill = FALSE) +
+  theme(axis.text.x = element_text(size = 6)) +
+  theme(axis.title.x = element_blank(),
+          axis.title.y = element_blank())
 
@@
 
-ggplot(points_in_veg_zones_forest_box) +
+for_box <- ggplot(points_in_veg_zones_forest_box) +
   aes(Zone,
@@
   geom_boxplot() +
-  geom_quasirandom()
+  geom_quasirandom() +
+  xlab("Ecoregion") + 
+  ylab("TCSA")
 
@@
 
+plot_grid(eco_plot, for_plot, eco_box, for_box, labels = c('A', 'C', 'B', 'D'))
 ```

### 2022-11-09 analysis/paper/paper.qmd
added references related to GIS map
tried Tukey's HSD

@@
 
-summary(aov(min_tcsa ~ as.factor(ecoregion), points_in_veg_zones_all))
+eco_aov <- aov(min_tcsa ~ as.factor(ecoregion), points_in_veg_zones_all)
+summary(eco_aov)
+eco_tur<- TukeyHSD(eco_aov)
+# plot(eco_tur)
 
@@
 
-summary(aov(min_tcsa ~ as.factor(Zone), points_in_veg_zones_forest_box))
+for_aov <- aov(min_tcsa ~ as.factor(Zone), points_in_veg_zones_forest_box)
+summary(for_aov)
+for_tur<- TukeyHSD(for_aov)
+# plot(for_tur)
 
@@
 ```
+
+Map A. @yim1975distribution, modified by @yi2010vegetation
+
+Map B. @lee2008relationship
 

### 2022-11-09 analysis/paper/paper.qmd
Edited discussion

@@
 Some sites show a wide range of TCSA values that represent multi-purpose usage of stemmed points.
-The temporal pattern of TCSA values is one of little change throughout the Late Paleolithic period.<!-- SP were mainly used before Last Glacial Maximum (LGM).
-SP was mainly located in certain latitude.
-why?
---> We conclude that stemmed points were  multi-functional tools, with many likely designed for javelin and stabbing spear tips.
+The temporal pattern of TCSA values is one of little change throughout the Late Paleolithic period but SP were mainly used before Last Glacial Maximum (LGM).
+We observed that SP was located in certain ecoregions in Korea but no clear functional pattern is confirmed.
+We conclude that stemmed points were  multi-functional tools, with many likely designed for javelin and stabbing spear tips.
+
 keywords: |
@@
 
-We summarized the distribution of TCSA values for each site to examine spatial patterns in TCSA values.
+We summarized the distribution of TCSA values for each assemblage to examine spatial patterns in TCSA values.
 We then explore possible environmental impact by connecting stemmed points on the map with eco-regional zones.
-We used QGIS 3.22.12 'Białowieża' (https://qgis.org/en/site/index.html) to draw a map based on the ecological study [CITE RESEARCH OF FINAL CHOICE] and used longitudes and latitudes of assemblages to locate the individual artifacts.
+We used QGIS 3.22.12 'Białowieża' (https://qgis.org/en/site/index.html) to draw a map based on the previous ecological studies [@yim1975distribution; @yi2010vegetation; @lee2008relationship] and located the individual artifacts using longitudes and latitudes of assemblages.
 Among `r n_sites` artifacts, we sat sites that contains fewer than 5 stemmed points aside and named them as "Other".
@@
 ```
+
 We then grouped stemmed points into three categories based on LGM event: before, LGM, and after to examine possibility of environmental impact on the use of the artifacts (@fig-tcsa-radio-carbon-box).
@@
 ```{r}
-#| label: fig-tcsa-sites
+#| label: fig-tcsa-per-sites
 #| fig-cap: "Histograms of TCSA values by archaeological site."
@@
 ```{r}
+#| label: spa-gis-tcsa-sites
+#| fig-cap: "TCSA values by assemblages and ecoregions.
+A.
+Modern vegetation map defined by temporature and precipitation (modified the map from @yi2010vegetation, but original source is @yim1975distribution).
+This map include 3 zones.
+Central Temperature Zone (CT), South Temperature Zone (ST), and Subtropical-warm Temperature Zone (SW).
+The individual points indicate assemblages.
+B.
+Boxplots for  vegitation zones.
+The individual points indicate artifacts.
+C.
+Ecoregion map based on geographical boundaries and ecological conditions (modified the map from @lee2008relationship).
+The individual points indicate assemblages.
+This map include 14 zones.
+Imjin river basin (IRB), Metropolitan (MP), Central inland (CI), Kangwon coastal (KC), Choongnam coastal (CC), Southwestern inland (SWI), Upper Nagdong river basin (UNRB), WoolYoung coastal (WYC), Western Cholla (WC), Southern mountain (SM), Southeastern inland (SEI), Hyungsan Taewha coastal (HTC), Western south coastal (WSC), and Eastern south coastal (ESC).
+The individual points indicate assemblages.
+D.
+Boxplots for  ecoregion zones, excludings zones with no artifacts.
+The individual points indicate artifacts."
+#| fig-width:   5
+#| fig-height:  4
 
@@
 
+library(sf)
 veg_all <- read_sf(here("analysis/data/gis_data/eco_re_two.shp"))
@@
 
+eco_lgm <- round(summary(eco_aov)[[1]][[5]][1], digits = 3)
+
 ############ explore tcsa with MODERN eco_regions & forest zones #######
@@
   geom_sf_text(data = veg_forest, aes(label = Zone), size =3.5)+
-  geom_sf(data = pts_forest,size = 3, aes(color = min_tcsa)) +
+  geom_sf(data = pts_forest, aes(color = min_tcsa)) +
   theme_minimal()+
@@
 
+for_lgm <- round(summary(for_aov)[[1]][[5]][1], digits = 3)
+
 plot_grid(eco_plot, for_plot, eco_box, for_box, labels = c('A', 'C', 'B', 'D'))
 ```
-
-Map A. @yim1975distribution, modified by @yi2010vegetation
-
-Map B. @lee2008relationship
-
-# Discussion ORDER OF TOPICS HERE MUST FOLLOW EXACTLY THE ORDER OF THE METHODS AND RESULTS SECTION
+We then located individual assemblages on maps that show ecoregional boundaries to explore spatial patterns.
+We used two maps from the previous studies that represents different zones based on vegetation and geography [@yi2010vegetation; @yim1975distribution; @lee2008relationship].
+Our results show that stemmed points are only located in certain zones, Central Temperature Zone (CT) or Central inland (CI) (@spa-gis-tcsa-sites), for example.
+We observed that TCSA values are stonely accociated with the zones (p-value for Vegetation zones = `r eco_lgm`, p-value for Ecoregion zones = `r for_lgm`).
+However, there is no clear functional pattern throughout the zones (@spa-gis-tcsa-sites panel B and D).
+There is a wide variation in TCSA regardless of the stemmed points density in each zone.
+
+# Discussion
 
@@
 There are a few stemmed points in our sample that may have been used as dart tips and un-poisoned arrow tips, but none of them have TCSA values in the range of poison arrow tips (@fig-tcsa-all-sp).
-We found a wide range of TCSA values in our sample, indicating that stemmed points played a diverse role in forager toolkits.
-Some sites, such as Nosanri and Sinbuk, likely used stemmed points as restricted-function tools.
-Meanwhile, stemmed points were used for a wider set of tasks at other sites, especially Suyanggae.
-In general, stemmed points played multiple roles in more sites (@fig-tcsa-sites).
-Through the Late Paleolithic period, TCSA values showed little directional change (@fig-tcsa-radio-carbon).
-The chronological sequence of TCSA varies from region to region with no clear pattern.
-Most stemmed points appear to be have designed for use as javelin and stabbing spear tips, but there are also large numbers of tools that likely could fulfill multiple roles as armatures.
-
-TCSA can, however, be impacted by other factors such as raw materials and portability.
+But in general, we found a wide range of TCSA in our sample, indicating that stemmed points played a diverse role in forager toolkits.
+TCSA can be impacted by other factors such as raw materials and portability.
 -@eren2022plains explain that smaller TCSA could be the result of pursuing production economy and transport efficiency.
 We explore the relationship between TCSA and raw material types and weight in order to examine these factors.
-Weight plays an important role in portability and penetration of projectiles [@hughes1998getting].
-According to our results, raw materials do not have a clear pattern, but weight is correlated with raw materials (@fig-tcsa-raw-materials and @fig-tcsa-size).
-Different clusters of weight are matched with different types of weapon tips (@fig-artifact-size-cluster).
-Based on these analyses, we assume that people were able to produce the ideal size tools regardless of raw materials.
+We observed no clear pattern for raw materials (@fig-tcsa-raw-materials) but weight (@fig-artifact-size-cluster).
+Interestingly, weight is correlated with raw materials (@fig-tcsa-size).
+Therefore we could assume that raw materials also associated with TCSA value.
+We found that different clusters of weight are matched with different types of weapon tips (@fig-artifact-size-cluster).
+The TCSA varies from assemblages to assemblages without any discernible temporal pattern in functional use (@fig-tcsa-radio-carbon).
+But the stemmed points mainly used before LGM (@fig-tcsa-radio-carbon-box).
+We found that stemmed points are only located in certain ecoregional zones such as Central Temperature zone without any clear pattern in function (@spa-gis-tcsa-sites).
+Both temporal and spatial analyses show that stemmed points were used for a particular environment, not so much harsh as LGM.
 
 Our models premise that the choice of weapon-tip types reflects the socio-environmental circumstances.
-To examine the possible connections between the roles of stemmed points and population dynamics or environmental change during the Late Paleolithic period, we compared the temporal pattern of TCSA to results of Summed probability distribution (SPD) and simulated Mean Annual Temperature (MAT) from @park2022did.
-In those two analyses, -@park2022did covered assemblages with and without stemmed points, which is useful for us to explore the overall socio-environmental dynamics.
-As we discussed, there is no clear pattern in the temporal or regional pattern of TCSA (@fig-tcsa-radio-carbon or @no-figure-no-more: A).
+To examine the possible connections between the roles of stemmed points and population dynamics or environmental change during the Late Paleolithic period, we compared the temporal and spatial patterns of TCSA to results of simulated Mean Annual Temperature (MAT) and Summed probability distribution (SPD) from @park2022did.
+We found clear correlation between stemmed points and temperature (@spa-gis-tcsa-sites and @fig-tcsa-radio-carbon-box, and Figure 9 from @park2022did)
+around 4 to 6 ℃
+
  We consider that TCSA, or the function of the stemmed points, are strongly dependent on the surrounding environment or their situation.
@@
 
-<--!
+\<--!
 In discussion we then compare those temporal patterns with demographic and environmental proxies based on summed probability distributions (SPD) and annual temperature changes from @park2022did.
--->
+--\>
 

### 2022-11-09 analysis/paper/paper.qmd
edit discussion

@@
 To examine the possible connections between the roles of stemmed points and population dynamics or environmental change during the Late Paleolithic period, we compared the temporal and spatial patterns of TCSA to results of simulated Mean Annual Temperature (MAT) and Summed probability distribution (SPD) from @park2022did.
-We found clear correlation between stemmed points and temperature (@spa-gis-tcsa-sites and @fig-tcsa-radio-carbon-box, and Figure 9 from @park2022did)
-around 4 to 6 ℃
-
+
+Stemmed points were most commonly used in environments with a temperature between 4 and 6 ℃ (@spa-gis-tcsa-sites and @fig-tcsa-radio-carbon-box, and Figure 9 from @park2022did).
+
+\<--!upto here --\>
  We consider that TCSA, or the function of the stemmed points, are strongly dependent on the surrounding environment or their situation.
@@
 
-\<--!
-In discussion we then compare those temporal patterns with demographic and environmental proxies based on summed probability distributions (SPD) and annual temperature changes from @park2022did.
---\>
-
 Our models premise that the choice of weapon-tip types reflects the socio-environmental circumstances.

### 2022-11-18 analysis/paper/paper.qmd
split code for tukey results from gis code chunk

@@
 ```{r}
-#| label: spa-gis-tcsa-sites
+#| label: fig-gis-tcsa-sites
 #| fig-cap: "TCSA values by assemblages and ecoregions.
@@
 
-eco_aov <- aov(min_tcsa ~ as.factor(ecoregion), points_in_veg_zones_all)
-summary(eco_aov)
-eco_tur<- TukeyHSD(eco_aov)
-# plot(eco_tur)
-
-eco_lgm <- round(summary(eco_aov)[[1]][[5]][1], digits = 3)
-
 ############ explore tcsa with MODERN eco_regions & forest zones #######
@@
 
+plot_grid(eco_plot, for_plot, eco_box, for_box, labels = c('A', 'C', 'B', 'D'))
+```
+```{r}
+#| label: fig-tukey-eco-region
+#| fig-cap: "."
+#| fig-width:   5
+#| fig-height:  4
+
+#vegetation region#
+eco_aov <- aov(min_tcsa ~ as.factor(ecoregion), points_in_veg_zones_all)
+summary(eco_aov)
+eco_tur<- TukeyHSD(eco_aov)
+# plot(eco_tur)
+
+eco_lgm <- round(summary(eco_aov)[[1]][[5]][1], digits = 3)
+
+#eco region#
 for_aov <- aov(min_tcsa ~ as.factor(Zone), points_in_veg_zones_forest_box)
@@
 
-plot_grid(eco_plot, for_plot, eco_box, for_box, labels = c('A', 'C', 'B', 'D'))
-```
+```
+
 We then located individual assemblages on maps that show ecoregional boundaries to explore spatial patterns.
 We used two maps from the previous studies that represents different zones based on vegetation and geography [@yi2010vegetation; @yim1975distribution; @lee2008relationship].
-Our results show that stemmed points are only located in certain zones, Central Temperature Zone (CT) or Central inland (CI) (@spa-gis-tcsa-sites), for example.
-We observed that TCSA values are stonely accociated with the zones (p-value for Vegetation zones = `r eco_lgm`, p-value for Ecoregion zones = `r for_lgm`).
-However, there is no clear functional pattern throughout the zones (@spa-gis-tcsa-sites panel B and D).
+Our results show that stemmed points are only located in certain zones, Central Temperature Zone (CT) or Central inland (CI) (@fig-gis-tcsa-sites), for example.
+We observed that TCSA values are stonely associated with the zones (p-value for Vegetation zones = `r eco_lgm`, p-value for Ecoregion zones = `r for_lgm`).
+However, there is no clear functional pattern throughout the zones (@fig-gis-tcsa-sites panel B and D).
 There is a wide variation in TCSA regardless of the stemmed points density in each zone.
@@
 But the stemmed points mainly used before LGM (@fig-tcsa-radio-carbon-box).
-We found that stemmed points are only located in certain ecoregional zones such as Central Temperature zone without any clear pattern in function (@spa-gis-tcsa-sites).
+We found that stemmed points are only located in certain ecoregional zones such as Central Temperature zone without any clear pattern in function (@fig-gis-tcsa-sites).
 Both temporal and spatial analyses show that stemmed points were used for a particular environment, not so much harsh as LGM.
@@
 To examine the possible connections between the roles of stemmed points and population dynamics or environmental change during the Late Paleolithic period, we compared the temporal and spatial patterns of TCSA to results of simulated Mean Annual Temperature (MAT) and Summed probability distribution (SPD) from @park2022did.
-
-Stemmed points were most commonly used in environments with a temperature between 4 and 6 ℃ (@spa-gis-tcsa-sites and @fig-tcsa-radio-carbon-box, and Figure 9 from @park2022did).
+We found no clear correlation between population trend and stemmed points.
+Population could possibly
+
+Stemmed points were most commonly used in environments with a temperature between 4 and 6 ℃ (@fig-gis-tcsa-sites and @fig-tcsa-radio-carbon-box, and Figure 9 from @park2022did).
+
+Figure 10 from @park2022did
 
@@
  We consider that TCSA, or the function of the stemmed points, are strongly dependent on the surrounding environment or their situation.
-@no-figure-no-more: B shows that the temperature slowly decreased until around 22 ka and then went up again with a little range of fluctuations.
+@no-figure-no-more: B shows that  slowly decreased until around 22 ka and then went up again with a little range of fluctuations.
 Temperatures differ significantly between sites, with a maximum difference of 5℃.
@@
 
-Our models premise that the choice of weapon-tip types reflects the socio-environmental circumstances.
-To examine the possible connections between the roles of stemmed points and population dynamics or environmental change during the Late Paleolithic period, we compared the temporal pattern of TCSA to results of Summed probability distribution (SPD) and simulated Mean Annual Temperature (MAT) from @park2022did.
-In those two analyses, -@park2022did covered assemblages with and without stemmed points, which is useful for us to explore the overall socio-environmental dynamics here.
-As we previously noted, there is no clear pattern in the temporal or regional pattern of TCSA (@fig-tcsa-radio-carbon).
 THIS DISCUSSION IS INCOMPLETE, WE NEED TO EXPLORE THESE RELATIONSHIPS ABOUT GEOGRAPHICAL VARIATION AND TEMPORAL VARIATION FURTHER, AND IN A NEW WAY THAT IS NOT THE SAME AS THE QSR PAPER, E.G.

### 2022-11-20 analysis/paper/paper.qmd
tukey test result ggplot

@@
 #vegetation region#
-eco_aov <- aov(min_tcsa ~ as.factor(ecoregion), points_in_veg_zones_all)
+points_in_veg_zones_all$ecoregion = as.factor(points_in_veg_zones_all$ecoregion)
+eco_aov <- aov(min_tcsa ~ ecoregion, points_in_veg_zones_all)
 summary(eco_aov)
 eco_tur<- TukeyHSD(eco_aov)
+eco_lgm <- round(summary(eco_aov)[[1]][[5]][1], digits = 3)
 # plot(eco_tur)
-
-eco_lgm <- round(summary(eco_aov)[[1]][[5]][1], digits = 3)
+# plot with ggplot
+eco_tur_df <- as.data.frame(eco_tur$ecoregion)
+
+ggplot(eco_tur_df, aes(row.names(eco_tur_df))) + 
+  geom_boxplot(aes(middle=diff, upper=upr, lower=lwr, ymin=lwr, ymax=upr), stat="identity")
 
@@
 for_tur<- TukeyHSD(for_aov)
+for_lgm <- round(summary(for_aov)[[1]][[5]][1], digits = 3)
 # plot(for_tur)
-
-for_lgm <- round(summary(for_aov)[[1]][[5]][1], digits = 3)
 

### 2022-11-20 analysis/paper/paper.qmd
change Fig 8 to boxplot

@@
  ggplot(tcsa_site,
-              aes(min_tcsa, 
-                  fill = n)) +
-  geom_histogram() +
-  theme_bw(base_size = 6) +
-  labs(x = "TCSA", 
-       y = "Count of artefacts") +
-  facet_wrap(~label_site,
-                 scales = "free_y",
-                  ncol = 1) +
-    scale_fill_viridis_c(name = "Number\nof artefacts") +
-  annotate("rect", 
-           xmin = 4,
-           xmax = 18,
-           ymin = 0,
-           ymax = 2.5,
-           fill = "green",
-           alpha = 0.2) +
-      annotate("text", 
-               size = point_function_group_label_size,
-               label = "Poisoned\narrow tips",
-               hjust = 0,
-               lineheight = 0.75,
-           y = (0+2.5)/2,
-           x = 5.5) +
-      annotate("rect", 
-           xmin = 17,
-           xmax = 47,
-           ymin = 2.5,
-           ymax = 5,
-           fill = "blue",
-           alpha = 0.2) +
-      annotate("text", 
-               size = point_function_group_label_size,
-               label = "Un-poisoned\narrow tips",
-               hjust = 0,
-               lineheight = 0.75,
-           y = (2.5+5)/2,
-           x = 28) +
-      annotate("rect", 
-           xmin = 40,
-           xmax = 76,
-           ymin = 5,
-           ymax = 7.5,
-           fill = "orange",
-           alpha = 0.2) +
-      annotate("text", 
-               size = point_function_group_label_size,
-               label = "Dart tips",
-               hjust = 0,
-           y = (5+7.5)/2,
-           x = 53) +
-      annotate("rect", 
-           xmin = 42,
-           xmax = 90,
-           ymin = 7.5,
-           ymax = 10,
-           fill = "purple",
-           alpha = 0.2) +
-      annotate("text", 
-               size = point_function_group_label_size,
-               label = "Javelin tips",
-               hjust = 0,
-           y = (7.5+10)/2,
-           x = 70) +
-     annotate("rect", 
-           xmin = 80,
-           xmax = 200,
-           ymin = 10,
-           ymax = 12.5,
-           fill = "yellow",
-           alpha = 0.2) +
-      annotate("text", 
-               label = "Stabbing spear tips",
-               size = point_function_group_label_size,
-               hjust = 0,
-           y = (10+12.5)/2,
-           x = 160)
+              aes(y = min_tcsa, 
+                  x = label_site)) +
+  geom_boxplot() +
+  theme_minimal() +
+  labs(x = "", 
+       y = "TCSA") +
+  theme(axis.text.x = element_text(size = 6)) +
+  scale_x_discrete(guide = guide_axis(angle = 45))+
+  geom_quasirandom(alpha = 0.4,
+                   size = 2.3)
+
 ```

### 2022-11-20 analysis/paper/paper.qmd
changed fig 4 to boxplot

@@
 #| label: fig-tcsa-raw-materials
-#| fig-cap: "Histograms of TCSA values by lithic raw material."
+#| fig-cap: "TCSA values by lithic raw material."
 #| fig-width:   4.5
@@
 # Facet by raw materials
+
  ggplot(tcsa_raw,
-              aes(min_tcsa, 
-                  fill = n)) +
-  geom_histogram() +
-  theme_bw(base_size = 6) +
-  labs(x = "TCSA", 
-       y = "Count of artefacts") +
-  facet_wrap(~label_raw,
-                  scales = "free_y",
-                  ncol = 1)  +
-  scale_fill_viridis_c(name = "Number\nof artefacts") +
-  annotate("rect", 
-           xmin = 4,
-           xmax = 18,
-           ymin = 0,
-           ymax = 2.5,
-           fill = "green",
-           alpha = 0.2) +
-      annotate("text", 
-               size = point_function_group_label_size,
-               label = "Poisoned\narrow tips",
-               hjust = 0,
-               lineheight = 0.75,
-           y = (0+2.5)/2,
-           x = 5.5) +
-      annotate("rect", 
-           xmin = 17,
-           xmax = 47,
-           ymin = 2.5,
-           ymax = 5,
-           fill = "blue",
-           alpha = 0.2) +
-      annotate("text", 
-               size = point_function_group_label_size,
-               label = "Un-poisoned\narrow tips",
-               hjust = 0,
-               lineheight = 0.75,
-           y = (2.5+5)/2,
-           x = 28) +
-      annotate("rect", 
-           xmin = 40,
-           xmax = 76,
-           ymin = 5,
-           ymax = 7.5,
-           fill = "orange",
-           alpha = 0.2) +
-      annotate("text", 
-               size = point_function_group_label_size,
-               label = "Dart tips",
-               hjust = 0,
-           y = (5+7.5)/2,
-           x = 53) +
-      annotate("rect", 
-           xmin = 42,
-           xmax = 90,
-           ymin = 7.5,
-           ymax = 10,
-           fill = "purple",
-           alpha = 0.2) +
-      annotate("text", 
-               size = point_function_group_label_size,
-               label = "Javelin tips",
-               hjust = 0,
-           y = (7.5+10)/2,
-           x = 70) +
-     annotate("rect", 
-           xmin = 80,
-           xmax = 200,
-           ymin = 10,
-           ymax = 12.5,
-           fill = "yellow",
-           alpha = 0.2) +
-      annotate("text", 
-               label = "Stabbing spear tips",
-               size = point_function_group_label_size,
-               hjust = 0,
-           y = (10+12.5)/2,
-           x = 160)
+              aes(y = min_tcsa, 
+                  x = label_raw)) +
+  geom_boxplot() +
+  theme_minimal() +
+  labs(x = "", 
+       y = "TCSA") +
+  theme(axis.text.x = element_text(size = 6)) +
+  scale_x_discrete(guide = guide_axis(angle = 45))+
+  geom_quasirandom(alpha = 0.4,
+                   size = 2.3)
 
@@
 #| label: fig-tcsa-per-sites
-#| fig-cap: "Histograms of TCSA values by archaeological site."
+#| fig-cap: "TCSA values by archaeological site."
 #| fig-width:   4.5

### 2022-11-20 analysis/paper/paper.qmd
updated methods

@@
 We summarized the distribution of TCSA values for each assemblage to examine spatial patterns in TCSA values.
-We then explore possible environmental impact by connecting stemmed points on the map with eco-regional zones.
-We used QGIS 3.22.12 'Białowieża' (https://qgis.org/en/site/index.html) to draw a map based on the previous ecological studies [@yim1975distribution; @yi2010vegetation; @lee2008relationship] and located the individual artifacts using longitudes and latitudes of assemblages.
 Among `r n_sites` artifacts, we sat sites that contains fewer than 5 stemmed points aside and named them as "Other".
+We also explore possible environmental impact by connecting stemmed points on the eco-regional zones.
+We chose two maps of modern eco-regional zones and modified them for this research.
+@yim1975distribution first created vegetation forest map defined temperature and precipitations.
+This map then modified by @yi2010vegetation.
+They divided South Korea, our research area, into three zones: Central Temperature Zone (CT), South Temperature Zone (ST), and Subtropical-warm Temperature Zone (SW).
+We also used a map from @lee2008relationship, which separated our research area into 16 different zones based geographical characteristics such as inland, coastal, big river, island, and big mountain range in addition to ecological information including temperature and precipitation.
+We only focused on 14 zones, excluding two islands that have no evidence of Paleolithic material culture: Imjin river basin (IRB), Metropolitan (MP), Central inland (CI), Kangwon coastal (KC), Choongnam coastal (CC), Southwestern inland (SWI), Upper Nagdong river basin (UNRB), WoolYoung coastal (WYC), Western Cholla (WC), Southern mountain (SM), Southeastern inland (SEI), Hyungsan Taewha coastal (HTC), Western south coastal (WSC), and Eastern south coastal (ESC).
+We then located individual assemblages on those two maps using longitudes and latitudes and explore TCSA values for each zone.
+While it is acknowledged that modern eco-regional maps may not be suitable for studying the Paleolithic period directly, they are the only types of maps that are currently available to us.
 
@@
 
+p_lgm <- round(summary(aov(min_tcsa ~ as.factor(mis_stage), tcsa_site_ass_mis))[[1]][[5]][1], digits = 2)
+
 # box plot for each MIS stage
@@
            y = (200-80) /2 + 80 )
-
-p_lgm <- round(summary(aov(min_tcsa ~ as.factor(mis_stage), tcsa_site_ass_mis))[[1]][[5]][1], digits = 2)
 
@@
   geom_sf(data = veg_all, aes(fill = "grey")) +
-   geom_sf_text(data = veg_all, aes(label = ecoregion), size =3.5)+
+   geom_sf_text(data = veg_all, aes(label = ecoregion), size =1.5)+
   geom_sf(data = pts_all, aes(color = min_tcsa)) +
@@
   guides(fill = FALSE) +
-  theme(axis.text.x = element_text(size = 6))+
+  theme(axis.text.x = element_text(angle = 45, size = 6))+
+  theme(axis.text.y = element_text(size = 6))+
   theme(axis.title.x = element_blank(),
-          axis.title.y = element_blank())
+          axis.title.y = element_blank()) +
+  theme(legend.position = "none")
 
@@
   geom_boxplot() +
-  geom_quasirandom() +
+  geom_quasirandom(alpha = 0.3,
+                   size = 2.3) +
   xlab("Vegetation") + 
@@
   geom_sf(data = veg_forest, aes(fill = "grey")) +
-  geom_sf_text(data = veg_forest, aes(label = Zone), size =3.5)+
+  geom_sf_text(data = veg_forest, aes(label = Zone), size =1.5)+
   geom_sf(data = pts_forest, aes(color = min_tcsa)) +
@@
   guides(fill = FALSE) +
-  theme(axis.text.x = element_text(size = 6)) +
+  theme(axis.text.x = element_text(angle = 45, size = 6)) +
+  theme(axis.text.y = element_text(size = 6)) +
   theme(axis.title.x = element_blank(),
@@
   geom_boxplot() +
-  geom_quasirandom() +
+  geom_quasirandom(alpha = 0.3,
+                   size = 2.3) +
   xlab("Ecoregion") + 
@@
 ```
+
 ```{r}
@@
 TCSA can be impacted by other factors such as raw materials and portability.
--@eren2022plains explain that smaller TCSA could be the result of pursuing production economy and transport efficiency.
+@eren2022plains explain that smaller TCSA could be the result of pursuing production economy and transport efficiency.
 We explore the relationship between TCSA and raw material types and weight in order to examine these factors.

### 2022-11-21 analysis/paper/paper.qmd
apply apa style for p-value

@@
 
-p_lgm <- round(summary(aov(min_tcsa ~ as.factor(mis_stage), tcsa_site_ass_mis))[[1]][[5]][1], digits = 2)
+# p-value
+# simple version
+# p_lgm <- round(summary(aov(min_tcsa ~ as.factor(mis_stage), tcsa_site_ass_mis))[[1]][[5]][1], digits = 2)
+
+library("car")
+# devtools::install_github('achetverikov/apastats',subdir='apastats')
+library("apastats")
+p_lgm_test <- Anova(lm(min_tcsa ~ as.factor(mis_stage), tcsa_site_ass_mis))
+# describe.Anova(p_lgm_test, 1, 2)
 
@@
 Each category shows wide variation in LGM, which indicates diverse uses of stemmed points.
-There is no strong correlation between the TCSA values and the time period (p-value = `r p_lgm`).
+There is no strong correlation between the TCSA values and the time period (`apastats::describe.Anova(p_lgm_test, 1, 2)`).
 
@@
 
-We computed artefact TCSA values for `r n_sites` sites that contain more than `r min_n_points` stemmed points to observe variation between sites (@fig-tcsa-per-sites) Sites with fewer than `r min_n_points` stemmed points were grouped under the category of 'other'.
+We computed artifact TCSA values for `r n_sites` sites that contain more than 5 stemmed points to observe variation between sites (@fig-tcsa-per-sites) Sites with fewer than 5 stemmed points were grouped under the category of 'other'.
 Among the sites in our sample, Suyanggae has the most stemmed points and the widest range of TCSA values.
 This suggests that people made stemmed points for a variety of different purposes in Suyanggae.
-The TCSA values for Nosanri and Sibuk are have a narrower range of lower values, which indicates less diverse functions for stemmed points.
-Wolpyeng and Yongsujaeul have similar amounts of stemmed points to Nosanri and Sibuk, but exhibit wider distributions of TCSA values.
+The TCSA values for Nosanri, Sibuk, and Wolpyeng are have a narrower range of lower values, which indicates less diverse functions for stemmed points.
+Yongsujaeul shows a narrower range but higher TCSA value.
 Yongsandong has the second most stemmed points and TCSA shows a similar distribution to Yongsakul.
@@
 Modern vegetation map defined by temporature and precipitation (modified the map from @yi2010vegetation, but original source is @yim1975distribution).
-This map include 3 zones.
-Central Temperature Zone (CT), South Temperature Zone (ST), and Subtropical-warm Temperature Zone (SW).
 The individual points indicate assemblages.
@@
 Ecoregion map based on geographical boundaries and ecological conditions (modified the map from @lee2008relationship).
-The individual points indicate assemblages.
-This map include 14 zones.
-Imjin river basin (IRB), Metropolitan (MP), Central inland (CI), Kangwon coastal (KC), Choongnam coastal (CC), Southwestern inland (SWI), Upper Nagdong river basin (UNRB), WoolYoung coastal (WYC), Western Cholla (WC), Southern mountain (SM), Southeastern inland (SEI), Hyungsan Taewha coastal (HTC), Western south coastal (WSC), and Eastern south coastal (ESC).
-The individual points indicate assemblages.
 D.
-Boxplots for  ecoregion zones, excludings zones with no artifacts.
-The individual points indicate artifacts."
+Boxplots for  ecoregion zones, excludings zones with no artifacts."
 #| fig-width:   5
@@
 eco_tur<- TukeyHSD(eco_aov)
-eco_lgm <- round(summary(eco_aov)[[1]][[5]][1], digits = 3)
+# p-value
+# simple version
+# eco_lgm <- round(summary(eco_aov)[[1]][[5]][1], digits = 3)
+
+eco_lgm_test <- Anova(lm(min_tcsa ~  ecoregion, points_in_veg_zones_all))
+# describe.Anova(eco_lgm_test, 1, 2)
 # plot(eco_tur)
@@
 for_tur<- TukeyHSD(for_aov)
-for_lgm <- round(summary(for_aov)[[1]][[5]][1], digits = 3)
+# p-value
+# simple version
+# for_lgm <- round(summary(for_aov)[[1]][[5]][1], digits = 3)
+for_lgm_test <- Anova(lm(min_tcsa ~ as.factor(Zone), points_in_veg_zones_forest_box))
+# describe.Anova(for_lgm_test, 1, 2)
 # plot(for_tur)
@@
 
-We then located individual assemblages on maps that show ecoregional boundaries to explore spatial patterns.
+We then located individual assemblages on maps that show eco-regional boundaries to explore spatial patterns.
+
 We used two maps from the previous studies that represents different zones based on vegetation and geography [@yi2010vegetation; @yim1975distribution; @lee2008relationship].
 Our results show that stemmed points are only located in certain zones, Central Temperature Zone (CT) or Central inland (CI) (@fig-gis-tcsa-sites), for example.
-We observed that TCSA values are stonely associated with the zones (p-value for Vegetation zones = `r eco_lgm`, p-value for Ecoregion zones = `r for_lgm`).
+We observed that TCSA values are stonely associated with the zones (Vegetation zones: `apastats::describe.Anova(eco_lgm_test, 1, 2)`, Ecoregion zones: `apastats::describe.Anova(for_lgm_test, 1, 2)`).
 However, there is no clear functional pattern throughout the zones (@fig-gis-tcsa-sites panel B and D).
 There is a wide variation in TCSA regardless of the stemmed points density in each zone.
+
+Tukey
 

### 2022-11-21 analysis/paper/paper.qmd
Edited discussion

@@
 
+# explore artifacts in certain zone, related to Tukey's test
+tukey_veg <- points_in_veg_zones_all %>% 
+  filter(ecoregion == "ST") %>% 
+  dplyr::select(ecoregion, Assemblage, min_tcsa, raw_materials_top, age_ka)
+
 ############ explore tcsa with MODERN eco_regions & forest zones #######
@@
 ```{r}
-#| label: fig-tukey-eco-region
+#| label: fig-tukey-region
 #| fig-cap: "."
@@
 
-ggplot(eco_tur_df, aes(row.names(eco_tur_df))) + 
-  geom_boxplot(aes(middle=diff, upper=upr, lower=lwr, ymin=lwr, ymax=upr), stat="identity")
+eco_tur_box <- ggplot(
+  data = eco_tur_df,
+  mapping = aes(
+    x = row.names(eco_tur_df),
+    lower = lwr,
+    middle = diff,
+    upper = upr,
+    ymin = lwr,
+    ymax = upr
+    )
+  ) +
+  geom_boxplot(stat = "identity")+
+  xlab("Vegetation zones")+
+  theme(axis.text.x = element_text(angle = 45, size = 6)) +
+  theme(axis.text.y = element_text(size = 6))
 
@@
 
-```
-
-We then located individual assemblages on maps that show eco-regional boundaries to explore spatial patterns.
-
-We used two maps from the previous studies that represents different zones based on vegetation and geography [@yi2010vegetation; @yim1975distribution; @lee2008relationship].
-Our results show that stemmed points are only located in certain zones, Central Temperature Zone (CT) or Central inland (CI) (@fig-gis-tcsa-sites), for example.
-We observed that TCSA values are stonely associated with the zones (Vegetation zones: `apastats::describe.Anova(eco_lgm_test, 1, 2)`, Ecoregion zones: `apastats::describe.Anova(for_lgm_test, 1, 2)`).
-However, there is no clear functional pattern throughout the zones (@fig-gis-tcsa-sites panel B and D).
-There is a wide variation in TCSA regardless of the stemmed points density in each zone.
-
-Tukey
+for_tur_df <- as.data.frame(for_tur$`as.factor(Zone)`)
+
+for_tur_box <- ggplot(
+  data = for_tur_df,
+  mapping = aes(
+    x = row.names(for_tur_df),
+    lower = lwr,
+    middle = diff,
+    upper = upr,
+    ymin = lwr,
+    ymax = upr
+    )
+  ) +
+  geom_boxplot(stat = "identity") +
+  xlab("Eco-region zones") +
+  theme(axis.text.x = element_text(angle = 45, size = 6)) +
+  theme(axis.text.y = element_text(size = 6))
+
+# ggplot(for_tur_df) +
+#   geom_errorbar(aes(
+#     x = row.names(for_tur_df),
+#     ymin = lwr,
+#     ymax = upr
+#     ))
+
+plot_grid(eco_tur_box, for_tur_box, labels = c('A', 'B'))
+
+```
+
+In order to explore spatial patterns as well as the relationship between the function of the tools and the environment, we located individual assemblages on maps that depict eco-regional boundaries.
+Our results show that stemmed points are only located in certain zones, Central Temperature Zone (CT) of vegetation map or Central inland (CI) (@fig-gis-tcsa-sites) of eco-region map.
+We observed that TCSA values are strongely associated with the zones (Vegetation zones: `apastats::describe.Anova(eco_lgm_test, 1, 2)`, Ecoregion zones: `apastats::describe.Anova(for_lgm_test, 1, 2)`).
+To explore the TCSA distribution within the eco-regional zones, we excluded zones in the ecoregion map that have no stemmed points (@fig-gis-tcsa-sites panel D).
+We found no clear functional pattern throughout the zones, showing a wide variation in TCSA regardless of the density of stemmed point in each zone (@fig-gis-tcsa-sites panel B and D).
+
+We then computed Tukey's for comparison tests between the eco-regional zones.
+Our results show that ST and CT, and SWT-ST have significantly different while SWT and CT are close to each other (@fig-tukey-region panel A).
+We found no significant difference within the zones of eco-region map (@fig-tukey-region panel B).
 
@@
 There are a few stemmed points in our sample that may have been used as dart tips and un-poisoned arrow tips, but none of them have TCSA values in the range of poison arrow tips (@fig-tcsa-all-sp).
-But in general, we found a wide range of TCSA in our sample, indicating that stemmed points played a diverse role in forager toolkits.
+But in general, we found a wide range of TCSA in our sample, indicating that stemmed points played a diverse role in foraging toolkits.
 TCSA can be impacted by other factors such as raw materials and portability.
@@
 We explore the relationship between TCSA and raw material types and weight in order to examine these factors.
-We observed no clear pattern for raw materials (@fig-tcsa-raw-materials) but weight (@fig-artifact-size-cluster).
-Interestingly, weight is correlated with raw materials (@fig-tcsa-size).
+We observed no clear pattern for raw materials (@fig-tcsa-raw-materials) but weight (@fig-artifact-size-cluster), though weight is correlated with raw materials (@fig-tcsa-size).
 Therefore we could assume that raw materials also associated with TCSA value.
@@
 But the stemmed points mainly used before LGM (@fig-tcsa-radio-carbon-box).
-We found that stemmed points are only located in certain ecoregional zones such as Central Temperature zone without any clear pattern in function (@fig-gis-tcsa-sites).
+We found that stemmed points are only located in certain eco-regional zones such as Central Temperature zone without any clear pattern in function (@fig-gis-tcsa-sites).
 Both temporal and spatial analyses show that stemmed points were used for a particular environment, not so much harsh as LGM.
+@prates2022changes claimed that fishtail projectile points in South America was used for megafauna hunting and contributed to their extinction by proving strong relationship between the temporal density and spatial distribution of megafauna and the projectile points.
+Our results support this claim by showing the dense distribution of stemmed point in certain environment.
 
@@
 
-THIS DISCUSSION IS INCOMPLETE, WE NEED TO EXPLORE THESE RELATIONSHIPS ABOUT GEOGRAPHICAL VARIATION AND TEMPORAL VARIATION FURTHER, AND IN A NEW WAY THAT IS NOT THE SAME AS THE QSR PAPER, E.G.
-FOCUS ON THE LGM AS I MENTIONED IN OUR LAST MEETING.
-
-SPATIAL ANAlysis @prates2022changes provde strong relationship between the temporal density and spatial distribution of megafauna and Fishtail projectile points, and proposed that this weapon was designed and used for megafauna hunting, contributing to their extinction by exploring changes in projectile point technology, as well as the body mass of potential megafaunal prey, and show that Fishtails were strongly related to the largest extinct megafaunal species.
-
 # Conclusion

### 2022-11-21 analysis/paper/paper.qmd
Edited discussion

@@
 
-## Spatial patterns
+## Spatial patterns of TCSA values
 
@@
 Our models premise that the choice of weapon-tip types reflects the socio-environmental circumstances.
-To examine the possible connections between the roles of stemmed points and population dynamics or environmental change during the Late Paleolithic period, we compared the temporal and spatial patterns of TCSA to results of simulated Mean Annual Temperature (MAT) and Summed probability distribution (SPD) from @park2022did.
-We found no clear correlation between population trend and stemmed points.
-Population could possibly
-
-Stemmed points were most commonly used in environments with a temperature between 4 and 6 ℃ (@fig-gis-tcsa-sites and @fig-tcsa-radio-carbon-box, and Figure 9 from @park2022did).
-
-Figure 10 from @park2022did
-
-\<--!upto here --\>
- We consider that TCSA, or the function of the stemmed points, are strongly dependent on the surrounding environment or their situation.
-@no-figure-no-more: B shows that  slowly decreased until around 22 ka and then went up again with a little range of fluctuations.
-Temperatures differ significantly between sites, with a maximum difference of 5℃.
-Results of SPD show several peaks, as indicated by the positive deviations from the three null models (shaded in red), and downs (shaded in blue), but no drastic growth or decline (@no-figure-no-more: C).
-From the climate and population study, we could observe that there is no dominating change but minor and often fluctuations throughout the Late Paleolithic and variations per site.
-Overall this pattern is well matched with the TCSA results, indicating the people produced stemmed points as multifunctional tools suitable for unexpected or various situations in their habitats similar to the case of Clovis points (@eren2022plains).
+To examine the possible connections between the roles of stemmed points and environmental change or population dynamics during the Late Paleolithic period, we compared the temporal and spatial patterns of TCSA to results of simulated Mean Annual Temperature (MAT) and Summed probability distribution (SPD) from @park2022did.
+The simulated temperature distribution in south Korea during the Late Paleolithic period (Figure 9: A from @park2022did) seems similar to the modern eco-regional zones that we used in this research (@fig-gis-tcsa-sites).
+Figure 9: C from @park2022did shows the slow decrease of temperature until the end of LGM, which could explain our results of temporal pattern, meaning stemmed points were used during the cooler environment with a temperature between 4 and 6 ℃ (@fig-tcsa-radio-carbon-box).
+Results of SPD (Figure 10 from @park2022did) show several peaks, as indicated by the positive deviations from the three null models (shaded in red), and downs (shaded in blue), but no drastic growth or decline during our research period, indicating no major impact from population dynamics on stemmed points.
+
+Overall, our temporal and spatial patterns of TCSA show a wide TCSA range throughout the Late Paleolithic period and between eco-regional zones.
+Especially, we confirmed the widest TCSA range in a single site, Suyanggae (@fig-tcsa-per-sites).
+As a consequence, our results are consistent with our second model in which stemmed points are described as multifunctional tools, suggesting that people created stemmed points in response to unexpected or varying circumstances in their habitats, similar to Clovis points (@eren2022plains).
 

### 2022-11-21 analysis/paper/paper.qmd
Render qmd file
minor edits to some figures

@@
   xlab("Vegetation") + 
-  ylab("TCSA")
+  ylab("TCSA") +
+  theme_minimal()
 
@@
   xlab("Ecoregion") + 
-  ylab("TCSA")
-
-plot_grid(eco_plot, for_plot, eco_box, for_box, labels = c('A', 'C', 'B', 'D'))
+  ylab("TCSA") +
+  theme_minimal()
+
 ```
@@
 eco_aov <- aov(min_tcsa ~ ecoregion, points_in_veg_zones_all)
-summary(eco_aov)
-eco_tur<- TukeyHSD(eco_aov)
-# p-value
-# simple version
-# eco_lgm <- round(summary(eco_aov)[[1]][[5]][1], digits = 3)
 
 eco_lgm_test <- Anova(lm(min_tcsa ~ ecoregion, points_in_veg_zones_all))
-# describe.Anova(eco_lgm_test, 1, 2)
-# plot(eco_tur)
-# plot with ggplot
-eco_tur_df <- as.data.frame(eco_tur$ecoregion)
-
-eco_tur_box <- ggplot(
-  data = eco_tur_df,
-  mapping = aes(
-    x = row.names(eco_tur_df),
-    lower = lwr,
-    middle = diff,
-    upper = upr,
-    ymin = lwr,
-    ymax = upr
-    )
-  ) +
-  geom_boxplot(stat = "identity")+
-  xlab("Vegetation zones")+
-  theme(axis.text.x = element_text(angle = 45, size = 6)) +
-  theme(axis.text.y = element_text(size = 6))
+
+eco_lgm_test_pretty <- apastats::describe.Anova(eco_lgm_test, 1, 2)
+
+eco_tur <- TukeyHSD(eco_aov)
+
+# plot Tukey HSD
+eco_tur_box <- 
+eco_tur$ecoregion %>% 
+  as_tibble(rownames = "zones") %>% 
+  ggplot() +
+  geom_linerange(aes(xmin = lwr, 
+                     xmax = upr,
+                     y = zones)) +
+  geom_point(aes(diff,
+                 zones),
+             size = 4) +
+  geom_vline(xintercept = 0, colour = "red") +
+  ylab("Pairwise comparison\nof TSCA in different\nvegetation zones") +
+  xlab("Differences in means") +
+  theme_minimal(base_size = 8)
 
@@
 for_aov <- aov(min_tcsa ~ as.factor(Zone), points_in_veg_zones_forest_box)
-summary(for_aov)
-for_tur<- TukeyHSD(for_aov)
-# p-value
-# simple version
-# for_lgm <- round(summary(for_aov)[[1]][[5]][1], digits = 3)
+# summary(for_aov)
 for_lgm_test <- Anova(lm(min_tcsa ~ as.factor(Zone), points_in_veg_zones_forest_box))
-# describe.Anova(for_lgm_test, 1, 2)
-# plot(for_tur)
-
-for_tur_df <- as.data.frame(for_tur$`as.factor(Zone)`)
-
-for_tur_box <- ggplot(
-  data = for_tur_df,
-  mapping = aes(
-    x = row.names(for_tur_df),
-    lower = lwr,
-    middle = diff,
-    upper = upr,
-    ymin = lwr,
-    ymax = upr
-    )
-  ) +
-  geom_boxplot(stat = "identity") +
-  xlab("Eco-region zones") +
-  theme(axis.text.x = element_text(angle = 45, size = 6)) +
-  theme(axis.text.y = element_text(size = 6))
-
-# ggplot(for_tur_df) +
-#   geom_errorbar(aes(
-#     x = row.names(for_tur_df),
-#     ymin = lwr,
-#     ymax = upr
-#     ))
-
-plot_grid(eco_tur_box, for_tur_box, labels = c('A', 'B'))
+
+for_lgm_test_pretty <- apastats::describe.Anova(for_lgm_test, 1, 2)
+
+for_tur <- TukeyHSD(for_aov)
+
+# plot Tukey HSD
+for_tur_box <- 
+for_tur$`as.factor(Zone)` %>% 
+  as_tibble(rownames = "zones") %>% 
+  mutate(zones = fct_reorder(zones, diff)) %>% 
+  ggplot() +
+  geom_linerange(aes(xmin = lwr, 
+                     xmax = upr,
+                     y = zones)) +
+  geom_point(aes(diff,
+                 zones),
+             size = 4) +
+  geom_vline(xintercept = 0, colour = "red") +
+  ylab("Pairwise comparison\nof TSCA in different\necological zones") +
+  xlab("Differences in means") +
+  theme_minimal(base_size = 8)
+
+plot_grid(eco_plot, 
+          for_plot, 
+          eco_box, 
+          for_box, 
+          eco_tur_box, 
+          for_tur_box,
+          labels = c('A', 'C', 'B', 'D', 'E', 'F'),
+          ncol = 2)
 
@@
 Our results show that stemmed points are only located in certain zones, Central Temperature Zone (CT) of vegetation map or Central inland (CI) (@fig-gis-tcsa-sites) of eco-region map.
-We observed that TCSA values are strongely associated with the zones (Vegetation zones: `apastats::describe.Anova(eco_lgm_test, 1, 2)`, Ecoregion zones: `apastats::describe.Anova(for_lgm_test, 1, 2)`).
+We observed that TCSA values are strongely associated with the zones (Vegetation zones: `r eco_lgm_test_pretty`, Ecoregion zones: `r for_lgm_test_pretty`).
 To explore the TCSA distribution within the eco-regional zones, we excluded zones in the ecoregion map that have no stemmed points (@fig-gis-tcsa-sites panel D).
@@
 
-Marlize Lombard: Conceptualization, Methodology, Writing - Review & Editing, Supervision
-
-Donghee Chong: Data curation, Review & Editing
+Marlize Lombard: Conceptualization, Methodology, Supervision
+
+Donghee Chong: Data curation
 

### 2022-11-26 analysis/paper/paper.qmd
Editing upto first para of the methods

@@
 abstract: |
- The introduction of blade technology, stemmed points, end scrapers, burins, denticulates, and finer grained materials characterises the transition from the Early to Late Paleolithic in Korea.
+ The introduction of blade technology, stemmed points, end scrapers, burins, denticulates, and finer grained materials characterizes the transition from the Early to Late Paleolithic in Korea.
 Stemmed points have been considered a representative tool that led this set of changes.
@@
 We measured tip cross-sectional area (TCSA) to discriminate different likely use classes of projectile points, for example, as poisoned arrow tips or as stabbing spears.
-We analyzed TCSA with other variables including raw materials, weight, site and radiocarbon dates.
+We analyzed TCSA with other variables including raw materials, weight, site, radiocarbon dates and locations.
 Our results show that the stemmed points were likely used  as javelin tips and stabbing spear tips, with smaller numbers as dart tips and un-poisoned arrow tips.
@@
 The temporal pattern of TCSA values is one of little change throughout the Late Paleolithic period but SP were mainly used before Last Glacial Maximum (LGM).
-We observed that SP was located in certain ecoregions in Korea but no clear functional pattern is confirmed.
+We observed that SP was only located in certain ecoregions in Korea but no clear functional pattern is confirmed.
 We conclude that stemmed points were multi-functional tools, with many likely designed for javelin and stabbing spear tips.
@@
 Despite the importance of stemmed points, only a few studies to date have examined their likely uses.
-Previous work has mostly debated their origin, the chronology of the Korean Late Paleolithic, and their relationship with Japan [@chang2013human; @Park_2013; @lee2019were; @Chong_2021].[@chang2013human; @Park_2013; @lee2019were; @Chong_2021].
-
-To understand the likely roles that stemmed points played in the technological transition, this study examines the possible uses of stemmed points.
+Previous work has mostly discussed their origin, the chronology of the Korean Late Paleolithic, and their relationship with the Japanese archipelago [@chang2013human; @Park_2013; @lee2019were; @Chong_2021].[@chang2013human; @Park_2013; @lee2019were; @Chong_2021].
+
+To understand the roles that stemmed points played in the technological transition, this study examines the possible uses of stemmed points.
 We use the tip cross-sectional area (TCSA) metric to infer likely artifact use, and compare the results to other archaeological and ethnographic assemblages based on the idea that different shapes correspond to different weapon types [@lombard2021variation].
@@
 Elsewhere in the world these types of artifacts are often called 'tanged points', but we prefer 'stemmed point' here to distinguish from Bronze Age stone projectile points that have long been called 'tanged points' in Korea [@park2022did].
-After the first discovery at the Suyanggae site, more than 450 stemmed points have been found in over 30 sites throughout South Korea [@Chong_2021].
-Understanding the emergence of stemmed points is relevant to general questions about the origins of projectile technology, the emergence of the Late Palaeolithic, and relationships to adjacent regions in East Asia, such as Japan.
-
-Stemmed points are the first composite tools in the Korean Paleolithic that required two different parts, consisting of a stone point and shaft, presumably made out of wood [@seong2008tanged].
+Understanding the appearance of stemmed points is relevant to general questions about the origins of projectile technology, the emergence of the Late Paleolithic, and relationships to adjacent regions in East Asia, such as Japan.
+
+Stemmed points are the first composite tools in the Korean Paleolithic that required two different parts to be one complete tool, consisting of a stone point and shaft, presumably made out of wood [@seong2008tanged].
 Using blades as the blank for the point enables mass production for this composite tool and its shape can become more standardized [@lee2015characteristics; @park2022did].
-Therefore, @o2018origins claims that understanding the emergence of projectile technology gives insights to greater cultural, evolutionary, and behavioral cognitive flexibility.
-
-Since stemmed points represent the beginning of the Korean Late Paleolithic, investigating their origins is important for understanding the technological transition from the Early Palaeolithic, modern human dispersals into the region, and claims for the existence of the 'Middle' Paleolithic in Korea [@bae2010origin; @bae2017late; @norton2009evolution; @seong2016eastern].
+Therefore, as @o2018origins claimed, understanding the emergence of projectile technology gives insights to greater cultural, evolutionary, and behavioral cognitive flexibility.
+
+Since stemmed points represent the beginning of the Korean Late Paleolithic, investigating their origins is important for understanding the technological transition from the Early Paleolithic, modern human dispersals into the region, and claims for the existence of the 'Middle' Paleolithic in Korea [@bae2010origin; @bae2017late; @norton2009evolution; @seong2016eastern].
 There has been a noteworthy debate about explaining the origin(s) of stemmed points which can be summarized into two competing models: a 'heterogenic' migration [@bae2010origin], and an *in situ* evolution model [@seong2009emergence].
-The migration model claims that the new blade industry including stemmed points, and the earlier coarse flake tradition including large cores, polyhedrons, choppers, and even handaxes, came from different origins (e.g.
+The migration model claims that the new blade industry including stemmed points, and the earlier coarse flake tradition including large cores, polyhedrons, choppers, and even handaxes, came from different origins (i.e.
 1: North route: Siberia, Mongolia, or other regions of northeast China, and 2: South route: southern China) as the result of a continuing influx of modern human migration from two routes.
@@
 The difference between the two models comes from how to understand a few early sites that contain stemmed points but no blades and blade cores.
-The *in situ* model claims that stemmed points appeared before blades by acknowledging those early sites, including Bonggok, Songamri, Yonghodong, and Hwadaeri, while the migration model supports the traditional definition of the Late or Upper Palaeolithic in Europe and Africa that started with emergence and establishment of blade technology [@seong2016eastern].
-
-Stemmed points are important proxies to understand human behaviors during the Late Palaeolithic.
+While the *in situ* model claims that stemmed points appeared before blades by acknowledging those early sites, including Bonggok, Songamri, Yonghodong, and Hwadaeri, the migration model supports the traditional definition of the Late or Upper Paleolithic in Europe and Africa that started with emergence and establishment of blade technology [@seong2016eastern].
+
+Stemmed points are important proxies to understand human behaviors during the Late Paleolithic.
 @park2022did examined mobility and site occupation patterns by applying concepts of human behavioral ecology to lithic assemblages and found that forager groups using stemmed points may have been associated with the occupation of marginal or extreme environments, compared to groups with no stemmed points.
@@
 
-Stemmed points in both Korea and Japan have been studied since the late 1980s as a part of evidence for long-distance/maritime cultural interchanges or social networks [@lee2015characteristics].
+The connection between stemmed points in both Korea and Japan have been studied since the late 1980s as a part of evidence for long-distance/maritime cultural interchanges or social networks [@lee2015characteristics; @chang2013human].
 Stemmed points from the Bonggok site in Korea are currently accepted as the oldest ones among Northeast Asia, dated to ca.
 41.5 ka, and made on elongated flakes [@seong2009emergence; @seong2015diversity; @bae2017origin].
-After their first appearance in Korea, stemmed points (Hakuhensenntouki in Japanese) appear in Kyushu, Japan during late MIS 3.
+After their first appearance in Korea, stemmed points (Hakuhensenntouki in Japanese) appear in Kyushu, Japan during late Marine isotope stages (MIS) 3.
 In addition to the stemmed points, there are similar artifacts found in both regions such as microblade cores, Moppule-seokgi (Kakusuijyosekki in Japanese), backed knife, bilateral points, bifacial points, and transport of obsidian [@lee2012characteristics; @chang2013human; @lee2015characteristics; @kim2021evidence].
-Stemmed points are thus important for understanding the connections and technological transitions between islands in Japan [@chang2013human].
 
@@
 
-## Uses of stemmed points
-
-Though stemmed points are generally assumed to have been hunting armatures [@lee2019were; @lee2011study; @seong2008tanged; @lee2002new; @chang2013human], it is difficult to understand the likely uses of the tools without knowing their complete shape when attached to other components.
+## Previous studies about the function of stemmed points
+
+Though stemmed points have been generally assumed to have been hunting armatures [@lee2019were; @lee2011study; @seong2008tanged; @lee2002new; @chang2013human], it is difficult to understand the likely uses of the tools without knowing their complete shape when attached to other components.
 Preserved wooden components of projectile tools are too few and rare to generalize the overall scale and variability of the tools used in prehistory [@shea2006origins].
 @Lee_Kong_2002 claimed stemmed points were 'stemmed tools' because of their diverse shapes, which could represent different functions, as well as the uncertainty of complete tool shape.
-Other researchers agree that some stemmed points that are not symmetrical and have retouch on one side of the artifacts or have denticulate blades on one side should not be called stemmed points but instead stemmed knives, stemmed side-scrapers, stemmed end-scrapers, or stemmed burins [@seong2008tanged; @Lee_Jang_2011; @Kim] @fig-sp-hand-made: 7-8.
+Other researchers agree that a few stemmed points that are not symmetrical and have retouch on one side of the artifacts or have denticulate blades on one side should not be called as stemmed points but instead stemmed knives, stemmed side-scrapers, stemmed end-scrapers, or stemmed burins [@seong2008tanged; @Lee_Jang_2011; @Kim] @fig-sp-hand-made: 7-8.
 
@@
 There are a high proportion of broken tips and stems and the reused tools were repaired based on keeping the symmetry [@Park_2013; @Kim].
-In addition to the morphological aspect of stemmed points, investigation of the whole site and tool composition of an assemblage shows that stemmed points or stemmed tools are strongly associated with hunting activities including peeling the animal skin after slaughtering or separating the bones from the flesh [@seong2008tanged; @Chong_2021].
+In addition to the morphological aspect of stemmed points, investigation of a whole site and tool composition of an assemblage shows that stemmed points or stemmed tools could be strongly associated with hunting activities including peeling the animal skin after slaughtering or separating the bones from the flesh [@seong2008tanged; @Chong_2021].
 
@@
 
-Tip cross-sectional area (TCSA) of stone artefacts has been used as a ballistically relevant standard to probabalistically discriminate different between functional classes of projectile armatures such as spearthrower (a.k.a.
+Tip cross-sectional area (TCSA) of stone artifacts has been used as a ballistically relevant standard to probabalistically discriminate different between functional classes of projectile armatures such as spearthrower (a.k.a.
 atlatl) dart tips, un-poisoned arrow tips and large stabbing/thrusting spears [@sisk2011african; @lombard2021variation; @lombard2020tip; @lombard2021did; @o2018origins].
 It is important to note that the TCSA metric alone cannot unambiguously determine artifact function, it only suggests a best-fit ballistic probability for the points if they were hafted as weapon tips.
-The TCSA metric indicates the part of the tool that cuts the target's hide and relates to weapon flight and penetration dynamics [@hughes1998getting; @sitton2020tip; @lombard2021variation].
+The TCSA metric represents the part of the tool that cuts the target's hide and relates to weapon flight and penetration dynamics [@hughes1998getting; @sitton2020tip; @lombard2021variation].
 This method was first proposed by @hughes1998getting and and tested by @shea2006origins by comparing archaeological examples to ethnographically collected samples of known use.
 @shea2009impact also applied this approach to compare projectile points from Africa, the Levant, and Europe, and claimed that projectile weapons first appeared in Africa.
-A key advantage of the TCSA metric is its convenience: regardless of point type, only maximum width and thickness measurements are needed for calculating the TCSA value (0.5 x maximum width x maximum thickness) [@sisk2011african; @lombard2020tip].
-Later, @sisk2011african proposed an alternate measure, tip cross-sectional perimeter (TCSP), for more accurate measure for the force needed to penetrate a target to a lethal depth while TCSA metric is more associated with cutting aspect.
+A key advantage of the TCSA metric is its convenience of application: regardless of point type, only maximum width and thickness measurements are needed for calculating the TCSA value (0.5 x maximum width x maximum thickness) [@sisk2011african; @lombard2020tip].
+Later, @sisk2011african proposed an alternate metric, tip cross-sectional perimeter (TCSP), for more accurate measure for the force needed to penetrate a target to a lethal depth while TCSA metric is more associated with cutting aspect.
 However, TCSP has a few disadvantages that limit its applicability to our case study of Korean stemmed points.
 The force and penetration depth are not only affected by the stone tip, but also by the mass of the shaft, which cannot be known for most archaeological stone-tipped weapons [@lombard2020tip].
-@sisk2011 also mentioned that TCSP cannot be applied to backed pieces that are hafted as projectile armatures.
+@sisk2011african also mentioned that TCSP cannot be applied to backed pieces that are hafted as projectile armatures.
 
 @lee2019were first applied TCSA to stemmed points from Korea along with use-wear analysis.
-In that study, stemmed points were analysed from Jingeuneul, located in southwest of Korea, which has the largest number of stemmed points (n = 99) ever found in a single site.
+They analyzed stemmed points from Jingeuneul, located in southwest of Korea, which has the largest number of stemmed points (n = 99) ever found in a single site.
 For TCSA, they were only able to use ten stemmed points because they needed to select the stemmed points that retained the widest and thickest part of the specimens for TCSA metric and also displayed diagnostic impact fractures for the use-wear analysis.
@@
 Their results show that the TCSA range of Jingeuneul stemmed points is relatively wide, overlapping with both North American dart tips and arrowheads.
-Inspired by @lee2019were and TCSA research from lithic assemblages in other parts of the world, we aimed to investigate TCSA values for a large number of stemmed points from all over South Korea to better understand their likely uses during the Late Paleolithic.
+Inspired by @lee2019were and TCSA research from lithic assemblages in other parts of the world, we aim to investigate TCSA values for stemmed points from all over South Korea to better understand their likely uses during the Late Paleolithic.
 
@@
 
-Since the first evidence of the Korean Late Paleolithic was discovered at Seokjangri site, including blades and a stemmed base, around 300 stemmed points have been found in Korea [@Sohn_1967; @lee2019were].
+After the first discovery at the Suyanggae site, more than 450 stemmed points have been found in over 30 sites throughout South Korea [@Sohn_1967; @lee2019were; @Chong_2021].
 Most sites contain only a few stemmed points and only a few sites have many more, such as Jingeuneul, Suyanggae (n = 55), and Yongsandong (n = 38) [@Kim_2017].
-Among these stemmed points, we selected those that retained their widest and thickest parts.
-Our dataset contains stemmed points that were found during field surveys as well as those were from sites that were never dated but were associtaed with other Late Palaeoltihic artifacts.
+Among these stemmed points, we selected those that retain their widest and thickest parts.
+Our dataset contains stemmed points that were found during field surveys as well as those were from sites that were never dated but were associtaed with other Late Paleoltihic artifacts.
 Applying these sampling criteria resulted in a sample of `r total_sp` stemmed points from `r total_ass` assemblages unearthed from `r total_site` sites spanning the period 44-10 ka (@fig-map).
@@
 Because Hajinri was from the sixth excavation area, which is 3.5 kilometers apart from the other spots and dated to much earlier (around 42-30 ka) than the other assemblages (around 31-15 ka).
-According to the excavation reports for Hajinri, stemmed points first appeared in Hajinri around 42 ka with the earliest blade and blade cores [@Lee_Woo_Lee_An_Yun_Park_Otani_Kim_Kim_Han_et_al._2018].
-We consider this assemblage as an unusual outlier because the stemmed points and blades from Hajinri are so early and also so highly standardized and refined, which are only found at other sites much later in time.
+According to the excavation reports for Hajinri, stemmed points first appeared in Hajinri around 42 ka with the earliest blade and blade cores [@Lee_Woo_Lee_An_Yun_Park_Otani_Kim_Kim_Han_et_al_2018].
+We consider this assemblage as an unusual outlier because the stemmed points and blades from Hajinri are highly standardized and refined, which are only found at other sites much later in time.
 We have more confidence in finds from Bonggok, which has the second-earliest dates (41.5 ka) in our collection, as the first appearance of stemmed points.
@@
 
-To answer our research questions about the likely uses of stemmed points, we calculated TCSA values for the stemmed points in our sample, and explored the interaction of TCSA values with raw materials, and artefact size.
+To answer our research questions about the likely uses of stemmed points, we calculated TCSA values for the stemmed points in our sample, and explored the interaction of TCSA values with raw materials, and artifact size.
 To examine the temporal patterns in TCSA values we investigated radiocarbon ages associated with the points.
-We distinguish the research time period into three based on the major climate event during the Late Pleistocene, Last Glacial Maximum (LGM), and compare with the temporal pattern.
-We explore the relationship between TCSA values and the location of assemblages to explore spatial pattern.
+We distinguished the research time period into three based on the major climate event during the Late Pleistocene, Last Glacial Maximum (LGM), and examined the TCSA pattern.
+We explored the relationship between TCSA values and the location of assemblages by applying eco-regional information.
 We followed the TCSA ranges for different weapon-delivery systems that @lombard2022tip and @lombard2021variation created by summarizing @wadley2008segment' study of backed microliths and calibrating existing ranges (@tbl-tcsa-ranges).
-We excluded `r n_omit` artifacts from our dataset with TCSA greater than 250 and thus outside of the range of the comparative data.
+Among our dataset, we excluded `r n_omit` artifacts from our dataset with TCSA greater than 250 and thus outside of the range of the comparative data.
 For this study we include TCSA values from a total of `r min_n_points` stemmed points from `r n_ass` assemblages unearthed from `r n_sites` sites.
@@
 
+possible role
+indicate people used stemmed poitns in different way
+examined their likely uses
+
 SUGGEST SOME POSSIBLE DIRECTIONS FOR FUTURE WORK, E.G. EXPERIMENTAL ARCHAEOLOGY, USEWEAR ANALYSIS

### 2022-11-26 analysis/paper/paper.qmd
edited by the end of methods

@@
          label_site,
-         n) %>% 
-  distinct_all()
+         n) 
+# %>% 
+#   distinct_all()
 
@@
 For this study we include TCSA values from a total of `r min_n_points` stemmed points from `r n_ass` assemblages unearthed from `r n_sites` sites.
+Each analysis uses slightly different number of artifacts due to data availability.
 
@@
 Selective use of raw material is a key characteristic of the technological transition of the Korean Late Paleolithic.
-Prior to the Late Paleolithic people used quartzite and vein quartz for core and flake tools.
+Prior to the Late Paleolithic, people used quartzite and vein quartz for core and flake tools.
 Then, finer grained materials were newly added to assemblages for producing the newly introduced tools [@seong2004quartzite].
@@
 Different sizes of stone artifacts can constrain or enable different functions.
-Overall size of stone points has been used as a potential proxy to separate different armature types [@thomas1978arrowheads; @sahle2019assessment].
+Overall size of stone points has been used as a potential proxy to distinguish different armature types [@thomas1978arrowheads; @sahle2019assessment].
 We chose weight as a proxy for overall size of the stemmed points.
@@
 
-### Temporal pattern
-
-We used radiocarbon ages to explore the temporal pattern of the function of stemmed points.
+### Temporal patterns
+
+We used radiocarbon ages to explore the temporal patterns of the likely uses of stemmed points.
 After excluding assemblages that have no radiocarbon dates, we arranged `r n_sites_with_radiocarbon_ages` assemblages dated from 45ka to 14.8ka to explore variation in TCSA over time.
@@
 We summarized the distribution of TCSA values for each assemblage to examine spatial patterns in TCSA values.
-Among `r n_sites` artifacts, we sat sites that contains fewer than 5 stemmed points aside and named them as "Other".
-We also explore possible environmental impact by connecting stemmed points on the eco-regional zones.
+Among `r n_sites` artifacts, we combined sites that contains fewer than 5 stemmed points aside and named them as "Other".
+We also explored possible environmental impact by connecting stemmed points to the eco-regional zones.
 We chose two maps of modern eco-regional zones and modified them for this research.
@@
 They divided South Korea, our research area, into three zones: Central Temperature Zone (CT), South Temperature Zone (ST), and Subtropical-warm Temperature Zone (SW).
-We also used a map from @lee2008relationship, which separated our research area into 16 different zones based geographical characteristics such as inland, coastal, big river, island, and big mountain range in addition to ecological information including temperature and precipitation.
-We only focused on 14 zones, excluding two islands that have no evidence of Paleolithic material culture: Imjin river basin (IRB), Metropolitan (MP), Central inland (CI), Kangwon coastal (KC), Choongnam coastal (CC), Southwestern inland (SWI), Upper Nagdong river basin (UNRB), WoolYoung coastal (WYC), Western Cholla (WC), Southern mountain (SM), Southeastern inland (SEI), Hyungsan Taewha coastal (HTC), Western south coastal (WSC), and Eastern south coastal (ESC).
+We also used a map from @lee2008relationship, which separated our research area into 16 different zones based geographical characteristics such as inland, coastal, major rivers, island, and major mountain ranges in addition to ecological information including temperature and precipitation.
+We only focused on 14 zones, excluding two islands that have no records of Paleolithic material culture: Imjin river basin (IRB), Metropolitan (MP), Central inland (CI), Kangwon coastal (KC), Choongnam coastal (CC), Southwestern inland (SWI), Upper Nagdong river basin (UNRB), WoolYoung coastal (WYC), Western Cholla (WC), Southern mountain (SM), Southeastern inland (SEI), Hyungsan Taewha coastal (HTC), Western south coastal (WSC), and Eastern south coastal (ESC).
 We then located individual assemblages on those two maps using longitudes and latitudes and explore TCSA values for each zone.
-While it is acknowledged that modern eco-regional maps may not be suitable for studying the Paleolithic period directly, they are the only types of maps that are currently available to us.
+It is acknowledged that modern eco-regional maps may not be suitable for studying the Paleolithic period directly, but they are the only maps of eco-regional seperation that are currently available to us.
 
@@
 
-The process of the introduction of new technologies can vary depending on the environmental and social contexts of the transmission of the techniques for making the new tools, and the new tools themselves can alter the typical contexts of tool use.
+The process of the introduction of new technologies can vary depending on the environmental and social contexts of the transmission of the manufacturing techniques for the new tools.
+Moreover, the new tools themselves can alter the typical contexts of tool use.
 For example, the introduction of the bow and arrow in American enabled individuals to work without hunting in teams and provided opportunities for non-elite hunters to produce their own subsistence or pursue individual wealth [@bettinger2013effects; @angelbeck2014faustian; @rorabaugh2015timing].
@@
 Clovis points were used as multifunctional tools that performed a wider range of tasks, including cutting and sawing.
-Folsom points show a narrower range of variation, indicating a they were more likely used for specific tasks.
-
-Inspired by @eren2022plains's approach, we hypothesized two basic scenarios that might explain the regional or/and temporal pattern of TCSA range.
-If stemmed points represent a narrow TCSA range, then people likely produced tools that performed a small set tasks.
-This may be due to low levels of uncertainty in the forager's social and physical environments.
+Folsom points, on the other hand, show a narrower range of variation, indicating a they were more likely used for specific tasks.
+
+Inspired by @eren2022plains's approach, we hypothesized two basic scenarios that might explain the temporal or/and spatial patterns of TCSA range.
+If stemmed points represent a narrow TCSA range, then people likely produced tools that performed specific tasks.
+This may be related to low levels of uncertainty in the forager's social and physical environments.
 On the other hand, a wide TCSA range may indicate that stemmed points were multifunctional tools, suggesting that people were responding to unfamiliar situations, such as moving into an unfamiliar landscape or unpredictable changes in patch productivity and travel times [@bird2006behavioral; @bettinger2016marginal; @kelly2007foraging].
-We expect to observe different TCSA ranges, reflecting morphologically and functionally different stemmed points, across different regions or environments in Korea, and over time.
+We expect to observe various TCSA ranges, reflecting morphologically and functionally different stemmed points, across different regions or environments in Korea, and over time.
 

### 2022-11-28 analysis/paper/paper.qmd
editing result part

@@
 library(gt)
+library(sass)
 read_csv(here::here("analysis/data/raw_data/tcsa-range-from-lombard-2022.csv")) %>% 
@@
 #| label: fig-tcsa-all-sp
-#| fig-cap: "Distribution of TCSA values for all Korean stemmed points in the current dataset. Shaded boxes show TSCA ranges for various weapon types."
+#| fig-cap: "Distribution of TCSA values for all Korean stemmed points in the current dataset. Shaded boxes indicate TSCA ranges for various weapon types."
 #| fig-width:   3
@@
 point_function_group_label_y <- 0.75
-point_function_group_label_size <- 1
+point_function_group_label_size <- 1.5
 
@@
                size = point_function_group_label_size,
-               label = "Poisoned\narrow tips",
+               label = "Poisoned arrow tips",
                hjust = 0,
@@
                size = point_function_group_label_size,
-               label = "Un-poisoned\narrow tips",
+               label = "Un-poisoned arrow tips",
                hjust = 0,
@@
       annotate("text", 
-               label = "Stabbing\nspear tips",
+               label = "Stabbing spear tips",
                size = point_function_group_label_size,
@@
   geom_quasirandom(alpha = 0.4,
-                   size = 1.5) +
+                   size = 1.1) +
   coord_cartesian(xlim = c(-0.1,1)) +
@@
 
-@fig-tcsa-all-sp shows TCSA values for all stemmed points in our sample, with shaded rectangles to assist in the interpretation of their likely function.
-The ranges indicated by the shaded rectangles were taken from @lombard2021variation.
+@fig-tcsa-all-sp shows TCSA values for all stemmed points in our sample, with shaded rectangles to assist in the interpretation of their likely uses.
 Overall we see a wide variation of TCSA values in this boxplot.
 The sample mean TCSA is `r round(mean(tcsa$min_tcsa), digits=1)`, and standard deviation of TCSA is `r round(sd(tcsa$min_tcsa), digits=1)`.
-According to the TCSA ranges presented in Table 1 of @lombard2021variation, Korean stemmed points are overall mostly in the categories of javelin tips and stabbing spear tips, with smaller numbers as dart tips and un-poisoned arrow tips.
+According to the TCSA ranges presented in @tbl-tcsa-ranges, Korean stemmed points are overall mostly in the categories of javelin tips and stabbing spear tips, with smaller numbers as dart tips and un-poisoned arrow tips.
 Among the weapon-tip types, only poison arrow tips appear to be absent from these Korean assemblages.
@@
 #| fig-cap: "TCSA values by lithic raw material."
-#| fig-width:   4.5
-#| fig-height:  4
+#| fig-width:   4
+#| fig-height:  3.5
 
@@
   theme_minimal() +
-  labs(x = "", 
+  labs(x = "Raw materials", 
        y = "TCSA") +
@@
   geom_quasirandom(alpha = 0.4,
-                   size = 2.3)
+                   size = 1.9) +
+   annotate("rect", 
+           xmin = 6.4,
+           xmax = 6.5,
+           ymin = 4,
+           ymax = 18,
+           fill = "green",
+           alpha = 0.2) +
+      annotate("text", 
+               size = point_function_group_label_size,
+               label = "Poisoned\narrow tips",
+               hjust = 0,
+               lineheight = 0.75,
+           x = 6.4,
+           y = (18-4) /2 + 4) +
+      annotate("rect", 
+           xmin = 6.5,
+           xmax = 6.6,
+           ymin = 17,
+           ymax = 47,
+           fill = "blue",
+           alpha = 0.2) +
+      annotate("text", 
+               size = point_function_group_label_size,
+               label = "Un-poisoned\narrow tips",
+               hjust = 0,
+               lineheight = 0.75,
+           x = 6.5,
+           y = (47-17) /2 + 17) +
+      annotate("rect", 
+           xmin = 6.6,
+           xmax = 6.7,
+           ymin = 40,
+           ymax = 76,
+           fill = "orange",
+           alpha = 0.2) +
+      annotate("text", 
+               size = point_function_group_label_size,
+               label = "Dart tips",
+               hjust = 0,
+           x = 6.6,
+           y = (76-40) /2 + 40) +
+      annotate("rect", 
+           xmin = 6.7,
+           xmax = 6.8,
+           ymin = 42,
+           ymax = 90,
+           fill = "purple",
+           alpha = 0.2) +
+      annotate("text", 
+               size = point_function_group_label_size,
+               label = "Javelin tips",
+               hjust = 0,
+           x = 6.7,
+           y = (90-42) /2 + 42) +
+     annotate("rect", 
+           xmin = 6.8,
+           xmax = 6.9,
+           ymin = 80,
+           ymax = 200,
+           fill = "yellow",
+           alpha = 0.2) +
+      annotate("text", 
+               label = "Stabbing\nspear tips",
+               size = point_function_group_label_size,
+               hjust = 0,
+               lineheight = 0.75,
+           x = 6.8,
+           y = (200-80) /2 + 80 )
 
@@
 We conducted a univariate cluster analysis [@song2020efficient; @wang2011ckmeans] of stemmed points by weight and it revealed three clusters of artifact sizes (A of @fig-artifact-size-cluster, mean = `r round(mean(tcsa_mass_no_na$'Weight(g)'), digits=1)`, SD = `r round(sd(tcsa_mass_no_na$'Weight(g)'), digits=1)`).
-Cluster 1, the smallest (lightest) artifacts, have a lower range of TCSA values, compared to Cluster 2 (B of @fig-artifact-size-cluster).
-TCSA values for Cluster 3 are the highest, except for one artifact, which is lower than 50 (A of @fig-artifact-size-cluster, mean = `r round(mean(tcsa_mass_no_na$'min_tcsa'), digits=1)`, SD = `r round(sd(tcsa_mass_no_na$'min_tcsa'), digits=1)`).
+Cluster 1, the smallest (lightest) artifacts, have a lower range of TCSA values, compared to Cluster 2 (B of @fig-artifact-size-cluster, mean = `r round(mean(tcsa_mass_no_na$'min_tcsa'), digits=1)`, SD = `r round(sd(tcsa_mass_no_na$'min_tcsa'), digits=1)`).
+TCSA values for Cluster 3 are the highest, except for one artifact, which is lower than 50.
 There are a few overlaps, but we can still assume that artifacts in different size clusters might be made for different purposes.
@@
 #| label: fig-artifact-size-cluster
-#| fig-cap: "A. Distribution of artefact weight showing three clusters. B. TCSA values for all artefacts with size classes indicated by data point shape."
+#| fig-cap: "A. Distribution of artifact weight showing three clusters. B. TCSA values for all artifacts with size classes indicated by data point shape."
 #| fig-width:   4
@@
 @fig-tcsa-size shows a positive relationship between TCSA, artifact weight and raw material.
-The relationship varies by raw material types, perhaps due to variation in the density of different lithic raw materials.
+We could observe variations by raw material types.
 There is a stronger correlation between TCSA and artifact weight for acidic volcanic points (i.e.
 points are closer to the regression line), whereas the correlation for shale stemmed points is weaker.
-The other raw materials show varied distribution patterns around the regression line, reconfirming that raw material appears not to have strongly influenced TCSA (see also @fig-tcsa-raw-materials).
+The other raw materials show various distribution patterns around the regression line, reconfirming that raw material appears not to have strongly influenced TCSA (see also @fig-tcsa-raw-materials).
 
@@
 
-We arranged `r n_sites_with_radiocarbon_ages` assemblages in chronological order (@fig-tcsa-radio-carbon).
-Our analysis shows that earlier assemblages before LGM tend to possess more stemmed points, such as Suyanggae, and these have a wide range of TCSA values, indicating multiple likely usages of stemmed points.
+We arranged TCSA values of `r n_sites_with_radiocarbon_ages` assemblages in chronological order (@fig-tcsa-radio-carbon panel A).
+Our analysis shows that earlier assemblages before LGM tend to possess more stemmed points such as Suyanggae.
+These assemblages have a wide range of TCSA values, indicating multiple likely usages of stemmed points.
 Overall, there is no clear pattern in these assemblages over time.
-Depending on the assemblage, there may be wide variation or narrow variation.
+The TCSA range varies depending on the assemblages.
 
@@
 #| label: fig-tcsa-radio-carbon
-#| fig-cap: "Distribution of TCSA values over time. The light gray shade indicates Marine isotope stages (MIS) 3, the gray shade indicates MIS 2 and the darkest shade indicates LGM period."
+#| fig-cap: "A: Distribution of TCSA values over time.
+The light gray shade indicates Marine isotope stages (MIS) 3, the gray shade indicates MIS 2 and the darkest shade indicates the duration of LGM.
+B: Distribution of TCSA values grouped by LGM event"
 #| fig-width:   4.5
@@
 
-temporal_tcsa +
+all_time <- temporal_tcsa +
  annotate("rect", 
@@
 
-```
-
-```{r}
-
-#| label: fig-tcsa-radio-carbon-box
-#| fig-cap: "Distribution of TCSA values grouped by LGM event"
-#| fig-width:   4.5
-#| fig-height:  4
-
 # aggregate all the MIS 3 sites, MIS 2 sites
@@
 
+three_time_p_pretty <- apastats::describe.Anova(p_lgm_test, 1, 2)
+
 # box plot for each MIS stage
-  ggplot(tcsa_site_ass_mis, 
+three_time <-  ggplot(tcsa_site_ass_mis, 
        aes(y = min_tcsa,
@@
 
-```
-
-We then grouped stemmed points into three categories based on LGM event: before, LGM, and after to examine possibility of environmental impact on the use of the artifacts (@fig-tcsa-radio-carbon-box).
+plot_grid(all_time, 
+          three_time, 
+          labels = c('A', 'B'),
+          ncol = 1)
+
+```
+
+We then grouped stemmed points into three categories based on LGM event: before, LGM, and after to examine possibility of environmental impact on the use of the artifacts (@fig-tcsa-radio-carbon panel B).
 Our results shows that stemmed points were made predominantly before LGM and only a few left after LGM.
 Each category shows wide variation in LGM, which indicates diverse uses of stemmed points.
-There is no strong correlation between the TCSA values and the time period (`apastats::describe.Anova(p_lgm_test, 1, 2)`).
+There is no strong correlation between the TCSA values and the time period (`r three_time_p_pretty`).
 
@@
   geom_quasirandom(alpha = 0.4,
-                   size = 2.3)
+                   size = 2.3)+
+   annotate("rect", 
+           xmin = 7.4,
+           xmax = 7.5,
+           ymin = 4,
+           ymax = 18,
+           fill = "green",
+           alpha = 0.2) +
+      annotate("text", 
+               size = point_function_group_label_size,
+               label = "Poisoned\narrow tips",
+               hjust = 0,
+               lineheight = 0.75,
+           x = 7.4,
+           y = (18-4) /2 + 4) +
+      annotate("rect", 
+           xmin = 7.5,
+           xmax = 7.6,
+           ymin = 17,
+           ymax = 47,
+           fill = "blue",
+           alpha = 0.2) +
+      annotate("text", 
+               size = point_function_group_label_size,
+               label = "Un-poisoned\narrow tips",
+               hjust = 0,
+               lineheight = 0.75,
+           x = 7.5,
+           y = (47-17) /2 + 17) +
+      annotate("rect", 
+           xmin = 7.6,
+           xmax = 7.7,
+           ymin = 40,
+           ymax = 76,
+           fill = "orange",
+           alpha = 0.2) +
+      annotate("text", 
+               size = point_function_group_label_size,
+               label = "Dart tips",
+               hjust = 0,
+           x = 7.6,
+           y = (76-40) /2 + 40) +
+      annotate("rect", 
+           xmin = 7.7,
+           xmax = 7.8,
+           ymin = 42,
+           ymax = 90,
+           fill = "purple",
+           alpha = 0.2) +
+      annotate("text", 
+               size = point_function_group_label_size,
+               label = "Javelin tips",
+               hjust = 0,
+           x = 7.7,
+           y = (90-42) /2 + 42) +
+     annotate("rect", 
+           xmin = 7.8,
+           xmax = 7.9,
+           ymin = 80,
+           ymax = 200,
+           fill = "yellow",
+           alpha = 0.2) +
+      annotate("text", 
+               label = "Stabbing\nspear tips",
+               size = point_function_group_label_size,
+               hjust = 0,
+               lineheight = 0.75,
+           x = 7.8,
+           y = (200-80) /2 + 80 )
 

### 2022-11-28 analysis/paper/paper.qmd
Wrote conclusion

@@
 We measured tip cross-sectional area (TCSA) to discriminate different likely use classes of projectile points, for example, as poisoned arrow tips or as stabbing spears.
-We analyzed TCSA with other variables including raw materials, weight, site, radiocarbon dates and locations.
+We analyzed TCSA with other variables including raw materials, weight, radiocarbon dates and locations.
 Our results show that the stemmed points were likely used  as javelin tips and stabbing spear tips, with smaller numbers as dart tips and un-poisoned arrow tips.
@@
 
-Tip cross-sectional area (TCSA) of stone artifacts has been used as a ballistically relevant standard to probabalistically discriminate different between functional classes of projectile armatures such as spearthrower (a.k.a.
+Tip cross-sectional area (TCSA) of stone artifacts has been used as a ballistically relevant standard to probabalistically discriminate difference between functional classes of projectile armatures such as spearthrower (a.k.a.
 atlatl) dart tips, un-poisoned arrow tips and large stabbing/thrusting spears [@sisk2011african; @lombard2021variation; @lombard2020tip; @lombard2021did; @o2018origins].
@@
 We explored the relationship between TCSA values and the location of assemblages by applying eco-regional information.
-We followed the TCSA ranges for different weapon-delivery systems that @lombard2022tip and @lombard2021variation created by summarizing @wadley2008segment' study of backed microliths and calibrating existing ranges (@tbl-tcsa-ranges).
-Among our dataset, we excluded `r n_omit` artifacts from our dataset with TCSA greater than 250 and thus outside of the range of the comparative data.
-For this study we include TCSA values from a total of `r min_n_points` stemmed points from `r n_ass` assemblages unearthed from `r n_sites` sites.
-Each analysis uses slightly different number of artifacts due to data availability.
 
@@
 
+We followed the TCSA ranges for different weapon-delivery systems that @lombard2022tip and @lombard2021variation created by summarizing @wadley2008segment' study of backed microliths and calibrating existing ranges (@tbl-tcsa-ranges).
+Among our dataset, we excluded `r n_omit` artifacts from our dataset with TCSA greater than 250 and thus outside of the range of the comparative data.
+For this study we include TCSA values from a total of `r min_n_points` stemmed points from `r n_ass` assemblages unearthed from `r n_sites` sites.
+Each analysis uses slightly different number of artifacts due to data availability.
+
 ### Raw materials
@@
 This map then modified by @yi2010vegetation.
-They divided South Korea, our research area, into three zones: Central Temperature Zone (CT), South Temperature Zone (ST), and Subtropical-warm Temperature Zone (SW).
+They divided South Korea, our research area, into three zones: Central Temperature Zone (CT), South Temperature Zone (ST), and Subtropical-warm Temperature Zone (SWT).
 We also used a map from @lee2008relationship, which separated our research area into 16 different zones based geographical characteristics such as inland, coastal, major rivers, island, and major mountain ranges in addition to ecological information including temperature and precipitation.
@@
 
-We computed artifact TCSA values for `r n_sites` sites that contain more than 5 stemmed points to observe variation between sites (@fig-tcsa-per-sites) Sites with fewer than 5 stemmed points were grouped under the category of 'other'.
+We computed artifact TCSA values for `r n_sites` sites that contain more than 5 stemmed points to observe variation between sites (@fig-tcsa-per-sites) Sites with fewer than 5 stemmed points were grouped under the category of "Other".
 Among the sites in our sample, Suyanggae has the most stemmed points and the widest range of TCSA values.
-This suggests that people made stemmed points for a variety of different purposes in Suyanggae.
-The TCSA values for Nosanri, Sibuk, and Wolpyeng are have a narrower range of lower values, which indicates less diverse functions for stemmed points.
+This suggests that people made stemmed points for accomplishing a variety of tasks at Suyanggae site.
+Nosanri, Sibuk, and Wolpyeng have a narrower range of lower values, which indicates less diverse likely uses for stemmed points.
 Yongsujaeul shows a narrower range but higher TCSA value.
@@
 #| fig-cap: "TCSA values by assemblages and ecoregions.
-A.
-Modern vegetation map defined by temporature and precipitation (modified the map from @yi2010vegetation, but original source is @yim1975distribution).
+A: Modern vegetation map defined by temporature and precipitation.
 The individual points indicate assemblages.
-B.
-Boxplots for  vegitation zones.
+B: TCSA distribution for vegitation zones.
 The individual points indicate artifacts.
-C.
-Ecoregion map based on geographical boundaries and ecological conditions (modified the map from @lee2008relationship).
-D.
-Boxplots for  ecoregion zones, excludings zones with no artifacts."
-#| fig-width:   5
-#| fig-height:  4
+C: Tukey test result for vegitation zones.
+D: Ecoregion map based on geographical boundaries and ecological conditions (modified the map from @lee2008relationship).
+E: TCSA distribution for ecoregion zones, excluding zones with no artifacts.
+F: Tukey test result for ecoregion zones"
+#| fig-width:   4.3
+#| fig-height:  5.8
 
@@
 
-```
-
-```{r}
-#| label: fig-tukey-region
-#| fig-cap: "."
-#| fig-width:   5
-#| fig-height:  4
-
+######Tukey test######
 #vegetation region#
@@
           for_tur_box,
-          labels = c('A', 'C', 'B', 'D', 'E', 'F'),
+          labels = c('A', 'D', 'B', 'E', 'C', 'F'),
           ncol = 2)
@@
 
-In order to explore spatial patterns as well as the relationship between the function of the tools and the environment, we located individual assemblages on maps that depict eco-regional boundaries.
-Our results show that stemmed points are only located in certain zones, Central Temperature Zone (CT) of vegetation map or Central inland (CI) (@fig-gis-tcsa-sites) of eco-region map.
+In order to explore spatial patterns as well as the relationship between the role of stemmed points and the environment, we located individual assemblages on maps that depict eco-regional boundaries.
+Our results show that stemmed points are only located in certain zones, Central Temperature Zone (CT) of vegetation map or Central inland (CI) (@fig-gis-tcsa-sites panel A & D) of eco-region map.
 We observed that TCSA values are strongely associated with the zones (Vegetation zones: `r eco_lgm_test_pretty`, Ecoregion zones: `r for_lgm_test_pretty`).
-To explore the TCSA distribution within the eco-regional zones, we excluded zones in the ecoregion map that have no stemmed points (@fig-gis-tcsa-sites panel D).
-We found no clear functional pattern throughout the zones, showing a wide variation in TCSA regardless of the density of stemmed point in each zone (@fig-gis-tcsa-sites panel B and D).
+To explore the TCSA distribution within the eco-regional zones, we excluded zones in the ecoregion map that have no stemmed points (@fig-gis-tcsa-sites panel E).
+We found no clear distribution pattern throughout the zones, showing a wide variation in TCSA regardless of the density of stemmed point in each zone.
+For example, explored the six artifacts from South Temperature Zone (ST) to examine the gap of TCSA values among the artifacts (i.e.
+Three artifacts are over 175 while the other three are lower around or lower than 100.
+@fig-gis-tcsa-sites panel B).
+We observed two artifacts from Bonggok site with no age difference, both made out of rhyolite, have huge difference in TCSA values, indicating no pattern for TCSA variation.
 
 We then computed Tukey's for comparison tests between the eco-regional zones.
-Our results show that ST and CT, and SWT-ST have significantly different while SWT and CT are close to each other (@fig-tukey-region panel A).
-We found no significant difference within the zones of eco-region map (@fig-tukey-region panel B).
+Our results show that ST and CT, and Subtropical-warm Temperature Zone (SWT)-ST have significantly different while SWT and CT are close to each other (@fig-gis-tcsa-sites panel C).
+However, we found no significant difference within the zones of eco-region map (@fig-gis-tcsa-sites panel F).
 
@@
 
-To investigate the role that stemmed points played in the technological transition during the Korean Late Paleolithic, we computed the TCSA metric to infer likely artifact uses by comparing the results to other archaeological and ethnographic studies Our main questions were: What were the best-fit ballistic probabilities for the stemmed points if they were hafted as weapon tips?
+To investigate the role that stemmed points played in the technological transition during the Korean Late Paleolithic, we computed the TCSA metric to infer likely uses by comparing the results to the TCSA range from other archaeological and ethnographic cases (@tbl-tcsa-ranges).
+Our main questions were: What were the best-fit ballistic probabilities for the stemmed points if they were hafted as weapon tips?
 How diverse were their likely uses?
@@
 
-Our results indicate that javelin tips and stabbing spear tips are the most probable ballistic uses for stemmed points.
+Our results show that javelin tips and stabbing spear tips are the most probable ballistic uses for stemmed points.
 There are a few stemmed points in our sample that may have been used as dart tips and un-poisoned arrow tips, but none of them have TCSA values in the range of poison arrow tips (@fig-tcsa-all-sp).
-But in general, we found a wide range of TCSA in our sample, indicating that stemmed points played a diverse role in foraging toolkits.
+But in general, we found a wide range of TCSA in our sample, indicating that stemmed points might play diverse roles in foraging toolkits.
 TCSA can be impacted by other factors such as raw materials and portability.
@@
 We found that different clusters of weight are matched with different types of weapon tips (@fig-artifact-size-cluster).
-The TCSA varies from assemblages to assemblages without any discernible temporal pattern in functional use (@fig-tcsa-radio-carbon).
-But the stemmed points mainly used before LGM (@fig-tcsa-radio-carbon-box).
+The TCSA varies from assemblages to assemblages without any discernible temporal pattern in functional use (@fig-tcsa-radio-carbon panel A).
+But the stemmed points mainly used before LGM (@fig-tcsa-radio-carbon panel B).
 We found that stemmed points are only located in certain eco-regional zones such as Central Temperature zone without any clear pattern in function (@fig-gis-tcsa-sites).
-Both temporal and spatial analyses show that stemmed points were used for a particular environment, not so much harsh as LGM.
+Both temporal and spatial analyses show that stemmed points were made for a particular environment, not so much harsh as LGM.
 @prates2022changes claimed that fishtail projectile points in South America was used for megafauna hunting and contributed to their extinction by proving strong relationship between the temporal density and spatial distribution of megafauna and the projectile points.
@@
 To examine the possible connections between the roles of stemmed points and environmental change or population dynamics during the Late Paleolithic period, we compared the temporal and spatial patterns of TCSA to results of simulated Mean Annual Temperature (MAT) and Summed probability distribution (SPD) from @park2022did.
-The simulated temperature distribution in south Korea during the Late Paleolithic period (Figure 9: A from @park2022did) seems similar to the modern eco-regional zones that we used in this research (@fig-gis-tcsa-sites).
-Figure 9: C from @park2022did shows the slow decrease of temperature until the end of LGM, which could explain our results of temporal pattern, meaning stemmed points were used during the cooler environment with a temperature between 4 and 6 ℃ (@fig-tcsa-radio-carbon-box).
+The simulated temperature distribution in south Korea during the Late Paleolithic period (Figure 9: A from @park2022did) seems similar to the modern eco-regional zones that we used in this research (@fig-gis-tcsa-sites panel A & D).
+Figure 9: C from @park2022did shows the slow decrease of temperature until the end of LGM, which could explain our results of temporal pattern, meaning stemmed points were used during the cooler environment with a temperature between 4 and 6 ℃.
 Results of SPD (Figure 10 from @park2022did) show several peaks, as indicated by the positive deviations from the three null models (shaded in red), and downs (shaded in blue), but no drastic growth or decline during our research period, indicating no major impact from population dynamics on stemmed points.
@@
 
-RESTATE QUESTIONS AND ANSWERS VERY BRIELFY
-
-possible role
-indicate people used stemmed poitns in different way
-examined their likely uses
-
-SUGGEST SOME POSSIBLE DIRECTIONS FOR FUTURE WORK, E.G. EXPERIMENTAL ARCHAEOLOGY, USEWEAR ANALYSIS
-
-We are aware of the arbitrary that discriminating the function of small projectile points could be arbitrary work [@erlandson2014darts].
-Because TCSA covers the critical elements of projectiles, flight and penetration dynamics (i.e.
-increased or decrease by shape of tip and cross section), we consider it a useful metric for discriminating between weapon tip functions [@hughes1998getting; @sitton2020tip; @lombard2021variation].
-Stemmed points have been considered so far as long-distance hunting amateurs pursuing large bodies of prey such as a deer [@lee2019were; @lee2011study; @seong2008tanged; @lee2002new; @chang2013human].
-TCSA enabled us to gain a more detailed understanding of the function of stemmed points.
+Considering the importance of stemmed points for the technological transition during the Late Paleolithic in Korea, our research examines the likely uses of stemmed points by asking three research questions: What were stemmed points used for?
+How diverse were their likely uses?
+What are the temporal and spatial patterns in stemmed point use?
+We applied tip cross-sectional area (TCSA) metric because it has been used as a ballistically relevant standard to discriminate different likely use classes of projectile points and it requires only a few measurements on a stone artifact (i.e width and thickness) to compute the metric.
+We calculated TCSA for a total of `r min_n_points` stemmed points from `r n_ass` assemblages unearthed from `r n_sites` sites.
+Then we examined the TCSA values with other variables including raw materials, weight, radiocarbon dates, and site locations.
+Based on the evolutionary approach, we premised that the way of using stemmed points is likely to reflect the socio-environmental circumstances that people encountered and managed.
+We examined the possible impact of LGM on stemmed points and their distribution in eco-regional maps.
+
+According to the different weapon-delivery systems that summarize TCSA of projectile points from archaeological and ethnographic cases, a majority of stemmed points might be used the most likely as javelin tips and stabbing spear tips.
+But overall we observed a wide range and also difference distribution of TCSA in each assemblage.
+Therefore, we conclude that stemmed points might play diverse roles during the Late Paleolithic.
+Based on our results, we assume that people were exposed to largely unfamiliar situation and created a stemmed point as multifunctional tool to performed multiple tasks.
+We are aware that discriminating the likely use of small projectile points could be arbitrary [@erlandson2014darts].
+Since TCSA covers the critical elements of projectiles, flight and penetration dynamics (i.e.
+increased or decrease by shape of tip and cross section), we still consider it a useful metric for discriminating between weapon tip tools [@hughes1998getting; @sitton2020tip; @lombard2021variation].
+In our future research, we will conduct usewear analysis on stemmed points and experimental research as another attempt to identify detailed function of stemmed points and the technological connection between hunter-gatherer groups.
 

### 2022-11-28 analysis/paper/paper.qmd
Edited conclusion

@@
 How diverse were their likely uses?
-What are the temporal and spatial patterns in stemmed point use?
-We applied tip cross-sectional area (TCSA) metric because it has been used as a ballistically relevant standard to discriminate different likely use classes of projectile points and it requires only a few measurements on a stone artifact (i.e width and thickness) to compute the metric.
+What are the temporal and spatial patterns of stemmed point use?
+We applied the tip cross-sectional area (TCSA) metric because it has been used as a ballistically relevant standard to discriminate different likely use classes of projectile points and it requires only a few measurements on a stone artifact (i.e.
+width and thickness) to compute the metric.
 We calculated TCSA for a total of `r min_n_points` stemmed points from `r n_ass` assemblages unearthed from `r n_sites` sites.
@@
 Based on the evolutionary approach, we premised that the way of using stemmed points is likely to reflect the socio-environmental circumstances that people encountered and managed.
-We examined the possible impact of LGM on stemmed points and their distribution in eco-regional maps.
+We examined the possible impact of LGM on stemmed points and their distribution on eco-regional maps.
 
 According to the different weapon-delivery systems that summarize TCSA of projectile points from archaeological and ethnographic cases, a majority of stemmed points might be used the most likely as javelin tips and stabbing spear tips.
-But overall we observed a wide range and also difference distribution of TCSA in each assemblage.
+In general, though, we noted a wide range and also differing distribution of TCSA in each assemblage.
 Therefore, we conclude that stemmed points might play diverse roles during the Late Paleolithic.
-Based on our results, we assume that people were exposed to largely unfamiliar situation and created a stemmed point as multifunctional tool to performed multiple tasks.
+From our findings, we infer that people were exposed to an unfamiliar situation and used a stemmed point as a multifunctional tool to carry out multiple tasks.
 We are aware that discriminating the likely use of small projectile points could be arbitrary [@erlandson2014darts].
@@
 increased or decrease by shape of tip and cross section), we still consider it a useful metric for discriminating between weapon tip tools [@hughes1998getting; @sitton2020tip; @lombard2021variation].
-In our future research, we will conduct usewear analysis on stemmed points and experimental research as another attempt to identify detailed function of stemmed points and the technological connection between hunter-gatherer groups.
+As part of our future research, we will conduct usewear analyses of stemmed points and experimental research in an effort to elucidate the detailed function of stemmed points and technological links between hunter-gatherer groups.
 

### 2022-11-28 analysis/paper/paper.qmd
Changed table code

@@
 #| label: tbl-tcsa-ranges
-#| tbl-cap: TCSA range from Lombard (2021 & 2022)
-
-library(gt)
-library(sass)
+#| tbl-cap: "TCSA range from Lombard (2021 & 2022)"
+
+# library(gt)
+# library(sass)
 read_csv(here::here("analysis/data/raw_data/tcsa-range-from-lombard-2022.csv")) %>% 
-  gt() %>% 
-  tab_options(row.striping.include_table_body = FALSE) %>%
-  tab_style(
-    style = cell_borders(
-      sides = c("top"),
-      color = "black",
-      weight = px(3),
-      style = "solid"
-    ),
-    locations = cells_body(
-      columns = everything(),
-      rows = c(1)
-    )
-  ) %>% 
-  tab_options(
-    table.border.top.style = "hidden",
-    table.border.bottom.style = "hidden"
-  )
+  knitr::kable(.)
+  # gt() %>% 
+  # tab_options(row.striping.include_table_body = FALSE) %>%
+  # tab_style(
+  #   style = cell_borders(
+  #     sides = c("top"),
+  #     color = "black",
+  #     weight = px(3),
+  #     style = "solid"
+  #   ),
+  #   locations = cells_body(
+  #     columns = everything(),
+  #     rows = c(1)
+  #   )
+  # ) %>% 
+  # tab_options(
+  #   table.border.top.style = "hidden",
+  #   table.border.bottom.style = "hidden"
+  # )
 ```

### 2022-12-02 analysis/paper/paper.qmd
Edited plots based on the look in Wordfile.

@@
            x = point_function_group_label_y,
-           y = (76-40) /2 + 40 ) +
+           y = (76-40) /2 + 40 - 5) +
       annotate("rect", 
@@
   geom_quasirandom(alpha = 0.4,
-                   size = 1.1) +
+                   size = 0.7) +
   coord_cartesian(xlim = c(-0.1,1)) +
@@
 #| fig-cap: "TCSA values by lithic raw material."
-#| fig-width:   4
-#| fig-height:  3.5
+#| fig-width:   4.4
+#| fig-height:  3.2
 
@@
        y = "TCSA") +
-  theme(axis.text.x = element_text(size = 6)) +
+  theme(axis.text.x = element_text(size = 6),
+        axis.text.y = element_text(size = 6),
+        text = element_text(size=8)) +
   scale_x_discrete(guide = guide_axis(angle = 45))+
   geom_quasirandom(alpha = 0.4,
-                   size = 1.9) +
-   annotate("rect", 
-           xmin = 6.4,
-           xmax = 6.5,
+                   size = 1.3) +
+   annotate("rect", 
+           xmin = 0.5,
+           xmax = 1.7,
            ymin = 4,
@@
            alpha = 0.2) +
-      annotate("text", 
-               size = point_function_group_label_size,
-               label = "Poisoned\narrow tips",
-               hjust = 0,
-               lineheight = 0.75,
-           x = 6.4,
-           y = (18-4) /2 + 4) +
-      annotate("rect", 
-           xmin = 6.5,
-           xmax = 6.6,
+      annotate("rect", 
+           xmin = 1.7,
+           xmax = 2.9,
            ymin = 17,
@@
            alpha = 0.2) +
-      annotate("text", 
-               size = point_function_group_label_size,
-               label = "Un-poisoned\narrow tips",
-               hjust = 0,
-               lineheight = 0.75,
-           x = 6.5,
-           y = (47-17) /2 + 17) +
-      annotate("rect", 
-           xmin = 6.6,
-           xmax = 6.7,
+      annotate("rect", 
+           xmin = 2.9,
+           xmax = 4.1,
            ymin = 40,
@@
            alpha = 0.2) +
-      annotate("text", 
-               size = point_function_group_label_size,
-               label = "Dart tips",
-               hjust = 0,
-           x = 6.6,
-           y = (76-40) /2 + 40) +
-      annotate("rect", 
-           xmin = 6.7,
-           xmax = 6.8,
+      annotate("rect", 
+           xmin = 4.1,
+           xmax = 5.3,
            ymin = 42,
@@
            alpha = 0.2) +
-      annotate("text", 
-               size = point_function_group_label_size,
-               label = "Javelin tips",
-               hjust = 0,
-           x = 6.7,
-           y = (90-42) /2 + 42) +
-     annotate("rect", 
-           xmin = 6.8,
-           xmax = 6.9,
+     annotate("rect", 
+           xmin = 5.3,
+           xmax = 6.5,
            ymin = 80,
@@
            fill = "yellow",
-           alpha = 0.2) +
-      annotate("text", 
-               label = "Stabbing\nspear tips",
-               size = point_function_group_label_size,
-               hjust = 0,
-               lineheight = 0.75,
-           x = 6.8,
-           y = (200-80) /2 + 80 )
+           alpha = 0.2) 
+ # + annotate("text", 
+ #               size = 2,
+ #               label = "Poisoned arrow tips",
+ #               hjust = 0,
+ #               lineheight = 0.75,
+ #           x = 0.75,
+   #         y = (18-4) /2 + 4) +
+   # annotate("text",
+   #          size = 2,
+   #             label = "Un-poisoned arrow tips",
+   #             hjust = 0,
+   #             lineheight = 0.75,
+   #         x = 1.85,
+   #         y = (47-17) /2 + 17) +
+   # annotate("text", 
+   #             size = 2,
+   #             label = "Dart tips",
+   #             hjust = 0,
+   #         x = 3.3,
+   #         y = (76-40) /2 + 40) +
+   # annotate("text", 
+   #             size = 2,
+   #             label = "Javelin tips",
+   #             hjust = 0,
+   #         x = 4.4,
+   #         y = (90-42) /2 + 42) +
+    # annotate("text", 
+    #            label = "Stabbing spear tips",
+    #            size = 2,
+    #            hjust = 0,
+    #            lineheight = 0.75,
+    #        x = 5.52,
+    #        y = (200-80) /2 + 80 )
 
@@
                    alpha = 0.4,
-                   size = 1.5) +
+                   size = 1.1) +
   labs(shape='Cluster',
@@
                    alpha = 0.4,
-                   size = 1.5) +
+                   size = 1.1) +
   labs(shape='Cluster') +
   theme_bw(base_size = 6) +
-  # show regions for tools from Table 3 of Lomabrd 2021
-  # https://www-sciencedirect-com.offcampus.lib.washington.edu/science/article/pii/S0277379121002286
     annotate("rect", 
@@
            x = point_function_group_label_y,
-           y = (76-40) /2 + 40) +
+           y = (76-40) /2 + 40 - 5) +
       annotate("rect", 
@@
 B: Distribution of TCSA values grouped by LGM event"
-#| fig-width:   4.5
+#| fig-width:   3.8
 #| fig-height:  4
@@
   geom_quasirandom(alpha = 0.4,
-                   size = 2.3) +
+                   size = 1.9) +
     scale_x_discrete(limits=rev) +
      annotate("rect", 
-           xmin = 3.4,
-           xmax = 3.5,
+           xmin = 0.3,
+           xmax = 1,
            ymin = 4,
@@
            alpha = 0.2) +
-      annotate("text", 
-               size = point_function_group_label_size,
-               label = "Poisoned\narrow tips",
-               hjust = 0,
-               lineheight = 0.75,
-           x = 3.4,
-           y = (18-4) /2 + 4) +
-      annotate("rect", 
-           xmin = 3.5,
-           xmax = 3.6,
+      # annotate("text", 
+      #          size = point_function_group_label_size,
+      #          label = "Poisoned\narrow tips",
+      #          hjust = 0,
+      #          lineheight = 0.75,
+      #      x = 3.4,
+      #      y = (18-4) /2 + 4) +
+      annotate("rect", 
+           xmin = 1,
+           xmax = 1.7,
            ymin = 17,
@@
            alpha = 0.2) +
-      annotate("text", 
-               size = point_function_group_label_size,
-               label = "Un-poisoned\narrow tips",
-               hjust = 0,
-               lineheight = 0.75,
-           x = 3.5,
-           y = (47-17) /2 + 17) +
-      annotate("rect", 
-           xmin = 3.6,
-           xmax = 3.7,
+      # annotate("text", 
+      #          size = point_function_group_label_size,
+      #          label = "Un-poisoned\narrow tips",
+      #          hjust = 0,
+      #          lineheight = 0.75,
+      #      x = 3.5,
+      #      y = (47-17) /2 + 17) +
+      annotate("rect", 
+           xmin = 1.7,
+           xmax = 2.4,
            ymin = 40,
@@
            alpha = 0.2) +
-      annotate("text", 
-               size = point_function_group_label_size,
-               label = "Dart tips",
-               hjust = 0,
-           x = 3.6,
-           y = (76-40) /2 + 40) +
-      annotate("rect", 
-           xmin = 3.7,
-           xmax = 3.8,
+      # annotate("text", 
+      #          size = point_function_group_label_size,
+      #          label = "Dart tips",
+      #          hjust = 0,
+      #      x = 3.6,
+      #      y = (76-40) /2 + 40) +
+      annotate("rect", 
+           xmin = 2.4,
+           xmax = 3.1,
            ymin = 42,
@@
            alpha = 0.2) +
-      annotate("text", 
-               size = point_function_group_label_size,
-               label = "Javelin tips",
-               hjust = 0,
-           x = 3.7,
-           y = (90-42) /2 + 42) +
-     annotate("rect", 
-           xmin = 3.8,
-           xmax = 3.9,
+      # annotate("text", 
+      #          size = point_function_group_label_size,
+      #          label = "Javelin tips",
+      #          hjust = 0,
+      #      x = 3.7,
+      #      y = (90-42) /2 + 42) +
+     annotate("rect", 
+           xmin = 3.1,
+           xmax = 3.8,
            ymin = 80,
@@
            fill = "yellow",
-           alpha = 0.2) +
-      annotate("text", 
-               label = "Stabbing\nspear tips",
-               size = point_function_group_label_size,
-               hjust = 0,
-               lineheight = 0.75,
-           x = 3.8,
-           y = (200-80) /2 + 80 )
+           alpha = 0.2) 
+# + annotate("text", 
+#                label = "Stabbing\nspear tips",
+#                size = point_function_group_label_size,
+           #     hjust = 0,
+           #     lineheight = 0.75,
+           # x = 3.8,
+           # y = (200-80) /2 + 80 )
 
@@
           labels = c('A', 'B'),
-          ncol = 1)
+          ncol = 1,
+          rel_heights = c(2.7, 1.5))
 

### 2022-12-02 analysis/paper/paper.qmd
Edited plots

@@
           labels = "AUTO",
+          label_size = 9,
           rel_widths = c(0.5, 1))
@@
 #| fig-cap: "Artifact size and TSCA values by lithic raw material"
-#| fig-width:   3.5
-#| fig-height:  3
+#| fig-height:  3.5
 
@@
            colour = raw_materials_top)) +
-  geom_point() +
+  geom_point(size = 1) +
     labs(x = "Weight (g)", 
@@
        colour='Raw material') +
-  geom_smooth(method=lm, se=FALSE, size = 0.4)+
+  geom_smooth(method=lm, se=FALSE, size = 0.3)+
   theme_bw(base_size = 6) +
   theme(
-  legend.title = element_text(size = 5),
-  legend.text = element_text(size = 4),
-  legend.position = c(.8,.3)
-  )
+  legend.title = element_text(size = 4),
+  legend.text = element_text(size = 3),
+  legend.position = c(.8,.3),
+  legend.key.height = unit(4.5, 'mm'),
+  legend.key.width = unit(2, 'mm'))
+
 ```
@@
 B: Distribution of TCSA values grouped by LGM event"
-#| fig-width:   3.8
-#| fig-height:  4
+#| fig-height:  5
 
@@
            x = reorder(label_site, -age_ka))) +
-   geom_boxplot() +
+   geom_boxplot(size = 0.2) +
    theme_minimal() +
@@
         x = "") +
-  theme(axis.text.x = element_text(size = 6)) +
+  theme(axis.text.x = element_text(size = 3.5),
+        axis.text.y = element_text(size = 3.5),
+        text = element_text(size=5)) +
   scale_x_discrete(guide = guide_axis(angle = 45))
@@
   annotate("text", 
-               size = 3.5,
+               size = 3,
                label = "MIS 3",
@@
   annotate("text", 
-               size = 3.5,
+               size = 3,
                label = "MIS 2",
@@
   annotate("text", 
-               size = 2.3,
+               size = 2.1,
                label = "LGM",
@@
            x = factor(mis_stage, level = c('After', 'LGM', 'Before')))) +
-   geom_boxplot() +
+   geom_boxplot(size = 0.4) +
    theme_minimal() +
-   theme(legend.position="none") +
+   theme(legend.position="none", 
+         axis.text.x = element_text(size = 3.5),
+        axis.text.y = element_text(size = 3.5),
+        text = element_text(size=5)) +
     labs(y = "TCSA", 
@@
   geom_quasirandom(alpha = 0.4,
-                   size = 1.9) +
+                   size = 1.1) +
     scale_x_discrete(limits=rev) +
@@
           three_time, 
-          labels = c('A', 'B'),
+          labels = "AUTO",
+          label_size = 9,
           ncol = 1,
-          rel_heights = c(2.7, 1.5))
+          rel_heights = c(2.7, 2))
 
@@
 #| fig-cap: "TCSA values by archaeological site."
-#| fig-width:   4.5
-#| fig-height:  4
+#| fig-height:  3
 
@@
        y = "TCSA") +
-  theme(axis.text.x = element_text(size = 6)) +
+  theme(axis.text.x = element_text(size = 5)) +
   scale_x_discrete(guide = guide_axis(angle = 45))+
   geom_quasirandom(alpha = 0.4,
-                   size = 2.3)+
-   annotate("rect", 
-           xmin = 7.4,
-           xmax = 7.5,
+                   size = 1.6)+
+   annotate("rect", 
+           xmin = 0.4,
+           xmax = 1.9,
            ymin = 4,
@@
            alpha = 0.2) +
-      annotate("text", 
-               size = point_function_group_label_size,
-               label = "Poisoned\narrow tips",
-               hjust = 0,
-               lineheight = 0.75,
-           x = 7.4,
-           y = (18-4) /2 + 4) +
-      annotate("rect", 
-           xmin = 7.5,
-           xmax = 7.6,
+      annotate("rect", 
+           xmin = 1.9,
+           xmax = 3.4,
            ymin = 17,
@@
            alpha = 0.2) +
-      annotate("text", 
-               size = point_function_group_label_size,
-               label = "Un-poisoned\narrow tips",
-               hjust = 0,
-               lineheight = 0.75,
-           x = 7.5,
-           y = (47-17) /2 + 17) +
-      annotate("rect", 
-           xmin = 7.6,
-           xmax = 7.7,
+      annotate("rect", 
+           xmin = 3.4,
+           xmax = 4.9,
            ymin = 40,
@@
            alpha = 0.2) +
-      annotate("text", 
-               size = point_function_group_label_size,
-               label = "Dart tips",
-               hjust = 0,
-           x = 7.6,
-           y = (76-40) /2 + 40) +
-      annotate("rect", 
-           xmin = 7.7,
-           xmax = 7.8,
+      annotate("rect", 
+           xmin = 4.9,
+           xmax = 6.4,
            ymin = 42,
@@
            alpha = 0.2) +
-      annotate("text", 
-               size = point_function_group_label_size,
-               label = "Javelin tips",
-               hjust = 0,
-           x = 7.7,
-           y = (90-42) /2 + 42) +
-     annotate("rect", 
-           xmin = 7.8,
+     annotate("rect", 
+           xmin = 6.4,
            xmax = 7.9,
@@
            fill = "yellow",
-           alpha = 0.2) +
-      annotate("text", 
-               label = "Stabbing\nspear tips",
-               size = point_function_group_label_size,
-               hjust = 0,
-               lineheight = 0.75,
-           x = 7.8,
-           y = (200-80) /2 + 80 )
+           alpha = 0.2)
 
@@
 F: Tukey test result for ecoregion zones"
-#| fig-width:   4.3
-#| fig-height:  5.8
+#| fig-height:  5
 
@@
    geom_sf_text(data = veg_all, aes(label = ecoregion), size =1.5)+
-  geom_sf(data = pts_all, aes(color = min_tcsa)) +
+  geom_sf(data = pts_all, aes(color = min_tcsa), size = 1.1) +
   theme_minimal() +
@@
   guides(fill = FALSE) +
-  theme(axis.text.x = element_text(angle = 45, size = 6))+
-  theme(axis.text.y = element_text(size = 6))+
+  theme(axis.text.x = element_text(angle = 45, size = 4.5))+
+  theme(axis.text.y = element_text(size = 4.5))+
   theme(axis.title.x = element_blank(),
@@
       min_tcsa) +
-  geom_boxplot() +
+  geom_boxplot(size = 0.2) +
   geom_quasirandom(alpha = 0.3,
-                   size = 2.3) +
+                   size = 1.3) +
   xlab("Vegetation") + 
   ylab("TCSA") +
-  theme_minimal()
+  theme_minimal() +
+    theme(axis.text.x = element_text(size = 4.5)) +
+  theme(axis.text.y = element_text(size = 4.5)) +
+   theme(axis.title.x = element_text(size = 6),
+          axis.title.y = element_text(size = 6))
 
@@
   geom_sf_text(data = veg_forest, aes(label = Zone), size =1.5)+
-  geom_sf(data = pts_forest, aes(color = min_tcsa)) +
+  geom_sf(data = pts_forest, aes(color = min_tcsa), size = 1.1) +
   theme_minimal()+
@@
   guides(fill = FALSE) +
-  theme(axis.text.x = element_text(angle = 45, size = 6)) +
-  theme(axis.text.y = element_text(size = 6)) +
+  theme(axis.text.x = element_text(angle = 45, size = 4.5)) +
+  theme(axis.text.y = element_text(size = 4.5)) +
   theme(axis.title.x = element_blank(),
-          axis.title.y = element_blank())
+          axis.title.y = element_blank()) +
+  theme(legend.title = element_text(size= 6),
+        legend.key.height = unit(4.5, 'mm'),
+        legend.key.width = unit(2, 'mm'),
+        legend.text = element_text(size = 4))
 
@@
       min_tcsa) +
-  geom_boxplot() +
+  geom_boxplot(size = 0.2) +
   geom_quasirandom(alpha = 0.3,
-                   size = 2.3) +
+                   size = 1.3) +
   xlab("Ecoregion") + 
   ylab("TCSA") +
-  theme_minimal()
+  theme_minimal() +
+      theme(axis.text.x = element_text(size = 4.5)) +
+  theme(axis.text.y = element_text(size = 4.5)) +
+   theme(axis.title.x = element_text(size = 6),
+          axis.title.y = element_text(size = 6))
 
@@
                      xmax = upr,
-                     y = zones)) +
+                     y = zones),
+                 size = 0.4) +
   geom_point(aes(diff,
                  zones),
-             size = 4) +
+             size = 1) +
   geom_vline(xintercept = 0, colour = "red") +
@@
   xlab("Differences in means") +
-  theme_minimal(base_size = 8)
+  theme_minimal(base_size = 8) +
+  theme(axis.text.x = element_text(size = 4.5)) +
+  theme(axis.text.y = element_text(size = 4.5)) +
+   theme(axis.title.x = element_text(size = 6),
+          axis.title.y = element_text(size = 6))
 
@@
                      xmax = upr,
-                     y = zones)) +
+                     y = zones), 
+                 size = 0.4) +
   geom_point(aes(diff,
                  zones),
-             size = 4) +
+             size = 1) +
   geom_vline(xintercept = 0, colour = "red") +
@@
           labels = c('A', 'D', 'B', 'E', 'C', 'F'),
+          label_size = 9,
           ncol = 2)

### 2022-12-03 analysis/paper/paper.qmd
stash

--- 2022-11-28 analysis/paper/paper.qmd
@@
 #| label: tbl-tcsa-ranges
-#| tbl-cap: "TCSA range from Lombard (2021 & 2022)"
-
-# library(gt)
-# library(sass)
+#| tbl-cap: TCSA range from Lombard (2021 & 2022)
+
+library(gt)
+library(sass)
 read_csv(here::here("analysis/data/raw_data/tcsa-range-from-lombard-2022.csv")) %>% 
-  knitr::kable(.)
-  # gt() %>% 
-  # tab_options(row.striping.include_table_body = FALSE) %>%
-  # tab_style(
-  #   style = cell_borders(
-  #     sides = c("top"),
-  #     color = "black",
-  #     weight = px(3),
-  #     style = "solid"
-  #   ),
-  #   locations = cells_body(
-  #     columns = everything(),
-  #     rows = c(1)
-  #   )
-  # ) %>% 
-  # tab_options(
-  #   table.border.top.style = "hidden",
-  #   table.border.bottom.style = "hidden"
-  # )
+  gt(caption = "TCSA range from Lombard (2021 & 2022)") %>% 
+  tab_options(row.striping.include_table_body = FALSE) %>%
+  tab_style(
+    style = cell_borders(
+      sides = c("top"),
+      color = "black",
+      weight = px(3),
+      style = "solid"
+    ),
+    locations = cells_body(
+      columns = everything(),
+      rows = c(1)
+    )
+  ) %>% 
+  tab_options(
+    table.border.top.style = "hidden",
+    table.border.bottom.style = "hidden"
+  )
 ```

### 2023-01-20 analysis/paper/paper.qmd
minor fixes to enable rendering

--- 2022-12-02 analysis/paper/paper.qmd
@@
 The temporal pattern of TCSA values is one of little change throughout the Late Paleolithic period but SP were mainly used before Last Glacial Maximum (LGM).
-We observed that SP was only located in certain ecoregions in Korea but no clear functional pattern is confirmed.
+We observed that stemmed points was only located in certain ecoregions in Korea but no clear functional pattern is confirmed.
 We conclude that stemmed points were multi-functional tools, with many likely designed for javelin and stabbing spear tips.
+<!-- BM: as I recall Marlize does not agree that TCSA says anything about 'use' so I recommend you either prepare for her negative feedback about that, or edit the text everywhere to follow her way of writing about TCSA and artefact fucntion -->
 
@@
 To understand the roles that stemmed points played in the technological transition, this study examines the possible uses of stemmed points.
-We use the tip cross-sectional area (TCSA) metric to infer likely artifact use, and compare the results to other archaeological and ethnographic assemblages based on the idea that different shapes correspond to different weapon types [@lombard2021variation].
+We use the tip cross-sectional area (TCSA) metric to infer likely artifact use <!-- BM: is is how Marlize writes about how TCSA is interpreted?
+I recommend you re-check her papers and study very carefully the phrasing she uses, and use that throughout this paper also.
+Otherwise she may have negative feedback that will delay progress on this paper -->, and compare the results to other archaeological and ethnographic assemblages based on the idea that different shapes correspond to different weapon types [@lombard2021variation].
 We then explore the relationship between likely use and raw materials, artifact size and discard location, and how these change over time.
@@
 As part of our discussion, we examine possible connections between the roles of stemmed points and population dynamics or environmental change during the Late Paleolithic period.
-In order to understand how a certain weapon-tip type was chosen, we apply an evolutionary perspective with the assumption that people chose their weapon tip types depending on their hunting strategies and socio-environmental circumstances.
+In order to understand how a certain weapon-tip type was chosen, we apply an evolutionary perspective with the assumption that people chose their weapon tip types as part of their adaptation to the socio-environmental circumstances they found themselves in.
 
@@
 
-# library(gt)
-# library(sass)
+library(gt)
+library(sass)
+
 read_csv(here::here("analysis/data/raw_data/tcsa-range-from-lombard-2022.csv")) %>% 
-  knitr::kable(.)
-  # gt() %>% 
-  # tab_options(row.striping.include_table_body = FALSE) %>%
-  # tab_style(
-  #   style = cell_borders(
-  #     sides = c("top"),
-  #     color = "black",
-  #     weight = px(3),
-  #     style = "solid"
-  #   ),
-  #   locations = cells_body(
-  #     columns = everything(),
-  #     rows = c(1)
-  #   )
-  # ) %>% 
-  # tab_options(
-  #   table.border.top.style = "hidden",
-  #   table.border.bottom.style = "hidden"
-  # )
+  knitr::kable(caption = "TCSA range from Lombard (2021 & 2022)")
 ```

### 2023-01-20 analysis/paper/paper.qmd
editing up to 'Tip Cross-Sectional Area' section

@@
 Some sites show a wide range of TCSA values that represent multi-purpose usage of stemmed points.
-The temporal pattern of TCSA values is one of little change throughout the Late Paleolithic period but SP were mainly used before Last Glacial Maximum (LGM).
-We observed that stemmed points was only located in certain ecoregions in Korea but no clear functional pattern is confirmed.
+The temporal pattern of TCSA values is one of little change throughout the Late Paleolithic period but points were mainly used before Last Glacial Maximum (LGM).
+We observed that stemmed points were mostly located in certain ecoregions in Korea but no clear spatial pattern is confirmed.
 We conclude that stemmed points were multi-functional tools, with many likely designed for javelin and stabbing spear tips.
-<!-- BM: as I recall Marlize does not agree that TCSA says anything about 'use' so I recommend you either prepare for her negative feedback about that, or edit the text everywhere to follow her way of writing about TCSA and artefact fucntion -->
 
@@
 Stemmed points are considered the first evidence of a suite of new technologies defining the Late Paleolithic in this region [@seong2016eastern; @seong2008tanged].
-This is related to the fact that stemmed points originate from Korea and spread throughout Northeast Asia, and they have a close association with mobility, site formation, and occupation diversity [@o2018origins; @Chong_2021; @park2022did].
+This is related to the fact that stemmed points appear to originate from Korea and spread throughout Northeast Asia, and they have a close association with mobility, site formation, and occupation diversity [@o2018origins; @Chong_2021; @park2022did].
 Despite the importance of stemmed points, only a few studies to date have examined their likely uses.
@@
 To understand the roles that stemmed points played in the technological transition, this study examines the possible uses of stemmed points.
-We use the tip cross-sectional area (TCSA) metric to infer likely artifact use <!-- BM: is is how Marlize writes about how TCSA is interpreted?
-I recommend you re-check her papers and study very carefully the phrasing she uses, and use that throughout this paper also.
-Otherwise she may have negative feedback that will delay progress on this paper -->, and compare the results to other archaeological and ethnographic assemblages based on the idea that different shapes correspond to different weapon types [@lombard2021variation].
+We use the tip cross-sectional area (TCSA) metric to infer likely artifact use, and compare the results to other archaeological and ethnographic assemblages based on the idea that different shapes correspond to different weapon types [@lombard2021variation].
 We then explore the relationship between likely use and raw materials, artifact size and discard location, and how these change over time.
@@
 Elsewhere in the world these types of artifacts are often called 'tanged points', but we prefer 'stemmed point' here to distinguish from Bronze Age stone projectile points that have long been called 'tanged points' in Korea [@park2022did].
-Understanding the appearance of stemmed points is relevant to general questions about the origins of projectile technology, the emergence of the Late Paleolithic, and relationships to adjacent regions in East Asia, such as Japan.
+Understanding the appearance of stemmed points is relevant to general questions about the origins of projectile technology, the emergence of the Late Paleolithic, and relationships between Korea and adjacent regions in East Asia, such as Japan.
 
@@
 Since stemmed points represent the beginning of the Korean Late Paleolithic, investigating their origins is important for understanding the technological transition from the Early Paleolithic, modern human dispersals into the region, and claims for the existence of the 'Middle' Paleolithic in Korea [@bae2010origin; @bae2017late; @norton2009evolution; @seong2016eastern].
-There has been a noteworthy debate about explaining the origin(s) of stemmed points which can be summarized into two competing models: a 'heterogenic' migration [@bae2010origin], and an *in situ* evolution model [@seong2009emergence].
+Debate about explaining the origin of stemmed points which can be summarized into two competing models: a 'heterogenic' migration [@bae2010origin], and an *in situ* evolution model [@seong2009emergence].
 The migration model claims that the new blade industry including stemmed points, and the earlier coarse flake tradition including large cores, polyhedrons, choppers, and even handaxes, came from different origins (i.e.
@@
 Preserved wooden components of projectile tools are too few and rare to generalize the overall scale and variability of the tools used in prehistory [@shea2006origins].
-@Lee_Kong_2002 claimed stemmed points were 'stemmed tools' because of their diverse shapes, which could represent different functions, as well as the uncertainty of complete tool shape.
-Other researchers agree that a few stemmed points that are not symmetrical and have retouch on one side of the artifacts or have denticulate blades on one side should not be called as stemmed points but instead stemmed knives, stemmed side-scrapers, stemmed end-scrapers, or stemmed burins [@seong2008tanged; @Lee_Jang_2011; @Kim] @fig-sp-hand-made: 7-8.
+@Lee_Kong_2002 claimed stemmed points were generalised 'stemmed tools' because of their diverse shapes, which could represent different functions, as well as the uncertainty of complete tool shape.
+Other researchers propose that non-symmetrical stemmed points, with retouch on one side of the artifact or denticulate blades on one side, should not be called as stemmed points but instead stemmed knives, stemmed side-scrapers, stemmed end-scrapers, or stemmed burins [@seong2008tanged; @Lee_Jang_2011; @Kim] @fig-sp-hand-made: 7-8.
 
@@
 Their results show that a considerable number of the stemmed points show diagnostic impact fractures (DIFs) on the surface, which likely occurred due to longitudinal forces from the shaft when a stone tip hits a target.
-In other words, their research indicates that stemmed points may have been used as spear-throwers or bows [@sano2015backed].
+They conclude that stemmed points may have been used as spear-throwers or bows [@sano2015backed].
 

### 2023-01-23 analysis/paper/paper.qmd
edited up to spatial results

@@
 The TCSA metric represents the part of the tool that cuts the target's hide and relates to weapon flight and penetration dynamics [@hughes1998getting; @sitton2020tip; @lombard2021variation].
-This method was first proposed by @hughes1998getting and and tested by @shea2006origins by comparing archaeological examples to ethnographically collected samples of known use.
+This method was first proposed by @hughes1998getting and and validated by @shea2006origins by comparing archaeological examples to ethnographically collected samples of known use.
 @shea2009impact also applied this approach to compare projectile points from Africa, the Levant, and Europe, and claimed that projectile weapons first appeared in Africa.
+
 A key advantage of the TCSA metric is its convenience of application: regardless of point type, only maximum width and thickness measurements are needed for calculating the TCSA value (0.5 x maximum width x maximum thickness) [@sisk2011african; @lombard2020tip].
-Later, @sisk2011african proposed an alternate metric, tip cross-sectional perimeter (TCSP), for more accurate measure for the force needed to penetrate a target to a lethal depth while TCSA metric is more associated with cutting aspect.
+Later, @sisk2011african proposed an alternate metric, tip cross-sectional perimeter (TCSP), for more accurate measure for the force needed to penetrate a target to a lethal depth, while TCSA metric is more associated with cutting aspect.
 However, TCSP has a few disadvantages that limit its applicability to our case study of Korean stemmed points.
-The force and penetration depth are not only affected by the stone tip, but also by the mass of the shaft, which cannot be known for most archaeological stone-tipped weapons [@lombard2020tip].
-@sisk2011african also mentioned that TCSP cannot be applied to backed pieces that are hafted as projectile armatures.
+The force and penetration depth are not only affected by the stone tip, but also by the mass of the shaft, which cannot be known for most archaeological stone-tipped weapons because they have not been preserved in the archaeological record [@lombard2020tip].
+@sisk2011african also mentioned that TCSP cannot be applied to backed pieces that were hafted as projectile armatures.
 
 @lee2019were first applied TCSA to stemmed points from Korea along with use-wear analysis.
-They analyzed stemmed points from Jingeuneul, located in southwest of Korea, which has the largest number of stemmed points (n = 99) ever found in a single site.
+They analyzed stemmed points from Jingeuneul, located in southwest of Korea, which has the largest number of stemmed points (n = 99) found so far in a single site.
 For TCSA, they were only able to use ten stemmed points because they needed to select the stemmed points that retained the widest and thickest part of the specimens for TCSA metric and also displayed diagnostic impact fractures for the use-wear analysis.
@@
 Among these stemmed points, we selected those that retain their widest and thickest parts.
-Our dataset contains stemmed points that were found during field surveys as well as those were from sites that were never dated but were associtaed with other Late Paleoltihic artifacts.
+We selected stemmed points that were found during field surveys as well as those were from sites that were never dated but were associated with other Late Palaeolithic artifacts.
 Applying these sampling criteria resulted in a sample of `r total_sp` stemmed points from `r total_ass` assemblages unearthed from `r total_site` sites spanning the period 44-10 ka (@fig-map).
@@
 We separated one assemblage among the four by using a different site name, Hajinri, following the convention established the excavators of that location.
-Because Hajinri was from the sixth excavation area, which is 3.5 kilometers apart from the other spots and dated to much earlier (around 42-30 ka) than the other assemblages (around 31-15 ka).
-According to the excavation reports for Hajinri, stemmed points first appeared in Hajinri around 42 ka with the earliest blade and blade cores [@Lee_Woo_Lee_An_Yun_Park_Otani_Kim_Kim_Han_et_al_2018].
-We consider this assemblage as an unusual outlier because the stemmed points and blades from Hajinri are highly standardized and refined, which are only found at other sites much later in time.
+Hajinri is the sixth excavation area at Suyanggae, which is 3.5 kilometers apart from the other areas and dated to much earlier (around 42-30 ka) than the other assemblages (around 31-15 ka).
+According to the excavation reports for Hajinri, stemmed points first appeared there around 42 ka with the earliest blade and blade cores [@Lee_Woo_Lee_An_Yun_Park_Otani_Kim_Kim_Han_et_al_2018].
+While we include data from Hajinri here, we consider this assemblage as an unusual outlier because the stemmed points and blades from Hajinri are highly standardized and refined, which are only found at other sites much later in time.
 We have more confidence in finds from Bonggok, which has the second-earliest dates (41.5 ka) in our collection, as the first appearance of stemmed points.
@@
 
-To answer our research questions about the likely uses of stemmed points, we calculated TCSA values for the stemmed points in our sample, and explored the interaction of TCSA values with raw materials, and artifact size.
+To answer our research questions about the likely uses of stemmed points, we calculated TCSA values for the stemmed points in our sample, and explored the interaction of TCSA values with raw materials and artifact size.
 To examine the temporal patterns in TCSA values we investigated radiocarbon ages associated with the points.
-We distinguished the research time period into three based on the major climate event during the Late Pleistocene, Last Glacial Maximum (LGM), and examined the TCSA pattern.
-We explored the relationship between TCSA values and the location of assemblages by applying eco-regional information.
+We separated the research time period into three phases based on the major climate event during the Late Pleistocene, Last Glacial Maximum (LGM), and examined the distribution of TCSA values before, during and after.
+We explored the relationship between TCSA values and the location of assemblages by comparing the distributions of TCSA values across ecological and vegetation zones.
 
@@
 
-We followed the TCSA ranges for different weapon-delivery systems that @lombard2022tip and @lombard2021variation created by summarizing @wadley2008segment' study of backed microliths and calibrating existing ranges (@tbl-tcsa-ranges).
-Among our dataset, we excluded `r n_omit` artifacts from our dataset with TCSA greater than 250 and thus outside of the range of the comparative data.
+To aid with interpreting our results, we followed the TCSA ranges for different weapon-delivery systems that @lombard2022tip and @lombard2021variation created by summarizing @wadley2008segment' study of backed microliths and calibrating existing ranges (@tbl-tcsa-ranges).
+We excluded `r n_omit` artifacts from our dataset with TCSA greater than 250 and thus outside of the range of the comparative data.
 For this study we include TCSA values from a total of `r min_n_points` stemmed points from `r n_ass` assemblages unearthed from `r n_sites` sites.
-Each analysis uses slightly different number of artifacts due to data availability.
 
@@
 Selective use of raw material is a key characteristic of the technological transition of the Korean Late Paleolithic.
-Prior to the Late Paleolithic, people used quartzite and vein quartz for core and flake tools.
+Prior to the Late Paleolithic, people mostly used quartzite and vein quartz for stone artefacts.
 Then, finer grained materials were newly added to assemblages for producing the newly introduced tools [@seong2004quartzite].
-We analyzed TCSA values for each raw material type with `r n_raw` artifacts to examine the interaction of raw material types and TCSA values.
+We analyzed TCSA values of `r n_raw` artifacts with raw material information to examine the interaction of raw material types and TCSA values.
 We categorized rare raw materials, which have less than 10 artifacts, as "Other".
@@
 We chose weight as a proxy for overall size of the stemmed points.
-Weight is a reliable discriminator between tools of different sizes and a descriptive attribute which can be determined rapidly and objectively and serve as a criterion of likeness or dissimilarity in projectile point description [@fenenga1953weights; @shea2006origins].
+Weight is a reliable discriminator between tools of different sizes and a descriptive attribute which can be measured rapidly and objectively and serve as a criterion of likeness or dissimilarity in projectile point description [@fenenga1953weights; @shea2006origins].
 We then explored the relationship between weight and raw materials on TCSA values.
@@
 We used radiocarbon ages to explore the temporal patterns of the likely uses of stemmed points.
-After excluding assemblages that have no radiocarbon dates, we arranged `r n_sites_with_radiocarbon_ages` assemblages dated from 45ka to 14.8ka to explore variation in TCSA over time.
-Then we divided the artifacts into three groups based on LGM: before LGM, LGM, and after LGM to examine the impact of this major climate event on the material culture.
+After excluding assemblages that have no radiocarbon dates, we used `r n_sites_with_radiocarbon_ages` assemblages dated from 45ka to 14.8ka to explore variation in TCSA over time.
+We divided the artifacts from dated assemblages into three groups based on the LGM: before LGM, LGM, and after LGM, to examine the impact of this major climate event on TCSA values.
 
@@
 Among `r n_sites` artifacts, we combined sites that contains fewer than 5 stemmed points aside and named them as "Other".
-We also explored possible environmental impact by connecting stemmed points to the eco-regional zones.
-We chose two maps of modern eco-regional zones and modified them for this research.
-@yim1975distribution first created vegetation forest map defined temperature and precipitations.
-This map then modified by @yi2010vegetation.
-They divided South Korea, our research area, into three zones: Central Temperature Zone (CT), South Temperature Zone (ST), and Subtropical-warm Temperature Zone (SWT).
-We also used a map from @lee2008relationship, which separated our research area into 16 different zones based geographical characteristics such as inland, coastal, major rivers, island, and major mountain ranges in addition to ecological information including temperature and precipitation.
-We only focused on 14 zones, excluding two islands that have no records of Paleolithic material culture: Imjin river basin (IRB), Metropolitan (MP), Central inland (CI), Kangwon coastal (KC), Choongnam coastal (CC), Southwestern inland (SWI), Upper Nagdong river basin (UNRB), WoolYoung coastal (WYC), Western Cholla (WC), Southern mountain (SM), Southeastern inland (SEI), Hyungsan Taewha coastal (HTC), Western south coastal (WSC), and Eastern south coastal (ESC).
+We explored possible effect of environmental variation on TCSA values by grouping stemmed points by the eco-regional zones that they were found in, and comparing the distributions of TCSA values across the different zones.
+We compared distributions across four vegetation zones, and also across 16 zones of different geographical characteristics.
+
+For the vegetation zones we used @yi2010vegetation's divisions of South Korea into three zones: Central Temperate Zone (CT), South Temperate Zone (ST), and Subtropical-warm Temperate Zone (SWT).
+These divisions are based on @yim1975distribution's forest vegetation map, defined by recent temperature and precipitation values.
+
+We also explored spatial patterns using 16 geographical zones based geographical characteristics such as inland, coastal, major rivers, island, and major mountain ranges in addition to ecological information including temperature and precipitation [@lee2008relationship].
+The sites in our sample occur in 14 of these zones: Imjin river basin (IRB), Metropolitan (MP), Central inland (CI), Kangwon coastal (KC), Choongnam coastal (CC), Southwestern inland (SWI), Upper Nagdong river basin (UNRB), WoolYoung coastal (WYC), Western Cholla (WC), Southern mountain (SM), Southeastern inland (SEI), Hyungsan Taewha coastal (HTC), Western south coastal (WSC), and Eastern south coastal (ESC).
 We then located individual assemblages on those two maps using longitudes and latitudes and explore TCSA values for each zone.
-It is acknowledged that modern eco-regional maps may not be suitable for studying the Paleolithic period directly, but they are the only maps of eco-regional seperation that are currently available to us.
+We recognise that modern eco-regional maps may not fully resemble the Pleistocene landscape in which the points were made and used, however these are the only maps of eco-regional zones that are currently available.
 
@@
 
-Inspired by @eren2022plains's approach, we hypothesized two basic scenarios that might explain the temporal or/and spatial patterns of TCSA range.
-If stemmed points represent a narrow TCSA range, then people likely produced tools that performed specific tasks.
+Inspired by @eren2022plains's approach, we hypothesized some simple scenarios that might explain the temporal and spatial patterns of TCSA range in Korea.
+If stemmed points have a narrow range of TCSA values, then people likely produced tools that performed a small set of specific tasks.
 This may be related to low levels of uncertainty in the forager's social and physical environments.
-On the other hand, a wide TCSA range may indicate that stemmed points were multifunctional tools, suggesting that people were responding to unfamiliar situations, such as moving into an unfamiliar landscape or unpredictable changes in patch productivity and travel times [@bird2006behavioral; @bettinger2016marginal; @kelly2007foraging].
-We expect to observe various TCSA ranges, reflecting morphologically and functionally different stemmed points, across different regions or environments in Korea, and over time.
+On the other hand, a wide range of TCSA values may indicate that stemmed points were multifunctional tools, suggesting that people were responding to unfamiliar situations, such as moving into an unfamiliar landscape or unpredictable variation in patch productivity and travel times [@bird2006behavioral; @bettinger2016marginal; @kelly2007foraging].
+
+We predict a temporal pattern of more variable TCSA values at the first appearance of stemmed points, suggesting that people were using the tools as part of an adaptation to moving into unfamiliar landscapes, with unpredictable variation in patch productivity and travel times.
+We predict a further increase in the variability of TCSA values in the LGM as lower temperatures alter the distribution of resources and lower the predictability of resource encounters.
+After the LGM we predict a reduction in the range of TCSA values, with higher temperatures and increased bioproductivity.
+
+We predict spatial patterns... <!-- Please GP can you finish this paragraph? We need some predictions that correspond to the vegetation and geographical zones -->
 
@@
 @fig-tcsa-all-sp shows TCSA values for all stemmed points in our sample, with shaded rectangles to assist in the interpretation of their likely uses.
-Overall we see a wide variation of TCSA values in this boxplot.
+Overall we see a wide variation of TCSA values.
 The sample mean TCSA is `r round(mean(tcsa$min_tcsa), digits=1)`, and standard deviation of TCSA is `r round(sd(tcsa$min_tcsa), digits=1)`.
@@
 Among the weapon-tip types, only poison arrow tips appear to be absent from these Korean assemblages.
+
+## Variation in TCSA values by artifact size
+
+```{r}
+
+# Facet by size class
+# https://cran.r-project.org/web/packages/Ckmeans.1d.dp/vignettes/Ckmeans.1d.dp.html
+library(Ckmeans.1d.dp)
+
+# find clusters
+result <- Ckmeans.1d.dp(tcsa_mass_no_na$`Weight(g)`)
+
+tcsa_mass_no_na <-
+tcsa_mass_no_na %>%
+  mutate(mass_cluster = factor(result$cluster))
+
+```
+
+Using weight as a size proxy we examined the relationship between size and likely use of the stemmed points inferred from TCSA values.
+We conducted a univariate cluster analysis [@song2020efficient; @wang2011ckmeans] of stemmed points by weight and it revealed three clusters of artifact sizes (A of @fig-artifact-size-cluster, mean = `r round(mean(tcsa_mass_no_na$'Weight(g)'), digits=1)`, SD = `r round(sd(tcsa_mass_no_na$'Weight(g)'), digits=1)`).
+Cluster 1, the smallest (lightest) artifacts, have a lower range of TCSA values, compared to Cluster 2 (B of @fig-artifact-size-cluster, mean = `r round(mean(tcsa_mass_no_na$'min_tcsa'), digits=1)`, SD = `r round(sd(tcsa_mass_no_na$'min_tcsa'), digits=1)`).
+TCSA values for Cluster 3 are the highest, except for one artifact, which is lower than 50.
+There are a few overlaps, but we can still assume that artifacts in different size clusters might be made for different purposes.
 
@@
    annotate("rect", 
-           xmin = 0.5,
-           xmax = 1.7,
+           xmin = 6.5,
+           xmax = 7,
            ymin = 4,
@@
       annotate("rect", 
-           xmin = 1.7,
-           xmax = 2.9,
+           xmin = 6.5,
+           xmax = 7,
            ymin = 17,
@@
       annotate("rect", 
-           xmin = 2.9,
-           xmax = 4.1,
+           xmin = 6.5,
+           xmax = 7,
            ymin = 40,
@@
       annotate("rect", 
-           xmin = 4.1,
-           xmax = 5.3,
+           xmin = 6.5,
+           xmax = 7,
            ymin = 42,
@@
      annotate("rect", 
-           xmin = 5.3,
-           xmax = 6.5,
+           xmin = 6.5,
+           xmax = 7,
            ymin = 80,
@@
            alpha = 0.2) 
- # + annotate("text", 
- #               size = 2,
- #               label = "Poisoned arrow tips",
- #               hjust = 0,
- #               lineheight = 0.75,
- #           x = 0.75,
-   #         y = (18-4) /2 + 4) +
-   # annotate("text",
-   #          size = 2,
-   #             label = "Un-poisoned arrow tips",
-   #             hjust = 0,
-   #             lineheight = 0.75,
-   #         x = 1.85,
-   #         y = (47-17) /2 + 17) +
-   # annotate("text", 
-   #             size = 2,
-   #             label = "Dart tips",
-   #             hjust = 0,
-   #         x = 3.3,
-   #         y = (76-40) /2 + 40) +
-   # annotate("text", 
-   #             size = 2,
-   #             label = "Javelin tips",
-   #             hjust = 0,
-   #         x = 4.4,
-   #         y = (90-42) /2 + 42) +
-    # annotate("text", 
-    #            label = "Stabbing spear tips",
-    #            size = 2,
-    #            hjust = 0,
-    #            lineheight = 0.75,
-    #        x = 5.52,
-    #        y = (200-80) /2 + 80 )
-
 ```
@@
 
-## Variation in TCSA values by artifact size
-
-```{r}
-
-# Facet by size class
-# https://cran.r-project.org/web/packages/Ckmeans.1d.dp/vignettes/Ckmeans.1d.dp.html
-library(Ckmeans.1d.dp)
-
-# find clusters
-result <- Ckmeans.1d.dp(tcsa_mass_no_na$`Weight(g)`)
-
-tcsa_mass_no_na <-
-tcsa_mass_no_na %>%
-  mutate(mass_cluster = factor(result$cluster))
-
-```
-
-Using weight as a size proxy we examined the relationship between size and likely use of the stemmed points inferred from TCSA values.
-We conducted a univariate cluster analysis [@song2020efficient; @wang2011ckmeans] of stemmed points by weight and it revealed three clusters of artifact sizes (A of @fig-artifact-size-cluster, mean = `r round(mean(tcsa_mass_no_na$'Weight(g)'), digits=1)`, SD = `r round(sd(tcsa_mass_no_na$'Weight(g)'), digits=1)`).
-Cluster 1, the smallest (lightest) artifacts, have a lower range of TCSA values, compared to Cluster 2 (B of @fig-artifact-size-cluster, mean = `r round(mean(tcsa_mass_no_na$'min_tcsa'), digits=1)`, SD = `r round(sd(tcsa_mass_no_na$'min_tcsa'), digits=1)`).
-TCSA values for Cluster 3 are the highest, except for one artifact, which is lower than 50.
-There are a few overlaps, but we can still assume that artifacts in different size clusters might be made for different purposes.
-
 ```{r}
 #| label: fig-artifact-size-cluster
-#| fig-cap: "A. Distribution of artifact weight showing three clusters. B. TCSA values for all artifacts with size classes indicated by data point shape."
+#| fig-cap: "A. Distribution of artifact weight showing three clusters. B. TCSA values for all artifacts. Artefact size classes indicated by the digits representing data point values"
 #| fig-width:   4
@@
   geom_quasirandom(aes(shape = mass_cluster),
-                   alpha = 0.4,
+                   alpha = 0.7,
                    size = 1.1) +
+  scale_shape_identity() +
   labs(shape='Cluster',
@@
   geom_quasirandom(aes(shape = mass_cluster),
-                   alpha = 0.4,
+                   alpha = 0.7,
                    size = 1.1) +
+  scale_shape_identity() +
   labs(shape='Cluster') +
@@
 
-@fig-tcsa-size shows a positive relationship between TCSA, artifact weight and raw material.
-We could observe variations by raw material types.
+@fig-tcsa-size shows a positive relationship between TCSA, raw material and artifact weight.
 There is a stronger correlation between TCSA and artifact weight for acidic volcanic points (i.e.
 points are closer to the regression line), whereas the correlation for shale stemmed points is weaker.
-The other raw materials show various distribution patterns around the regression line, reconfirming that raw material appears not to have strongly influenced TCSA (see also @fig-tcsa-raw-materials).
+The other raw materials show various distribution patterns around the regression line, confirming that raw material appears not to have strongly influenced TCSA (see also @fig-tcsa-raw-materials).
 
@@
            colour = raw_materials_top)) +
-  geom_point(size = 1) +
+  geom_point(size = 3) +
     labs(x = "Weight (g)", 
@@
        colour='Raw material') +
-  geom_smooth(method=lm, se=FALSE, size = 0.3)+
-  theme_bw(base_size = 6) +
+  geom_smooth(method=lm, se=FALSE, size = 1)+
+  theme_bw(base_size = 10) +
   theme(
-  legend.title = element_text(size = 4),
-  legend.text = element_text(size = 3),
+  legend.title = element_text(size = 6),
+  legend.text = element_text(size = 5),
   legend.position = c(.8,.3),
@@
 
-We arranged TCSA values of `r n_sites_with_radiocarbon_ages` assemblages in chronological order (@fig-tcsa-radio-carbon panel A).
-Our analysis shows that earlier assemblages before LGM tend to possess more stemmed points such as Suyanggae.
+@fig-tcsa-radio-carbon shows `r n_sites_with_radiocarbon_ages` assemblages with stemmed points in chronological order (panel A).
+Assemblages discarded before LGM tend to possess more stemmed points, such as Suyanggae.
 These assemblages have a wide range of TCSA values, indicating multiple likely usages of stemmed points.
@@
         x = "") +
-  theme(axis.text.x = element_text(size = 3.5),
-        axis.text.y = element_text(size = 3.5),
-        text = element_text(size=5)) +
-  scale_x_discrete(guide = guide_axis(angle = 45))
+  theme(axis.text.x = element_text(size = 5),
+        axis.text.y = element_text(size = 5),
+        text = element_text(size=8)) +
+  scale_x_discrete(guide = guide_axis(angle = 45)) +
+  scale_y_continuous(breaks=seq(50, 200, 50))
 
@@
 
-all_time <- temporal_tcsa +
+all_time <- 
+  temporal_tcsa +
  annotate("rect", 
@@
            xmax = 12.2,
-           ymin = 0,
-           ymax = 300,
+           ymin = 240,
+           ymax = 260,
            fill = "lightgray",
@@
            x = 5,
-           y = 260) +
+           y = 250) +
    annotate("rect", 
@@
            xmax = 24.5,
-           ymin = 0,
-           ymax = 300,
+           ymin = 240,
+           ymax = 260,
            alpha = 0.2) +
@@
            x = 17,
-           y = 260) +
+           y = 250) +
    annotate("rect", 
@@
            xmax = 21,
-           ymin = 0,
-           ymax = 230,
-           alpha = 0.2) +
+           ymin = 220,
+           ymax = 240,
+           alpha = 0.4) +
   annotate("text", 
@@
            x = 17.3,
-           y = 210)
+           y = 230)
 
@@
      annotate("rect", 
-           xmin = 0.3,
-           xmax = 1,
+           xmin = 3.5,
+           xmax = 3.8,
            ymin = 4,
@@
            alpha = 0.2) +
-      # annotate("text", 
-      #          size = point_function_group_label_size,
-      #          label = "Poisoned\narrow tips",
-      #          hjust = 0,
-      #          lineheight = 0.75,
-      #      x = 3.4,
-      #      y = (18-4) /2 + 4) +
-      annotate("rect", 
-           xmin = 1,
-           xmax = 1.7,
+      annotate("rect", 
+           xmin = 3.5,
+           xmax = 3.8,
            ymin = 17,
@@
            alpha = 0.2) +
-      # annotate("text", 
-      #          size = point_function_group_label_size,
-      #          label = "Un-poisoned\narrow tips",
-      #          hjust = 0,
-      #          lineheight = 0.75,
-      #      x = 3.5,
-      #      y = (47-17) /2 + 17) +
-      annotate("rect", 
-           xmin = 1.7,
-           xmax = 2.4,
+      annotate("rect", 
+           xmin = 3.5,
+           xmax = 3.8,
            ymin = 40,
@@
            alpha = 0.2) +
-      # annotate("text", 
-      #          size = point_function_group_label_size,
-      #          label = "Dart tips",
-      #          hjust = 0,
-      #      x = 3.6,
-      #      y = (76-40) /2 + 40) +
-      annotate("rect", 
-           xmin = 2.4,
-           xmax = 3.1,
+      annotate("rect", 
+           xmin = 3.5,
+           xmax = 3.8,
            ymin = 42,
@@
            alpha = 0.2) +
-      # annotate("text", 
-      #          size = point_function_group_label_size,
-      #          label = "Javelin tips",
-      #          hjust = 0,
-      #      x = 3.7,
-      #      y = (90-42) /2 + 42) +
-     annotate("rect", 
-           xmin = 3.1,
+     annotate("rect", 
+           xmin = 3.5,
            xmax = 3.8,
@@
            alpha = 0.2)
-# + annotate("text", 
-#                label = "Stabbing\nspear tips",
-#                size = point_function_group_label_size,
-           #     hjust = 0,
-           #     lineheight = 0.75,
-           # x = 3.8,
-           # y = (200-80) /2 + 80 )
 
@@
 
-We then grouped stemmed points into three categories based on LGM event: before, LGM, and after to examine possibility of environmental impact on the use of the artifacts (@fig-tcsa-radio-carbon panel B).
-Our results shows that stemmed points were made predominantly before LGM and only a few left after LGM.
+Panel B of @fig-tcsa-radio-carbon shows the distribution of TSCA values from assemblages dating to before the LGM, during the LGM, and after, to explore the impact of climate change on the probably function of the artifacts.
+Our results show that stemmed points were made predominantly before LGM and only a few left after LGM.
 Each category shows wide variation in LGM, which indicates diverse uses of stemmed points.
-There is no strong correlation between the TCSA values and the time period (`r three_time_p_pretty`).
+While the median TSCA value for assemblages discarded during the LGM is higher than assemblages from earlier and later, there is no statistically significant difference in TCSA values across the three periods (`r three_time_p_pretty`).
 
@@
    annotate("rect", 
-           xmin = 0.4,
-           xmax = 1.9,
+           xmin = 7.5,
+           xmax = 8,
            ymin = 4,
@@
       annotate("rect", 
-           xmin = 1.9,
-           xmax = 3.4,
+           xmin = 7.5,
+           xmax = 8,
            ymin = 17,
@@
       annotate("rect", 
-           xmin = 3.4,
-           xmax = 4.9,
+           xmin = 7.5,
+           xmax = 8,
            ymin = 40,
@@
       annotate("rect", 
-           xmin = 4.9,
-           xmax = 6.4,
+           xmin = 7.5,
+           xmax = 8,
            ymin = 42,
@@
      annotate("rect", 
-           xmin = 6.4,
-           xmax = 7.9,
+           xmin = 7.5,
+           xmax = 8,
            ymin = 80,

### 2023-01-24 analysis/paper/paper.qmd
finished editing the results

@@
   labs(color="TCSA") +
-  guides(fill = FALSE) +
+  guides(fill = "none") +
   theme(axis.text.x = element_text(angle = 45, size = 4.5))+
@@
 
-veg_forest <- read_sf(here("analysis/data/gis_data/eco_forest_two.shp"))
-
-veg_forest <- st_transform(veg_forest, crs = st_crs(4326))
+veg_forest <- 
+  read_sf(here("analysis/data/gis_data/eco_forest_two.shp"))
+
+veg_forest <- 
+  st_transform(veg_forest, crs = st_crs(4326))
 
@@
 
-pts_forest <- st_as_sf(korean_archaeological_site_locations_tcsa,
+pts_forest <- 
+  st_as_sf(korean_archaeological_site_locations_tcsa,
                 coords = c("long_dd.y", "lat_dd.y"),
@@
   labs(color="TCSA") +
-  guides(fill = FALSE) +
+  guides(fill = "none") +
   theme(axis.text.x = element_text(angle = 45, size = 4.5)) +
@@
 # get rid of zones that has no artifact
-points_in_veg_zones_forest_box <- points_in_veg_zones_forest %>% 
+points_in_veg_zones_forest_box <- 
+  points_in_veg_zones_forest %>% 
   drop_na(Region)
@@
 #eco region#
-for_aov <- aov(min_tcsa ~ as.factor(Zone), points_in_veg_zones_forest_box)
+for_aov <- 
+  aov(min_tcsa ~ as.factor(Zone),
+      points_in_veg_zones_forest_box)
+
 # summary(for_aov)
-for_lgm_test <- Anova(lm(min_tcsa ~ as.factor(Zone), points_in_veg_zones_forest_box))
-
-for_lgm_test_pretty <- apastats::describe.Anova(for_lgm_test, 1, 2)
+for_lgm_test <- 
+  Anova(lm(min_tcsa ~ as.factor(Zone),
+           points_in_veg_zones_forest_box))
+
+for_lgm_test_pretty <- 
+  apastats::describe.Anova(for_lgm_test, 1, 2)
 
@@
 
-In order to explore spatial patterns as well as the relationship between the role of stemmed points and the environment, we located individual assemblages on maps that depict eco-regional boundaries.
+In order to explore spatial patterns and the relationship between the role of stemmed points and the environment, we located individual assemblages on maps that depict eco-regional zones.
 Our results show that stemmed points are only located in certain zones, Central Temperature Zone (CT) of vegetation map or Central inland (CI) (@fig-gis-tcsa-sites panel A & D) of eco-region map.
-We observed that TCSA values are strongely associated with the zones (Vegetation zones: `r eco_lgm_test_pretty`, Ecoregion zones: `r for_lgm_test_pretty`).
-To explore the TCSA distribution within the eco-regional zones, we excluded zones in the ecoregion map that have no stemmed points (@fig-gis-tcsa-sites panel E).
-We found no clear distribution pattern throughout the zones, showing a wide variation in TCSA regardless of the density of stemmed point in each zone.
-For example, explored the six artifacts from South Temperature Zone (ST) to examine the gap of TCSA values among the artifacts (i.e.
-Three artifacts are over 175 while the other three are lower around or lower than 100.
-@fig-gis-tcsa-sites panel B).
-We observed two artifacts from Bonggok site with no age difference, both made out of rhyolite, have huge difference in TCSA values, indicating no pattern for TCSA variation.
-
-We then computed Tukey's for comparison tests between the eco-regional zones.
-Our results show that ST and CT, and Subtropical-warm Temperature Zone (SWT)-ST have significantly different while SWT and CT are close to each other (@fig-gis-tcsa-sites panel C).
-However, we found no significant difference within the zones of eco-region map (@fig-gis-tcsa-sites panel F).
+We observed that there are statistically significant differences in the distributions of TCSA values between some zones (Vegetation zones: `r eco_lgm_test_pretty`, Ecoregion zones: `r for_lgm_test_pretty`).
+The Tukey's HSD results show that TCSA values from the South Temperature Zone (ST) zone are significantly different from the other two vegetation zones (@fig-gis-tcsa-sites panel C).
+Looking into this further, we see that there are only six artifacts from the ST zone, with only three of these having TSCA values above 175.
+The other three have TCSA values lower than 100, similar to the other zones (@fig-gis-tcsa-sites panel B).
+Given the small sample size of artefacts with extreme TCSA values in the ST zone, we hesitate to conclude this is an archaeologically significant pattern of difference TCSA values.
+The Tukey HSD pairwise differences show no pairs of eco-regions with significant differences in TCSA values.
+Overall, our results indicate no clear patterns in the distribution of TCSA values across the vegetation and eco-regional zones.
 

### 2023-01-26 analysis/paper/paper.qmd
update for meeting

@@
 The other three have TCSA values lower than 100, similar to the other zones (@fig-gis-tcsa-sites panel B).
-Given the small sample size of artefacts with extreme TCSA values in the ST zone, we hesitate to conclude this is an archaeologically significant pattern of difference TCSA values.
+Given the small number of artefacts with extreme TCSA values in the ST zone, we hesitate to conclude this is an archaeologically significant pattern of different TCSA values.
 The Tukey HSD pairwise differences show no pairs of eco-regions with significant differences in TCSA values.
@@
 Our results show that javelin tips and stabbing spear tips are the most probable ballistic uses for stemmed points.
-There are a few stemmed points in our sample that may have been used as dart tips and un-poisoned arrow tips, but none of them have TCSA values in the range of poison arrow tips (@fig-tcsa-all-sp).
+There are a few stemmed points in our sample that may have been used as dart tips and un-poisoned arrow tips.
+None have TCSA values in the range of poison arrow tips (@fig-tcsa-all-sp).
 But in general, we found a wide range of TCSA in our sample, indicating that stemmed points might play diverse roles in foraging toolkits.

### 2023-02-03 analysis/paper/paper.qmd
add more citations

@@
 In this research, we examine the possible role that stemmed points played during the technological transition, as well as throughout the Late Paleolithic.
-Our main questions are: What were stemmed points used for?
+Our main questions are: What were the best-fit ballistic probabilities for the stemmed points if they were hafted as weapon tips?
 How diverse were their likely uses?
@@
 We analyzed TCSA with other variables including raw materials, weight, radiocarbon dates and locations.
-Our results show that the stemmed points were likely used  as javelin tips and stabbing spear tips, with smaller numbers as dart tips and un-poisoned arrow tips.
+Our results show that the stemmed points were could have served as javelin tips and stabbing spear tips, with smaller numbers as dart tips and un-poisoned arrow tips.
 TCSA values are controlled mostly by size rather than raw material types.
-We found different usage of stemmed points at different sites, which could indicate people used stemmed points in different ways depending on the local environment.
+We found different TCSA range of stemmed points at different sites, which could indicate people used stemmed points in different ways depending on the local environment.
 Some sites show a wide range of TCSA values that represent multi-purpose usage of stemmed points.
-The temporal pattern of TCSA values is one of little change throughout the Late Paleolithic period but points were mainly used before Last Glacial Maximum (LGM).
+The temporal pattern of TCSA values is one of little change throughout the Late Paleolithic period but points were mainly produced before Last Glacial Maximum (LGM).
 We observed that stemmed points were mostly located in certain ecoregions in Korea but no clear spatial pattern is confirmed.
@@
 To understand the roles that stemmed points played in the technological transition, this study examines the possible uses of stemmed points.
-We use the tip cross-sectional area (TCSA) metric to infer likely artifact use, and compare the results to other archaeological and ethnographic assemblages based on the idea that different shapes correspond to different weapon types [@lombard2021variation].
-We then explore the relationship between likely use and raw materials, artifact size and discard location, and how these change over time.
+We use the tip cross-sectional area (TCSA) metric to infer potential variability in weapon-use strategies, and compare the results to other archaeological and ethnographic assemblages based on the idea that different shapes correspond to different weapon types [@lombard2021variation].
+We then explore the relationship between the TCSA range and raw materials, artifact size and discard location, and how these change over time.
 Our main questions are: What were the best-fit ballistic probabilities for the stemmed points if they were hafted as weapon tips?
@@
 What are the temporal and spatial patterns in stemmed point uses?
-As part of our discussion, we examine possible connections between the roles of stemmed points and population dynamics or environmental change during the Late Paleolithic period.
+As part of our discussion, we examine possible connections between the roles of stemmed points and environmental change or population dynamics during the Late Paleolithic period.
 In order to understand how a certain weapon-tip type was chosen, we apply an evolutionary perspective with the assumption that people chose their weapon tip types as part of their adaptation to the socio-environmental circumstances they found themselves in.
@@
 Though stemmed points have been generally assumed to have been hunting armatures [@lee2019were; @lee2011study; @seong2008tanged; @lee2002new; @chang2013human], it is difficult to understand the likely uses of the tools without knowing their complete shape when attached to other components.
-Preserved wooden components of projectile tools are too few and rare to generalize the overall scale and variability of the tools used in prehistory [@shea2006origins].
-@Lee_Kong_2002 claimed stemmed points were generalised 'stemmed tools' because of their diverse shapes, which could represent different functions, as well as the uncertainty of complete tool shape.
+Preserved wooden components of projectile tools are too few and rare to standardize the overall scale and variability of the tools used in prehistory [@shea2006origins].
+@Lee_Kong_2002 claimed stemmed points were generalized 'stemmed tools' because of their diverse shapes, which could represent different functions, as well as the uncertainty of complete tool shape.
 Other researchers propose that non-symmetrical stemmed points, with retouch on one side of the artifact or denticulate blades on one side, should not be called as stemmed points but instead stemmed knives, stemmed side-scrapers, stemmed end-scrapers, or stemmed burins [@seong2008tanged; @Lee_Jang_2011; @Kim] @fig-sp-hand-made: 7-8.
@@
 Most stemmed points are symmetrical from the tip to the tang, having the central axis function as a line of symmetry [@Lee_Jang_2011].
+There are a high proportion of broken tips and stems and the reused tools were repaired based on keeping the symmetry [@Park_2013; @Kim_2017].
 Studies of stemmed point manufacturing processes and the patterns of broken pieces show that stemmed points may have been used mainly as spear tips [@Lee_1985; @Chang_2002].
-There are a high proportion of broken tips and stems and the reused tools were repaired based on keeping the symmetry [@Park_2013; @Kim].
+On Yongsandong site, for example, the percentage of complete tools is only 10%, while the percentage of missing tips is 33%.
+On Jingeuneul site, the percentages are 16% and 50%, respectively [@Park_2013].
 In addition to the morphological aspect of stemmed points, investigation of a whole site and tool composition of an assemblage shows that stemmed points or stemmed tools could be strongly associated with hunting activities including peeling the animal skin after slaughtering or separating the bones from the flesh [@seong2008tanged; @Chong_2021].
@@
 
-Tip cross-sectional area (TCSA) of stone artifacts has been used as a ballistically relevant standard to probabalistically discriminate difference between functional classes of projectile armatures such as spearthrower (a.k.a.
-atlatl) dart tips, un-poisoned arrow tips and large stabbing/thrusting spears [@sisk2011african; @lombard2021variation; @lombard2020tip; @lombard2021did; @o2018origins].
+Tip cross-sectional area (TCSA) of stone artifacts has been used as a ballistically relevant standard to probabalistically discriminate different likely use classes of projectile armatures such as spearthrower (a.k.a.
+atlatl) dart tips, un-poisoned arrow tips and large stabbing/thrusting spears [@sisk2011african; @lombard2021variation; @lombard2020tip; @lombard2021did; @o2018origins; @lombard2022standardized].
 It is important to note that the TCSA metric alone cannot unambiguously determine artifact function, it only suggests a best-fit ballistic probability for the points if they were hafted as weapon tips.
@@
 
-Considering the importance of stemmed points for the technological transition during the Late Paleolithic in Korea, our research examines the likely uses of stemmed points by asking three research questions: What were stemmed points used for?
+Considering the importance of stemmed points for the technological transition during the Late Paleolithic in Korea, our research examines the likely uses of stemmed points by asking three research questions: What were the best-fit ballistic probabilities for the stemmed points if they were hafted as weapon tips?
 How diverse were their likely uses?

### 2023-02-03 analysis/paper/paper.qmd
prediction done

@@
 Similarly, we identified multiple assemblages in a single excavation or even a trench where archaeological deposits were separated by culturally sterile deposits, or where distinct artifact-bearing stratigraphic units could be identified by major differences in the texture, color, and composition of the sedimentary deposits.
-As a result, in this research we used four assemblages from the Suyanggae site.
+As a result, in this research we used four assemblages from Suyanggae.
 We separated one assemblage among the four by using a different site name, Hajinri, following the convention established the excavators of that location.
-Hajinri is the sixth excavation area at Suyanggae, which is 3.5 kilometers apart from the other areas and dated to much earlier (around 42-30 ka) than the other assemblages (around 31-15 ka).
+Hajinri is the sixth excavation location at Suyanggae, which is 3.5 kilometers apart from the other areas and dated to much earlier (around 42-30 ka) than the other assemblages (around 31-15 ka).
 According to the excavation reports for Hajinri, stemmed points first appeared there around 42 ka with the earliest blade and blade cores [@Lee_Woo_Lee_An_Yun_Park_Otani_Kim_Kim_Han_et_al_2018].
@@
 Selective use of raw material is a key characteristic of the technological transition of the Korean Late Paleolithic.
-Prior to the Late Paleolithic, people mostly used quartzite and vein quartz for stone artefacts.
+Prior to the Late Paleolithic, people mostly used quartzite and vein quartz for stone artifacts.
 Then, finer grained materials were newly added to assemblages for producing the newly introduced tools [@seong2004quartzite].
@@
 After excluding assemblages that have no radiocarbon dates, we used `r n_sites_with_radiocarbon_ages` assemblages dated from 45ka to 14.8ka to explore variation in TCSA over time.
-We divided the artifacts from dated assemblages into three groups based on the LGM: before LGM, LGM, and after LGM, to examine the impact of this major climate event on TCSA values.
+We divided the artifacts from dated assemblages into three groups based on the LGM: before, during and after, to examine the impact of this major climate event on TCSA values.
 
@@
 We then located individual assemblages on those two maps using longitudes and latitudes and explore TCSA values for each zone.
-We recognise that modern eco-regional maps may not fully resemble the Pleistocene landscape in which the points were made and used, however these are the only maps of eco-regional zones that are currently available.
+We recognize that modern eco-regional maps may not fully resemble the Pleistocene landscape in which the points were made and used, however these are the only maps of eco-regional zones that are currently available.
 
@@
 
-We predict spatial patterns... <!-- Please GP can you finish this paragraph? We need some predictions that correspond to the vegetation and geographical zones -->
+We predict that TCSA values vary across vegetation types and geographical zones in response to heterogeneously distributed food resources.
+In harsh environments with lower patch productivity, similar to the case of LGM duration, we predict a further increase in the variability of TCSA values.
+On the other had, we predict a reduction in the range of TCSA values in affluent patches with predictable types of prey.
+In their study of hunter-gatherer mobility strategies, @hamilton2016ecological show that resources are abundant and predictable along coasts or lake shores, which is highly dependent on temperature and precipitation and thus hunter-gatherers often become effectively sedentary.
+Our prediction is that the range of TCSA values will be narrower in coastal areas such as ST and SW of vegetation zones than in inland areas such as CT.
 
@@
 
-# all artefacts overall
+# all artifacts overall
 
@@
 #| label: fig-artifact-size-cluster
-#| fig-cap: "A. Distribution of artifact weight showing three clusters. B. TCSA values for all artifacts. Artefact size classes indicated by the digits representing data point values"
+#| fig-cap: "A. Distribution of artifact weight showing three clusters. B. TCSA values for all artifacts. Artifact size classes indicated by the digits representing data point values"
 #| fig-width:   4
@@
 The other three have TCSA values lower than 100, similar to the other zones (@fig-gis-tcsa-sites panel B).
-Given the small number of artefacts with extreme TCSA values in the ST zone, we hesitate to conclude this is an archaeologically significant pattern of different TCSA values.
+Given the small number of artifacts with extreme TCSA values in the ST zone, we hesitate to conclude this is an archaeologically significant pattern of different TCSA values.
 The Tukey HSD pairwise differences show no pairs of eco-regions with significant differences in TCSA values.

### 2023-02-03 analysis/paper/paper.qmd
Finished editing results

@@
 
+@fig-tcsa-all-sp shows TCSA values for all stemmed points in our sample, with shaded rectangles to assist in the interpretation of their likely uses.
+Overall we see a wide variation of TCSA values.
+The sample mean TCSA is `r round(mean(tcsa$min_tcsa), digits=1)`, and standard deviation of TCSA is `r round(sd(tcsa$min_tcsa), digits=1)`.
+According to the TCSA ranges presented in @tbl-tcsa-ranges, Korean stemmed points are overall mostly in the categories of javelin tips and stabbing spear tips, with smaller numbers as dart tips and un-poisoned arrow tips.
+Among the weapon-tip types, only poison arrow tips appear to be absent from these Korean assemblages.
+
 ```{r}
@@
 
-@fig-tcsa-all-sp shows TCSA values for all stemmed points in our sample, with shaded rectangles to assist in the interpretation of their likely uses.
-Overall we see a wide variation of TCSA values.
-The sample mean TCSA is `r round(mean(tcsa$min_tcsa), digits=1)`, and standard deviation of TCSA is `r round(sd(tcsa$min_tcsa), digits=1)`.
-According to the TCSA ranges presented in @tbl-tcsa-ranges, Korean stemmed points are overall mostly in the categories of javelin tips and stabbing spear tips, with smaller numbers as dart tips and un-poisoned arrow tips.
-Among the weapon-tip types, only poison arrow tips appear to be absent from these Korean assemblages.
-
 ## Variation in TCSA values by artifact size
@@
 There are a few overlaps, but we can still assume that artifacts in different size clusters might be made for different purposes.
-
-## Variation in TCSA values by raw material
-
-```{r}
-#| label: fig-tcsa-raw-materials
-#| fig-cap: "TCSA values by lithic raw material."
-#| fig-width:   4.4
-#| fig-height:  3.2
-
-# Facet by raw materials
-
- ggplot(tcsa_raw,
-              aes(y = min_tcsa, 
-                  x = label_raw)) +
-  geom_boxplot() +
-  theme_minimal() +
-  labs(x = "Raw materials", 
-       y = "TCSA") +
-  theme(axis.text.x = element_text(size = 6),
-        axis.text.y = element_text(size = 6),
-        text = element_text(size=8)) +
-  scale_x_discrete(guide = guide_axis(angle = 45))+
-  geom_quasirandom(alpha = 0.4,
-                   size = 1.3) +
-   annotate("rect", 
-           xmin = 6.5,
-           xmax = 7,
-           ymin = 4,
-           ymax = 18,
-           fill = "green",
-           alpha = 0.2) +
-      annotate("rect", 
-           xmin = 6.5,
-           xmax = 7,
-           ymin = 17,
-           ymax = 47,
-           fill = "blue",
-           alpha = 0.2) +
-      annotate("rect", 
-           xmin = 6.5,
-           xmax = 7,
-           ymin = 40,
-           ymax = 76,
-           fill = "orange",
-           alpha = 0.2) +
-      annotate("rect", 
-           xmin = 6.5,
-           xmax = 7,
-           ymin = 42,
-           ymax = 90,
-           fill = "purple",
-           alpha = 0.2) +
-     annotate("rect", 
-           xmin = 6.5,
-           xmax = 7,
-           ymin = 80,
-           ymax = 200,
-           fill = "yellow",
-           alpha = 0.2) 
-```
-
-@fig-tcsa-raw-materials shows that about half of the stemmed points were made from shale and its TCSA range is widely distributed.
-Acidic volcanic rocks tend to have skewed lower TCSA values.
-The other raw materials including hornfels, rhyolite, and tuff show a wide distribution, similar to shale.
-The category of "Other" raw materials include porphyry, trachyte, felsite, chert, quartz, quartzite, granite, mudstone, and unidentified rocks.
-Overall, there is no clear pattern of TCSA among raw materials.
 
@@
 
+## Variation in TCSA values by raw material
+
+@fig-tcsa-raw-materials shows that about half of the stemmed points were made from shale and its TCSA range is widely distributed.
+Acidic volcanic rocks tend to have skewed lower TCSA values.
+The other raw materials including hornfels, rhyolite, and tuff show a wide distribution, similar to shale.
+The category of "Other" raw materials include porphyry, trachyte, felsite, chert, quartz, quartzite, granite, mudstone, and unidentified rocks.
+Overall, there is no clear pattern of TCSA among raw materials.
+
+```{r}
+#| label: fig-tcsa-raw-materials
+#| fig-cap: "TCSA values by lithic raw material."
+#| fig-width:   4.4
+#| fig-height:  3.2
+
+# Facet by raw materials
+
+ ggplot(tcsa_raw,
+              aes(y = min_tcsa, 
+                  x = label_raw)) +
+  geom_boxplot() +
+  theme_minimal() +
+  labs(x = "Raw materials", 
+       y = "TCSA") +
+  theme(axis.text.x = element_text(size = 6),
+        axis.text.y = element_text(size = 6),
+        text = element_text(size=8)) +
+  scale_x_discrete(guide = guide_axis(angle = 45))+
+  geom_quasirandom(alpha = 0.4,
+                   size = 1.3) +
+   annotate("rect", 
+           xmin = 6.5,
+           xmax = 7,
+           ymin = 4,
+           ymax = 18,
+           fill = "green",
+           alpha = 0.2) +
+      annotate("rect", 
+           xmin = 6.5,
+           xmax = 7,
+           ymin = 17,
+           ymax = 47,
+           fill = "blue",
+           alpha = 0.2) +
+      annotate("rect", 
+           xmin = 6.5,
+           xmax = 7,
+           ymin = 40,
+           ymax = 76,
+           fill = "orange",
+           alpha = 0.2) +
+      annotate("rect", 
+           xmin = 6.5,
+           xmax = 7,
+           ymin = 42,
+           ymax = 90,
+           fill = "purple",
+           alpha = 0.2) +
+     annotate("rect", 
+           xmin = 6.5,
+           xmax = 7,
+           ymin = 80,
+           ymax = 200,
+           fill = "yellow",
+           alpha = 0.2) 
+```
+
 @fig-tcsa-size shows a positive relationship between TCSA, raw material and artifact weight.
@@
            colour = raw_materials_top)) +
-  geom_point(size = 3) +
+  geom_point(size = 1.5) +
     labs(x = "Weight (g)", 
@@
        colour='Raw material') +
-  geom_smooth(method=lm, se=FALSE, size = 1)+
+  geom_smooth(method=lm, se=FALSE, size = .7)+
   theme_bw(base_size = 10) +
   theme(
-  legend.title = element_text(size = 6),
-  legend.text = element_text(size = 5),
+  legend.title = element_text(size = 6.3),
+  legend.text = element_text(size = 5.3),
   legend.position = c(.8,.3),
-  legend.key.height = unit(4.5, 'mm'),
-  legend.key.width = unit(2, 'mm'))
+  legend.key.height = unit(4.8, 'mm'),
+  legend.key.width = unit(2.5, 'mm'))
 
@@
 The TCSA range varies depending on the assemblages.
+
+Panel B of @fig-tcsa-radio-carbon shows the distribution of TSCA values from assemblages dating to before the LGM, during the LGM, and after, to explore the impact of climate change on the probably function of the artifacts.
+Our results show that stemmed points were made predominantly before LGM and only a few left after LGM.
+Each category shows wide variation in LGM, which indicates diverse uses of stemmed points.
+While the median TSCA value for assemblages discarded during the LGM is higher than assemblages from earlier and later, there is no statistically significant difference in TCSA values across the three periods (`r three_time_p_pretty`).
 
@@
 
-Panel B of @fig-tcsa-radio-carbon shows the distribution of TSCA values from assemblages dating to before the LGM, during the LGM, and after, to explore the impact of climate change on the probably function of the artifacts.
-Our results show that stemmed points were made predominantly before LGM and only a few left after LGM.
-Each category shows wide variation in LGM, which indicates diverse uses of stemmed points.
-While the median TSCA value for assemblages discarded during the LGM is higher than assemblages from earlier and later, there is no statistically significant difference in TCSA values across the three periods (`r three_time_p_pretty`).
-
 ## Spatial patterns of TCSA values
+
+We computed artifact TCSA values for `r n_sites` sites that contain more than 5 stemmed points to observe variation between sites (@fig-tcsa-per-sites).
+Sites with fewer than 5 stemmed points were grouped under the category of "Other".
+Among the sites in our sample, Suyanggae has the most stemmed points and the widest range of TCSA values.
+This suggests that people made stemmed points for accomplishing a variety of tasks at Suyanggae site.
+Nosanri, Sibuk, and Wolpyeng have a narrower range of lower values, which indicates less diverse likely uses for stemmed points.
+Yongsujaeul shows a narrower range but higher TCSA value.
+Yongsandong has the second most stemmed points and shows roughly two different clusters of TCSA values.
 
@@
 
-We computed artifact TCSA values for `r n_sites` sites that contain more than 5 stemmed points to observe variation between sites (@fig-tcsa-per-sites) Sites with fewer than 5 stemmed points were grouped under the category of "Other".
-Among the sites in our sample, Suyanggae has the most stemmed points and the widest range of TCSA values.
-This suggests that people made stemmed points for accomplishing a variety of tasks at Suyanggae site.
-Nosanri, Sibuk, and Wolpyeng have a narrower range of lower values, which indicates less diverse likely uses for stemmed points.
-Yongsujaeul shows a narrower range but higher TCSA value.
-Yongsandong has the second most stemmed points and TCSA shows a similar distribution to Yongsakul.
+In order to explore spatial patterns and the relationship between the role of stemmed points and the environment, we located individual assemblages on maps that depict eco-regional zones.
+Our results show that stemmed points are only located in certain zones, Central Temperature Zone (CT) of vegetation map or Central inland (CI) (@fig-gis-tcsa-sites panel A & D) of eco-region map.
+We observed that there are statistically significant differences in the distributions of TCSA values between some zones (Vegetation zones: `r eco_lgm_test_pretty`, Ecoregion zones: `r for_lgm_test_pretty`).
+The Tukey's HSD results show that TCSA values from the South Temperature Zone (ST) zone are significantly different from the other two vegetation zones (@fig-gis-tcsa-sites panel C).
+Looking into this further, we see that there are only six artifacts from the ST zone, with only three of these having TSCA values above 175.
+The other three have TCSA values lower than 100, similar to the other zones (@fig-gis-tcsa-sites panel B).
+Given the small number of artifacts with extreme TCSA values in the ST zone, we hesitate to conclude this is an archaeologically significant pattern of different TCSA values.
+The Tukey HSD pairwise differences show no pairs of eco-regions with significant differences in TCSA values (@fig-gis-tcsa-sites panel F).
+Overall, our results indicate no clear patterns in the distribution of TCSA values across the vegetation and eco-regional zones.
 
@@
 
-In order to explore spatial patterns and the relationship between the role of stemmed points and the environment, we located individual assemblages on maps that depict eco-regional zones.
-Our results show that stemmed points are only located in certain zones, Central Temperature Zone (CT) of vegetation map or Central inland (CI) (@fig-gis-tcsa-sites panel A & D) of eco-region map.
-We observed that there are statistically significant differences in the distributions of TCSA values between some zones (Vegetation zones: `r eco_lgm_test_pretty`, Ecoregion zones: `r for_lgm_test_pretty`).
-The Tukey's HSD results show that TCSA values from the South Temperature Zone (ST) zone are significantly different from the other two vegetation zones (@fig-gis-tcsa-sites panel C).
-Looking into this further, we see that there are only six artifacts from the ST zone, with only three of these having TSCA values above 175.
-The other three have TCSA values lower than 100, similar to the other zones (@fig-gis-tcsa-sites panel B).
-Given the small number of artifacts with extreme TCSA values in the ST zone, we hesitate to conclude this is an archaeologically significant pattern of different TCSA values.
-The Tukey HSD pairwise differences show no pairs of eco-regions with significant differences in TCSA values.
-Overall, our results indicate no clear patterns in the distribution of TCSA values across the vegetation and eco-regional zones.
-
 # Discussion

### 2023-02-03 analysis/paper/paper.qmd
Finished editing the draft!

@@
 
-A recent study conducted use-wear analysis on 95 stemmed points from the Jingeuneul site, along with examining fracture patterns, TCSA, tip cross-sectional perimeter (TCSP), and neck width to understand the use of the tools [@lee2019were].
-Their results show that a considerable number of the stemmed points show diagnostic impact fractures (DIFs) on the surface, which likely occurred due to longitudinal forces from the shaft when a stone tip hits a target.
-They conclude that stemmed points may have been used as spear-throwers or bows [@sano2015backed].
-
 ## Tip Cross-Sectional Area
@@
 They analyzed stemmed points from Jingeuneul, located in southwest of Korea, which has the largest number of stemmed points (n = 99) found so far in a single site.
-For TCSA, they were only able to use ten stemmed points because they needed to select the stemmed points that retained the widest and thickest part of the specimens for TCSA metric and also displayed diagnostic impact fractures for the use-wear analysis.
+For TCSA, they were only able to use 10 stemmed points because they needed to select the stemmed points that retained the widest and thickest part of the specimens for TCSA metric and also displayed diagnostic impact fractures for the use-wear analysis.
 The purpose of using TCSA in their research was to compare the values to North American dart tips and arrowheads.
 Their results show that the TCSA range of Jingeuneul stemmed points is relatively wide, overlapping with both North American dart tips and arrowheads.
+According to their use-wear analysis, a significant number of the stemmed points have diagnostic impact fractures (DIFs) on the surface, likely caused by longitudinal forces from the shaft.
+They conclude that stemmed points may have been used as spear-throwers or bows.
 Inspired by @lee2019were and TCSA research from lithic assemblages in other parts of the world, we aim to investigate TCSA values for stemmed points from all over South Korea to better understand their likely uses during the Late Paleolithic.
@@
 In their study of hunter-gatherer mobility strategies, @hamilton2016ecological show that resources are abundant and predictable along coasts or lake shores, which is highly dependent on temperature and precipitation and thus hunter-gatherers often become effectively sedentary.
-Our prediction is that the range of TCSA values will be narrower in coastal areas such as ST and SW of vegetation zones than in inland areas such as CT.
+Our prediction is that the range of TCSA values will be narrower in coastal areas such as ST and SW of vegetation zones and ESC of geographical zones than in inland areas such as CT of vegetation zones and CI of geographical zones.
 
@@
   legend.key.width = unit(2.5, 'mm'))
+```
+
+```{r}
+# p-value
+# simple version
+# p_lgm <- round(summary(aov(min_tcsa ~ as.factor(mis_stage), tcsa_site_ass_mis))[[1]][[5]][1], digits = 2)
+
+# aggregate all the MIS 3 sites, MIS 2 sites
+tcsa_site_ass_mis <- tcsa_site_ass %>% 
+                      mutate(mis_stage = case_when (
+                        age_ka > 26.5 ~ 'Before' , age_ka > 19 ~ 'LGM', age_ka < 19 ~ 'After'))
+
+library("car")
+# devtools::install_github('achetverikov/apastats',subdir='apastats')
+library("apastats")
+p_lgm_test <- Anova(lm(min_tcsa ~ as.factor(mis_stage), tcsa_site_ass_mis))
+# describe.Anova(p_lgm_test, 1, 2)
+
+three_time_p_pretty <- apastats::describe.Anova(p_lgm_test, 1, 2)
 
@@
            y = 230)
-
-# aggregate all the MIS 3 sites, MIS 2 sites
-tcsa_site_ass_mis <- tcsa_site_ass %>% 
-                      mutate(mis_stage = case_when (
-                        age_ka > 26.5 ~ 'Before' , age_ka > 19 ~ 'LGM', age_ka < 19 ~ 'After'))
-
-# p-value
-# simple version
-# p_lgm <- round(summary(aov(min_tcsa ~ as.factor(mis_stage), tcsa_site_ass_mis))[[1]][[5]][1], digits = 2)
-
-library("car")
-# devtools::install_github('achetverikov/apastats',subdir='apastats')
-library("apastats")
-p_lgm_test <- Anova(lm(min_tcsa ~ as.factor(mis_stage), tcsa_site_ass_mis))
-# describe.Anova(p_lgm_test, 1, 2)
-
-three_time_p_pretty <- apastats::describe.Anova(p_lgm_test, 1, 2)
 
@@
            alpha = 0.2)
-
-```
-
-In order to explore spatial patterns and the relationship between the role of stemmed points and the environment, we located individual assemblages on maps that depict eco-regional zones.
-Our results show that stemmed points are only located in certain zones, Central Temperature Zone (CT) of vegetation map or Central inland (CI) (@fig-gis-tcsa-sites panel A & D) of eco-region map.
-We observed that there are statistically significant differences in the distributions of TCSA values between some zones (Vegetation zones: `r eco_lgm_test_pretty`, Ecoregion zones: `r for_lgm_test_pretty`).
-The Tukey's HSD results show that TCSA values from the South Temperature Zone (ST) zone are significantly different from the other two vegetation zones (@fig-gis-tcsa-sites panel C).
-Looking into this further, we see that there are only six artifacts from the ST zone, with only three of these having TSCA values above 175.
-The other three have TCSA values lower than 100, similar to the other zones (@fig-gis-tcsa-sites panel B).
-Given the small number of artifacts with extreme TCSA values in the ST zone, we hesitate to conclude this is an archaeologically significant pattern of different TCSA values.
-The Tukey HSD pairwise differences show no pairs of eco-regions with significant differences in TCSA values (@fig-gis-tcsa-sites panel F).
-Overall, our results indicate no clear patterns in the distribution of TCSA values across the vegetation and eco-regional zones.
-
-```{r}
-#| label: fig-gis-tcsa-sites
-#| fig-cap: "TCSA values by assemblages and ecoregions.
-A: Modern vegetation map defined by temporature and precipitation.
-The individual points indicate assemblages.
-B: TCSA distribution for vegitation zones.
-The individual points indicate artifacts.
-C: Tukey test result for vegitation zones.
-D: Ecoregion map based on geographical boundaries and ecological conditions (modified the map from @lee2008relationship).
-E: TCSA distribution for ecoregion zones, excluding zones with no artifacts.
-F: Tukey test result for ecoregion zones"
-#| fig-height:  5
+```
+
+```{r}
+######Tukey test######
+#vegetation region#
 
@@
 
+library(sf)
+
 ############ explore tcsa with MODERN eco_regions #######
-
 library(sf)
 veg_all <- read_sf(here("analysis/data/gis_data/eco_re_two.shp"))
-
 veg_all <- st_transform(veg_all, crs = st_crs(4326))
-
 # ggplot() +
@@
 #   coord_sf()
-
 pts_all <- st_as_sf(korean_archaeological_site_locations_tcsa,
@@
 #   geom_sf(data = pts_all)
-
 sf_use_s2(FALSE) # to prevent error. Check here for more info: https://github.com/r-spatial/sf/issues/1762
-
 points_in_veg_zones_all <- 
@@
         join = st_contains)
-
 eco_plot <- ggplot() +
@@
           axis.title.y = element_text(size = 6)) 
-
 # explore artifacts in certain zone, related to Tukey's test
@@
 ############ explore tcsa with MODERN eco_regions & forest zones #######
-
 veg_forest <- 
   read_sf(here("analysis/data/gis_data/eco_forest_two.shp"))
-
 veg_forest <- 
@@
 #   coord_sf()
-
 pts_forest <- 
@@
 #   geom_sf(data = pts_forest)
-
 #sf_use_s2(FALSE)
-
 points_in_veg_zones_forest <- 
@@
         join = st_contains)
-
 for_plot <- ggplot() +
@@
   drop_na(Region) 
-
 for_box <- ggplot(points_in_veg_zones_forest_box) +
@@
 eco_aov <- aov(min_tcsa ~ ecoregion, points_in_veg_zones_all)
-
 eco_lgm_test <- Anova(lm(min_tcsa ~ ecoregion, points_in_veg_zones_all))
-
 eco_lgm_test_pretty <- apastats::describe.Anova(eco_lgm_test, 1, 2)
-
 eco_tur <- TukeyHSD(eco_aov)
@@
           axis.title.y = element_text(size = 6))
-
 #eco region#
@@
       points_in_veg_zones_forest_box)
-
 # summary(for_aov)
@@
            points_in_veg_zones_forest_box))
-
 for_lgm_test_pretty <- 
@@
 for_tur <- TukeyHSD(for_aov)
-
 # plot Tukey HSD
@@
 
+```
+
+In order to explore spatial patterns and the relationship between the role of stemmed points and the environment, we located individual assemblages on maps that depict eco-regional zones.
+Our results show that stemmed points are only located in certain zones, Central Temperature Zone (CT) of vegetation map or Central inland (CI) (@fig-gis-tcsa-sites panel A & D) of eco-region map.
+We observed that there are statistically significant differences in the distributions of TCSA values between some zones (Vegetation zones: `r eco_lgm_test_pretty`, Ecoregion zones: `r for_lgm_test_pretty`).
+The Tukey's HSD results show that TCSA values from the South Temperature Zone (ST) zone are significantly different from the other two vegetation zones (@fig-gis-tcsa-sites panel C).
+Looking into this further, we see that there are only six artifacts from the ST zone, with only three of these having TSCA values above 175.
+The other three have TCSA values lower than 100, similar to the other zones (@fig-gis-tcsa-sites panel B).
+Given the small number of artifacts with extreme TCSA values in the ST zone, we hesitate to conclude this is an archaeologically significant pattern of different TCSA values.
+The Tukey HSD pairwise differences show no pairs of eco-regions with significant differences in TCSA values (@fig-gis-tcsa-sites panel F).
+Overall, our results indicate no clear patterns in the distribution of TCSA values across the vegetation and eco-regional zones.
+
+```{r}
+#| label: fig-gis-tcsa-sites
+#| fig-cap: "TCSA values by assemblages and ecoregions.
+A: Modern vegetation map defined by temporature and precipitation.
+The individual points indicate assemblages.
+B: TCSA distribution for vegitation zones.
+The individual points indicate artifacts.
+C: Tukey test result for vegitation zones.
+D: Ecoregion map based on geographical boundaries and ecological conditions (modified the map from @lee2008relationship).
+E: TCSA distribution for ecoregion zones, excluding zones with no artifacts.
+F: Tukey test result for ecoregion zones"
+#| fig-height:  5
+
 plot_grid(eco_plot, 
@@
 None have TCSA values in the range of poison arrow tips (@fig-tcsa-all-sp).
-But in general, we found a wide range of TCSA in our sample, indicating that stemmed points might play diverse roles in foraging toolkits.
+@lee2019were computed TCSA for 10 stemmed points from Jingeuneul site, which we were not able to include in this research due to data availability, and concluded that their TCSA values fall within the range of North American dart tips and arrowheads.
+In general, we found a wide range of TCSA in our sample, indicating that stemmed points might play diverse roles in foraging toolkits.
+
 TCSA can be impacted by other factors such as raw materials and portability.
@@
 We found that different clusters of weight are matched with different types of weapon tips (@fig-artifact-size-cluster).
+
 The TCSA varies from assemblages to assemblages without any discernible temporal pattern in functional use (@fig-tcsa-radio-carbon panel A).
 But the stemmed points mainly used before LGM (@fig-tcsa-radio-carbon panel B).
-We found that stemmed points are only located in certain eco-regional zones such as Central Temperature zone without any clear pattern in function (@fig-gis-tcsa-sites).
+While we predicted the greatest variation in TCSA values during the LGM, we found that variability decreased during the LGM in comparison to the previous period.
+This could be explained by the overall massive decline in stemmed points usage.
+We found that stemmed points are only located in certain eco-regional zones such as Central Temperature zone without any clear pattern in TCSA function (@fig-gis-tcsa-sites).
+However, as predicted, TCSA values in inland areas such as CT of vegetation zones show higher variability.
+Based on the results, we assume that stemmed points performed a wider range of tasks in low-productivity patches.
 Both temporal and spatial analyses show that stemmed points were made for a particular environment, not so much harsh as LGM.
@@
 Especially, we confirmed the widest TCSA range in a single site, Suyanggae (@fig-tcsa-per-sites).
-As a consequence, our results are consistent with our second model in which stemmed points are described as multifunctional tools, suggesting that people created stemmed points in response to unexpected or varying circumstances in their habitats, similar to Clovis points (@eren2022plains).
+As a consequence, our results are consistent with our second scenario in which stemmed points are described as multifunctional tools, suggesting that people created stemmed points in response to unexpected or varying circumstances in their habitats, similar to Clovis points (@eren2022plains).
 
@@
 increased or decrease by shape of tip and cross section), we still consider it a useful metric for discriminating between weapon tip tools [@hughes1998getting; @sitton2020tip; @lombard2021variation].
-As part of our future research, we will conduct usewear analyses of stemmed points and experimental research in an effort to elucidate the detailed function of stemmed points and technological links between hunter-gatherer groups.
+As part of our future research, we will conduct use-wear analyses of stemmed points and experimental research in an effort to elucidate the detailed function of stemmed points and technological links between hunter-gatherer groups.
 

### 2023-02-10 analysis/paper/paper.qmd
editing up to the discussion

@@
 They analyzed stemmed points from Jingeuneul, located in southwest of Korea, which has the largest number of stemmed points (n = 99) found so far in a single site.
-For TCSA, they were only able to use 10 stemmed points because they needed to select the stemmed points that retained the widest and thickest part of the specimens for TCSA metric and also displayed diagnostic impact fractures for the use-wear analysis.
+For TCSA, they were only able to use 10 stemmed points because they chose to select the stemmed points that retained the widest and thickest part of the specimens for TCSA metric and also displayed diagnostic impact fractures for the use-wear analysis.
 The purpose of using TCSA in their research was to compare the values to North American dart tips and arrowheads.
-Their results show that the TCSA range of Jingeuneul stemmed points is relatively wide, overlapping with both North American dart tips and arrowheads.
+Their results show that the range of TCSA values for Jingeuneul stemmed points is relatively wide, overlapping with both North American dart tips and arrowheads.
 According to their use-wear analysis, a significant number of the stemmed points have diagnostic impact fractures (DIFs) on the surface, likely caused by longitudinal forces from the shaft.
 They conclude that stemmed points may have been used as spear-throwers or bows.
-Inspired by @lee2019were and TCSA research from lithic assemblages in other parts of the world, we aim to investigate TCSA values for stemmed points from all over South Korea to better understand their likely uses during the Late Paleolithic.
+Inspired by @lee2019were and TCSA research from lithic assemblages in other parts of the world, we aim to investigate TCSA values for a much larger sample of stemmed points from all over South Korea to better understand their likely uses during the Late Paleolithic.
 
@@
 Most sites contain only a few stemmed points and only a few sites have many more, such as Jingeuneul, Suyanggae (n = 55), and Yongsandong (n = 38) [@Kim_2017].
-Among these stemmed points, we selected those that retain their widest and thickest parts.
+Among these stemmed points, we selected those that retained their widest and thickest parts.
 We selected stemmed points that were found during field surveys as well as those were from sites that were never dated but were associated with other Late Palaeolithic artifacts.
@@
 
-To aid with interpreting our results, we followed the TCSA ranges for different weapon-delivery systems that @lombard2022tip and @lombard2021variation created by summarizing @wadley2008segment' study of backed microliths and calibrating existing ranges (@tbl-tcsa-ranges).
+To aid with interpreting our results, we followed the TCSA ranges for different weapon-delivery systems that @lombard2022tip and @lombard2021variation created by summarizing the analysis by @wadley2008segment of backed microliths (@tbl-tcsa-ranges).
 We excluded `r n_omit` artifacts from our dataset with TCSA greater than 250 and thus outside of the range of the comparative data.
-For this study we include TCSA values from a total of `r min_n_points` stemmed points from `r n_ass` assemblages unearthed from `r n_sites` sites.
+For this comparison with weapon-delivery systems we include TCSA values from a total of `r min_n_points` stemmed points from `r n_ass` assemblages unearthed from `r n_sites` sites.
 
@@
 We chose weight as a proxy for overall size of the stemmed points.
-Weight is a reliable discriminator between tools of different sizes and a descriptive attribute which can be measured rapidly and objectively and serve as a criterion of likeness or dissimilarity in projectile point description [@fenenga1953weights; @shea2006origins].
+Weight is a reliable discriminator between tools of different sizes and a descriptive attribute which can be measured rapidly and objectively [@fenenga1953weights; @shea2006origins].
 We then explored the relationship between weight and raw materials on TCSA values.
-Excluding points that we were not able to measure or obtain the records of their weight, we explored the relationship between TCSA and weight for `r n_size` artifacts.
+Excluding points that we were not able to directly measure or obtain the records of their weight, we explored the relationship between TCSA and weight for `r n_size` artifacts.
 
@@
 We summarized the distribution of TCSA values for each assemblage to examine spatial patterns in TCSA values.
-Among `r n_sites` artifacts, we combined sites that contains fewer than 5 stemmed points aside and named them as "Other".
+Among `r n_sites` sites, we combined those that contain fewer than five stemmed points and named them as "Other".
 We explored possible effect of environmental variation on TCSA values by grouping stemmed points by the eco-regional zones that they were found in, and comparing the distributions of TCSA values across the different zones.
@@
 In their study of hunter-gatherer mobility strategies, @hamilton2016ecological show that resources are abundant and predictable along coasts or lake shores, which is highly dependent on temperature and precipitation and thus hunter-gatherers often become effectively sedentary.
-Our prediction is that the range of TCSA values will be narrower in coastal areas such as ST and SW of vegetation zones and ESC of geographical zones than in inland areas such as CT of vegetation zones and CI of geographical zones.
+Our prediction is that the range of TCSA values will be narrower in coastal areas such as vegetation zones ST and SW and geographical zone ESC, compared to inland areas such as vegetation zone CT and geographical zone CI.
 
 The entire R code [@core2021team] and data files used for all the analysis and visualizations contained in this paper is openly available at https://doi.org/10.17605/osf.io/dqna8 to enable re-use of materials and improve reproducibility and transparency [@marwick2017computational].
-All of the figures, tables, statistical test results presented here in addition to the raw data can be independently reproduced with the code and data in this repository.
+All of the figures, tables, statistical test results presented here can be independently reproduced with the code and data in this repository.
 The code is released under the MIT license, the data as CC-0, and figures as CC-BY, to enable maximum re-use.
@@
 Using weight as a size proxy we examined the relationship between size and likely use of the stemmed points inferred from TCSA values.
-We conducted a univariate cluster analysis [@song2020efficient; @wang2011ckmeans] of stemmed points by weight and it revealed three clusters of artifact sizes (A of @fig-artifact-size-cluster, mean = `r round(mean(tcsa_mass_no_na$'Weight(g)'), digits=1)`, SD = `r round(sd(tcsa_mass_no_na$'Weight(g)'), digits=1)`).
-Cluster 1, the smallest (lightest) artifacts, have a lower range of TCSA values, compared to Cluster 2 (B of @fig-artifact-size-cluster, mean = `r round(mean(tcsa_mass_no_na$'min_tcsa'), digits=1)`, SD = `r round(sd(tcsa_mass_no_na$'min_tcsa'), digits=1)`).
+We conducted a univariate cluster analysis [@song2020efficient; @wang2011ckmeans] of stemmed points by weight, revealing three clusters of artifact sizes (A of @fig-artifact-size-cluster, mean = `r round(mean(tcsa_mass_no_na$'Weight(g)'), digits=1)`, SD = `r round(sd(tcsa_mass_no_na$'Weight(g)'), digits=1)`).
+Cluster 1, the smallest (lightest) artifacts, has a lower range of TCSA values, compared to Cluster 2 (B of @fig-artifact-size-cluster, mean = `r round(mean(tcsa_mass_no_na$'min_tcsa'), digits=1)`, SD = `r round(sd(tcsa_mass_no_na$'min_tcsa'), digits=1)`).
 TCSA values for Cluster 3 are the highest, except for one artifact, which is lower than 50.
-There are a few overlaps, but we can still assume that artifacts in different size clusters might be made for different purposes.
+There are a few overlaps, but we can still assume that artifacts in different size clusters might have been made for different purposes.
 
@@
 The category of "Other" raw materials include porphyry, trachyte, felsite, chert, quartz, quartzite, granite, mudstone, and unidentified rocks.
-Overall, there is no clear pattern of TCSA among raw materials.
+Overall, there is no clear pattern of TCSA values among different raw materials.
 
@@
 
-@fig-tcsa-size shows a positive relationship between TCSA, raw material and artifact weight.
+@fig-tcsa-size shows positive relationships between TCSA, raw material and artifact weight.
 There is a stronger correlation between TCSA and artifact weight for acidic volcanic points (i.e.
@@
 
-We computed artifact TCSA values for `r n_sites` sites that contain more than 5 stemmed points to observe variation between sites (@fig-tcsa-per-sites).
-Sites with fewer than 5 stemmed points were grouped under the category of "Other".
+We computed artifact TCSA values for `r n_sites` sites that contain more than five stemmed points to observe variation between sites (@fig-tcsa-per-sites).
+Sites with fewer than five stemmed points were grouped under the category of "Other".
 Among the sites in our sample, Suyanggae has the most stemmed points and the widest range of TCSA values.
@@
 Nosanri, Sibuk, and Wolpyeng have a narrower range of lower values, which indicates less diverse likely uses for stemmed points.
-Yongsujaeul shows a narrower range but higher TCSA value.
-Yongsandong has the second most stemmed points and shows roughly two different clusters of TCSA values.
+Yongsujaeul shows a narrower range but higher TCSA values.
+Yongsandong has the second highest number of stemmed points and shows two clusters of TCSA values.
 
@@
 In order to explore spatial patterns and the relationship between the role of stemmed points and the environment, we located individual assemblages on maps that depict eco-regional zones.
-Our results show that stemmed points are only located in certain zones, Central Temperature Zone (CT) of vegetation map or Central inland (CI) (@fig-gis-tcsa-sites panel A & D) of eco-region map.
+Our results show that stemmed points are only located in certain zones, Central Temperature Zone (CT) of the vegetation map or Central inland (CI) (@fig-gis-tcsa-sites panel A & D) of the eco-region map.
 We observed that there are statistically significant differences in the distributions of TCSA values between some zones (Vegetation zones: `r eco_lgm_test_pretty`, Ecoregion zones: `r for_lgm_test_pretty`).
@@
 Given the small number of artifacts with extreme TCSA values in the ST zone, we hesitate to conclude this is an archaeologically significant pattern of different TCSA values.
-The Tukey HSD pairwise differences show no pairs of eco-regions with significant differences in TCSA values (@fig-gis-tcsa-sites panel F).
+The Tukey HSD pairwise differences show no pairs of eco-regions with significant differences in TCSA values, likely due to the small number of points in several zones (@fig-gis-tcsa-sites panel F).
 Overall, our results indicate no clear patterns in the distribution of TCSA values across the vegetation and eco-regional zones.

### 2023-02-13 analysis/paper/paper.qmd
continuing editing

@@
 None have TCSA values in the range of poison arrow tips (@fig-tcsa-all-sp).
-@lee2019were computed TCSA for 10 stemmed points from Jingeuneul site, which we were not able to include in this research due to data availability, and concluded that their TCSA values fall within the range of North American dart tips and arrowheads.
-In general, we found a wide range of TCSA in our sample, indicating that stemmed points might play diverse roles in foraging toolkits.
+@lee2019were computed TCSA for 10 stemmed points from Jingeuneul site, which we were not able to include in this research due to the unavailability of their data, and concluded that their TCSA values fall within the range of North American dart tips and arrowheads.
+In general, we found a wider range of TCSA in our sample, indicating that stemmed points might play diverse roles in foraging toolkits.
 
@@
 @eren2022plains explain that smaller TCSA could be the result of pursuing production economy and transport efficiency.
-We explore the relationship between TCSA and raw material types and weight in order to examine these factors.
+We explored the relationship between TCSA and raw material types and weight in order to examine these factors.
 We observed no clear pattern for raw materials (@fig-tcsa-raw-materials) but weight (@fig-artifact-size-cluster), though weight is correlated with raw materials (@fig-tcsa-size).
+We speculate this might be due to the higher availability of raw materials such as shale on hornfels on the landscape.
 Therefore we could assume that raw materials also associated with TCSA value.
@@
 
-The TCSA varies from assemblages to assemblages without any discernible temporal pattern in functional use (@fig-tcsa-radio-carbon panel A).
-But the stemmed points mainly used before LGM (@fig-tcsa-radio-carbon panel B).
-While we predicted the greatest variation in TCSA values during the LGM, we found that variability decreased during the LGM in comparison to the previous period.
-This could be explained by the overall massive decline in stemmed points usage.
+The TCSA varies between assemblages without any discernible temporal pattern in functional use (@fig-tcsa-radio-carbon panel A).
+Stemmed points were mainly used before the LGM (@fig-tcsa-radio-carbon panel B).
+While we predicted the greatest variation in TCSA values during the LGM, our results show that variability decreased during the LGM, in comparison to the previous period.
+This could be explained by the overall decline in stemmed points usage.
 We found that stemmed points are only located in certain eco-regional zones such as Central Temperature zone without any clear pattern in TCSA function (@fig-gis-tcsa-sites).

### 2023-02-13 analysis/paper/paper.qmd
minor edits while waiting for GP

@@
 
-The TCSA varies between assemblages without any discernible temporal pattern in functional use (@fig-tcsa-radio-carbon panel A).
-Stemmed points were mainly used before the LGM (@fig-tcsa-radio-carbon panel B).
+Our results shows that TCSA values vary between assemblages with few discernible temporal patterns in function (@fig-tcsa-radio-carbon panel A).
+Stemmed points were used most frequently before the LGM (@fig-tcsa-radio-carbon panel B).
 While we predicted the greatest variation in TCSA values during the LGM, our results show that variability decreased during the LGM, in comparison to the previous period.
 This could be explained by the overall decline in stemmed points usage.
-We found that stemmed points are only located in certain eco-regional zones such as Central Temperature zone without any clear pattern in TCSA function (@fig-gis-tcsa-sites).
-However, as predicted, TCSA values in inland areas such as CT of vegetation zones show higher variability.
-Based on the results, we assume that stemmed points performed a wider range of tasks in low-productivity patches.
-Both temporal and spatial analyses show that stemmed points were made for a particular environment, not so much harsh as LGM.
+
+We found that stemmed points are most frequently located in a small number of eco-regional zones such as the Central Temperature zone with few clears pattern in TCSA function across the zones (@fig-gis-tcsa-sites).
+As predicted, TCSA values in inland areas, such as vegetation zone CT, show higher variability.
+Based on these results, we assume that stemmed points performed a wider range of tasks in low-productivity patches.
+Both temporal and spatial analyses show that stemmed points were likely more adapted for particular environments, rather than specific climate contexts.
 @prates2022changes claimed that fishtail projectile points in South America was used for megafauna hunting and contributed to their extinction by proving strong relationship between the temporal density and spatial distribution of megafauna and the projectile points.
-Our results support this claim by showing the dense distribution of stemmed point in certain environment.
-
-Our models premise that the choice of weapon-tip types reflects the socio-environmental circumstances.
-To examine the possible connections between the roles of stemmed points and environmental change or population dynamics during the Late Paleolithic period, we compared the temporal and spatial patterns of TCSA to results of simulated Mean Annual Temperature (MAT) and Summed probability distribution (SPD) from @park2022did.
-The simulated temperature distribution in south Korea during the Late Paleolithic period (Figure 9: A from @park2022did) seems similar to the modern eco-regional zones that we used in this research (@fig-gis-tcsa-sites panel A & D).
-Figure 9: C from @park2022did shows the slow decrease of temperature until the end of LGM, which could explain our results of temporal pattern, meaning stemmed points were used during the cooler environment with a temperature between 4 and 6 ℃.
-Results of SPD (Figure 10 from @park2022did) show several peaks, as indicated by the positive deviations from the three null models (shaded in red), and downs (shaded in blue), but no drastic growth or decline during our research period, indicating no major impact from population dynamics on stemmed points.
-
-Overall, our temporal and spatial patterns of TCSA show a wide TCSA range throughout the Late Paleolithic period and between eco-regional zones.
-Especially, we confirmed the widest TCSA range in a single site, Suyanggae (@fig-tcsa-per-sites).
-As a consequence, our results are consistent with our second scenario in which stemmed points are described as multifunctional tools, suggesting that people created stemmed points in response to unexpected or varying circumstances in their habitats, similar to Clovis points (@eren2022plains).
+We similarly found that stemmed points are more densly distributed in certain environments in South Korea, perhaps also relating to the distribution of fauna.
+
+Overall, our temporal and spatial patterns of TCSA show a wide range of TCSA values throughout the Late Paleolithic period and between eco-regional zones.
+We confirmed the widest TCSA range in a single site, Suyanggae (@fig-tcsa-per-sites).
+As a consequence, our results are consistent with our second scenario in which stemmed points are described as multifunctional tools, suggesting that people created stemmed points in response to unexpected or varying circumstances in their habitats, similar to how @eren2022plains explain Clovis points.
 
@@
 Then we examined the TCSA values with other variables including raw materials, weight, radiocarbon dates, and site locations.
-Based on the evolutionary approach, we premised that the way of using stemmed points is likely to reflect the socio-environmental circumstances that people encountered and managed.
-We examined the possible impact of LGM on stemmed points and their distribution on eco-regional maps.
-
-According to the different weapon-delivery systems that summarize TCSA of projectile points from archaeological and ethnographic cases, a majority of stemmed points might be used the most likely as javelin tips and stabbing spear tips.
-In general, though, we noted a wide range and also differing distribution of TCSA in each assemblage.
-Therefore, we conclude that stemmed points might play diverse roles during the Late Paleolithic.
-From our findings, we infer that people were exposed to an unfamiliar situation and used a stemmed point as a multifunctional tool to carry out multiple tasks.
-We are aware that discriminating the likely use of small projectile points could be arbitrary [@erlandson2014darts].
+Drawing on evolutionary theory, we premised that the way of using stemmed points is likely to reflect the socio-environmental circumstances that people encountered and managed.
+We examined the possible impact of LGM on stemmed points and their distribution across eco-regional zones.
+
+According to the different weapon-delivery systems that can be inferred from TCSA values, the majority of stemmed points from the Korean Palaeolithic were probably used as javelin tips and stabbing spear tips.
+In general, though, we noted a wide range and also differing distributions of TCSA values in each assemblage.
+Therefore, we conclude that stemmed points had diverse roles during the Late Paleolithic.
+We conclude that people frequently encountered unfamiliar situations and used stemmed points as a multifunctional tool to carry out multiple tasks.
+We are aware that discriminating the likely use of small numbers of projectile points could be arbitrary [@erlandson2014darts].
 Since TCSA covers the critical elements of projectiles, flight and penetration dynamics (i.e.

### 2023-02-17 analysis/paper/paper.qmd
Edited until methods

--- 2023-02-10 analysis/paper/paper.qmd
@@
 abstract: |
- The introduction of blade technology, stemmed points, end scrapers, burins, denticulates, and finer grained materials characterizes the transition from the Early to Late Paleolithic in Korea.
+ The transition from the Early to Late Paleolithic in Korea is characterized by the introduction of blade technology, stemmed points, end scrapers, burins, denticulates, and finer grained materials.
 Stemmed points have been considered a representative tool that led this set of changes.
-In this research, we examine the possible role that stemmed points played during the technological transition, as well as throughout the Late Paleolithic.
+In this study, we examine the possible role that stemmed points played during the technological transition, as well as throughout the Late Paleolithic period.
 Our main questions are: What were the best-fit ballistic probabilities for the stemmed points if they were hafted as weapon tips?
 How diverse were their likely uses?
-What are the temporal and spatial patterns in stemmed point use?
+What are the temporal and spatial patterns of stemmed point use?
 We measured tip cross-sectional area (TCSA) to discriminate different likely use classes of projectile points, for example, as poisoned arrow tips or as stabbing spears.
 We analyzed TCSA with other variables including raw materials, weight, radiocarbon dates and locations.
-Our results show that the stemmed points were could have served as javelin tips and stabbing spear tips, with smaller numbers as dart tips and un-poisoned arrow tips.
+Our results show that the stemmed points could have served as javelin tips and stabbing spear tips, with smaller numbers as dart tips and un-poisoned arrow tips.
 TCSA values are controlled mostly by size rather than raw material types.
-We found different TCSA range of stemmed points at different sites, which could indicate people used stemmed points in different ways depending on the local environment.
+We found different TCSA ranges of stemmed points at different sites, which could indicate people used stemmed points in different ways depending on the local environment.
 Some sites show a wide range of TCSA values that represent multi-purpose usage of stemmed points.
-The temporal pattern of TCSA values is one of little change throughout the Late Paleolithic period but points were mainly produced before Last Glacial Maximum (LGM).
-We observed that stemmed points were mostly located in certain ecoregions in Korea but no clear spatial pattern is confirmed.
+The temporal pattern of TCSA values is one of little change throughout the Late Paleolithic period but points were predominantly produced before the Last Glacial Maximum (LGM).
+We observed that stemmed points were mostly located in certain ecoregions in Korea but no clear spatial pattern was confirmed.
 We conclude that stemmed points were multi-functional tools, with many likely designed for javelin and stabbing spear tips.
@@
 The introduction of new stone artifact technologies marks a major transition in the Korean Paleolithic from the Early to the Late Paleolithic periods.
-The transition includes the appearance blade technology, stemmed points, end scrapers, burins, denticulates, etc [@bae2017origin; @nakazawa2018quaternary; @bae2017late; @lee2017korean; @seong2016eastern].
-Stemmed points are considered the first evidence of a suite of new technologies defining the Late Paleolithic in this region [@seong2016eastern; @seong2008tanged].
+The transition includes blade technology, stemmed points, end scrapers, burins, denticulates, etc.
+[@bae2017origin; @nakazawa2018quaternary; @bae2017late; @lee2017korean; @seong2016eastern].
+Stemmed points are considered to be the first evidence of a suite of new technologies defining the Late Paleolithic period in this region [@seong2016eastern; @seong2008tanged].
 This is related to the fact that stemmed points appear to originate from Korea and spread throughout Northeast Asia, and they have a close association with mobility, site formation, and occupation diversity [@o2018origins; @Chong_2021; @park2022did].
 Despite the importance of stemmed points, only a few studies to date have examined their likely uses.
-Previous work has mostly discussed their origin, the chronology of the Korean Late Paleolithic, and their relationship with the Japanese archipelago [@chang2013human; @Park_2013; @lee2019were; @Chong_2021].[@chang2013human; @Park_2013; @lee2019were; @Chong_2021].
-
-To understand the roles that stemmed points played in the technological transition, this study examines the possible uses of stemmed points.
+Previous work has mostly discussed their origin, the chronology of the Korean Late Paleolithic, and their relationship with the Japanese archipelago [@chang2013human; @Park_2013; @lee2019were; @Chong_2021].
+
+The purpose of this study is to examine the possible uses of stemmed points in order to understand how stemmed points played a role in the technological transition.
 We use the tip cross-sectional area (TCSA) metric to infer potential variability in weapon-use strategies, and compare the results to other archaeological and ethnographic assemblages based on the idea that different shapes correspond to different weapon types [@lombard2021variation].
@@
 How diverse were their likely uses?
-What are the temporal and spatial patterns in stemmed point uses?
-As part of our discussion, we examine possible connections between the roles of stemmed points and environmental change or population dynamics during the Late Paleolithic period.
+What are the temporal and spatial patterns of stemmed point uses?
+As part of our discussion, we examine possible links between the roles of stemmed points and environmental change or population dynamics during the Late Paleolithic period.
 In order to understand how a certain weapon-tip type was chosen, we apply an evolutionary perspective with the assumption that people chose their weapon tip types as part of their adaptation to the socio-environmental circumstances they found themselves in.
@@
 Stemmed points (Sumbejjirugae in Korean) are projectile points made on an elongated blade-like flake or blade with two parallel facets and a single or two arises that converge to form an inverted "Y" [@pratt2020circum] (@fig-sp-hand-made).
-Slight retouch is typically located on the proximal end to shape an acute tip and on the distal end to make a stem, which connects to a wooden shaft.
-Elsewhere in the world these types of artifacts are often called 'tanged points', but we prefer 'stemmed point' here to distinguish from Bronze Age stone projectile points that have long been called 'tanged points' in Korea [@park2022did].
-Understanding the appearance of stemmed points is relevant to general questions about the origins of projectile technology, the emergence of the Late Paleolithic, and relationships between Korea and adjacent regions in East Asia, such as Japan.
-
-Stemmed points are the first composite tools in the Korean Paleolithic that required two different parts to be one complete tool, consisting of a stone point and shaft, presumably made out of wood [@seong2008tanged].
-Using blades as the blank for the point enables mass production for this composite tool and its shape can become more standardized [@lee2015characteristics; @park2022did].
-Therefore, as @o2018origins claimed, understanding the emergence of projectile technology gives insights to greater cultural, evolutionary, and behavioral cognitive flexibility.
-
-Since stemmed points represent the beginning of the Korean Late Paleolithic, investigating their origins is important for understanding the technological transition from the Early Paleolithic, modern human dispersals into the region, and claims for the existence of the 'Middle' Paleolithic in Korea [@bae2010origin; @bae2017late; @norton2009evolution; @seong2016eastern].
+Slight retouch is typically performed on the proximal end to shape an acute tip and on the distal end to make a stem, which connects to a wooden shaft.
+Elsewhere in the world these types of artifacts are often called 'tanged points', but we prefer referring to them as 'stemmed points' here to distinguish them from Bronze Age stone projectile points which have long been called 'tanged points' in Korea [@park2022did].
+Understanding the appearance of stemmed points is relevant to general questions about the origins of projectile technology, the technological transition of the Late Paleolithic, and relationships between Korea and adjacent regions in East Asia, such as Japan.
+
+Stemmed points are the first composite tools in the Korean Paleolithic that required two different parts to be one complete tool, a stone point and a shaft, presumably made out of wood [@seong2008tanged].
+Using blades as the blank for the point enables mass production of this composite tool and its shape can become more standardized [@lee2015characteristics; @park2022did].
+Therefore, as @o2018origins claimed, understanding the emergence of projectile technology provides insights into greater cultural, evolutionary, and behavioral cognitive flexibility.
+
+Since stemmed points represent the beginning of the Korean Late Paleolithic, investigating their origins is critical for understanding the technological transition from the Early Paleolithic, modern human dispersals into the region, and claims for the existence of the 'Middle' Paleolithic in Korea [@bae2010origin; @bae2017late; @norton2009evolution; @seong2016eastern].
 Debate about explaining the origin of stemmed points which can be summarized into two competing models: a 'heterogenic' migration [@bae2010origin], and an *in situ* evolution model [@seong2009emergence].
@@
 1: North route: Siberia, Mongolia, or other regions of northeast China, and 2: South route: southern China) as the result of a continuing influx of modern human migration from two routes.
-The *in situ* model argues that stemmed points and other Late Paleolithic technologies, including blade industries, autonomously emerged in the south of the Korean peninsula, with no apparent external influence.
-The difference between the two models comes from how to understand a few early sites that contain stemmed points but no blades and blade cores.
-While the *in situ* model claims that stemmed points appeared before blades by acknowledging those early sites, including Bonggok, Songamri, Yonghodong, and Hwadaeri, the migration model supports the traditional definition of the Late or Upper Paleolithic in Europe and Africa that started with emergence and establishment of blade technology [@seong2016eastern].
+The *in situ* model argues that stemmed points and other Late Paleolithic technologies, such as blade industries, autonomously emerged in the south of the Korean peninsula, with no apparent external influence.
+The difference between the two models comes from how they explain a few early sites that contain stemmed points but no blades and blade cores.
+While the *in situ* model claims that stemmed points appeared before blades by acknowledging those early sites, including Bonggok, Songamri, Yonghodong, and Hwadaeri, the migration model supports the traditional definition of the Late or Upper Paleolithic in Europe and Africa that started with the emergence and establishment of blade technology [@seong2016eastern].
 
 Stemmed points are important proxies to understand human behaviors during the Late Paleolithic.
-@park2022did examined mobility and site occupation patterns by applying concepts of human behavioral ecology to lithic assemblages and found that forager groups using stemmed points may have been associated with the occupation of marginal or extreme environments, compared to groups with no stemmed points.
+@park2022did examined mobility and site occupation patterns by applying concepts of human behavioral ecology to lithic assemblages.
+The study found that forager groups using stemmed points may have been associated with the occupation of marginal or extreme environments, compared to groups with no stemmed points.
 Also, stemmed points were more frequently associated with expedient technologies, indicating residential and less mobile behaviors.
-@Chong_2021 claims that the morphological variation of the stemmed points along with tool types in lithic assemblages, assemblage size, use of raw materials, and types of blanks could represent specific characteristics of occupation such as a "limited activity station" and a "residential base camp." For example, stemmed points with high morphological variations in a tool size, shape of edge, degrees of damage, and types of edge retouching from Yongsandong site may indicate that the site was used for specific or limited activities such as hunting [@kim2004yongsan; @seong2015diversity; @Chong_2021].
-
-The connection between stemmed points in both Korea and Japan have been studied since the late 1980s as a part of evidence for long-distance/maritime cultural interchanges or social networks [@lee2015characteristics; @chang2013human].
+@Chong_2021 claims that the morphological variation of the stemmed points along with tool types in lithic assemblages, assemblage size, use of raw materials, and types of blanks could represent specific characteristics of occupation such as a "limited activity station" and a "residential base camp." For example, stemmed points with high morphological variations in tool size, shape of edge, degrees of damage, and types of edge retouching from Yongsandong site may indicate that the site was used for specific or limited activities such as hunting [@kim2004yongsan; @seong2015diversity; @Chong_2021].
+
+The connection between stemmed points in both Korea and Japan has been studied since the late 1980s as a part of evidence for long-distance/maritime cultural interchanges or social networks [@lee2015characteristics; @chang2013human].
 Stemmed points from the Bonggok site in Korea are currently accepted as the oldest ones among Northeast Asia, dated to ca.
-41.5 ka, and made on elongated flakes [@seong2009emergence; @seong2015diversity; @bae2017origin].
-After their first appearance in Korea, stemmed points (Hakuhensenntouki in Japanese) appear in Kyushu, Japan during late Marine isotope stages (MIS) 3.
-In addition to the stemmed points, there are similar artifacts found in both regions such as microblade cores, Moppule-seokgi (Kakusuijyosekki in Japanese), backed knife, bilateral points, bifacial points, and transport of obsidian [@lee2012characteristics; @chang2013human; @lee2015characteristics; @kim2021evidence].
+41.5 ka, and made out of elongated flakes [@seong2009emergence; @seong2015diversity; @bae2017origin].
+After their first appearance in Korea, stemmed points (Hakuhensenntouki in Japanese) appeared in Kyushu, Japan during the late Marine isotope stage (MIS) 3.
+In addition to the stemmed points, there are similar artifacts found in both regions such as microblade cores, Moppule-seokgi (Kakusuijyosekki  in Japanese), backed knives, bilateral points, bifacial points, and transport of obsidian [@lee2012characteristics; @chang2013human; @lee2015characteristics; @kim2021evidence].
 
@@
 
-Though stemmed points have been generally assumed to have been hunting armatures [@lee2019were; @lee2011study; @seong2008tanged; @lee2002new; @chang2013human], it is difficult to understand the likely uses of the tools without knowing their complete shape when attached to other components.
-Preserved wooden components of projectile tools are too few and rare to standardize the overall scale and variability of the tools used in prehistory [@shea2006origins].
-@Lee_Kong_2002 claimed stemmed points were generalized 'stemmed tools' because of their diverse shapes, which could represent different functions, as well as the uncertainty of complete tool shape.
-Other researchers propose that non-symmetrical stemmed points, with retouch on one side of the artifact or denticulate blades on one side, should not be called as stemmed points but instead stemmed knives, stemmed side-scrapers, stemmed end-scrapers, or stemmed burins [@seong2008tanged; @Lee_Jang_2011; @Kim] @fig-sp-hand-made: 7-8.
-
-Most stemmed points are symmetrical from the tip to the tang, having the central axis function as a line of symmetry [@Lee_Jang_2011].
-There are a high proportion of broken tips and stems and the reused tools were repaired based on keeping the symmetry [@Park_2013; @Kim_2017].
+Though stemmed points have generally been assumed to have been hunting armatures [@lee2019were; @lee2011study; @seong2008tanged; @lee2002new; @chang2013human], it is difficult to determine their likely uses without knowing their complete shape when attached to other components.
+Preserved wooden components of prehistoric projectile tools are too few and rare to standardize their overall scale and variability [@shea2006origins].
+@Lee_Kong_2002 claimed stemmed points should be considered more generally as 'stemmed tools' because of their diverse shapes, which could represent different functions, and the uncertainty of their complete shape.
+Other researchers propose that non-symmetrical stemmed points, with retouch on one side or denticulate blades on one side, should not be referred to as stemmed points but rather as stemmed knives, stemmed side-scrapers, stemmed end-scrapers, or stemmed burins (@fig-sp-hand-made: 7-8) [@seong2008tanged; @Lee_Jang_2011; @Kim].
+
+Stemmed points are typically symmetrical from tip to tang, with the central axis serving as a line of symmetry [@Lee_Jang_2011].
+There is a high percentage of broken tips and stems, and the reused tools were repaired in accordance with symmetry [@Park_2013; @Kim_2017].
 Studies of stemmed point manufacturing processes and the patterns of broken pieces show that stemmed points may have been used mainly as spear tips [@Lee_1985; @Chang_2002].
-On Yongsandong site, for example, the percentage of complete tools is only 10%, while the percentage of missing tips is 33%.
-On Jingeuneul site, the percentages are 16% and 50%, respectively [@Park_2013].
-In addition to the morphological aspect of stemmed points, investigation of a whole site and tool composition of an assemblage shows that stemmed points or stemmed tools could be strongly associated with hunting activities including peeling the animal skin after slaughtering or separating the bones from the flesh [@seong2008tanged; @Chong_2021].
+For example, at Yongsandong site, only 10% of the tools are complete, while 33% of the tips are missing.
+In the case of Jingeuneul, the percentages are 16% and 50%, respectively [Park_2013].
+In addition to the morphological aspect of stemmed points, investigations of a whole site and the tool composition of an assemblage suggest that stemmed points or stemmed tools could be strongly associated with hunting activities including peeling the animal skin after slaughtering or separating the bones from the flesh [@seong2008tanged; @Chong_2021].
 
@@
 
-Tip cross-sectional area (TCSA) of stone artifacts has been used as a ballistically relevant standard to probabalistically discriminate different likely use classes of projectile armatures such as spearthrower (a.k.a.
+Tip cross-sectional area (TCSA) of stone artifacts has been used as a ballistically relevant standard to probabilistically discriminate different likely use classes of projectile armatures such as spearthrower (a.k.a.
 atlatl) dart tips, un-poisoned arrow tips and large stabbing/thrusting spears [@sisk2011african; @lombard2021variation; @lombard2020tip; @lombard2021did; @o2018origins; @lombard2022standardized].
-It is important to note that the TCSA metric alone cannot unambiguously determine artifact function, it only suggests a best-fit ballistic probability for the points if they were hafted as weapon tips.
-The TCSA metric represents the part of the tool that cuts the target's hide and relates to weapon flight and penetration dynamics [@hughes1998getting; @sitton2020tip; @lombard2021variation].
-This method was first proposed by @hughes1998getting and and validated by @shea2006origins by comparing archaeological examples to ethnographically collected samples of known use.
+It is critical to note that the TCSA metric alone cannot unambiguously determine artifact function; it only suggests a best-fit ballistic probability for the points if they were hafted as weapon tips.
+The TCSA metric represents the part of the tool that cuts the target's hide and is related to weapon flight and penetration dynamics [@hughes1998getting; @sitton2020tip; @lombard2021variation].
+This method was first proposed by @hughes1998getting and validated by @shea2006origins through the comparison of archaeological examples with ethnographically collected samples of known use.
 @shea2009impact also applied this approach to compare projectile points from Africa, the Levant, and Europe, and claimed that projectile weapons first appeared in Africa.
 
-A key advantage of the TCSA metric is its convenience of application: regardless of point type, only maximum width and thickness measurements are needed for calculating the TCSA value (0.5 x maximum width x maximum thickness) [@sisk2011african; @lombard2020tip].
-Later, @sisk2011african proposed an alternate metric, tip cross-sectional perimeter (TCSP), for more accurate measure for the force needed to penetrate a target to a lethal depth, while TCSA metric is more associated with cutting aspect.
+One of the key advantages of the TCSA metric is its convenience of application: regardless of the point type, only the maximum width and thickness measurements are required to calculate the TCSA value (0.5 x maximum width x maximum thickness) [@sisk2011african; @lombard2020tip].
+Later, @sisk2011african proposed an alternate metric, tip cross-sectional perimeter (TCSP), for a more precise measure of the force required to penetrate a target to a lethal depth, whereas the TCSA metric is more associated with cutting.
 However, TCSP has a few disadvantages that limit its applicability to our case study of Korean stemmed points.
@@
 @lee2019were first applied TCSA to stemmed points from Korea along with use-wear analysis.
-They analyzed stemmed points from Jingeuneul, located in southwest of Korea, which has the largest number of stemmed points (n = 99) found so far in a single site.
-For TCSA, they were only able to use 10 stemmed points because they chose to select the stemmed points that retained the widest and thickest part of the specimens for TCSA metric and also displayed diagnostic impact fractures for the use-wear analysis.
+They analyzed stemmed points from Jingeuneul, located in the southwest of Korea, which has the largest number of stemmed points (n = 99) found at a single site to date.
+For the TCSA, they were only able to use ten stemmed points since they selected stemmed points that retained the widest and thickest parts of the specimens and showed diagnostic impact fractures for the use-wear analysis.
 The purpose of using TCSA in their research was to compare the values to North American dart tips and arrowheads.
@@
 According to their use-wear analysis, a significant number of the stemmed points have diagnostic impact fractures (DIFs) on the surface, likely caused by longitudinal forces from the shaft.
-They conclude that stemmed points may have been used as spear-throwers or bows.
+Based on the results, they conclude that stemmed points may have been used as spear-throwers or bows.
 Inspired by @lee2019were and TCSA research from lithic assemblages in other parts of the world, we aim to investigate TCSA values for a much larger sample of stemmed points from all over South Korea to better understand their likely uses during the Late Paleolithic.
@@
 
-After the first discovery at the Suyanggae site, more than 450 stemmed points have been found in over 30 sites throughout South Korea [@Sohn_1967; @lee2019were; @Chong_2021].
+After the first discovery at the Seokjangri site at the 1960s, more than 450 stemmed points have been found in over 30 sites throughout South Korea [@Sohn_1967; @lee2019were; @Chong_2021].
 Most sites contain only a few stemmed points and only a few sites have many more, such as Jingeuneul, Suyanggae (n = 55), and Yongsandong (n = 38) [@Kim_2017].
@@
 How diverse were their likely uses?
-What are the temporal and spatial patterns in stemmed point uses?
+What are the temporal and spatial patterns of stemmed point uses?
 

### 2023-02-17 analysis/paper/paper.qmd

@@
 
-The TCSA varies from assemblages to assemblages without any discernible temporal pattern in functional use (@fig-tcsa-radio-carbon panel A).
-But the stemmed points mainly used before LGM (@fig-tcsa-radio-carbon panel B).
-While we predicted the greatest variation in TCSA values during the LGM, we found that variability decreased during the LGM in comparison to the previous period.
-This could be explained by the overall massive decline in stemmed points usage.
-We found that stemmed points are only located in certain eco-regional zones such as Central Temperature zone without any clear pattern in TCSA function (@fig-gis-tcsa-sites).
-However, as predicted, TCSA values in inland areas such as CT of vegetation zones show higher variability.
-Based on the results, we assume that stemmed points performed a wider range of tasks in low-productivity patches.
-Both temporal and spatial analyses show that stemmed points were made for a particular environment, not so much harsh as LGM.
+Our results shows that TCSA values vary between assemblages with few discernible temporal patterns in function (@fig-tcsa-radio-carbon panel A).
+Stemmed points were used most frequently before the LGM (@fig-tcsa-radio-carbon panel B).
+While we predicted the greatest variation in TCSA values during the LGM, our results show that variability decreased during the LGM, in comparison to the previous period.
+This could be explained by the overall decline in stemmed points usage.
+
+We found that stemmed points are most frequently located in a small number of eco-regional zones such as the Central Temperature zone with few clears pattern in TCSA function across the zones (@fig-gis-tcsa-sites).
+As predicted, TCSA values in inland areas, such as vegetation zone CT, show higher variability.
+Based on these results, we assume that stemmed points performed a wider range of tasks in low-productivity patches.
+Both temporal and spatial analyses show that stemmed points were likely more adapted for particular environments, rather than specific climate contexts.
 @prates2022changes claimed that fishtail projectile points in South America was used for megafauna hunting and contributed to their extinction by proving strong relationship between the temporal density and spatial distribution of megafauna and the projectile points.
-Our results support this claim by showing the dense distribution of stemmed point in certain environment.
-
-Our models premise that the choice of weapon-tip types reflects the socio-environmental circumstances.
-To examine the possible connections between the roles of stemmed points and environmental change or population dynamics during the Late Paleolithic period, we compared the temporal and spatial patterns of TCSA to results of simulated Mean Annual Temperature (MAT) and Summed probability distribution (SPD) from @park2022did.
-The simulated temperature distribution in south Korea during the Late Paleolithic period (Figure 9: A from @park2022did) seems similar to the modern eco-regional zones that we used in this research (@fig-gis-tcsa-sites panel A & D).
-Figure 9: C from @park2022did shows the slow decrease of temperature until the end of LGM, which could explain our results of temporal pattern, meaning stemmed points were used during the cooler environment with a temperature between 4 and 6 ℃.
-Results of SPD (Figure 10 from @park2022did) show several peaks, as indicated by the positive deviations from the three null models (shaded in red), and downs (shaded in blue), but no drastic growth or decline during our research period, indicating no major impact from population dynamics on stemmed points.
-
-Overall, our temporal and spatial patterns of TCSA show a wide TCSA range throughout the Late Paleolithic period and between eco-regional zones.
-Especially, we confirmed the widest TCSA range in a single site, Suyanggae (@fig-tcsa-per-sites).
-As a consequence, our results are consistent with our second scenario in which stemmed points are described as multifunctional tools, suggesting that people created stemmed points in response to unexpected or varying circumstances in their habitats, similar to Clovis points (@eren2022plains).
+We similarly found that stemmed points are more densly distributed in certain environments in South Korea, perhaps also relating to the distribution of fauna.
+
+Overall, our temporal and spatial patterns of TCSA show a wide range of TCSA values throughout the Late Paleolithic period and between eco-regional zones.
+We confirmed the widest TCSA range in a single site, Suyanggae (@fig-tcsa-per-sites).
+As a consequence, our results are consistent with our second scenario in which stemmed points are described as multifunctional tools, suggesting that people created stemmed points in response to unexpected or varying circumstances in their habitats, similar to how @eren2022plains explain Clovis points.
 

### 2023-02-17 analysis/paper/paper.qmd
Edited by temporal pattern results & tried CV

@@
 
-After the first discovery at the Seokjangri site at the 1960s, more than 450 stemmed points have been found in over 30 sites throughout South Korea [@Sohn_1967; @lee2019were; @Chong_2021].
+After the first discovery at the Seokjangri site in the 1960s, more than 450 stemmed points have been found at over 30 sites throughout South Korea [@Sohn_1967; @lee2019were; @Chong_2021].
 Most sites contain only a few stemmed points and only a few sites have many more, such as Jingeuneul, Suyanggae (n = 55), and Yongsandong (n = 38) [@Kim_2017].
 Among these stemmed points, we selected those that retained their widest and thickest parts.
-We selected stemmed points that were found during field surveys as well as those were from sites that were never dated but were associated with other Late Palaeolithic artifacts.
+We included stemmed points discovered during field surveys as well as those found at sites that were never dated but were associated with other Late Palaeolithic artifacts.
 Applying these sampling criteria resulted in a sample of `r total_sp` stemmed points from `r total_ass` assemblages unearthed from `r total_site` sites spanning the period 44-10 ka (@fig-map).
@@
 
-We distinguished between multiple assemblages at a single site where numerous excavations have been conducted in different locations at the site, and by different institutions under the same site name.
+We distinguished between multiple assemblages at a single site where numerous excavations have taken place in different locations at the site, and by different institutions under the same site name.
 For example, Suyanggae site, a registered Korean National Heritage site, has been excavated more than ten times since 1980 by the local university museum and later by archaeological research institutions.
@@
 As a result, in this research we used four assemblages from Suyanggae.
-We separated one assemblage among the four by using a different site name, Hajinri, following the convention established the excavators of that location.
-Hajinri is the sixth excavation location at Suyanggae, which is 3.5 kilometers apart from the other areas and dated to much earlier (around 42-30 ka) than the other assemblages (around 31-15 ka).
-According to the excavation reports for Hajinri, stemmed points first appeared there around 42 ka with the earliest blade and blade cores [@Lee_Woo_Lee_An_Yun_Park_Otani_Kim_Kim_Han_et_al_2018].
+We separated one assemblage from the four by using a different site name, Hajinri, following the convention established by the excavators at that location.
+Hajinri is the sixth excavation location at Suyanggae, which is 3.5 kilometers apart from the other areas and dated much earlier (around 42-30 ka) than the other assemblages (around 31-15 ka).
+According to the excavation reports for Hajinri, stemmed points first appeared there around 42 ka with the earliest blades and blade cores [@Lee_Woo_Lee_An_Yun_Park_Otani_Kim_Kim_Han_et_al_2018].
 While we include data from Hajinri here, we consider this assemblage as an unusual outlier because the stemmed points and blades from Hajinri are highly standardized and refined, which are only found at other sites much later in time.
-We have more confidence in finds from Bonggok, which has the second-earliest dates (41.5 ka) in our collection, as the first appearance of stemmed points.
+We have more confidence in finds from Bonggok, which has the second-earliest dates (around 41.5 ka) in our collection, as the first appearance of stemmed points.
 Bonggok includes blades or elongate flakes, but without any accompanying blade cores [@park2022did].
@@
 
-To answer our research questions about the likely uses of stemmed points, we calculated TCSA values for the stemmed points in our sample, and explored the interaction of TCSA values with raw materials and artifact size.
-To examine the temporal patterns in TCSA values we investigated radiocarbon ages associated with the points.
-We separated the research time period into three phases based on the major climate event during the Late Pleistocene, Last Glacial Maximum (LGM), and examined the distribution of TCSA values before, during and after.
+To answer our research questions about the likely uses for stemmed points, we calculated TCSA values for the stemmed points in our sample.
+We also explored the interaction of TCSA values with raw materials and artifact size, which we used weight as a proxy.
+Because the shape of stone artifacts is highly influenced by the raw materials used [@mcpherron2014experimental].
+It is also assumed that raw materials are highly correlated with tool size.
+We investigated radiocarbon ages associated with the points in order to determine temporal patterns in TCSA values.
+We separated the research time period into three phases based on the major climate event during the Late Pleistocene, the Last Glacial Maximum (LGM), and examined the distribution of TCSA values before, during, and after the LGM.
 We explored the relationship between TCSA values and the location of assemblages by comparing the distributions of TCSA values across ecological and vegetation zones.
@@
 
-Selective use of raw material is a key characteristic of the technological transition of the Korean Late Paleolithic.
+Selective use of raw materials is a key characteristic of the technological transition of the Korean Late Paleolithic.
 Prior to the Late Paleolithic, people mostly used quartzite and vein quartz for stone artifacts.
-Then, finer grained materials were newly added to assemblages for producing the newly introduced tools [@seong2004quartzite].
+Then, finer grained materials were added to assemblages for producing the newly introduced tools [@seong2004quartzite].
 We analyzed TCSA values of `r n_raw` artifacts with raw material information to examine the interaction of raw material types and TCSA values.
@@
 Different sizes of stone artifacts can constrain or enable different functions.
-Overall size of stone points has been used as a potential proxy to distinguish different armature types [@thomas1978arrowheads; @sahle2019assessment].
-We chose weight as a proxy for overall size of the stemmed points.
-Weight is a reliable discriminator between tools of different sizes and a descriptive attribute which can be measured rapidly and objectively [@fenenga1953weights; @shea2006origins].
+Overall size of stone points has been used as a potential proxy for identifying different types of armatures [@thomas1978arrowheads; @sahle2019assessment].
+We chose weight as a proxy for the overall size of the stemmed points.
+As a reliable discriminator between tools of different sizes as well as a descriptive attribute, weight can be measured rapidly and objectively [@fenenga1953weights; @shea2006origins].
 We then explored the relationship between weight and raw materials on TCSA values.
-Excluding points that we were not able to directly measure or obtain the records of their weight, we explored the relationship between TCSA and weight for `r n_size` artifacts.
+Excluding points that we were not able to directly measure or obtain records of their weight, we explored the relationship between TCSA and weight for `r n_size` artifacts.
 
@@
 
-We used radiocarbon ages to explore the temporal patterns of the likely uses of stemmed points.
+We used radiocarbon ages to investigate the temporal patterns of the likely uses of stemmed points.
 After excluding assemblages that have no radiocarbon dates, we used `r n_sites_with_radiocarbon_ages` assemblages dated from 45ka to 14.8ka to explore variation in TCSA over time.
@@
 
-We summarized the distribution of TCSA values for each assemblage to examine spatial patterns in TCSA values.
-Among `r n_sites` sites, we combined those that contain fewer than five stemmed points and named them as "Other".
-We explored possible effect of environmental variation on TCSA values by grouping stemmed points by the eco-regional zones that they were found in, and comparing the distributions of TCSA values across the different zones.
-We compared distributions across four vegetation zones, and also across 16 zones of different geographical characteristics.
-
-For the vegetation zones we used @yi2010vegetation's divisions of South Korea into three zones: Central Temperate Zone (CT), South Temperate Zone (ST), and Subtropical-warm Temperate Zone (SWT).
+We summarized the distribution of TCSA values for each assemblage to analyze spatial patterns in TCSA values.
+Among `r n_sites` sites, we combined those that contain fewer than five stemmed points and named them "Other".
+We explored the possible effect of environmental variation on TCSA values by grouping stemmed points by the eco-regional zones that they were found in, and comparing the distributions of TCSA values across the different zones.
+We compared the distributions across four vegetation zones and 16 zones that differed in terms of their geographical characteristics.
+
+The vegetation zones are based on @yi2010vegetation's classification of South Korea into three zones: Central Temperate Zone (CT), South Temperate Zone (ST), and Subtropical-warm Temperate Zone (SWT).
 These divisions are based on @yim1975distribution's forest vegetation map, defined by recent temperature and precipitation values.
 
-We also explored spatial patterns using 16 geographical zones based geographical characteristics such as inland, coastal, major rivers, island, and major mountain ranges in addition to ecological information including temperature and precipitation [@lee2008relationship].
+We also explored spatial patterns using 16 geographical zones based on geographical characteristics including inland, coastal areas, major rivers, islands, and major mountain ranges in addition to ecological information including temperature and precipitation [@lee2008relationship].
 The sites in our sample occur in 14 of these zones: Imjin river basin (IRB), Metropolitan (MP), Central inland (CI), Kangwon coastal (KC), Choongnam coastal (CC), Southwestern inland (SWI), Upper Nagdong river basin (UNRB), WoolYoung coastal (WYC), Western Cholla (WC), Southern mountain (SM), Southeastern inland (SEI), Hyungsan Taewha coastal (HTC), Western south coastal (WSC), and Eastern south coastal (ESC).
-We then located individual assemblages on those two maps using longitudes and latitudes and explore TCSA values for each zone.
-We recognize that modern eco-regional maps may not fully resemble the Pleistocene landscape in which the points were made and used, however these are the only maps of eco-regional zones that are currently available.
+Then, we used longitudes and latitudes to locate individual assemblages and examined TCSA values for each zone.
+We fully acknowledge that modern eco-regional maps may not fully resemble the Pleistocene landscape in which the points were made and used, however these are the only maps of eco-regional zones that are currently available.
 
@@
 
-The process of the introduction of new technologies can vary depending on the environmental and social contexts of the transmission of the manufacturing techniques for the new tools.
-Moreover, the new tools themselves can alter the typical contexts of tool use.
-For example, the introduction of the bow and arrow in American enabled individuals to work without hunting in teams and provided opportunities for non-elite hunters to produce their own subsistence or pursue individual wealth [@bettinger2013effects; @angelbeck2014faustian; @rorabaugh2015timing].
+The process of introducing novel technologies can vary based on the social and environmental context in which the transmission of manufacturing techniques for the new tools takes place.
+Furthermore, the new tools themselves can alter the typical contexts of tool use.
+In America, for example, the advent of bows and arrows provided non-elite hunters with the opportunity to produce their own subsistence or pursue individual wealth without the necessity of hunting in teams [@bettinger2013effects; @angelbeck2014faustian; @rorabaugh2015timing].
 Taking an evolutionary approach, we assume that, given an opportunity to explore alternative technologies, human groups carefully selected a specific stone tool technology based on its advantages over other alternatives, according to their performance in a variety of domains, such as physical and social functions [@lombard2022tip].
 Thus, the selection of weapon-tip types is likely to reflect the socio-environmental circumstances that people encountered and managed.
-For example, @eren2022plains used TCSA to compare the morphological variance of Clovis and Folsom points and claimed that Clovis points were more variable in shape than Folsom points because Clovis foragers were exposed to largely unfamiliar landscape.
+In one example, @eren2022plains compared the morphological variance of Clovis and Folsom points and claimed that Clovis points were more variable in shape than Folsom points because Clovis foragers were exposed to largely unfamiliar landscapes.
 Clovis points were used as multifunctional tools that performed a wider range of tasks, including cutting and sawing.
-Folsom points, on the other hand, show a narrower range of variation, indicating a they were more likely used for specific tasks.
+Folsom points, on the other hand, show a narrower range of variation, indicating they were more likely used for specific tasks.
 
@@
 
-We predict a temporal pattern of more variable TCSA values at the first appearance of stemmed points, suggesting that people were using the tools as part of an adaptation to moving into unfamiliar landscapes, with unpredictable variation in patch productivity and travel times.
-We predict a further increase in the variability of TCSA values in the LGM as lower temperatures alter the distribution of resources and lower the predictability of resource encounters.
-After the LGM we predict a reduction in the range of TCSA values, with higher temperatures and increased bioproductivity.
-
-We predict that TCSA values vary across vegetation types and geographical zones in response to heterogeneously distributed food resources.
-In harsh environments with lower patch productivity, similar to the case of LGM duration, we predict a further increase in the variability of TCSA values.
-On the other had, we predict a reduction in the range of TCSA values in affluent patches with predictable types of prey.
+We predict a temporal pattern of more variable TCSA values at the first appearance of stemmed points, suggesting that the tools were being used as part of an adaptation to moving into unfamiliar landscapes, with unpredictable variations in patch productivity and travel times.
+The variability of TCSA values in the LGM is predicted to increase as lower temperatures alter the distribution of resources and reduce the predictability of resource encounters.
+After the LGM we anticipate a reduction in the range of TCSA values, with higher temperatures and increased bioproductivity.
+
+In response to heterogeneously distributed food resources, TCSA values are predicted to vary across vegetation types and geographical zones.
+In harsh environments with lower patch productivity, similar to the case of LGM duration, we expect to observe a further increase in the variability of TCSA values.
+On the other hand, we predict a reduction in the range of TCSA values in affluent patches with predictable types of prey.
 In their study of hunter-gatherer mobility strategies, @hamilton2016ecological show that resources are abundant and predictable along coasts or lake shores, which is highly dependent on temperature and precipitation and thus hunter-gatherers often become effectively sedentary.
-Our prediction is that the range of TCSA values will be narrower in coastal areas such as vegetation zones ST and SW and geographical zone ESC, compared to inland areas such as vegetation zone CT and geographical zone CI.
-
-The entire R code [@core2021team] and data files used for all the analysis and visualizations contained in this paper is openly available at https://doi.org/10.17605/osf.io/dqna8 to enable re-use of materials and improve reproducibility and transparency [@marwick2017computational].
+Therefore, our prediction is that the range of TCSA values will be narrower in coastal areas such as vegetation zones ST and SW and geographical zone ESC, compared to inland areas such as vegetation zone CT and geographical zone CI.
+
+The entire R code [@core2021team] and data files used for all the analyses and visualizations contained in this paper are openly available at https://doi.org/10.17605/osf.io/dqna8 to enable re-use of materials and improve reproducibility and transparency [@marwick2017computational].
 All of the figures, tables, statistical test results presented here can be independently reproduced with the code and data in this repository.
-The code is released under the MIT license, the data as CC-0, and figures as CC-BY, to enable maximum re-use.
+The code is released under the MIT license, the data as CC-0, and the figures as CC-BY, to enable maximum re-use.
 
@@
 @fig-tcsa-all-sp shows TCSA values for all stemmed points in our sample, with shaded rectangles to assist in the interpretation of their likely uses.
-Overall we see a wide variation of TCSA values.
-The sample mean TCSA is `r round(mean(tcsa$min_tcsa), digits=1)`, and standard deviation of TCSA is `r round(sd(tcsa$min_tcsa), digits=1)`.
-According to the TCSA ranges presented in @tbl-tcsa-ranges, Korean stemmed points are overall mostly in the categories of javelin tips and stabbing spear tips, with smaller numbers as dart tips and un-poisoned arrow tips.
+Overall we see a wide variation in TCSA values.
+The sample mean of TCSA is `r round(mean(tcsa$min_tcsa), digits=1)`, and the standard deviation of TCSA is `r round(sd(tcsa$min_tcsa), digits=1)`.
+According to the TCSA ranges presented in @tbl-tcsa-ranges, Korean stemmed points are mostly found in the categories of javelin tips and stabbing spear tips, with smaller numbers as dart tips and un-poisoned arrow tips.
 Among the weapon-tip types, only poison arrow tips appear to be absent from these Korean assemblages.
@@
 @fig-tcsa-raw-materials shows that about half of the stemmed points were made from shale and its TCSA range is widely distributed.
-Acidic volcanic rocks tend to have skewed lower TCSA values.
-The other raw materials including hornfels, rhyolite, and tuff show a wide distribution, similar to shale.
-The category of "Other" raw materials include porphyry, trachyte, felsite, chert, quartz, quartzite, granite, mudstone, and unidentified rocks.
+The TCSA values of acidic volcanic rocks are commonly skewed lower.
+Other raw materials including hornfels, rhyolite, and tuff show a wide distribution, similar to shale.
+The category of "Other" raw materials includes porphyry, trachyte, felsite, chert, quartz, quartzite, granite, mudstone, and unidentified rocks.
 Overall, there is no clear pattern of TCSA values among different raw materials.
@@
 
+############ test CV ########### to present
+library(raster)
+# cv for LGM for example
+data_lgm <- tcsa_site_ass_mis %>% 
+  filter(mis_stage=='LGM')
+
+cv_lgm <- cv(data_lgm$min_tcsa)
+
 ```
@@
 Assemblages discarded before LGM tend to possess more stemmed points, such as Suyanggae.
-These assemblages have a wide range of TCSA values, indicating multiple likely usages of stemmed points.
+These assemblages have a wide range of TCSA values, indicating multiple likely uses for stemmed points.
 Overall, there is no clear pattern in these assemblages over time.
-The TCSA range varies depending on the assemblages.
-
-Panel B of @fig-tcsa-radio-carbon shows the distribution of TSCA values from assemblages dating to before the LGM, during the LGM, and after, to explore the impact of climate change on the probably function of the artifacts.
+The TCSA range varies depending on the assemblage.
+
+In panel B, @fig-tcsa-radio-carbon shows the distribution of TSCA values from assemblages dating to before, during, and after the LGM, in order to explore the impact of climate change on the likely uses of the artifacts.
 Our results show that stemmed points were made predominantly before LGM and only a few left after LGM.
@@
 
-Considering the importance of stemmed points for the technological transition during the Late Paleolithic in Korea, our research examines the likely uses of stemmed points by asking three research questions: What were the best-fit ballistic probabilities for the stemmed points if they were hafted as weapon tips?
+Considering the importance of stemmed points for the technological transition during the Late Paleolithic in Korea, our research examines the likely uses for stemmed points by asking three research questions: What were the best-fit ballistic probabilities for the stemmed points if they were hafted as weapon tips?
 How diverse were their likely uses?

### 2023-02-17 analysis/paper/paper.qmd
Edit done by discussion

@@
 ```
-
 In order to explore spatial patterns and the relationship between the role of stemmed points and the environment, we located individual assemblages on maps that depict eco-regional zones.
 Our results show that stemmed points are only located in certain zones, Central Temperature Zone (CT) of the vegetation map or Central inland (CI) (@fig-gis-tcsa-sites panel A & D) of the eco-region map.
-We observed that there are statistically significant differences in the distributions of TCSA values between some zones (Vegetation zones: `r eco_lgm_test_pretty`, Ecoregion zones: `r for_lgm_test_pretty`).
+We observed that there are statistically significant differences in the distribution of TCSA values between some zones (Vegetation zones: `r eco_lgm_test_pretty`, Ecoregion zones: `r for_lgm_test_pretty`).
 The Tukey's HSD results show that TCSA values from the South Temperature Zone (ST) zone are significantly different from the other two vegetation zones (@fig-gis-tcsa-sites panel C).
@@
 The Tukey HSD pairwise differences show no pairs of eco-regions with significant differences in TCSA values, likely due to the small number of points in several zones (@fig-gis-tcsa-sites panel F).
-Overall, our results indicate no clear patterns in the distribution of TCSA values across the vegetation and eco-regional zones.
+Overall, our results indicate no clear patterns in the distribution of TCSA values across vegetation and eco-regional zones.
 
@@
 
-To investigate the role that stemmed points played in the technological transition during the Korean Late Paleolithic, we computed the TCSA metric to infer likely uses by comparing the results to the TCSA range from other archaeological and ethnographic cases (@tbl-tcsa-ranges).
-Our main questions were: What were the best-fit ballistic probabilities for the stemmed points if they were hafted as weapon tips?
+By comparing the results with the TCSA range from other archaeological and ethnographic cases (@tbl-tcsa-ranges), we were able to determine what role stemmed points played in the technological transition during the Korean Late Paleolithic.
+Our main questions were: What were the best-fit ballistic probabilities for stemmed points if they were hafted as weapon tips?
 How diverse were their likely uses?
@@
 
-Our results show that javelin tips and stabbing spear tips are the most probable ballistic uses for stemmed points.
+Our results indicate that javelin tips and stabbing spear tips are the most probable ballistic uses for stemmed points.
 There are a few stemmed points in our sample that may have been used as dart tips and un-poisoned arrow tips.
 None have TCSA values in the range of poison arrow tips (@fig-tcsa-all-sp).
-@lee2019were computed TCSA for 10 stemmed points from Jingeuneul site, which we were not able to include in this research due to the unavailability of their data, and concluded that their TCSA values fall within the range of North American dart tips and arrowheads.
-In general, we found a wider range of TCSA in our sample, indicating that stemmed points might play diverse roles in foraging toolkits.
+In their study, @lee2019 calculated TCSA values for ten stemmed points from Jingeuneul site, which were not included in the research due to data availability issues, and concluded that these values are within the range of North American dart tips and arrowheads.
+In general, we found a wide range of TCSA in our sample, suggesting that stemmed points might play diverse roles in foraging toolkits.
 
@@
 We observed no clear pattern for raw materials (@fig-tcsa-raw-materials) but weight (@fig-artifact-size-cluster), though weight is correlated with raw materials (@fig-tcsa-size).
-We speculate this might be due to the higher availability of raw materials such as shale on hornfels on the landscape.
-Therefore we could assume that raw materials also associated with TCSA value.
+We speculate that this might be due to the higher availability of raw materials, such as shale and hornfels on the landscape.
+Therefore, we could assume that raw materials also associated with TCSA value.
 We found that different clusters of weight are matched with different types of weapon tips (@fig-artifact-size-cluster).
 
-Our results shows that TCSA values vary between assemblages with few discernible temporal patterns in function (@fig-tcsa-radio-carbon panel A).
+Our results show that TCSA values vary between assemblages with few discernible temporal patterns in function (@fig-tcsa-radio-carbon panel A).
 Stemmed points were used most frequently before the LGM (@fig-tcsa-radio-carbon panel B).
@@
 
-We found that stemmed points are most frequently located in a small number of eco-regional zones such as the Central Temperature zone with few clears pattern in TCSA function across the zones (@fig-gis-tcsa-sites).
+We found that stemmed points are primarily located in a small number of eco-regional zones such as the Central Temperature zone with few clear patterns in TCSA function across the zones (@fig-gis-tcsa-sites).
 As predicted, TCSA values in inland areas, such as vegetation zone CT, show higher variability.
-Based on these results, we assume that stemmed points performed a wider range of tasks in low-productivity patches.
-Both temporal and spatial analyses show that stemmed points were likely more adapted for particular environments, rather than specific climate contexts.
-@prates2022changes claimed that fishtail projectile points in South America was used for megafauna hunting and contributed to their extinction by proving strong relationship between the temporal density and spatial distribution of megafauna and the projectile points.
-We similarly found that stemmed points are more densly distributed in certain environments in South Korea, perhaps also relating to the distribution of fauna.
-
-Overall, our temporal and spatial patterns of TCSA show a wide range of TCSA values throughout the Late Paleolithic period and between eco-regional zones.
-We confirmed the widest TCSA range in a single site, Suyanggae (@fig-tcsa-per-sites).
-As a consequence, our results are consistent with our second scenario in which stemmed points are described as multifunctional tools, suggesting that people created stemmed points in response to unexpected or varying circumstances in their habitats, similar to how @eren2022plains explain Clovis points.
+These results suggest that stemmed points performed a wider range of tasks in low-productivity patches.
+Both temporal and spatial analyses show that stemmed points were more likely to have adapted to particular environments, rather than specific climate contexts.
+@prates2022changes claimed that fishtail projectile points in South America were used to hunt megafauna and contributed to their extinction by demonstrating a strong correlation between the spatial and temporal distribution of megafauna and the projectile points.
+We similarly found that stemmed points are more densely distributed in certain environments in South Korea, perhaps also related to faunal distribution.
+
+Overall, our results show a wide range of TCSA values throughout the Late Paleolithic period and between eco-regional zones.
+We confirmed the widest TCSA range at a single site, Suyanggae (@fig-tcsa-per-sites).
+Our foundings are therefore consistent with our second scenario, in which stemmed points are described as multifunctional tools, suggesting that people created stemmed points as a response to unexpected or varying circumstances in their habitats, similar to how @eren2022plains explain Clovis points.
 

### 2023-02-17 analysis/paper/paper.qmd
Finish editing the text

@@
 We confirmed the widest TCSA range at a single site, Suyanggae (@fig-tcsa-per-sites).
-Our foundings are therefore consistent with our second scenario, in which stemmed points are described as multifunctional tools, suggesting that people created stemmed points as a response to unexpected or varying circumstances in their habitats, similar to how @eren2022plains explain Clovis points.
+Our findings are therefore consistent with our second scenario, in which stemmed points are described as multifunctional tools, suggesting that people created stemmed points as a response to unexpected or varying circumstances in their habitats, similar to how @eren2022plains explain Clovis points.
 
@@
 Then we examined the TCSA values with other variables including raw materials, weight, radiocarbon dates, and site locations.
-Drawing on evolutionary theory, we premised that the way of using stemmed points is likely to reflect the socio-environmental circumstances that people encountered and managed.
+Drawing on evolutionary theory, we premised that the way of using stemmed points are likely to reflect the socio-environmental circumstances that people encountered and managed.
 We examined the possible impact of LGM on stemmed points and their distribution across eco-regional zones.
@@
 In general, though, we noted a wide range and also differing distributions of TCSA values in each assemblage.
-Therefore, we conclude that stemmed points had diverse roles during the Late Paleolithic.
-We conclude that people frequently encountered unfamiliar situations and used stemmed points as a multifunctional tool to carry out multiple tasks.
+Therefore, we conclude that stemmed points served diverse roles during the Late Paleolithic.
+People might frequently encounter unfamiliar situations and use stemmed points as a multifunctional tool to carry out multiple tasks.
 We are aware that discriminating the likely use of small numbers of projectile points could be arbitrary [@erlandson2014darts].
 Since TCSA covers the critical elements of projectiles, flight and penetration dynamics (i.e.
-increased or decrease by shape of tip and cross section), we still consider it a useful metric for discriminating between weapon tip tools [@hughes1998getting; @sitton2020tip; @lombard2021variation].
+increase or decrease by shape of tip and cross section), we still consider it a useful metric for discriminating between weapon tip tools [@hughes1998getting; @sitton2020tip; @lombard2021variation].
 As part of our future research, we will conduct use-wear analyses of stemmed points and experimental research in an effort to elucidate the detailed function of stemmed points and technological links between hunter-gatherer groups.

### 2023-02-18 analysis/paper/paper.qmd
Done with dealing with BM's comments except for adding CV and ANOVA

@@
 In their study, @lee2019 calculated TCSA values for ten stemmed points from Jingeuneul site, which were not included in the research due to data availability issues, and concluded that these values are within the range of North American dart tips and arrowheads.
-In general, we found a wide range of TCSA in our sample, suggesting that stemmed points might play diverse roles in foraging toolkits.
+We found a wide range of TCSA in our sample covering the range of dart tips and un-poisoned arrow tips, which are assumed to be equivalent to arrowheads.
+In general, the wide range of TCSA suggest that stemmed points might play diverse roles in foraging toolkits.
 
@@
 @eren2022plains explain that smaller TCSA could be the result of pursuing production economy and transport efficiency.
-We explored the relationship between TCSA and raw material types and weight in order to examine these factors.
-We observed no clear pattern for raw materials (@fig-tcsa-raw-materials) but weight (@fig-artifact-size-cluster), though weight is correlated with raw materials (@fig-tcsa-size).
+We explored the relationship between TCSA, and weight and raw material types in order to examine these factors.
+We found that different clusters of weight are matched with different types of weapon tips (@fig-artifact-size-cluster).
+There was no clear pattern for raw materials (@fig-tcsa-raw-materials).
+However, combined with weight, we observed a positive relationship between raw material type and TCSA values (@fig-tcsa-size).
 We speculate that this might be due to the higher availability of raw materials, such as shale and hornfels on the landscape.
 Therefore, we could assume that raw materials also associated with TCSA value.
-We found that different clusters of weight are matched with different types of weapon tips (@fig-artifact-size-cluster).
 

### 2023-02-24 analysis/paper/paper.qmd
Edited based on friend's proofreading and added CV and more ANOVA tests

@@
 What are the temporal and spatial patterns of stemmed point use?
-We measured tip cross-sectional area (TCSA) to discriminate different likely use classes of projectile points, for example, as poisoned arrow tips or as stabbing spears.
-We analyzed TCSA with other variables including raw materials, weight, radiocarbon dates and locations.
+We measured tip cross-sectional area (TCSA) to distinguish different likely use classes of projectile points, for example, as poisoned arrow tips or as stabbing spears.
+We analyzed TCSA with other variables, including raw materials, weight, radiocarbon dates and locations.
 Our results show that the stemmed points could have served as javelin tips and stabbing spear tips, with smaller numbers as dart tips and un-poisoned arrow tips.
@@
 Some sites show a wide range of TCSA values that represent multi-purpose usage of stemmed points.
-The temporal pattern of TCSA values is one of little change throughout the Late Paleolithic period but points were predominantly produced before the Last Glacial Maximum (LGM).
+The temporal pattern of TCSA values is one of little change throughout the Late Paleolithic period, but points were predominantly produced before the Last Glacial Maximum (LGM).
 We observed that stemmed points were mostly located in certain ecoregions in Korea but no clear spatial pattern was confirmed.
@@
 
-The introduction of new stone artifact technologies marks a major transition in the Korean Paleolithic from the Early to the Late Paleolithic periods.
+The introduction of new stone artifact technologies marked a major transition in the Korean Paleolithic from the Early to the Late Paleolithic periods.
 The transition includes blade technology, stemmed points, end scrapers, burins, denticulates, etc.
@@
 We use the tip cross-sectional area (TCSA) metric to infer potential variability in weapon-use strategies, and compare the results to other archaeological and ethnographic assemblages based on the idea that different shapes correspond to different weapon types [@lombard2021variation].
-We then explore the relationship between the TCSA range and raw materials, artifact size and discard location, and how these change over time.
+We then explore the relationship between the TCSA range and raw materials, artifact size and discard location, and how these changed over time.
 Our main questions are: What were the best-fit ballistic probabilities for the stemmed points if they were hafted as weapon tips?
@@
 Slight retouch is typically performed on the proximal end to shape an acute tip and on the distal end to make a stem, which connects to a wooden shaft.
-Elsewhere in the world these types of artifacts are often called 'tanged points', but we prefer referring to them as 'stemmed points' here to distinguish them from Bronze Age stone projectile points which have long been called 'tanged points' in Korea [@park2022did].
+Elsewhere in the world these types of artifacts are often called "tanged points", but we prefer referring to them as "stemmed points" here to distinguish them from Bronze Age stone projectile points which have long been called "tanged points" in Korea [@park2022did].
 Understanding the appearance of stemmed points is relevant to general questions about the origins of projectile technology, the technological transition of the Late Paleolithic, and relationships between Korea and adjacent regions in East Asia, such as Japan.
 
-Stemmed points are the first composite tools in the Korean Paleolithic that required two different parts to be one complete tool, a stone point and a shaft, presumably made out of wood [@seong2008tanged].
+Stemmed points are the first composite tools in the Korean Paleolithic that require two different parts to be form one complete tool, a stone point and a shaft, presumably made out of wood [@seong2008tanged].
 Using blades as the blank for the point enables mass production of this composite tool and its shape can become more standardized [@lee2015characteristics; @park2022did].
@@
 
-Since stemmed points represent the beginning of the Korean Late Paleolithic, investigating their origins is critical for understanding the technological transition from the Early Paleolithic, modern human dispersals into the region, and claims for the existence of the 'Middle' Paleolithic in Korea [@bae2010origin; @bae2017late; @norton2009evolution; @seong2016eastern].
-Debate about explaining the origin of stemmed points which can be summarized into two competing models: a 'heterogenic' migration [@bae2010origin], and an *in situ* evolution model [@seong2009emergence].
-The migration model claims that the new blade industry including stemmed points, and the earlier coarse flake tradition including large cores, polyhedrons, choppers, and even handaxes, came from different origins (i.e.
-1: North route: Siberia, Mongolia, or other regions of northeast China, and 2: South route: southern China) as the result of a continuing influx of modern human migration from two routes.
+Since stemmed points represent the beginning of the Korean Late Paleolithic, investigating their origins is critical to understanding the technological transition from the Early Paleolithic, modern human dispersals into the region, and claims for the existence of the 'Middle' Paleolithic in Korea [@bae2010origin; @bae2017late; @norton2009evolution; @seong2016eastern].
+The debate around explaining the origin of stemmed points which can be summarized into two competing models: (1) a 'heterogenic' migration [@bae2010origin], and (2) an *in situ* evolution model [@seong2009emergence].
+The migration model claims that the new blade industry - including stemmed points - and the earlier coarse flake tradition - including large cores, polyhedrons, choppers, and even handaxes - came from different origins (i.e.
+(1) North route: Siberia, Mongolia, or other regions of northeast China, and (2) South route: southern China) as the result of a continuing influx of modern human migration from two routes.
 The *in situ* model argues that stemmed points and other Late Paleolithic technologies, such as blade industries, autonomously emerged in the south of the Korean peninsula, with no apparent external influence.
@@
 @park2022did examined mobility and site occupation patterns by applying concepts of human behavioral ecology to lithic assemblages.
-The study found that forager groups using stemmed points may have been associated with the occupation of marginal or extreme environments, compared to groups with no stemmed points.
+The study found that forager groups using stemmed points may have been associated with the occupation of marginal or extreme environments, in contrast to groups with no stemmed points.
 Also, stemmed points were more frequently associated with expedient technologies, indicating residential and less mobile behaviors.
-@Chong_2021 claims that the morphological variation of the stemmed points along with tool types in lithic assemblages, assemblage size, use of raw materials, and types of blanks could represent specific characteristics of occupation such as a "limited activity station" and a "residential base camp." For example, stemmed points with high morphological variations in tool size, shape of edge, degrees of damage, and types of edge retouching from Yongsandong site may indicate that the site was used for specific or limited activities such as hunting [@kim2004yongsan; @seong2015diversity; @Chong_2021].
+@Chong_2021 claims that the morphological variation of the stemmed points along with tool types in lithic assemblages, assemblage size, use of raw materials, and types of blanks could represent specific characteristics of occupation, such as a "limited activity station" and a "residential base camp." For example, stemmed points with high morphological variations in tool size, shape of edge, degrees of damage, and types of edge retouching from the Yongsandong site may indicate that the site was used for specific or limited activities, such as hunting [@kim2004yongsan; @seong2015diversity; @Chong_2021].
 
 The connection between stemmed points in both Korea and Japan has been studied since the late 1980s as a part of evidence for long-distance/maritime cultural interchanges or social networks [@lee2015characteristics; @chang2013human].
-Stemmed points from the Bonggok site in Korea are currently accepted as the oldest ones among Northeast Asia, dated to ca.
-41.5 ka, and made out of elongated flakes [@seong2009emergence; @seong2015diversity; @bae2017origin].
-After their first appearance in Korea, stemmed points (Hakuhensenntouki in Japanese) appeared in Kyushu, Japan during the late Marine isotope stage (MIS) 3.
-In addition to the stemmed points, there are similar artifacts found in both regions such as microblade cores, Moppule-seokgi (Kakusuijyosekki  in Japanese), backed knives, bilateral points, bifacial points, and transport of obsidian [@lee2012characteristics; @chang2013human; @lee2015characteristics; @kim2021evidence].
+Stemmed points from the Bonggok site in Korea are currently accepted as the oldest ones within Northeast Asia, dating to ca.
+41.5 ka, and are made from elongated flakes [@seong2009emergence; @seong2015diversity; @bae2017origin].
+After their first appearance in Korea, stemmed points (Hakuhensenntouki in Japanese) appeared in Kyushu, Japan during the late Marine Isotope Stage (MIS) 3.
+In addition to the stemmed points, there are similar artifacts found in both regions, such as microblade cores, Moppule-seokgi (Kakusuijyosekki  in Japanese), backed knives, bilateral points, bifacial points, and transport of obsidian [@lee2012characteristics; @chang2013human; @lee2015characteristics; @kim2021evidence].
 
@@
 #| label: fig-sp-hand-made
-#| fig-cap: "Stemmed points from Yongsandong site. 1-6: plain stemmed points, 7-8: one side denticulate stemmed points."
+#| fig-cap: "Stemmed points from Yongsandong site. 1-6: plain stemmed points, 7-8: one-sided denticulate stemmed points."
 
@@
 
-## Previous studies about the function of stemmed points
+## Previous Studies about the Function of Stemmed Points
 
@@
 Preserved wooden components of prehistoric projectile tools are too few and rare to standardize their overall scale and variability [@shea2006origins].
-@Lee_Kong_2002 claimed stemmed points should be considered more generally as 'stemmed tools' because of their diverse shapes, which could represent different functions, and the uncertainty of their complete shape.
-Other researchers propose that non-symmetrical stemmed points, with retouch on one side or denticulate blades on one side, should not be referred to as stemmed points but rather as stemmed knives, stemmed side-scrapers, stemmed end-scrapers, or stemmed burins (@fig-sp-hand-made: 7-8) [@seong2008tanged; @Lee_Jang_2011; @Kim].
+@Lee_Kong_2002 claim stemmed points should be considered more generally as 'stemmed tools' because of the uncertainty of their complete shape.
+Other researchers propose that non-symmetrical stemmed points - with retouch on one side or denticulate blades on one side - should not be referred to as stemmed points but rather as stemmed knives, stemmed side-scrapers, stemmed end-scrapers, or stemmed burins (@fig-sp-hand-made: 7-8) [@seong2008tanged; @Lee_Jang_2011; @Kim].
 
@@
 For example, at Yongsandong site, only 10% of the tools are complete, while 33% of the tips are missing.
-In the case of Jingeuneul, the percentages are 16% and 50%, respectively [Park_2013].
+In the case of Jingeuneul, the percentages are 16% and 50%, respectively [@Park_2013].
 In addition to the morphological aspect of stemmed points, investigations of a whole site and the tool composition of an assemblage suggest that stemmed points or stemmed tools could be strongly associated with hunting activities including peeling the animal skin after slaughtering or separating the bones from the flesh [@seong2008tanged; @Chong_2021].
@@
 
-Tip cross-sectional area (TCSA) of stone artifacts has been used as a ballistically relevant standard to probabilistically discriminate different likely use classes of projectile armatures such as spearthrower (a.k.a.
+Tip cross-sectional area (TCSA) of stone artifacts has been used as a ballistically relevant standard to probabilistically discriminate different likely use classes of projectile armatures, such as spear-thrower (a.k.a.
 atlatl) dart tips, un-poisoned arrow tips and large stabbing/thrusting spears [@sisk2011african; @lombard2021variation; @lombard2020tip; @lombard2021did; @o2018origins; @lombard2022standardized].
@@
 This method was first proposed by @hughes1998getting and validated by @shea2006origins through the comparison of archaeological examples with ethnographically collected samples of known use.
-@shea2009impact also applied this approach to compare projectile points from Africa, the Levant, and Europe, and claimed that projectile weapons first appeared in Africa.
+@shea2009impact also applied this approach to compare projectile points from Africa, the Levant, and Europe, claiming that projectile weapons first appeared in Africa.
 
@@
 
-## Archaeological sites
+## Archaeological Sites
 
@@
 We distinguished between multiple assemblages at a single site where numerous excavations have taken place in different locations at the site, and by different institutions under the same site name.
-For example, Suyanggae site, a registered Korean National Heritage site, has been excavated more than ten times since 1980 by the local university museum and later by archaeological research institutions.
+For example, Suyanggae site, a registered Korean National Heritage site, has been excavated more than ten times since 1980 by the local university museum, and later, by archaeological research institutions.
 There are six different excavation locations that range from a few meters apart to a few kilometers.
@@
 While we include data from Hajinri here, we consider this assemblage as an unusual outlier because the stemmed points and blades from Hajinri are highly standardized and refined, which are only found at other sites much later in time.
-We have more confidence in finds from Bonggok, which has the second-earliest dates (around 41.5 ka) in our collection, as the first appearance of stemmed points.
+We have more confidence in finds from Bonggok, which has the second-earliest dates (around 41.5 ka) in our collection as the first appearance of stemmed points.
 Bonggok includes blades or elongate flakes, but without any accompanying blade cores [@park2022did].
 
-## Investigating patterns in TCSA values
+## Investigating Patterns in TCSA Values
 
@@
 We also explored the interaction of TCSA values with raw materials and artifact size, which we used weight as a proxy.
-Because the shape of stone artifacts is highly influenced by the raw materials used [@mcpherron2014experimental].
-It is also assumed that raw materials are highly correlated with tool size.
+Because the shape of stone artifacts is highly influenced by the raw materials [@mcpherron2014experimental], it is assumed that raw materials could be highly correlated with tool size.
 We investigated radiocarbon ages associated with the points in order to determine temporal patterns in TCSA values.
 We separated the research time period into three phases based on the major climate event during the Late Pleistocene, the Last Glacial Maximum (LGM), and examined the distribution of TCSA values before, during, and after the LGM.
-We explored the relationship between TCSA values and the location of assemblages by comparing the distributions of TCSA values across ecological and vegetation zones.
+We explored the relationship between TCSA values and the location of assemblages by comparing the distribution of TCSA values across ecological and vegetation zones.
 
@@
 
-To aid with interpreting our results, we followed the TCSA ranges for different weapon-delivery systems that @lombard2022tip and @lombard2021variation created by summarizing the analysis by @wadley2008segment of backed microliths (@tbl-tcsa-ranges).
-We excluded `r n_omit` artifacts from our dataset with TCSA greater than 250 and thus outside of the range of the comparative data.
-For this comparison with weapon-delivery systems we include TCSA values from a total of `r min_n_points` stemmed points from `r n_ass` assemblages unearthed from `r n_sites` sites.
-
-### Raw materials
+To aid in interpreting our results, we followed the TCSA ranges for different weapon-delivery systems that @lombard2022tip and @lombard2021variation created by summarizing the analysis by @wadley2008segment of backed microliths (@tbl-tcsa-ranges).
+We excluded `r n_omit` artifacts from our dataset with TCSA greater than 250, which were outside of our range of the comparative data.
+For this comparison with weapon-delivery systems, we included TCSA values from a total of `r min_n_points` stemmed points from `r n_ass` assemblages unearthed from `r n_sites` sites.
+
+### Raw Materials
 
@@
 We analyzed TCSA values of `r n_raw` artifacts with raw material information to examine the interaction of raw material types and TCSA values.
-We categorized rare raw materials, which have less than 10 artifacts, as "Other".
+We categorized rare raw materials, which have less than 10 artifacts, as "Other."
 
@@
 
-### Temporal patterns
+### Temporal Patterns
 
@@
 We summarized the distribution of TCSA values for each assemblage to analyze spatial patterns in TCSA values.
-Among `r n_sites` sites, we combined those that contain fewer than five stemmed points and named them "Other".
-We explored the possible effect of environmental variation on TCSA values by grouping stemmed points by the eco-regional zones that they were found in, and comparing the distributions of TCSA values across the different zones.
+Among the `r n_sites` sites, we combined those that contained fewer than five stemmed points and named them "Other." We explored the possible effect of environmental variation on TCSA values by grouping stemmed points by the eco-regional zones that they were found in, and comparing the distributions of TCSA values across the different zones.
 We compared the distributions across four vegetation zones and 16 zones that differed in terms of their geographical characteristics.
@@
 The sites in our sample occur in 14 of these zones: Imjin river basin (IRB), Metropolitan (MP), Central inland (CI), Kangwon coastal (KC), Choongnam coastal (CC), Southwestern inland (SWI), Upper Nagdong river basin (UNRB), WoolYoung coastal (WYC), Western Cholla (WC), Southern mountain (SM), Southeastern inland (SEI), Hyungsan Taewha coastal (HTC), Western south coastal (WSC), and Eastern south coastal (ESC).
-Then, we used longitudes and latitudes to locate individual assemblages and examined TCSA values for each zone.
-We fully acknowledge that modern eco-regional maps may not fully resemble the Pleistocene landscape in which the points were made and used, however these are the only maps of eco-regional zones that are currently available.
-
-## Modelling weapon-tip type selection
+Then, we used longitude and latitude to locate individual assemblages and examined TCSA values for each zone.
+We fully acknowledge that modern eco-regional maps may not fully resemble the Pleistocene landscape in which the points were made and used; however, these are the only maps of eco-regional zones that are currently available.
+
+## Modelling Weapon-Tip Type Selection
 
@@
 Clovis points were used as multifunctional tools that performed a wider range of tasks, including cutting and sawing.
-Folsom points, on the other hand, show a narrower range of variation, indicating they were more likely used for specific tasks.
+On the other hand, Folsom points show a narrower range of variation, indicating they were more likely used for specific tasks.
 
@@
 In response to heterogeneously distributed food resources, TCSA values are predicted to vary across vegetation types and geographical zones.
-In harsh environments with lower patch productivity, similar to the case of LGM duration, we expect to observe a further increase in the variability of TCSA values.
+In harsh environments with lower patch productivity, similar to the case of the LGM duration, we expect to observe a further increase in the variability of TCSA values.
 On the other hand, we predict a reduction in the range of TCSA values in affluent patches with predictable types of prey.
@@
 
-## TCSA range of Korean stemmed points
+## TCSA Range of Korean Stemmed Points
 
@@
 #| label: fig-tcsa-all-sp
-#| fig-cap: "Distribution of TCSA values for all Korean stemmed points in the current dataset. Shaded boxes indicate TSCA ranges for various weapon types."
+#| fig-cap: "Distribution of TCSA values for all Korean stemmed points in the current dataset. Shaded boxes indicate TCSA ranges for various weapon types."
 #| fig-width:   3
@@
 
-## Variation in TCSA values by artifact size
+## Variation in TCSA Values by Artifact Size
 
@@
 
-Using weight as a size proxy we examined the relationship between size and likely use of the stemmed points inferred from TCSA values.
+Using weight as a size proxy, we examined the relationship between size and likely use of the stemmed points inferred from TCSA values.
 We conducted a univariate cluster analysis [@song2020efficient; @wang2011ckmeans] of stemmed points by weight, revealing three clusters of artifact sizes (A of @fig-artifact-size-cluster, mean = `r round(mean(tcsa_mass_no_na$'Weight(g)'), digits=1)`, SD = `r round(sd(tcsa_mass_no_na$'Weight(g)'), digits=1)`).
-Cluster 1, the smallest (lightest) artifacts, has a lower range of TCSA values, compared to Cluster 2 (B of @fig-artifact-size-cluster, mean = `r round(mean(tcsa_mass_no_na$'min_tcsa'), digits=1)`, SD = `r round(sd(tcsa_mass_no_na$'min_tcsa'), digits=1)`).
+Cluster 1, the smallest (lightest) artifacts, has a lower range of TCSA values compared to Cluster 2 (B of @fig-artifact-size-cluster, mean = `r round(mean(tcsa_mass_no_na$'min_tcsa'), digits=1)`, SD = `r round(sd(tcsa_mass_no_na$'min_tcsa'), digits=1)`).
 TCSA values for Cluster 3 are the highest, except for one artifact, which is lower than 50.
@@
 
-## Variation in TCSA values by raw material
+## Variation in TCSA Values by Raw Material
+
+```{r}
+#ANOVA test
+
+library("car")
+# devtools::install_github('achetverikov/apastats',subdir='apastats')
+library("apastats")
+p_raw_test <- Anova(lm(min_tcsa ~ as.factor(raw_materials_top), tcsa_raw))
+
+raw_p_pretty <- apastats::describe.Anova(p_raw_test, 1, 2)
+```
 
@@
 The TCSA values of acidic volcanic rocks are commonly skewed lower.
-Other raw materials including hornfels, rhyolite, and tuff show a wide distribution, similar to shale.
+Other raw materials, including hornfels, rhyolite, and tuff, show a wide distribution, similar to shale.
 The category of "Other" raw materials includes porphyry, trachyte, felsite, chert, quartz, quartzite, granite, mudstone, and unidentified rocks.
-Overall, there is no clear pattern of TCSA values among different raw materials.
+Overall, there is no clear pattern of TCSA values among different raw materials (`r raw_p_pretty`).
 
@@
            alpha = 0.2)
-```
-
-@fig-tcsa-size shows positive relationships between TCSA, raw material and artifact weight.
+
+```
+
+@fig-tcsa-size shows positive relationships between TCSA for raw material and artifact weight.
 There is a stronger correlation between TCSA and artifact weight for acidic volcanic points (i.e.
@@
 
-library("car")
-# devtools::install_github('achetverikov/apastats',subdir='apastats')
-library("apastats")
+#ANOVA test
 p_lgm_test <- Anova(lm(min_tcsa ~ as.factor(mis_stage), tcsa_site_ass_mis))
@@
 library(raster)
-# cv for LGM for example
+# cv for before 
+data_bef <- tcsa_site_ass_mis %>% 
+  filter(mis_stage=='Before')
+
+cv_bef <- round(cv(data_bef$min_tcsa),2)
+
+# cv for LGM 
 data_lgm <- tcsa_site_ass_mis %>% 
@@
 
-cv_lgm <- cv(data_lgm$min_tcsa)
-
-```
-
-## Temporal patterns of TCSA values
+cv_lgm <- round(cv(data_lgm$min_tcsa),2)
+
+# cv for after LGM 
+data_aft <- tcsa_site_ass_mis %>% 
+  filter(mis_stage=='After')
+
+cv_aft <- round(cv(data_aft$min_tcsa),2)
+
+```
+
+## Temporal Patterns of TCSA Values
 
@@
 
-In panel B, @fig-tcsa-radio-carbon shows the distribution of TSCA values from assemblages dating to before, during, and after the LGM, in order to explore the impact of climate change on the likely uses of the artifacts.
-Our results show that stemmed points were made predominantly before LGM and only a few left after LGM.
+In panel B, @fig-tcsa-radio-carbon shows the distribution of TCSA values from assemblages dating to before, during, and after the LGM in order to explore the impact of climate change on the likely uses of the artifacts.
+Our results show that stemmed points were made predominantly before LGM and only a few after LGM.
 Each category shows wide variation in LGM, which indicates diverse uses of stemmed points.
-While the median TSCA value for assemblages discarded during the LGM is higher than assemblages from earlier and later, there is no statistically significant difference in TCSA values across the three periods (`r three_time_p_pretty`).
+While the median TCSA value for assemblages discarded during the LGM is higher than assemblages from earlier and later, there is no statistically significant difference in TCSA values across the three periods (`r three_time_p_pretty`).
 
@@
 #| fig-cap: "A: Distribution of TCSA values over time.
-The light gray shade indicates Marine isotope stages (MIS) 3, the gray shade indicates MIS 2 and the darkest shade indicates the duration of LGM.
-B: Distribution of TCSA values grouped by LGM event"
+The light gray shade indicates Marine Isotope Stages (MIS) 3; the gray shade indicates MIS 2; and the darkest shade indicates the duration of LGM.
+B: Distribution of TCSA values grouped by LGM event."
 #| fig-height:  5
@@
 
-## Spatial patterns of TCSA values
-
-We computed artifact TCSA values for `r n_sites` sites that contain more than five stemmed points to observe variation between sites (@fig-tcsa-per-sites).
-Sites with fewer than five stemmed points were grouped under the category of "Other".
-Among the sites in our sample, Suyanggae has the most stemmed points and the widest range of TCSA values.
-This suggests that people made stemmed points for accomplishing a variety of tasks at Suyanggae site.
+## Spatial Patterns of TCSA Values
+
+```{r}
+#ANOVA test
+
+p_site_test <- Anova(lm(min_tcsa ~ as.factor(label_site), tcsa_site))
+
+site_p_pretty <- apastats::describe.Anova(p_site_test, 1, 2)
+```
+
+We computed artifact TCSA values for `r n_sites` sites that contain more than five stemmed points to observe variation between sites (`r site_p_pretty`) (@fig-tcsa-per-sites).
+Sites with fewer than five stemmed points were grouped under the category of "Other." Among the sites in our sample, Suyanggae has the most stemmed points and the widest range of TCSA values.
+This suggests that people made stemmed points for accomplishing a variety of tasks at the Suyanggae site.
 Nosanri, Sibuk, and Wolpyeng have a narrower range of lower values, which indicates less diverse likely uses for stemmed points.
@@
 #| label: fig-tcsa-per-sites
-#| fig-cap: "TCSA values by archaeological site."
+#| fig-cap: "TCSA values by archaeological site"
 #| fig-height:  3
@@
 We observed that there are statistically significant differences in the distribution of TCSA values between some zones (Vegetation zones: `r eco_lgm_test_pretty`, Ecoregion zones: `r for_lgm_test_pretty`).
-The Tukey's HSD results show that TCSA values from the South Temperature Zone (ST) zone are significantly different from the other two vegetation zones (@fig-gis-tcsa-sites panel C).
-Looking into this further, we see that there are only six artifacts from the ST zone, with only three of these having TSCA values above 175.
+The Tukey's HSD results show that TCSA values from the South Temperature Zone (ST) are significantly different from the other two vegetation zones (@fig-gis-tcsa-sites panel C).
+Looking into this further, we see that there are only six artifacts from the ST zone, with only three of these having TCSA values above 175.
 The other three have TCSA values lower than 100, similar to the other zones (@fig-gis-tcsa-sites panel B).
-Given the small number of artifacts with extreme TCSA values in the ST zone, we hesitate to conclude this is an archaeologically significant pattern of different TCSA values.
-The Tukey HSD pairwise differences show no pairs of eco-regions with significant differences in TCSA values, likely due to the small number of points in several zones (@fig-gis-tcsa-sites panel F).
+Given the small number of artifacts with extreme TCSA values in the ST zone, we hesitate to conclude that this is an archaeologically significant pattern of different TCSA values.
+The Tukey HSD pairwise differences show no pairs of eco-regions with significant differences in TCSA values; this was likely due to the small number of points in several zones (@fig-gis-tcsa-sites panel F).
 Overall, our results indicate no clear patterns in the distribution of TCSA values across vegetation and eco-regional zones.
@@
 #| fig-cap: "TCSA values by assemblages and ecoregions.
-A: Modern vegetation map defined by temporature and precipitation.
+A: Modern vegetation map defined by temperature and precipitation.
 The individual points indicate assemblages.
-B: TCSA distribution for vegitation zones.
+B: TCSA distribution for vegetation zones.
 The individual points indicate artifacts.
-C: Tukey test result for vegitation zones.
-D: Ecoregion map based on geographical boundaries and ecological conditions (modified the map from @lee2008relationship).
+C: Tukey test result for vegetation zones.
+D: Eco-region map based on geographical boundaries and ecological conditions (modified the map from @lee2008relationship).
 E: TCSA distribution for ecoregion zones, excluding zones with no artifacts.
-F: Tukey test result for ecoregion zones"
+F: Tukey test result for eco-region zones"
 #| fig-height:  5
@@
 None have TCSA values in the range of poison arrow tips (@fig-tcsa-all-sp).
-In their study, @lee2019 calculated TCSA values for ten stemmed points from Jingeuneul site, which were not included in the research due to data availability issues, and concluded that these values are within the range of North American dart tips and arrowheads.
+In their study, @lee2019were calculated TCSA values for ten stemmed points from Jingeuneul site, which were not included in this research due to data availability issues.
+They claimed that these values are within the range of North American dart tips and arrowheads.
 We found a wide range of TCSA in our sample covering the range of dart tips and un-poisoned arrow tips, which are assumed to be equivalent to arrowheads.
@@
 @eren2022plains explain that smaller TCSA could be the result of pursuing production economy and transport efficiency.
-We explored the relationship between TCSA, and weight and raw material types in order to examine these factors.
+We explored the relationship between TCSA for weight and raw material types in order to examine these factors.
 We found that different clusters of weight are matched with different types of weapon tips (@fig-artifact-size-cluster).
@@
 However, combined with weight, we observed a positive relationship between raw material type and TCSA values (@fig-tcsa-size).
-We speculate that this might be due to the higher availability of raw materials, such as shale and hornfels on the landscape.
-Therefore, we could assume that raw materials also associated with TCSA value.
+We speculate that this might be due to the higher availability of raw materials, such as shale and hornfels within the landscape.
+Therefore, we could assume that raw materials are also associated with TCSA value.
 
@@
 Stemmed points were used most frequently before the LGM (@fig-tcsa-radio-carbon panel B).
-While we predicted the greatest variation in TCSA values during the LGM, our results show that variability decreased during the LGM, in comparison to the previous period.
+While we predicted the greatest variation in TCSA values during the LGM, our results show that variability decreased during the LGM (CV = `r cv_lgm`), in comparison to the previous period (CV = `r cv_bef`).
 This could be explained by the overall decline in stemmed points usage.
 
-We found that stemmed points are primarily located in a small number of eco-regional zones such as the Central Temperature zone with few clear patterns in TCSA function across the zones (@fig-gis-tcsa-sites).
+We found that stemmed points are primarily located in a small number of eco-regional zones such as the Central Temperature (CT) zone with few clear patterns in TCSA function across the zones (@fig-gis-tcsa-sites).
 As predicted, TCSA values in inland areas, such as vegetation zone CT, show higher variability.
 These results suggest that stemmed points performed a wider range of tasks in low-productivity patches.
-Both temporal and spatial analyses show that stemmed points were more likely to have adapted to particular environments, rather than specific climate contexts.
+Both temporal and spatial analyses show that stemmed points were more likely to have adapted to particular environments rather than specific climate contexts.
 @prates2022changes claimed that fishtail projectile points in South America were used to hunt megafauna and contributed to their extinction by demonstrating a strong correlation between the spatial and temporal distribution of megafauna and the projectile points.
@@
 We confirmed the widest TCSA range at a single site, Suyanggae (@fig-tcsa-per-sites).
-Our findings are therefore consistent with our second scenario, in which stemmed points are described as multifunctional tools, suggesting that people created stemmed points as a response to unexpected or varying circumstances in their habitats, similar to how @eren2022plains explain Clovis points.
+Our findings are therefore consistent with our second scenario, in which stemmed points are described as multifunctional tools, suggesting that people created stemmed points as a response to unexpected or varying circumstances in their habitats, similar to how @eren2022plains explained Clovis points.
 
@@
 We calculated TCSA for a total of `r min_n_points` stemmed points from `r n_ass` assemblages unearthed from `r n_sites` sites.
-Then we examined the TCSA values with other variables including raw materials, weight, radiocarbon dates, and site locations.
-Drawing on evolutionary theory, we premised that the way of using stemmed points are likely to reflect the socio-environmental circumstances that people encountered and managed.
+Then we examined the TCSA values with other variables, including raw materials, weight, radiocarbon dates, and site locations.
+Drawing on evolutionary theory, we premised that the way of using stemmed points likely reflects the socio-environmental circumstances that people encountered and managed.
 We examined the possible impact of LGM on stemmed points and their distribution across eco-regional zones.
@@
 Therefore, we conclude that stemmed points served diverse roles during the Late Paleolithic.
-People might frequently encounter unfamiliar situations and use stemmed points as a multifunctional tool to carry out multiple tasks.
+People might have frequently encounter unfamiliar situations and use stemmed points as a multifunctional tool to carry out multiple tasks.
 We are aware that discriminating the likely use of small numbers of projectile points could be arbitrary [@erlandson2014darts].

### 2023-02-24 analysis/paper/paper.qmd
created png files for creating slide

@@
   labs(y = "TCSA")
+
+# ggsave(here::here("analysis/figures/fig-tcsa-all-sp.png"),
+#        width = 4.45,
+#        height = 4.5,
+#        units = "in")
+
 ```
@@
 
+# ggsave(here::here("analysis/figures/fig-artifact-size-cluster.png"),
+#        width = 5,
+#        height = 4,
+#        units = "in")
+
 ```
@@
            alpha = 0.2)
+
+# ggsave(here::here("analysis/figures/fig-tcsa-raw-materials.png"),
+#        width = 4,
+#        height = 3,
+#        units = "in")
 

### 2023-02-24 analysis/paper/paper.qmd
created png files & fixed bg color of the png files

@@
 # ggsave(here::here("analysis/figures/fig-artifact-size-cluster.png"),
-#        width = 5,
+#        width = 4,
 #        height = 4,
@@
 # ggsave(here::here("analysis/figures/fig-tcsa-raw-materials.png"),
-#        width = 4,
-#        height = 3,
-#        units = "in")
+#        width = 4.3,
+#        height = 2.7,
+#        units = "in",
+#        bg = "white")
 
@@
   legend.key.width = unit(2.5, 'mm'))
+
+# ggsave(here::here("analysis/figures/fig-tcsa-size.png"),
+#        width = 4,
+#        height = 3.3,
+#        units = "in")
+
 ```
@@
 
+# ggsave(here::here("analysis/figures/fig-tcsa-radio-carbon.png"),
+#        width = 4,
+#        height = 4.5,
+#        units = "in",
+#        bg = "white")
+
 ```
@@
            alpha = 0.2)
+
+# ggsave(here::here("analysis/figures/fig-tcsa-per-sites.png"),
+#        width = 4.3,
+#        height = 3,
+#        units = "in",
+#        bg = "white")
 ```
@@
 
+# ggsave(here::here("analysis/figures/fig-gis-tcsa-sitess.png"),
+#        width = 4,
+#        height = 6,
+#        units = "in",
+#        bg = "white")
+
 ```

### 2023-02-24 analysis/paper/paper.qmd
add new citation & create png file for map

@@
 Tip cross-sectional area (TCSA) of stone artifacts has been used as a ballistically relevant standard to probabilistically discriminate different likely use classes of projectile armatures, such as spear-thrower (a.k.a.
-atlatl) dart tips, un-poisoned arrow tips and large stabbing/thrusting spears [@sisk2011african; @lombard2021variation; @lombard2020tip; @lombard2021did; @o2018origins; @lombard2022standardized].
+atlatl) dart tips, un-poisoned arrow tips and large stabbing/thrusting spears [@sisk2011african; @lombard2021variation; @lombard2020tip; @lombard2021did; @o2018origins; @lombard2022standardized; @metz2023bow].
 It is critical to note that the TCSA metric alone cannot unambiguously determine artifact function; it only suggests a best-fit ballistic probability for the points if they were hafted as weapon tips.
@@
 
+# ggsave(here::here("analysis/figures/fig-map.png"),
+#        width = 4,
+#        height = 4.5,
+#        units = "in")
 ```

### 2023-03-09 analysis/paper/paper.qmd
Adding lombard's new article

@@
 Tip cross-sectional area (TCSA) of stone artifacts has been used as a ballistically relevant standard to probabilistically discriminate different likely use classes of projectile armatures, such as spear-thrower (a.k.a.
-atlatl) dart tips, un-poisoned arrow tips and large stabbing/thrusting spears [@sisk2011african; @lombard2021variation; @lombard2020tip; @lombard2021did; @o2018origins; @lombard2022standardized; @metz2023bow].
+atlatl) dart tips, un-poisoned arrow tips and large stabbing/thrusting spears [@sisk2011african; @lombard2021variation; @lombard2020tip; @lombard2021did; @o2018origins; @lombard2022standardized; @metz2023bow; @lombard2023neanderthal].
 It is critical to note that the TCSA metric alone cannot unambiguously determine artifact function; it only suggests a best-fit ballistic probability for the points if they were hafted as weapon tips.
@@
 ```
+
 In order to explore spatial patterns and the relationship between the role of stemmed points and the environment, we located individual assemblages on maps that depict eco-regional zones.

### 2023-03-10 analysis/paper/paper.qmd
stash

--- 2023-02-24 analysis/paper/paper.qmd
@@
   labs(y = "TCSA")
-
-# ggsave(here::here("analysis/figures/fig-tcsa-all-sp.png"),
-#        width = 4.45,
-#        height = 4.5,
-#        units = "in")
-
 ```
@@
 
-# ggsave(here::here("analysis/figures/fig-artifact-size-cluster.png"),
-#        width = 4,
-#        height = 4,
-#        units = "in")
-
 ```
@@
            alpha = 0.2)
-
-# ggsave(here::here("analysis/figures/fig-tcsa-raw-materials.png"),
-#        width = 4.3,
-#        height = 2.7,
-#        units = "in",
-#        bg = "white")
 
@@
   legend.key.width = unit(2.5, 'mm'))
-
-# ggsave(here::here("analysis/figures/fig-tcsa-size.png"),
-#        width = 4,
-#        height = 3.3,
-#        units = "in")
-
 ```
@@
 
-# ggsave(here::here("analysis/figures/fig-tcsa-radio-carbon.png"),
-#        width = 4,
-#        height = 4.5,
-#        units = "in",
-#        bg = "white")
-
 ```
@@
            alpha = 0.2)
-
-# ggsave(here::here("analysis/figures/fig-tcsa-per-sites.png"),
-#        width = 4.3,
-#        height = 3,
-#        units = "in",
-#        bg = "white")
 ```
@@
 
-# ggsave(here::here("analysis/figures/fig-gis-tcsa-sitess.png"),
-#        width = 4,
-#        height = 6,
-#        units = "in",
-#        bg = "white")
-
 ```

### 2023-03-12 analysis/paper/paper.qmd
editing abstract

--- 2023-03-09 analysis/paper/paper.qmd
@@
 abstract: |
- The transition from the Early to Late Paleolithic in Korea is characterized by the introduction of blade technology, stemmed points, end scrapers, burins, denticulates, and finer grained materials.
+ The transition from the Early to Late Paleolithic in Korea is characterized by the introduction of blade technology, stemmed points, end scrapers, burins, denticulates, and higher proportions of finer grained materials.
 Stemmed points have been considered a representative tool that led this set of changes.
-In this study, we examine the possible role that stemmed points played during the technological transition, as well as throughout the Late Paleolithic period.
+In this study, we examine the possible role that stemmed points played during this technological transition, as well as throughout the Late Paleolithic period.
 Our main questions are: What were the best-fit ballistic probabilities for the stemmed points if they were hafted as weapon tips?
@@
 We analyzed TCSA with other variables, including raw materials, weight, radiocarbon dates and locations.
-Our results show that the stemmed points could have served as javelin tips and stabbing spear tips, with smaller numbers as dart tips and un-poisoned arrow tips.
-TCSA values are controlled mostly by size rather than raw material types.
+Our results show that the stemmed points likely served as javelin tips and stabbing spear tips, with smaller numbers as dart tips and un-poisoned arrow tips.
+TCSA values were controlled mostly by size rather than raw material types.
 We found different TCSA ranges of stemmed points at different sites, which could indicate people used stemmed points in different ways depending on the local environment.
@@
 The temporal pattern of TCSA values is one of little change throughout the Late Paleolithic period, but points were predominantly produced before the Last Glacial Maximum (LGM).
-We observed that stemmed points were mostly located in certain ecoregions in Korea but no clear spatial pattern was confirmed.
-We conclude that stemmed points were multi-functional tools, with many likely designed for javelin and stabbing spear tips.
+We observed that stemmed points were mostly located in certain ecoregions in Korea, but no clear spatial pattern was apparent We conclude that stemmed points were multi-functional tools, with many likely designed for use as javelin and stabbing spear tips.
 
@@
 #        bg = "white")
-
 ```

### 2023-03-12 analysis/paper/paper.qmd
editing intro

@@
 
-The introduction of new stone artifact technologies marked a major transition in the Korean Paleolithic from the Early to the Late Paleolithic periods.
+The introduction of new stone artifact technologies marked a major transition in the Korean Paleolithic, from the Early to the Late Paleolithic periods.
 The transition includes blade technology, stemmed points, end scrapers, burins, denticulates, etc.
@@
 
-The purpose of this study is to examine the possible uses of stemmed points in order to understand how stemmed points played a role in the technological transition.
-We use the tip cross-sectional area (TCSA) metric to infer potential variability in weapon-use strategies, and compare the results to other archaeological and ethnographic assemblages based on the idea that different shapes correspond to different weapon types [@lombard2021variation].
-We then explore the relationship between the TCSA range and raw materials, artifact size and discard location, and how these changed over time.
+The purpose of this study is to examine the possible uses of stemmed points in order to understand what role stemmed points may have played in the technological transition from Early to the Late Paleolithic in Korea.
+We use the tip cross-sectional area (TCSA) metric to infer weapon-use strategies based on comparison with other archaeological and ethnographic assemblages, assuming that different ranges of TCSA values correspond to different weapon types [@lombard2021variation].
+We then explore the relationship between the TCSA range and raw materials, artifact size and discard location, and how these changed over time in Korea.
 Our main questions are: What were the best-fit ballistic probabilities for the stemmed points if they were hafted as weapon tips?
@@
 What are the temporal and spatial patterns of stemmed point uses?
-As part of our discussion, we examine possible links between the roles of stemmed points and environmental change or population dynamics during the Late Paleolithic period.
-In order to understand how a certain weapon-tip type was chosen, we apply an evolutionary perspective with the assumption that people chose their weapon tip types as part of their adaptation to the socio-environmental circumstances they found themselves in.
+We examine possible links between the roles of stemmed points and environmental change or population dynamics during the Late Paleolithic period.
+In order to understand how a certain weapon-tip type was chosen, we apply an evolutionary perspective with the assumption that people chose their weapon tip types as part of their adaptation to their socio-environmental circumstances.
 

### 2023-03-25 analysis/paper/paper.qmd
finished editing

@@
 What are the temporal and spatial patterns of stemmed point uses?
-We examine possible links between the roles of stemmed points and environmental change or population dynamics during the Late Paleolithic period.
+We examine possible links between the roles of stemmed points and environmental change, especially the Last Glacial Maximum, or population dynamics during the Late Paleolithic period.
 In order to understand how a certain weapon-tip type was chosen, we apply an evolutionary perspective with the assumption that people chose their weapon tip types as part of their adaptation to their socio-environmental circumstances.
@@
 Slight retouch is typically performed on the proximal end to shape an acute tip and on the distal end to make a stem, which connects to a wooden shaft.
-Elsewhere in the world these types of artifacts are often called "tanged points", but we prefer referring to them as "stemmed points" here to distinguish them from Bronze Age stone projectile points which have long been called "tanged points" in Korea [@park2022did].
+Elsewhere in the world these types of artifacts are often called "tanged points", but we prefer "stemmed points" here to distinguish them from Bronze Age stone projectile points which have long been called "tanged points" in Korea [@park2022did].
 Understanding the appearance of stemmed points is relevant to general questions about the origins of projectile technology, the technological transition of the Late Paleolithic, and relationships between Korea and adjacent regions in East Asia, such as Japan.
 
-Stemmed points are the first composite tools in the Korean Paleolithic that require two different parts to be form one complete tool, a stone point and a shaft, presumably made out of wood [@seong2008tanged].
+Stemmed points are the first composite tools in the Korean Paleolithic.
+They require two different parts to be form one complete tool: a stone point and a shaft, presumably made out of wood [@seong2008tanged].
 Using blades as the blank for the point enables mass production of this composite tool and its shape can become more standardized [@lee2015characteristics; @park2022did].
@@
 Since stemmed points represent the beginning of the Korean Late Paleolithic, investigating their origins is critical to understanding the technological transition from the Early Paleolithic, modern human dispersals into the region, and claims for the existence of the 'Middle' Paleolithic in Korea [@bae2010origin; @bae2017late; @norton2009evolution; @seong2016eastern].
-The debate around explaining the origin of stemmed points which can be summarized into two competing models: (1) a 'heterogenic' migration [@bae2010origin], and (2) an *in situ* evolution model [@seong2009emergence].
-The migration model claims that the new blade industry - including stemmed points - and the earlier coarse flake tradition - including large cores, polyhedrons, choppers, and even handaxes - came from different origins (i.e.
-(1) North route: Siberia, Mongolia, or other regions of northeast China, and (2) South route: southern China) as the result of a continuing influx of modern human migration from two routes.
+The debate around the origin of stemmed points can be summarized into two competing models: (1) a 'heterogenic' migration [@bae2010origin], and (2) an *in situ* evolution model [@seong2009emergence].
+The migration model claims that the new blade industry - including stemmed points - and the earlier coarse flake tradition - including large cores, polyhedrons, choppers, and even handaxes - came from different origins.
+These are (1) a Northern route: Siberia, Mongolia, or other regions of northeast China, and (2) a Southern route: southern China) as the result of a continuing influx of modern human migration from two routes.
+
 The *in situ* model argues that stemmed points and other Late Paleolithic technologies, such as blade industries, autonomously emerged in the south of the Korean peninsula, with no apparent external influence.
@@
 While the *in situ* model claims that stemmed points appeared before blades by acknowledging those early sites, including Bonggok, Songamri, Yonghodong, and Hwadaeri, the migration model supports the traditional definition of the Late or Upper Paleolithic in Europe and Africa that started with the emergence and establishment of blade technology [@seong2016eastern].
+<!-- BM: I think the last two sentences of this paragraph are unclear and confusing.
+I expect others will also struggle to follow the logic here and recommend your expand and clarify in detail -->
 
@@
 For the TCSA, they were only able to use ten stemmed points since they selected stemmed points that retained the widest and thickest parts of the specimens and showed diagnostic impact fractures for the use-wear analysis.
-The purpose of using TCSA in their research was to compare the values to North American dart tips and arrowheads.
+Their purpose in using TCSA was to compare the values to North American dart tips and arrowheads.
 Their results show that the range of TCSA values for Jingeuneul stemmed points is relatively wide, overlapping with both North American dart tips and arrowheads.
@@
 To answer our research questions about the likely uses for stemmed points, we calculated TCSA values for the stemmed points in our sample.
-We also explored the interaction of TCSA values with raw materials and artifact size, which we used weight as a proxy.
-Because the shape of stone artifacts is highly influenced by the raw materials [@mcpherron2014experimental], it is assumed that raw materials could be highly correlated with tool size.
+We also explored the interaction of TCSA values with raw materials and artifact size, using weight as a proxy.
+Because the shape of stone artifacts is highly influenced by the raw materials [@mcpherron2014experimental], we assumed that raw materials may be highly correlated with tool size.
 We investigated radiocarbon ages associated with the points in order to determine temporal patterns in TCSA values.
@@
 
-To aid in interpreting our results, we followed the TCSA ranges for different weapon-delivery systems that @lombard2022tip and @lombard2021variation created by summarizing the analysis by @wadley2008segment of backed microliths (@tbl-tcsa-ranges).
-We excluded `r n_omit` artifacts from our dataset with TCSA greater than 250, which were outside of our range of the comparative data.
+To aid in interpreting our results, we referenced the TCSA ranges for different weapon-delivery systems that @lombard2022tip and @lombard2021variation created by summarizing the analysis by @wadley2008segment of backed microliths (@tbl-tcsa-ranges).
+We excluded `r n_omit` artifacts from our dataset with TCSA greater than 250, which were outside of the range of our comparative data.
 For this comparison with weapon-delivery systems, we included TCSA values from a total of `r min_n_points` stemmed points from `r n_ass` assemblages unearthed from `r n_sites` sites.
@@
 Then, we used longitude and latitude to locate individual assemblages and examined TCSA values for each zone.
-We fully acknowledge that modern eco-regional maps may not fully resemble the Pleistocene landscape in which the points were made and used; however, these are the only maps of eco-regional zones that are currently available.
+We acknowledge that modern eco-regional maps may not fully resemble the Pleistocene landscape in which the points were made and used; however, these are the only maps of eco-regional zones that are currently available.
 
@@
 In America, for example, the advent of bows and arrows provided non-elite hunters with the opportunity to produce their own subsistence or pursue individual wealth without the necessity of hunting in teams [@bettinger2013effects; @angelbeck2014faustian; @rorabaugh2015timing].
-Taking an evolutionary approach, we assume that, given an opportunity to explore alternative technologies, human groups carefully selected a specific stone tool technology based on its advantages over other alternatives, according to their performance in a variety of domains, such as physical and social functions [@lombard2022tip].
+Taking an evolutionary approach, we assume that, given an opportunity to explore alternative technologies, human groups selected a specific stone tool technology based on its advantages over other alternatives, according to their performance in a variety of domains, such as physical and social functions [@lombard2022tip].
 Thus, the selection of weapon-tip types is likely to reflect the socio-environmental circumstances that people encountered and managed.
@@
 Clovis points were used as multifunctional tools that performed a wider range of tasks, including cutting and sawing.
-On the other hand, Folsom points show a narrower range of variation, indicating they were more likely used for specific tasks.
+On the other hand, Folsom points show a narrower range of variation, indicating they were more likely used for a small set of specific tasks.
 
@@
 @fig-tcsa-radio-carbon shows `r n_sites_with_radiocarbon_ages` assemblages with stemmed points in chronological order (panel A).
-Assemblages discarded before LGM tend to possess more stemmed points, such as Suyanggae.
 These assemblages have a wide range of TCSA values, indicating multiple likely uses for stemmed points.
@@
 
-In panel B, @fig-tcsa-radio-carbon shows the distribution of TCSA values from assemblages dating to before, during, and after the LGM in order to explore the impact of climate change on the likely uses of the artifacts.
+To explore the impact of climate change on the likely uses of the artifacts, panel B of @fig-tcsa-radio-carbon shows the distribution of TCSA values from assemblages aggregated into three periods: before, during, and after the LGM.
 Our results show that stemmed points were made predominantly before LGM and only a few after LGM.
@@
 None have TCSA values in the range of poison arrow tips (@fig-tcsa-all-sp).
-In their study, @lee2019were calculated TCSA values for ten stemmed points from Jingeuneul site, which were not included in this research due to data availability issues.
+In their study, @lee2019were calculated TCSA values for ten stemmed points from Jingeuneul site, which were not included in this research due to the unavailability of their data.
 They claimed that these values are within the range of North American dart tips and arrowheads.
@@
 We speculate that this might be due to the higher availability of raw materials, such as shale and hornfels within the landscape.
-Therefore, we could assume that raw materials are also associated with TCSA value.
+Therefore, we find that raw materials, via nodule size, were influential on TCSA values.
 
 Our results show that TCSA values vary between assemblages with few discernible temporal patterns in function (@fig-tcsa-radio-carbon panel A).
-Stemmed points were used most frequently before the LGM (@fig-tcsa-radio-carbon panel B).
+Stemmed points were produced and discarded most frequently before the LGM (@fig-tcsa-radio-carbon panel B).
 While we predicted the greatest variation in TCSA values during the LGM, our results show that variability decreased during the LGM (CV = `r cv_lgm`), in comparison to the previous period (CV = `r cv_bef`).
-This could be explained by the overall decline in stemmed points usage.
-
-We found that stemmed points are primarily located in a small number of eco-regional zones such as the Central Temperature (CT) zone with few clear patterns in TCSA function across the zones (@fig-gis-tcsa-sites).
-As predicted, TCSA values in inland areas, such as vegetation zone CT, show higher variability.
+This could be explained by the overall decline in stemmed points usage, perhaps due to a retreat into refugia during the LGM, rather than increased mobility into unfamiliar areas.
+
+We found that stemmed points are primarily located in a small number of eco-regional zones such as the Central Temperature (CT) Zone with few clear patterns in TCSA function across the zones (@fig-gis-tcsa-sites).
+As predicted, TCSA values in inland areas, such as the CT Zone, show higher variability.
 These results suggest that stemmed points performed a wider range of tasks in low-productivity patches.
-Both temporal and spatial analyses show that stemmed points were more likely to have adapted to particular environments rather than specific climate contexts.
 @prates2022changes claimed that fishtail projectile points in South America were used to hunt megafauna and contributed to their extinction by demonstrating a strong correlation between the spatial and temporal distribution of megafauna and the projectile points.
-We similarly found that stemmed points are more densely distributed in certain environments in South Korea, perhaps also related to faunal distribution.
+We similarly found that stemmed points are more densely distributed in certain environments in South Korea.
+Future work should investigate more specifically patterns in the distribution of stemmed points and ranges of faunal taxa.
 
 Overall, our results show a wide range of TCSA values throughout the Late Paleolithic period and between eco-regional zones.
-We confirmed the widest TCSA range at a single site, Suyanggae (@fig-tcsa-per-sites).
-Our findings are therefore consistent with our second scenario, in which stemmed points are described as multifunctional tools, suggesting that people created stemmed points as a response to unexpected or varying circumstances in their habitats, similar to how @eren2022plains explained Clovis points.
+The widest TCSA range was found at a single site, Suyanggae (@fig-tcsa-per-sites).
+Our findings are therefore consistent with our second scenario, in which stemmed points are best described as multifunctional tools, suggesting that in most cases people created stemmed points as a response to unexpected or varying circumstances in their habitats, similar to how @eren2022plains explained Clovis points.
+The LGM period may be the exception here, with a reduction in variability during this time.
 
@@
 In general, though, we noted a wide range and also differing distributions of TCSA values in each assemblage.
-Therefore, we conclude that stemmed points served diverse roles during the Late Paleolithic.
-People might have frequently encounter unfamiliar situations and use stemmed points as a multifunctional tool to carry out multiple tasks.
+Therefore, we conclude that stemmed points served diverse and not highly patterned roles during the Late Paleolithic.
+Prior to the LGM, people may have frequently encountered unfamiliar situations and use stemmed points as a multifunctional tool to carry out multiple tasks.
 We are aware that discriminating the likely use of small numbers of projectile points could be arbitrary [@erlandson2014darts].
 Since TCSA covers the critical elements of projectiles, flight and penetration dynamics (i.e.
-increase or decrease by shape of tip and cross section), we still consider it a useful metric for discriminating between weapon tip tools [@hughes1998getting; @sitton2020tip; @lombard2021variation].
-As part of our future research, we will conduct use-wear analyses of stemmed points and experimental research in an effort to elucidate the detailed function of stemmed points and technological links between hunter-gatherer groups.
+increase or decrease by shape of tip and cross section), we nevertheless consider it a useful metric for discriminating between weapon tip tools [@hughes1998getting; @sitton2020tip; @lombard2021variation].
+As part of our future research, we plan to conduct use-wear analyses of stemmed points and experiments in an effort to elucidate in more detail the function of stemmed points and technological links between hunter-gatherer groups.
 

### 2023-03-26 analysis/paper/paper.qmd
minor edits
finish editing

@@
 We predict a temporal pattern of more variable TCSA values at the first appearance of stemmed points, suggesting that the tools were being used as part of an adaptation to moving into unfamiliar landscapes, with unpredictable variations in patch productivity and travel times.
-The variability of TCSA values in the LGM is predicted to increase as lower temperatures alter the distribution of resources and reduce the predictability of resource encounters.
+The variability of TCSA values, as measured by coefficients of variation (CV), in the LGM is predicted to increase as lower temperatures alter the distribution of resources and reduce the predictability of resource encounters.
 After the LGM we anticipate a reduction in the range of TCSA values, with higher temperatures and increased bioproductivity.
@@
 C: Tukey test result for vegetation zones.
-D: Eco-region map based on geographical boundaries and ecological conditions (modified the map from @lee2008relationship).
+D: Eco-region map based on geographical boundaries and ecological conditions (modified from @lee2008relationship).
 E: TCSA distribution for ecoregion zones, excluding zones with no artifacts.
@@
 Our results show that TCSA values vary between assemblages with few discernible temporal patterns in function (@fig-tcsa-radio-carbon panel A).
-Stemmed points were produced and discarded most frequently before the LGM (@fig-tcsa-radio-carbon panel B).
-While we predicted the greatest variation in TCSA values during the LGM, our results show that variability decreased during the LGM (CV = `r cv_lgm`), in comparison to the previous period (CV = `r cv_bef`).
+Contrary to our prediction, the earliest stemmed points do not have more variable TSCA values than subsequent ones.
+The most striking temporal pattern is simply that stemmed points were produced and discarded most frequently before the LGM (@fig-tcsa-radio-carbon panel B).
+While we predicted the greatest variation in TCSA values during the LGM, our results show the opposite: variability decreased during the LGM (CV = `r cv_lgm`), in comparison to the previous period (CV = `r cv_bef`).
 This could be explained by the overall decline in stemmed points usage, perhaps due to a retreat into refugia during the LGM, rather than increased mobility into unfamiliar areas.
@@
 The widest TCSA range was found at a single site, Suyanggae (@fig-tcsa-per-sites).
-Our findings are therefore consistent with our second scenario, in which stemmed points are best described as multifunctional tools, suggesting that in most cases people created stemmed points as a response to unexpected or varying circumstances in their habitats, similar to how @eren2022plains explained Clovis points.
+Our findings are therefore consistent with our second scenario, in which stemmed points are best described as multifunctional tools, suggesting that in most cases people created stemmed points as a response to unexpected or varying circumstances in their specific habitats, similar to how @eren2022plains explained Clovis points.
 The LGM period may be the exception here, with a reduction in variability during this time.
@@
 Therefore, we conclude that stemmed points served diverse and not highly patterned roles during the Late Paleolithic.
-Prior to the LGM, people may have frequently encountered unfamiliar situations and use stemmed points as a multifunctional tool to carry out multiple tasks.
+Prior to the LGM, people may have encountered unfamiliar situations and use stemmed points as a multifunctional tool to carry out multiple tasks.
+During the LGM and following we find production and discard of stemmed points declined, but their ballistic properties remained largely unchanged.
+Considering that many of our predictions about TCSA were not supported by the data, we speculate that the ballistic aspects of stemmed points were not highly sensitive to global climate dynamics and local geographic variation, but instead were optimal for the wide range of conditions encountered by people during the Late Palaeolithic in Korea.
+
 We are aware that discriminating the likely use of small numbers of projectile points could be arbitrary [@erlandson2014darts].

### 2023-03-29 analysis/paper/paper.qmd
Minor spelling change

@@
 #| label: fig-tcsa-size
-#| fig-cap: "Artifact size and TSCA values by lithic raw material"
+#| fig-cap: "Artifact size and TCSA values by lithic raw material"
 #| fig-height:  3.5
@@
 
-According to the different weapon-delivery systems that can be inferred from TCSA values, the majority of stemmed points from the Korean Palaeolithic were probably used as javelin tips and stabbing spear tips.
+According to the different weapon-delivery systems that can be inferred from TCSA values, the majority of stemmed points from the Korean Paleolithic were probably used as javelin tips and stabbing spear tips.
 In general, though, we noted a wide range and also differing distributions of TCSA values in each assemblage.

### 2023-04-21 analysis/paper/paper.qmd
add sample size & cv plot

@@
 
+###### regression of CV with sample size ####
+
+# Calculate the CV for each group
+
+cv_tcsa_site_ass_mis <- tcsa_site_ass_mis %>%
+  dplyr::group_by(mis_stage) %>%
+  dplyr::summarize(CV = cv(min_tcsa), count = n()) %>%
+  dplyr::rename(sample_size = count)
+
+cv_tcsa_site_ass_mis$mis_stage <- factor(cv_tcsa_site_ass_mis$mis_stage, levels = c("Before", "LGM", "After"))
+
+ggplot(cv_tcsa_site_ass_mis, aes(x = CV, y = sample_size, color = mis_stage)) +
+  geom_point() +
+  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, color = "red") +
+  labs(x = "CV of Min TCSA", y = "Sample size", title = "CV per each period")
+
+### combine with raw material info ##
+
+tcsa_site_ass_mis_raw <- tcsa_site_ass %>% 
+                          left_join(tcsa_raw, by = "min_tcsa") %>% 
+                      mutate(mis_stage = case_when (
+                        age_ka > 26.5 ~ 'Before' , age_ka > 19 ~ 'LGM', age_ka < 19 ~ 'After'))
+
+cv_tcsa_site_ass_mis_raw <- tcsa_site_ass_mis_raw %>%
+  dplyr::group_by(raw_materials_top) %>%
+  dplyr::summarize(CV = cv(min_tcsa), count = n()) %>%
+  dplyr::rename(sample_size = count)
+
+ggplot(cv_tcsa_site_ass_mis_raw, aes(x = CV, y = sample_size, color = raw_materials_top)) +
+  geom_point() +
+  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, color = "red") +
+  labs(x = "CV of Min TCSA", y = "Sample size", title = "CV per each raw material")
+
+ggplot(cv_tcsa_site_ass_mis_raw, aes(x = CV, y = raw_materials_top)) +
+  geom_point() +
+  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, color = "red") +
+  labs(x = "CV of Min TCSA", y = "Raw Materials", title = "CV per each raw material")
+
 ```

### 2023-04-21 analysis/paper/paper.qmd
CV plots for each variables

--- 2023-03-29 analysis/paper/paper.qmd
@@
 cv_aft <- round(cv(data_aft$min_tcsa),2)
-
 ```
@@
 ```
+```{r}
+
+###### regression of CV with sample size ####
+
+####combine all four variables ###
+
+library(dplyr)
+library(purrr)
+
+# Define the list of data frames to be joined
+dfs <- list(tcsa_site_ass, tcsa_raw, points_in_veg_zones_all, points_in_veg_zones_forest)
+
+# Define the common key column for joining
+join_col <- "min_tcsa"
+
+# Use reduce() to iteratively join the data frames
+tcsa_all <- reduce(dfs, left_join, by = join_col) %>%
+            dplyr::select(age_ka.x, min_tcsa, Assemblage.x, raw_materials_top.x, ecoregion, Zone) %>%             mutate(mis_stage = case_when (age_ka.x > 26.5 ~ 'Before' , age_ka.x > 19 ~ 'LGM', age_ka.x < 19 ~ 'After'))
+
+# Calculate the CV for each group : LGM stage
+
+tcsa_all_stage <- tcsa_all%>%
+  dplyr::group_by(mis_stage) %>%
+  dplyr::summarize(CV = cv(min_tcsa), count = n()) %>%
+  dplyr::rename(sample_size = count)
+
+tcsa_all_stage$mis_stage <- factor(tcsa_all_stage$mis_stage, levels = c("Before", "LGM", "After"))
+
+tcsa_all_stage_plot <- ggplot(tcsa_all_stage, aes(x = CV, y = sample_size, color = mis_stage)) +
+  geom_point() +
+  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, color = "red") +
+  labs(x = "CV of Min TCSA", y = "Sample size", title = "CV per each period")+
+  theme(legend.key.size = unit(0.2, 'cm'), #change legend key size
+        legend.key.height = unit(0.2, 'cm'), #change legend key height
+        legend.key.width = unit(0.2, 'cm'), #change legend key width
+        legend.title = element_text(size=4), #change legend title font size
+        legend.text = element_text(size=2)) #change legend text font size
+
+# Calculate the CV for each group : raw materials
+
+tcsa_all_raw <- tcsa_all %>% 
+    dplyr::group_by(raw_materials_top.x) %>%
+  dplyr::summarize(CV = cv(min_tcsa), count = n()) %>%
+  dplyr::rename(sample_size = count)
+
+tcsa_all_raw_plot <- ggplot(tcsa_all_raw, aes(x = CV, y = sample_size, color = raw_materials_top.x)) +
+  geom_point() +
+  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, color = "red") +
+  labs(x = "CV of Min TCSA", y = "Sample size", title = "CV per each raw material")+
+  theme(legend.key.size = unit(0.2, 'cm'), #change legend key size
+        legend.key.height = unit(0.2, 'cm'), #change legend key height
+        legend.key.width = unit(0.2, 'cm'), #change legend key width
+        legend.title = element_text(size=4), #change legend title font size
+        legend.text = element_text(size=2)) #change legend text font size
+
+# Calculate the CV for each group : assemblages
+
+tcsa_all_ass <- tcsa_all %>% 
+    dplyr::group_by(Assemblage.x) %>%
+  dplyr::summarize(CV = cv(min_tcsa), count = n()) %>%
+  dplyr::rename(sample_size = count)
+
+tcsa_all_ass_plot <- ggplot(tcsa_all_ass, aes(x = CV, y = sample_size, color = Assemblage.x)) +
+  geom_point() +
+  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, color = "red") +
+  labs(x = "CV of Min TCSA", y = "Sample size", title = "CV per each asseblage(Location)") +
+  theme(legend.key.size = unit(0.2, 'cm'), #change legend key size
+        legend.key.height = unit(0.2, 'cm'), #change legend key height
+        legend.key.width = unit(0.2, 'cm'), #change legend key width
+        legend.title = element_text(size=4), #change legend title font size
+        legend.text = element_text(size=2)) #change legend text font size
+
+# Calculate the CV for each group : vegetation zone
+
+tcsa_all_veg <- tcsa_all %>% 
+    dplyr::group_by(ecoregion) %>%
+  dplyr::summarize(CV = cv(min_tcsa), count = n()) %>%
+  dplyr::rename(sample_size = count)
+
+tcsa_all_veg_plot <- ggplot(tcsa_all_veg, aes(x = CV, y = sample_size, color = ecoregion)) +
+  geom_point() +
+  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, color = "red") +
+  labs(x = "CV of Min TCSA", y = "Sample size", title = "CV per each vegetation zone")+
+  theme(legend.key.size = unit(0.2, 'cm'), #change legend key size
+        legend.key.height = unit(0.2, 'cm'), #change legend key height
+        legend.key.width = unit(0.2, 'cm'), #change legend key width
+        legend.title = element_text(size=4), #change legend title font size
+        legend.text = element_text(size=2)) #change legend text font size
+
+# Calculate the CV for each group : geographic zone
+
+tcsa_all_geo <- tcsa_all %>% 
+    dplyr::group_by(Zone) %>%
+  dplyr::summarize(CV = cv(min_tcsa), count = n()) %>%
+  dplyr::rename(sample_size = count)
+
+tcsa_all_geo_plot <- ggplot(tcsa_all_geo, aes(x = CV, y = sample_size, color = Zone)) +
+  geom_point() +
+  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, color = "red") +
+  labs(x = "CV of Min TCSA", y = "Sample size", title = "CV per each geographic zone")+
+  theme(legend.key.size = unit(0.2, 'cm'), #change legend key size
+        legend.key.height = unit(0.2, 'cm'), #change legend key height
+        legend.key.width = unit(0.2, 'cm'), #change legend key width
+        legend.title = element_text(size=4), #change legend title font size
+        legend.text = element_text(size=2)) #change legend text font size
+
+plot_grid(  tcsa_all_stage_plot,
+  tcsa_all_raw_plot,
+tcsa_all_ass_plot,
+ tcsa_all_veg_plot, 
+  tcsa_all_geo_plot,
+ncol =2,
+label_size = 1)
+
+```
 

### 2023-05-01 analysis/paper/paper.qmd
Edited based on Chong&Lomabrd's comments

@@
 
-The introduction of new stone artifact technologies marked a major transition in the Korean Paleolithic, from the Early to the Late Paleolithic periods.
+The introduction of new stone artifact technologies marked a major transition in the Korean Paleolithic, from the Early (approx.350~40 ka) to the Late Paleolithic periods (approx.40~12 ka).
 The transition includes blade technology, stemmed points, end scrapers, burins, denticulates, etc.
@@
 Despite the importance of stemmed points, only a few studies to date have examined their likely uses.
-Previous work has mostly discussed their origin, the chronology of the Korean Late Paleolithic, and their relationship with the Japanese archipelago [@chang2013human; @Park_2013; @lee2019were; @Chong_2021].
-
-The purpose of this study is to examine the possible uses of stemmed points in order to understand what role stemmed points may have played in the technological transition from Early to the Late Paleolithic in Korea.
+Previous work mostly discussed their origin, the chronology of the Korean Late Paleolithic, and their relationship with the Japanese archipelago [@chang2013human; @Park_2013; @lee2019were; @Chong_2021].
+
+The purpose of this study is to examine the possible uses of stemmed points to understand what role they may have played in the technological transition from the Early to the Late Paleolithic in Korea.
 We use the tip cross-sectional area (TCSA) metric to infer weapon-use strategies based on comparison with other archaeological and ethnographic assemblages, assuming that different ranges of TCSA values correspond to different weapon types [@lombard2021variation].
@@
 We examine possible links between the roles of stemmed points and environmental change, especially the Last Glacial Maximum, or population dynamics during the Late Paleolithic period.
-In order to understand how a certain weapon-tip type was chosen, we apply an evolutionary perspective with the assumption that people chose their weapon tip types as part of their adaptation to their socio-environmental circumstances.
+To understand how a certain weapon-tip type was chosen, we apply an evolutionary perspective with the assumption that people chose their weapon tip types as part of their adaptation to their socio-environmental circumstances.
 
@@
 
-Stemmed points (Sumbejjirugae in Korean) are projectile points made on an elongated blade-like flake or blade with two parallel facets and a single or two arises that converge to form an inverted "Y" [@pratt2020circum] (@fig-sp-hand-made).
+Stemmed points (Sumbejjirugae in Korean) are projectile points made on an elongated blade-like flake or blade with two parallel facets and a single or two ridges that converge to form an inverted "Y" [@pratt2020circum] (@fig-sp-hand-made).
 Slight retouch is typically performed on the proximal end to shape an acute tip and on the distal end to make a stem, which connects to a wooden shaft.
-Elsewhere in the world these types of artifacts are often called "tanged points", but we prefer "stemmed points" here to distinguish them from Bronze Age stone projectile points which have long been called "tanged points" in Korea [@park2022did].
-Understanding the appearance of stemmed points is relevant to general questions about the origins of projectile technology, the technological transition of the Late Paleolithic, and relationships between Korea and adjacent regions in East Asia, such as Japan.
+Elsewhere in the world these types of artifacts are often called "tanged points", but we prefer "stemmed points" to distinguish them from Bronze Age stone projectile points known as "tanged points" in Korea [@park2022did].
+Understanding the appearance of stemmed points is relevant to general questions about the direction of projectile technology, the technological transition into the Late Paleolithic, and relationships between Korea and adjacent regions in East Asia, such as Japan.
 
@@
 Since stemmed points represent the beginning of the Korean Late Paleolithic, investigating their origins is critical to understanding the technological transition from the Early Paleolithic, modern human dispersals into the region, and claims for the existence of the 'Middle' Paleolithic in Korea [@bae2010origin; @bae2017late; @norton2009evolution; @seong2016eastern].
-The debate around the origin of stemmed points can be summarized into two competing models: (1) a 'heterogenic' migration [@bae2010origin], and (2) an *in situ* evolution model [@seong2009emergence].
+The debate around the origin of stemmed points can be summarized into two competing models [@bae2013early;@lee2016patterns] : (1) a 'heterogenic' migration [@bae2010origin], and (2) an *in situ* evolution model [@Seong_2006].
 The migration model claims that the new blade industry - including stemmed points - and the earlier coarse flake tradition - including large cores, polyhedrons, choppers, and even handaxes - came from different origins.
 These are (1) a Northern route: Siberia, Mongolia, or other regions of northeast China, and (2) a Southern route: southern China) as the result of a continuing influx of modern human migration from two routes.
-
 The *in situ* model argues that stemmed points and other Late Paleolithic technologies, such as blade industries, autonomously emerged in the south of the Korean peninsula, with no apparent external influence.
-The difference between the two models comes from how they explain a few early sites that contain stemmed points but no blades and blade cores.
-While the *in situ* model claims that stemmed points appeared before blades by acknowledging those early sites, including Bonggok, Songamri, Yonghodong, and Hwadaeri, the migration model supports the traditional definition of the Late or Upper Paleolithic in Europe and Africa that started with the emergence and establishment of blade technology [@seong2016eastern].
-<!-- BM: I think the last two sentences of this paragraph are unclear and confusing.
-I expect others will also struggle to follow the logic here and recommend your expand and clarify in detail -->
+The difference between the two models comes from how they explain a few early sites and stemmed points made out of flakes.
+The *in situ* model argues that stemmed points appeared before blades, and identifies early sites such as Bonggok, Songamri, Yonghodong, and Hwadaeri as evidence that the Korean Late Paleolithic began with the emergence of stemmed points.
+In contrast, the migration model contends that the Late Paleolithic began with the introduction and widespread use of blade technology, similar to the traditional definition of the Late or Upper Paleolithic in Europe and the Later Stone Age in Africa.
+[@seong2016eastern].
+Besides these two main models, there are combination of migration and trade interaction model [@bae2012nature] and complex and non-directional model [@lee2016patterns].
 
@@
 
-Tip cross-sectional area (TCSA) of stone artifacts has been used as a ballistically relevant standard to probabilistically discriminate different likely use classes of projectile armatures, such as spear-thrower (a.k.a.
-atlatl) dart tips, un-poisoned arrow tips and large stabbing/thrusting spears [@sisk2011african; @lombard2021variation; @lombard2020tip; @lombard2021did; @o2018origins; @lombard2022standardized; @metz2023bow; @lombard2023neanderthal].
+The tip cross-sectional area (TCSA) of stone artifacts has been used as a ballistically relevant standard to probabilistically discriminate between likely weapon-tip types, such as spear-thrower (a.k.a.
+atlatl) dart tips, un-poisoned arrow tips and large stabbing/thrusting spears [@sisk2011african; @lombard2021variation; @lombard2020tip; @lombard2021did; @o2018origins; @lombard2022standardized; @metz2023bow; @lombard2023neanderthal; @lombard2022tip].
 It is critical to note that the TCSA metric alone cannot unambiguously determine artifact function; it only suggests a best-fit ballistic probability for the points if they were hafted as weapon tips.
@@
 This method was first proposed by @hughes1998getting and validated by @shea2006origins through the comparison of archaeological examples with ethnographically collected samples of known use.
-@shea2009impact also applied this approach to compare projectile points from Africa, the Levant, and Europe, claiming that projectile weapons first appeared in Africa.
+@shea2009impact also applied the approach to compare projectile points from Africa, the Levant, and Europe, claiming that projectile weapons first appeared in Africa.
 
@@
 Among these stemmed points, we selected those that retained their widest and thickest parts.
-We included stemmed points discovered during field surveys as well as those found at sites that were never dated but were associated with other Late Palaeolithic artifacts.
+We included stemmed points discovered during field surveys as well as those found at sites that were never dated but were associated with other Late Paleolithic artifacts.
 Applying these sampling criteria resulted in a sample of `r total_sp` stemmed points from `r total_ass` assemblages unearthed from `r total_site` sites spanning the period 44-10 ka (@fig-map).
@@
 #| label: tbl-tcsa-ranges
-#| tbl-cap: "TCSA range from Lombard (2021 & 2022)"
+#| tbl-cap: "TCSA ranges from Lombard et al. (2022)"
 
@@
 read_csv(here::here("analysis/data/raw_data/tcsa-range-from-lombard-2022.csv")) %>% 
-  knitr::kable(caption = "TCSA range from Lombard (2021 & 2022)")
+  knitr::kable(caption = "TCSA ranges from Lombard et al. (2022)")
 ```
@@
 
-The vegetation zones are based on @yi2010vegetation's classification of South Korea into three zones: Central Temperate Zone (CT), South Temperate Zone (ST), and Subtropical-warm Temperate Zone (SWT).
-These divisions are based on @yim1975distribution's forest vegetation map, defined by recent temperature and precipitation values.
+The vegetation zones are based on Yi and Kim’s [-@yi2010vegetation] classification of South Korea into three zones: Central Temperate Zone (CT), South Temperate Zone (ST), and Subtropical-warm Temperate Zone (SWT).
+These divisions are based on Yim and Kira's [-@yim1975distribution] forest vegetation map, defined by recent temperature and precipitation values.
 
@@
 
-In order to explore spatial patterns and the relationship between the role of stemmed points and the environment, we located individual assemblages on maps that depict eco-regional zones.
+To explore spatial patterns and the relationship between the role of stemmed points and the environment, we located individual assemblages on maps that depict eco-regional zones.
 Our results show that stemmed points are only located in certain zones, Central Temperature Zone (CT) of the vegetation map or Central inland (CI) (@fig-gis-tcsa-sites panel A & D) of the eco-region map.
@@
 While we predicted the greatest variation in TCSA values during the LGM, our results show the opposite: variability decreased during the LGM (CV = `r cv_lgm`), in comparison to the previous period (CV = `r cv_bef`).
-This could be explained by the overall decline in stemmed points usage, perhaps due to a retreat into refugia during the LGM, rather than increased mobility into unfamiliar areas.
+This could be explained by the overall decline in stemmed point usage, perhaps due to a retreat into refugia during the LGM, rather than increased mobility into unfamiliar areas.
 
@@
 What are the temporal and spatial patterns of stemmed point use?
-We applied the tip cross-sectional area (TCSA) metric because it has been used as a ballistically relevant standard to discriminate different likely use classes of projectile points and it requires only a few measurements on a stone artifact (i.e.
-width and thickness) to compute the metric.
-We calculated TCSA for a total of `r min_n_points` stemmed points from `r n_ass` assemblages unearthed from `r n_sites` sites.
+We applied the tip cross-sectional area (TCSA) metric because it has been used as a ballistically relevant standard to discriminate different likely use classes of projectile points and it requires only a few measurements on a stone artifact (i.e., maximum width and thickness) to compute the metric.
+We calculated TCSA for a total of `r min_n_points` stemmed points from `r n_ass` assemblages excavated from `r n_sites` sites.
 Then we examined the TCSA values with other variables, including raw materials, weight, radiocarbon dates, and site locations.
@@
 During the LGM and following we find production and discard of stemmed points declined, but their ballistic properties remained largely unchanged.
-Considering that many of our predictions about TCSA were not supported by the data, we speculate that the ballistic aspects of stemmed points were not highly sensitive to global climate dynamics and local geographic variation, but instead were optimal for the wide range of conditions encountered by people during the Late Palaeolithic in Korea.
+Considering that many of our predictions about TCSA were not supported by the data, we speculate that composite projectile points made with micro-blades were newly introduced in response to global climate dynamics and stemmed points were still optimal for the wide range of conditions encountered by people during the Late Paleolithic in Korea.
 
@@
 Since TCSA covers the critical elements of projectiles, flight and penetration dynamics (i.e.
-increase or decrease by shape of tip and cross section), we nevertheless consider it a useful metric for discriminating between weapon tip tools [@hughes1998getting; @sitton2020tip; @lombard2021variation].
+increase or decrease by shape of tip and cross section), we nevertheless consider it a useful metric for hypothesizing about different weapon-delivery systems [@hughes1998getting; @sitton2020tip; @lombard2021variation].
 As part of our future research, we plan to conduct use-wear analyses of stemmed points and experiments in an effort to elucidate in more detail the function of stemmed points and technological links between hunter-gatherer groups.

### 2023-05-01 analysis/paper/paper.qmd
Edited based on Lape's comments

@@
 
-Since stemmed points represent the beginning of the Korean Late Paleolithic, investigating their origins is critical to understanding the technological transition from the Early Paleolithic, modern human dispersals into the region, and claims for the existence of the 'Middle' Paleolithic in Korea [@bae2010origin; @bae2017late; @norton2009evolution; @seong2016eastern].
+Since the first appearance of stemmed points defines the beginning of the Korean Late Paleolithic, investigating their origins is critical to understanding the technological transition from the Early Paleolithic, modern human dispersals into the region, and claims for the existence of the 'Middle' Paleolithic in Korea [@bae2010origin; @bae2017late; @norton2009evolution; @seong2016eastern].
 The debate around the origin of stemmed points can be summarized into two competing models [@bae2013early;@lee2016patterns] : (1) a 'heterogenic' migration [@bae2010origin], and (2) an *in situ* evolution model [@Seong_2006].
@@
 We investigated radiocarbon ages associated with the points in order to determine temporal patterns in TCSA values.
-We separated the research time period into three phases based on the major climate event during the Late Pleistocene, the Last Glacial Maximum (LGM), and examined the distribution of TCSA values before, during, and after the LGM.
-We explored the relationship between TCSA values and the location of assemblages by comparing the distribution of TCSA values across ecological and vegetation zones.
+We separated the research time period into three phases, based on the major climate event during the Late Pleistocene, the Last Glacial Maximum (LGM).
+Previous research on the LGM climate in Korea, using age-controlled pollen records [@yi2010vegetation] and computational simulation models [@kim2015last;@park2022did], indicates that the climate was colder and drier than the preceding period.
+Surface temperature cooling ranged from 5 to 6°C, and there was a precipitation decrease of approximately 14%.
+Using this chronology, we examined the distribution of TCSA values before, during, and after the LGM.
+We then explored the relationship between TCSA values and the location of assemblages by comparing the distribution of TCSA values across ecological and vegetation zones.
 
@@
 #| label: fig-tcsa-all-sp
-#| fig-cap: "Distribution of TCSA values for all Korean stemmed points in the current dataset. Shaded boxes indicate TCSA ranges for various weapon types."
+#| fig-cap: "Distribution of TCSA values for all Korean stemmed points in the current dataset. The shaded boxes in color indicate TCSA ranges for different weapon types based on @tbl-tcsa-ranges."
 #| fig-width:   3
@@
             min_tcsa)) +
-  # show regions for tools from Lomabrd 2021 & 2022
+  # show regions for tools from Lomabrd et al. 2022
   # https://www-sciencedirect-com.offcampus.lib.washington.edu/science/article/pii/S0277379121002286
@@
         join = st_contains)
+
 eco_plot <- ggplot() +
   geom_sf(data = veg_all, aes(fill = "grey")) +
-   geom_sf_text(data = veg_all, aes(label = ecoregion), size =1.5)+
-  geom_sf(data = pts_all, aes(color = min_tcsa), size = 1.1) +
+   geom_sf_text(data = veg_all, aes(label = ecoregion), size =1.2)+
+  geom_sf(data = pts_all, aes(color = min_tcsa), size = 0.8) +
   theme_minimal() +
@@
   geom_sf(data = veg_forest, aes(fill = "grey")) +
-  geom_sf_text(data = veg_forest, aes(label = Zone), size =1.5)+
-  geom_sf(data = pts_forest, aes(color = min_tcsa), size = 1.1) +
+  geom_sf_text(data = veg_forest, aes(label = Zone), size =1.2)+
+  geom_sf(data = pts_forest, aes(color = min_tcsa), size = 0.8) +
   theme_minimal()+
@@
 
-By comparing the results with the TCSA range from other archaeological and ethnographic cases (@tbl-tcsa-ranges), we were able to determine what role stemmed points played in the technological transition during the Korean Late Paleolithic.
+By comparing the results with the TCSA ranges from other archaeological and ethnographic cases (@tbl-tcsa-ranges), we were able to examine what role stemmed points played in the technological transition during the Korean Late Paleolithic.
 Our main questions were: What were the best-fit ballistic probabilities for stemmed points if they were hafted as weapon tips?
@@
 During the LGM and following we find production and discard of stemmed points declined, but their ballistic properties remained largely unchanged.
-Considering that many of our predictions about TCSA were not supported by the data, we speculate that composite projectile points made with micro-blades were newly introduced in response to global climate dynamics and stemmed points were still optimal for the wide range of conditions encountered by people during the Late Paleolithic in Korea.
+Considering that many of our predictions about TCSA were not supported by the data, we speculate that composite projectile points made with microblades were newly introduced [@chang2013human], in response to global climate dynamics and stemmed points were still optimal for the wide range of conditions encountered by people during the Late Paleolithic in Korea.
 
@@
 increase or decrease by shape of tip and cross section), we nevertheless consider it a useful metric for hypothesizing about different weapon-delivery systems [@hughes1998getting; @sitton2020tip; @lombard2021variation].
-As part of our future research, we plan to conduct use-wear analyses of stemmed points and experiments in an effort to elucidate in more detail the function of stemmed points and technological links between hunter-gatherer groups.
+As part of our future research, we plan to conduct use-wear analyses and experiments in an effort to elucidate in more detail the function of stemmed points and technological links between hunter-gatherer groups.
 

### 2023-05-06 analysis/paper/paper.qmd
Edited based on Fitzhugh's comments

@@
 Stemmed points (Sumbejjirugae in Korean) are projectile points made on an elongated blade-like flake or blade with two parallel facets and a single or two ridges that converge to form an inverted "Y" [@pratt2020circum] (@fig-sp-hand-made).
-Slight retouch is typically performed on the proximal end to shape an acute tip and on the distal end to make a stem, which connects to a wooden shaft.
+Slight retouch is typically performed on the distal end to shape an acute tip and on the proximal end to make a stem, which connects to a wooden shaft.
 Elsewhere in the world these types of artifacts are often called "tanged points", but we prefer "stemmed points" to distinguish them from Bronze Age stone projectile points known as "tanged points" in Korea [@park2022did].
@@
 These are (1) a Northern route: Siberia, Mongolia, or other regions of northeast China, and (2) a Southern route: southern China) as the result of a continuing influx of modern human migration from two routes.
+
 The *in situ* model argues that stemmed points and other Late Paleolithic technologies, such as blade industries, autonomously emerged in the south of the Korean peninsula, with no apparent external influence.
@@
 The *in situ* model argues that stemmed points appeared before blades, and identifies early sites such as Bonggok, Songamri, Yonghodong, and Hwadaeri as evidence that the Korean Late Paleolithic began with the emergence of stemmed points.
-In contrast, the migration model contends that the Late Paleolithic began with the introduction and widespread use of blade technology, similar to the traditional definition of the Late or Upper Paleolithic in Europe and the Later Stone Age in Africa.
+In contrast, the migration model contends that the Late Paleolithic began with the introduction and widespread use of blade technology similar to the traditional definition of the Late or Upper Paleolithic in Europe and the Later Stone Age in Africa, without counting the stemmed points made out of non-blade materials.
 [@seong2016eastern].
@@
 This method was first proposed by @hughes1998getting and validated by @shea2006origins through the comparison of archaeological examples with ethnographically collected samples of known use.
+Lombard et al.
+demonstrated the efficacy of TCSA by analyzing Middle Stone Age points from Sibudu Cave in KwaZulu-Natal, South Africa, which are dated to approximately 35-50 ka.
+They compared the results obtained from experimental and use-trace studies such as organic residue analysis, macrofracture analysis and use-wear analysis with TCSA values [@lombard2021variation;@lombard2004distribution;@lombard2005evidence).
 @shea2009impact also applied the approach to compare projectile points from Africa, the Levant, and Europe, claiming that projectile weapons first appeared in Africa.
@@
 Then, we used longitude and latitude to locate individual assemblages and examined TCSA values for each zone.
-We acknowledge that modern eco-regional maps may not fully resemble the Pleistocene landscape in which the points were made and used; however, these are the only maps of eco-regional zones that are currently available.
+We are fully acknowledged that modern eco-regional maps do not fully resemble the Pleistocene landscape in which the points were made and used.
+However, in the absence of detailed paleoenvironmental information, we thought it is still meaningful to adapt the modern information of vegetation and geography by assuming that some environmental factors such as  altitude, slope, drainage, hydrology, and bedrock are relatively stable over time.
+And those factors enable us to think that the modern vegetation patterns can be used to make reasonable references about the past.
 
@@
 Our results indicate that javelin tips and stabbing spear tips are the most probable ballistic uses for stemmed points.
-There are a few stemmed points in our sample that may have been used as dart tips and un-poisoned arrow tips.
+There are a few stemmed points in our sample that fall within the ranges of dart tips and un-poisoned arrow tips.
 None have TCSA values in the range of poison arrow tips (@fig-tcsa-all-sp).

### 2023-05-08 analysis/paper/paper.qmd
hided the cv tests for impact of sample size
Edit repetitive word.

@@
 Then, we used longitude and latitude to locate individual assemblages and examined TCSA values for each zone.
-We are fully acknowledged that modern eco-regional maps do not fully resemble the Pleistocene landscape in which the points were made and used.
+We are well aware that modern eco-regional maps do not fully resemble the Pleistocene landscape in which the points were made and used.
 However, in the absence of detailed paleoenvironmental information, we thought it is still meaningful to adapt the modern information of vegetation and geography by assuming that some environmental factors such as  altitude, slope, drainage, hydrology, and bedrock are relatively stable over time.
@@
 
-plot_grid(  tcsa_all_stage_plot,
-  tcsa_all_raw_plot,
-tcsa_all_ass_plot,
- tcsa_all_veg_plot, 
-  tcsa_all_geo_plot,
-ncol =2,
-label_size = 1)
+# plot_grid(  tcsa_all_stage_plot,
+#   tcsa_all_raw_plot,
+# tcsa_all_ass_plot,
+#  tcsa_all_veg_plot, 
+#   tcsa_all_geo_plot,
+# ncol =2,
+# label_size = 1)
 

### 2023-05-11 analysis/paper/paper.qmd
stash

--- 2023-05-01 analysis/paper/paper.qmd
@@
 
-The introduction of new stone artifact technologies marked a major transition in the Korean Paleolithic, from the Early (approx.350~40 ka) to the Late Paleolithic periods (approx.40~12 ka).
+The introduction of new stone artifact technologies marked a major transition in the Korean Paleolithic, from the Early to the Late Paleolithic periods.
 The transition includes blade technology, stemmed points, end scrapers, burins, denticulates, etc.
@@
 Despite the importance of stemmed points, only a few studies to date have examined their likely uses.
-Previous work mostly discussed their origin, the chronology of the Korean Late Paleolithic, and their relationship with the Japanese archipelago [@chang2013human; @Park_2013; @lee2019were; @Chong_2021].
-
-The purpose of this study is to examine the possible uses of stemmed points to understand what role they may have played in the technological transition from the Early to the Late Paleolithic in Korea.
+Previous work has mostly discussed their origin, the chronology of the Korean Late Paleolithic, and their relationship with the Japanese archipelago [@chang2013human; @Park_2013; @lee2019were; @Chong_2021].
+
+The purpose of this study is to examine the possible uses of stemmed points in order to understand what role stemmed points may have played in the technological transition from Early to the Late Paleolithic in Korea.
 We use the tip cross-sectional area (TCSA) metric to infer weapon-use strategies based on comparison with other archaeological and ethnographic assemblages, assuming that different ranges of TCSA values correspond to different weapon types [@lombard2021variation].
@@
 We examine possible links between the roles of stemmed points and environmental change, especially the Last Glacial Maximum, or population dynamics during the Late Paleolithic period.
-To understand how a certain weapon-tip type was chosen, we apply an evolutionary perspective with the assumption that people chose their weapon tip types as part of their adaptation to their socio-environmental circumstances.
+In order to understand how a certain weapon-tip type was chosen, we apply an evolutionary perspective with the assumption that people chose their weapon tip types as part of their adaptation to their socio-environmental circumstances.
 
@@
 
-Stemmed points (Sumbejjirugae in Korean) are projectile points made on an elongated blade-like flake or blade with two parallel facets and a single or two ridges that converge to form an inverted "Y" [@pratt2020circum] (@fig-sp-hand-made).
+Stemmed points (Sumbejjirugae in Korean) are projectile points made on an elongated blade-like flake or blade with two parallel facets and a single or two arises that converge to form an inverted "Y" [@pratt2020circum] (@fig-sp-hand-made).
 Slight retouch is typically performed on the proximal end to shape an acute tip and on the distal end to make a stem, which connects to a wooden shaft.
-Elsewhere in the world these types of artifacts are often called "tanged points", but we prefer "stemmed points" to distinguish them from Bronze Age stone projectile points known as "tanged points" in Korea [@park2022did].
-Understanding the appearance of stemmed points is relevant to general questions about the direction of projectile technology, the technological transition into the Late Paleolithic, and relationships between Korea and adjacent regions in East Asia, such as Japan.
+Elsewhere in the world these types of artifacts are often called "tanged points", but we prefer "stemmed points" here to distinguish them from Bronze Age stone projectile points which have long been called "tanged points" in Korea [@park2022did].
+Understanding the appearance of stemmed points is relevant to general questions about the origins of projectile technology, the technological transition of the Late Paleolithic, and relationships between Korea and adjacent regions in East Asia, such as Japan.
 
@@
 
-Since the first appearance of stemmed points defines the beginning of the Korean Late Paleolithic, investigating their origins is critical to understanding the technological transition from the Early Paleolithic, modern human dispersals into the region, and claims for the existence of the 'Middle' Paleolithic in Korea [@bae2010origin; @bae2017late; @norton2009evolution; @seong2016eastern].
-The debate around the origin of stemmed points can be summarized into two competing models [@bae2013early;@lee2016patterns] : (1) a 'heterogenic' migration [@bae2010origin], and (2) an *in situ* evolution model [@Seong_2006].
+Since stemmed points represent the beginning of the Korean Late Paleolithic, investigating their origins is critical to understanding the technological transition from the Early Paleolithic, modern human dispersals into the region, and claims for the existence of the 'Middle' Paleolithic in Korea [@bae2010origin; @bae2017late; @norton2009evolution; @seong2016eastern].
+The debate around the origin of stemmed points can be summarized into two competing models: (1) a 'heterogenic' migration [@bae2010origin], and (2) an *in situ* evolution model [@seong2009emergence].
 The migration model claims that the new blade industry - including stemmed points - and the earlier coarse flake tradition - including large cores, polyhedrons, choppers, and even handaxes - came from different origins.
 These are (1) a Northern route: Siberia, Mongolia, or other regions of northeast China, and (2) a Southern route: southern China) as the result of a continuing influx of modern human migration from two routes.
+
 The *in situ* model argues that stemmed points and other Late Paleolithic technologies, such as blade industries, autonomously emerged in the south of the Korean peninsula, with no apparent external influence.
-The difference between the two models comes from how they explain a few early sites and stemmed points made out of flakes.
-The *in situ* model argues that stemmed points appeared before blades, and identifies early sites such as Bonggok, Songamri, Yonghodong, and Hwadaeri as evidence that the Korean Late Paleolithic began with the emergence of stemmed points.
-In contrast, the migration model contends that the Late Paleolithic began with the introduction and widespread use of blade technology, similar to the traditional definition of the Late or Upper Paleolithic in Europe and the Later Stone Age in Africa.
-[@seong2016eastern].
-Besides these two main models, there are combination of migration and trade interaction model [@bae2012nature] and complex and non-directional model [@lee2016patterns].
+The difference between the two models comes from how they explain a few early sites that contain stemmed points but no blades and blade cores.
+While the *in situ* model claims that stemmed points appeared before blades by acknowledging those early sites, including Bonggok, Songamri, Yonghodong, and Hwadaeri, the migration model supports the traditional definition of the Late or Upper Paleolithic in Europe and Africa that started with the emergence and establishment of blade technology [@seong2016eastern].
+<!-- BM: I think the last two sentences of this paragraph are unclear and confusing.
+I expect others will also struggle to follow the logic here and recommend your expand and clarify in detail -->
 
@@
 
-The tip cross-sectional area (TCSA) of stone artifacts has been used as a ballistically relevant standard to probabilistically discriminate between likely weapon-tip types, such as spear-thrower (a.k.a.
-atlatl) dart tips, un-poisoned arrow tips and large stabbing/thrusting spears [@sisk2011african; @lombard2021variation; @lombard2020tip; @lombard2021did; @o2018origins; @lombard2022standardized; @metz2023bow; @lombard2023neanderthal; @lombard2022tip].
+Tip cross-sectional area (TCSA) of stone artifacts has been used as a ballistically relevant standard to probabilistically discriminate different likely use classes of projectile armatures, such as spear-thrower (a.k.a.
+atlatl) dart tips, un-poisoned arrow tips and large stabbing/thrusting spears [@sisk2011african; @lombard2021variation; @lombard2020tip; @lombard2021did; @o2018origins; @lombard2022standardized; @metz2023bow; @lombard2023neanderthal].
 It is critical to note that the TCSA metric alone cannot unambiguously determine artifact function; it only suggests a best-fit ballistic probability for the points if they were hafted as weapon tips.
@@
 This method was first proposed by @hughes1998getting and validated by @shea2006origins through the comparison of archaeological examples with ethnographically collected samples of known use.
-@shea2009impact also applied the approach to compare projectile points from Africa, the Levant, and Europe, claiming that projectile weapons first appeared in Africa.
+@shea2009impact also applied this approach to compare projectile points from Africa, the Levant, and Europe, claiming that projectile weapons first appeared in Africa.
 
@@
 Among these stemmed points, we selected those that retained their widest and thickest parts.
-We included stemmed points discovered during field surveys as well as those found at sites that were never dated but were associated with other Late Paleolithic artifacts.
+We included stemmed points discovered during field surveys as well as those found at sites that were never dated but were associated with other Late Palaeolithic artifacts.
 Applying these sampling criteria resulted in a sample of `r total_sp` stemmed points from `r total_ass` assemblages unearthed from `r total_site` sites spanning the period 44-10 ka (@fig-map).
@@
 We investigated radiocarbon ages associated with the points in order to determine temporal patterns in TCSA values.
-We separated the research time period into three phases, based on the major climate event during the Late Pleistocene, the Last Glacial Maximum (LGM).
-Previous research on the LGM climate in Korea, using age-controlled pollen records [@yi2010vegetation] and computational simulation models [@kim2015last;@park2022did], indicates that the climate was colder and drier than the preceding period.
-Surface temperature cooling ranged from 5 to 6°C, and there was a precipitation decrease of approximately 14%.
-Using this chronology, we examined the distribution of TCSA values before, during, and after the LGM.
-We then explored the relationship between TCSA values and the location of assemblages by comparing the distribution of TCSA values across ecological and vegetation zones.
+We separated the research time period into three phases based on the major climate event during the Late Pleistocene, the Last Glacial Maximum (LGM), and examined the distribution of TCSA values before, during, and after the LGM.
+We explored the relationship between TCSA values and the location of assemblages by comparing the distribution of TCSA values across ecological and vegetation zones.
 
@@
 #| label: tbl-tcsa-ranges
-#| tbl-cap: "TCSA ranges from Lombard et al. (2022)"
+#| tbl-cap: "TCSA range from Lombard (2021 & 2022)"
 
@@
 read_csv(here::here("analysis/data/raw_data/tcsa-range-from-lombard-2022.csv")) %>% 
-  knitr::kable(caption = "TCSA ranges from Lombard et al. (2022)")
+  knitr::kable(caption = "TCSA range from Lombard (2021 & 2022)")
 ```
@@
 
-The vegetation zones are based on Yi and Kim’s [-@yi2010vegetation] classification of South Korea into three zones: Central Temperate Zone (CT), South Temperate Zone (ST), and Subtropical-warm Temperate Zone (SWT).
-These divisions are based on Yim and Kira's [-@yim1975distribution] forest vegetation map, defined by recent temperature and precipitation values.
+The vegetation zones are based on @yi2010vegetation's classification of South Korea into three zones: Central Temperate Zone (CT), South Temperate Zone (ST), and Subtropical-warm Temperate Zone (SWT).
+These divisions are based on @yim1975distribution's forest vegetation map, defined by recent temperature and precipitation values.
 
@@
 #| label: fig-tcsa-all-sp
-#| fig-cap: "Distribution of TCSA values for all Korean stemmed points in the current dataset. The shaded boxes in color indicate TCSA ranges for different weapon types based on @tbl-tcsa-ranges."
+#| fig-cap: "Distribution of TCSA values for all Korean stemmed points in the current dataset. Shaded boxes indicate TCSA ranges for various weapon types."
 #| fig-width:   3
@@
             min_tcsa)) +
-  # show regions for tools from Lomabrd et al. 2022
+  # show regions for tools from Lomabrd 2021 & 2022
   # https://www-sciencedirect-com.offcampus.lib.washington.edu/science/article/pii/S0277379121002286
@@
         join = st_contains)
-
 eco_plot <- ggplot() +
   geom_sf(data = veg_all, aes(fill = "grey")) +
-   geom_sf_text(data = veg_all, aes(label = ecoregion), size =1.2)+
-  geom_sf(data = pts_all, aes(color = min_tcsa), size = 0.8) +
+   geom_sf_text(data = veg_all, aes(label = ecoregion), size =1.5)+
+  geom_sf(data = pts_all, aes(color = min_tcsa), size = 1.1) +
   theme_minimal() +
@@
   geom_sf(data = veg_forest, aes(fill = "grey")) +
-  geom_sf_text(data = veg_forest, aes(label = Zone), size =1.2)+
-  geom_sf(data = pts_forest, aes(color = min_tcsa), size = 0.8) +
+  geom_sf_text(data = veg_forest, aes(label = Zone), size =1.5)+
+  geom_sf(data = pts_forest, aes(color = min_tcsa), size = 1.1) +
   theme_minimal()+
@@
 
-To explore spatial patterns and the relationship between the role of stemmed points and the environment, we located individual assemblages on maps that depict eco-regional zones.
+In order to explore spatial patterns and the relationship between the role of stemmed points and the environment, we located individual assemblages on maps that depict eco-regional zones.
 Our results show that stemmed points are only located in certain zones, Central Temperature Zone (CT) of the vegetation map or Central inland (CI) (@fig-gis-tcsa-sites panel A & D) of the eco-region map.
@@
 ```
-```{r}
+
+```{r}
+#| eval: false
 
@@
 # Define the list of data frames to be joined
-dfs <- list(tcsa_site_ass, tcsa_raw, points_in_veg_zones_all, points_in_veg_zones_forest)
+dfs <- list(tcsa_site_ass, 
+            tcsa_raw, 
+            points_in_veg_zones_all, 
+            points_in_veg_zones_forest)
 
@@
 tcsa_all <- reduce(dfs, left_join, by = join_col) %>%
-            dplyr::select(age_ka.x, min_tcsa, Assemblage.x, raw_materials_top.x, ecoregion, Zone) %>%             mutate(mis_stage = case_when (age_ka.x > 26.5 ~ 'Before' , age_ka.x > 19 ~ 'LGM', age_ka.x < 19 ~ 'After'))
+            dplyr::select(age_ka.x, 
+                          min_tcsa, 
+                          Assemblage.x, 
+                          raw_materials_top.x, 
+                          ecoregion, 
+                          Zone) %>%             
+  mutate(mis_stage = case_when (
+    age_ka.x > 26.5 ~ 'Before' , 
+    age_ka.x > 19 ~ 'LGM', 
+    age_ka.x < 19 ~ 'After'))
 
@@
 
-tcsa_all_stage$mis_stage <- factor(tcsa_all_stage$mis_stage, levels = c("Before", "LGM", "After"))
-
-tcsa_all_stage_plot <- ggplot(tcsa_all_stage, aes(x = CV, y = sample_size, color = mis_stage)) +
+tcsa_all_stage$mis_stage <- 
+  factor(tcsa_all_stage$mis_stage, 
+         levels = c("Before", "LGM", "After"))
+
+tcsa_all_stage_plot <- 
+  ggplot(tcsa_all_stage, 
+         aes(x = CV, 
+             y = sample_size, 
+             color = mis_stage)) +
   geom_point() +
@@
 
-tcsa_all_raw_plot <- ggplot(tcsa_all_raw, aes(x = CV, y = sample_size, color = raw_materials_top.x)) +
+tcsa_all_raw_plot <- 
+  ggplot(tcsa_all_raw, 
+         aes(x = CV, 
+             y = sample_size, 
+             color = raw_materials_top.x)) +
   geom_point() +
@@
 
-tcsa_all_ass <- tcsa_all %>% 
+tcsa_all_ass <- 
+  tcsa_all %>% 
   dplyr::group_by(Assemblage.x) %>%
@@
 
-tcsa_all_ass_plot <- ggplot(tcsa_all_ass, aes(x = CV, y = sample_size, color = Assemblage.x)) +
+tcsa_all_ass_plot <- 
+  ggplot(tcsa_all_ass, 
+         aes(x = CV, 
+             y = sample_size, 
+             color = Assemblage.x)) +
   geom_point() +
@@
 
-tcsa_all_veg_plot <- ggplot(tcsa_all_veg, aes(x = CV, y = sample_size, color = ecoregion)) +
+tcsa_all_veg_plot <- 
+  ggplot(tcsa_all_veg, 
+         aes(x = CV, 
+             y = sample_size, 
+             color = ecoregion)) +
   geom_point() +
@@
 
-tcsa_all_geo_plot <- ggplot(tcsa_all_geo, aes(x = CV, y = sample_size, color = Zone)) +
+tcsa_all_geo_plot <- 
+  ggplot(tcsa_all_geo, 
+         aes(x = CV, 
+             y = sample_size, 
+             color = Zone)) +
   geom_point() +
@@
 
-plot_grid(  tcsa_all_stage_plot,
+plot_grid(
+  tcsa_all_stage_plot,
   tcsa_all_raw_plot,
@@
   tcsa_all_geo_plot,
-ncol =2,
-label_size = 1)
+  ncol = 2,
+  label_size = 1
+)
+
+# BM combine data frames so we can put them all on a single plot panel
+
+library(ggrepel)
+bind_rows(tcsa_all_stage %>% rename(var = mis_stage),
+          tcsa_all_raw   %>% rename(var = raw_materials_top.x),
+          tcsa_all_ass   %>% rename(var = Assemblage.x ),
+          tcsa_all_veg   %>% rename(var = ecoregion ),
+          tcsa_all_geo   %>% rename(var = Zone )) %>% 
+  ggplot() + 
+         aes(x = sample_size, 
+             y = CV, 
+             label = var) +
+  geom_point() +
+  geom_text_repel() +
+  geom_smooth(method = "lm",
+              se = FALSE) +
+  labs(y = "CV of Min TCSA", 
+       x = "Sample size") +
+  theme_minimal()
 
@@
 
-By comparing the results with the TCSA ranges from other archaeological and ethnographic cases (@tbl-tcsa-ranges), we were able to examine what role stemmed points played in the technological transition during the Korean Late Paleolithic.
+By comparing the results with the TCSA range from other archaeological and ethnographic cases (@tbl-tcsa-ranges), we were able to determine what role stemmed points played in the technological transition during the Korean Late Paleolithic.
 Our main questions were: What were the best-fit ballistic probabilities for stemmed points if they were hafted as weapon tips?
@@
 While we predicted the greatest variation in TCSA values during the LGM, our results show the opposite: variability decreased during the LGM (CV = `r cv_lgm`), in comparison to the previous period (CV = `r cv_bef`).
-This could be explained by the overall decline in stemmed point usage, perhaps due to a retreat into refugia during the LGM, rather than increased mobility into unfamiliar areas.
+This could be explained by the overall decline in stemmed points usage, perhaps due to a retreat into refugia during the LGM, rather than increased mobility into unfamiliar areas.
 
@@
 What are the temporal and spatial patterns of stemmed point use?
-We applied the tip cross-sectional area (TCSA) metric because it has been used as a ballistically relevant standard to discriminate different likely use classes of projectile points and it requires only a few measurements on a stone artifact (i.e., maximum width and thickness) to compute the metric.
-We calculated TCSA for a total of `r min_n_points` stemmed points from `r n_ass` assemblages excavated from `r n_sites` sites.
+We applied the tip cross-sectional area (TCSA) metric because it has been used as a ballistically relevant standard to discriminate different likely use classes of projectile points and it requires only a few measurements on a stone artifact (i.e.
+width and thickness) to compute the metric.
+We calculated TCSA for a total of `r min_n_points` stemmed points from `r n_ass` assemblages unearthed from `r n_sites` sites.
 Then we examined the TCSA values with other variables, including raw materials, weight, radiocarbon dates, and site locations.
@@
 During the LGM and following we find production and discard of stemmed points declined, but their ballistic properties remained largely unchanged.
-Considering that many of our predictions about TCSA were not supported by the data, we speculate that composite projectile points made with microblades were newly introduced [@chang2013human], in response to global climate dynamics and stemmed points were still optimal for the wide range of conditions encountered by people during the Late Paleolithic in Korea.
+Considering that many of our predictions about TCSA were not supported by the data, we speculate that the ballistic aspects of stemmed points were not highly sensitive to global climate dynamics and local geographic variation, but instead were optimal for the wide range of conditions encountered by people during the Late Palaeolithic in Korea.
 
@@
 Since TCSA covers the critical elements of projectiles, flight and penetration dynamics (i.e.
-increase or decrease by shape of tip and cross section), we nevertheless consider it a useful metric for hypothesizing about different weapon-delivery systems [@hughes1998getting; @sitton2020tip; @lombard2021variation].
-As part of our future research, we plan to conduct use-wear analyses and experiments in an effort to elucidate in more detail the function of stemmed points and technological links between hunter-gatherer groups.
+increase or decrease by shape of tip and cross section), we nevertheless consider it a useful metric for discriminating between weapon tip tools [@hughes1998getting; @sitton2020tip; @lombard2021variation].
+As part of our future research, we plan to conduct use-wear analyses of stemmed points and experiments in an effort to elucidate in more detail the function of stemmed points and technological links between hunter-gatherer groups.
 

### 2023-05-11 analysis/paper/paper.qmd
minor edits throughout

--- 2023-05-06 analysis/paper/paper.qmd
@@
 @park2022did examined mobility and site occupation patterns by applying concepts of human behavioral ecology to lithic assemblages.
-The study found that forager groups using stemmed points may have been associated with the occupation of marginal or extreme environments, in contrast to groups with no stemmed points.
+They found that forager groups using stemmed points may have been associated with the occupation of marginal or extreme environments, in contrast to groups with no stemmed points.
 Also, stemmed points were more frequently associated with expedient technologies, indicating residential and less mobile behaviors.
@@
 
-Though stemmed points have generally been assumed to have been hunting armatures [@lee2019were; @lee2011study; @seong2008tanged; @lee2002new; @chang2013human], it is difficult to determine their likely uses without knowing their complete shape when attached to other components.
+Though stemmed points have generally been assumed to have been hunting armatures [@lee2019were; @lee2011study; @seong2008tanged; @lee2002new; @chang2013human], it has been difficult to determine their likely uses without knowing their complete shape when attached to other components.
 Preserved wooden components of prehistoric projectile tools are too few and rare to standardize their overall scale and variability [@shea2006origins].
@@
 demonstrated the efficacy of TCSA by analyzing Middle Stone Age points from Sibudu Cave in KwaZulu-Natal, South Africa, which are dated to approximately 35-50 ka.
-They compared the results obtained from experimental and use-trace studies such as organic residue analysis, macrofracture analysis and use-wear analysis with TCSA values [@lombard2021variation;@lombard2004distribution;@lombard2005evidence).
+They compared the results obtained from experimental and use-trace studies such as organic residue analysis, macrofracture analysis and use-wear analysis with TCSA values [@lombard2021variation; @lombard2004distribution; @lombard2005evidence).
 @shea2009impact also applied the approach to compare projectile points from Africa, the Levant, and Europe, claiming that projectile weapons first appeared in Africa.
@@
 Then, we used longitude and latitude to locate individual assemblages and examined TCSA values for each zone.
-We are fully acknowledged that modern eco-regional maps do not fully resemble the Pleistocene landscape in which the points were made and used.
-However, in the absence of detailed paleoenvironmental information, we thought it is still meaningful to adapt the modern information of vegetation and geography by assuming that some environmental factors such as  altitude, slope, drainage, hydrology, and bedrock are relatively stable over time.
-And those factors enable us to think that the modern vegetation patterns can be used to make reasonable references about the past.
+We note that modern eco-regional maps may not fully resemble the Pleistocene landscape in which the points were made and used.
+Unfortunately, reconstructions of Pleistocene landscapes are not available for our study area.
+We assume the underlying environmental factors such as altitude, slope, drainage, hydrology, and bedrock are relatively stable over time, and are suitable as an analogy for initial hypothesizing about relationships between Pleistocene geographical variation and stone artefact technology.
 
@@
 
-Inspired by @eren2022plains's approach, we hypothesized some simple scenarios that might explain the temporal and spatial patterns of TCSA range in Korea.
+Inspired by @eren2022plains's approach, we hypothesized simple scenarios that might explain the temporal and spatial patterns of TCSA range in Korea.
 If stemmed points have a narrow range of TCSA values, then people likely produced tools that performed a small set of specific tasks.
@@
 TCSA values for Cluster 3 are the highest, except for one artifact, which is lower than 50.
-There are a few overlaps, but we can still assume that artifacts in different size clusters might have been made for different purposes.
 
@@
 #| label: fig-artifact-size-cluster
-#| fig-cap: "A. Distribution of artifact weight showing three clusters. B. TCSA values for all artifacts. Artifact size classes indicated by the digits representing data point values"
+#| fig-cap: "A.
+Distribution of artifact weight showing three clusters.
+B.
+TCSA values for all artifacts.
+Artifact size classes indicated by the digits representing data point values.
+The shaded boxes in color indicate TCSA ranges for different weapon types based on Table 1"
 #| fig-width:   4
@@
 #| label: fig-tcsa-raw-materials
-#| fig-cap: "TCSA values by lithic raw material."
+#| fig-cap: "TCSA values by lithic raw material. The shaded boxes in color indicate TCSA ranges for different weapon types based on Table 1."
 #| fig-width:   4.4
@@
 The light gray shade indicates Marine Isotope Stages (MIS) 3; the gray shade indicates MIS 2; and the darkest shade indicates the duration of LGM.
-B: Distribution of TCSA values grouped by LGM event."
+B: Distribution of TCSA values grouped by LGM event.
+The shaded boxes in color indicate TCSA ranges for different weapon types based on Table 1."
 #| fig-height:  5
@@
 #| label: fig-tcsa-per-sites
-#| fig-cap: "TCSA values by archaeological site"
+#| fig-cap: "TCSA values by archaeological site. The shaded boxes in color indicate TCSA ranges for different weapon types based on Table 1."
 #| fig-height:  3
@@
 To explore spatial patterns and the relationship between the role of stemmed points and the environment, we located individual assemblages on maps that depict eco-regional zones.
-Our results show that stemmed points are only located in certain zones, Central Temperature Zone (CT) of the vegetation map or Central inland (CI) (@fig-gis-tcsa-sites panel A & D) of the eco-region map.
-We observed that there are statistically significant differences in the distribution of TCSA values between some zones (Vegetation zones: `r eco_lgm_test_pretty`, Ecoregion zones: `r for_lgm_test_pretty`).
+Our results show that stemmed points are only located in certain zones, Central Temperature Zone (CT) of the vegetation map and Central inland (CI) (@fig-gis-tcsa-sites panel A & D) of the eco-region map.
+We observed some statistically significant differences in the distribution of TCSA values between some zones (Vegetation zones: `r eco_lgm_test_pretty`, Ecoregion zones: `r for_lgm_test_pretty`).
 The Tukey's HSD results show that TCSA values from the South Temperature Zone (ST) are significantly different from the other two vegetation zones (@fig-gis-tcsa-sites panel C).
@@
 ```
-```{r}
+
+```{r}
+#| eval: false
 
 ###### regression of CV with sample size ####
+# To address the question that CV is controlled more by sample size than anything else, 
+# this does not produce a figure in the paper currently, just for our satisfaction
 
@@
 
-tcsa_all_stage$mis_stage <- factor(tcsa_all_stage$mis_stage, levels = c("Before", "LGM", "After"))
-
-tcsa_all_stage_plot <- ggplot(tcsa_all_stage, aes(x = CV, y = sample_size, color = mis_stage)) +
-  geom_point() +
-  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, color = "red") +
-  labs(x = "CV of Min TCSA", y = "Sample size", title = "CV per each period")+
-  theme(legend.key.size = unit(0.2, 'cm'), #change legend key size
-        legend.key.height = unit(0.2, 'cm'), #change legend key height
-        legend.key.width = unit(0.2, 'cm'), #change legend key width
-        legend.title = element_text(size=4), #change legend title font size
-        legend.text = element_text(size=2)) #change legend text font size
+tcsa_all_stage$mis_stage <- 
+  factor(tcsa_all_stage$mis_stage, 
+         levels = c("Before", "LGM", "After"))
 
@@
 
-tcsa_all_raw_plot <- ggplot(tcsa_all_raw, aes(x = CV, y = sample_size, color = raw_materials_top.x)) +
-  geom_point() +
-  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, color = "red") +
-  labs(x = "CV of Min TCSA", y = "Sample size", title = "CV per each raw material")+
-  theme(legend.key.size = unit(0.2, 'cm'), #change legend key size
-        legend.key.height = unit(0.2, 'cm'), #change legend key height
-        legend.key.width = unit(0.2, 'cm'), #change legend key width
-        legend.title = element_text(size=4), #change legend title font size
-        legend.text = element_text(size=2)) #change legend text font size
-
 # Calculate the CV for each group : assemblages
 
-tcsa_all_ass <- tcsa_all %>% 
+tcsa_all_ass <- 
+  tcsa_all %>% 
   dplyr::group_by(Assemblage.x) %>%
@@
   dplyr::rename(sample_size = count)
-
-tcsa_all_ass_plot <- ggplot(tcsa_all_ass, aes(x = CV, y = sample_size, color = Assemblage.x)) +
-  geom_point() +
-  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, color = "red") +
-  labs(x = "CV of Min TCSA", y = "Sample size", title = "CV per each asseblage(Location)") +
-  theme(legend.key.size = unit(0.2, 'cm'), #change legend key size
-        legend.key.height = unit(0.2, 'cm'), #change legend key height
-        legend.key.width = unit(0.2, 'cm'), #change legend key width
-        legend.title = element_text(size=4), #change legend title font size
-        legend.text = element_text(size=2)) #change legend text font size
 
@@
 
-tcsa_all_veg_plot <- ggplot(tcsa_all_veg, aes(x = CV, y = sample_size, color = ecoregion)) +
-  geom_point() +
-  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, color = "red") +
-  labs(x = "CV of Min TCSA", y = "Sample size", title = "CV per each vegetation zone")+
-  theme(legend.key.size = unit(0.2, 'cm'), #change legend key size
-        legend.key.height = unit(0.2, 'cm'), #change legend key height
-        legend.key.width = unit(0.2, 'cm'), #change legend key width
-        legend.title = element_text(size=4), #change legend title font size
-        legend.text = element_text(size=2)) #change legend text font size
-
 # Calculate the CV for each group : geographic zone
@@
 
-tcsa_all_geo_plot <- ggplot(tcsa_all_geo, aes(x = CV, y = sample_size, color = Zone)) +
+# BM combine data frames so we can put them all on a single plot panel
+
+library(ggrepel)
+bind_rows(tcsa_all_stage %>% rename(var = mis_stage),
+          tcsa_all_raw   %>% rename(var = raw_materials_top.x),
+          tcsa_all_ass   %>% rename(var = Assemblage.x ),
+          tcsa_all_veg   %>% rename(var = ecoregion ),
+          tcsa_all_geo   %>% rename(var = Zone )) %>% 
+  ggplot() + 
+         aes(x = sample_size, 
+             y = CV, 
+             label = var) +
   geom_point() +
-  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, color = "red") +
-  labs(x = "CV of Min TCSA", y = "Sample size", title = "CV per each geographic zone")+
-  theme(legend.key.size = unit(0.2, 'cm'), #change legend key size
-        legend.key.height = unit(0.2, 'cm'), #change legend key height
-        legend.key.width = unit(0.2, 'cm'), #change legend key width
-        legend.title = element_text(size=4), #change legend title font size
-        legend.text = element_text(size=2)) #change legend text font size
-
-plot_grid(  tcsa_all_stage_plot,
-  tcsa_all_raw_plot,
-tcsa_all_ass_plot,
- tcsa_all_veg_plot, 
-  tcsa_all_geo_plot,
-ncol =2,
-label_size = 1)
+  geom_text_repel() +
+  geom_smooth(method = "lm",
+              se = FALSE) +
+  labs(y = "CV of Min TCSA", 
+       x = "Sample size") +
+  theme_minimal()
 
@@
 They claimed that these values are within the range of North American dart tips and arrowheads.
-We found a wide range of TCSA in our sample covering the range of dart tips and un-poisoned arrow tips, which are assumed to be equivalent to arrowheads.
+We found a wider range of TCSA in our sample covering the range of dart tips and un-poisoned arrow tips, which are assumed to be equivalent to arrowheads.
 In general, the wide range of TCSA suggest that stemmed points might play diverse roles in foraging toolkits.
@@
 TCSA can be impacted by other factors such as raw materials and portability.
-@eren2022plains explain that smaller TCSA could be the result of pursuing production economy and transport efficiency.
+@eren2022plains explain that lower TCSA values could be the result of pursuing production economy and transport efficiency.
 We explored the relationship between TCSA for weight and raw material types in order to examine these factors.
@@
 
-We found that stemmed points are primarily located in a small number of eco-regional zones such as the Central Temperature (CT) Zone with few clear patterns in TCSA function across the zones (@fig-gis-tcsa-sites).
+We found that stemmed points are primarily located in a small number of eco-regional zones such as the Central Temperature (CT) Zone with few clear patterns in TCSA function evident across the zones (@fig-gis-tcsa-sites).
 As predicted, TCSA values in inland areas, such as the CT Zone, show higher variability.
@@
 How diverse were their likely uses?
-What are the temporal and spatial patterns of stemmed point use?
+What were the temporal and spatial patterns of stemmed point use?
 We applied the tip cross-sectional area (TCSA) metric because it has been used as a ballistically relevant standard to discriminate different likely use classes of projectile points and it requires only a few measurements on a stone artifact (i.e., maximum width and thickness) to compute the metric.
@@
 Then we examined the TCSA values with other variables, including raw materials, weight, radiocarbon dates, and site locations.
-Drawing on evolutionary theory, we premised that the way of using stemmed points likely reflects the socio-environmental circumstances that people encountered and managed.
+Drawing on evolutionary theory, we premised that the uses of stemmed points likely reflects the socio-environmental circumstances that people encountered.
 We examined the possible impact of LGM on stemmed points and their distribution across eco-regional zones.
@@
 During the LGM and following we find production and discard of stemmed points declined, but their ballistic properties remained largely unchanged.
-Considering that many of our predictions about TCSA were not supported by the data, we speculate that composite projectile points made with microblades were newly introduced [@chang2013human], in response to global climate dynamics and stemmed points were still optimal for the wide range of conditions encountered by people during the Late Paleolithic in Korea.
+Considering that many of our predictions about TCSA were not supported by the data, we speculate that composite projectile points made with microblades were introduced [@chang2013human] in response to global climate dynamics, and stemmed points were optimal for the wide range of conditions encountered by people during the Late Paleolithic in Korea.
 
@@
 increase or decrease by shape of tip and cross section), we nevertheless consider it a useful metric for hypothesizing about different weapon-delivery systems [@hughes1998getting; @sitton2020tip; @lombard2021variation].
-As part of our future research, we plan to conduct use-wear analyses and experiments in an effort to elucidate in more detail the function of stemmed points and technological links between hunter-gatherer groups.
+Use-wear analyses and experiments should be part of future research into the function of stemmed points to investigate the validity of our current results.
 

### 2023-05-11 analysis/paper/paper.qmd
Minor edit

@@
 The temporal pattern of TCSA values is one of little change throughout the Late Paleolithic period, but points were predominantly produced before the Last Glacial Maximum (LGM).
-We observed that stemmed points were mostly located in certain ecoregions in Korea, but no clear spatial pattern was apparent We conclude that stemmed points were multi-functional tools, with many likely designed for use as javelin and stabbing spear tips.
+We observed that stemmed points were mostly located in certain ecoregions in Korea, but no clear spatial pattern was apparent.
+We conclude that stemmed points were multi-functional tools, with many likely designed for use as javelin and stabbing spear tips.
 
@@
 Stemmed points are the first composite tools in the Korean Paleolithic.
-They require two different parts to be form one complete tool: a stone point and a shaft, presumably made out of wood [@seong2008tanged].
+They require two different parts to form one complete tool: a stone point and a shaft, presumably made out of wood [@seong2008tanged].
 Using blades as the blank for the point enables mass production of this composite tool and its shape can become more standardized [@lee2015characteristics; @park2022did].
@@
 [@seong2016eastern].
-Besides these two main models, there are combination of migration and trade interaction model [@bae2012nature] and complex and non-directional model [@lee2016patterns].
+Besides these two main models, there are combination of migration and trade interaction models [@bae2012nature] and complex and non-directional models [@lee2016patterns].
 
@@
 Unfortunately, reconstructions of Pleistocene landscapes are not available for our study area.
-We assume the underlying environmental factors such as altitude, slope, drainage, hydrology, and bedrock are relatively stable over time, and are suitable as an analogy for initial hypothesizing about relationships between Pleistocene geographical variation and stone artefact technology.
+We assume the underlying environmental factors such as altitude, slope, drainage, hydrology, and bedrock are relatively stable over time, and are suitable as an analogy for initial hypothesizing about relationships between Pleistocene geographical variation and stone artifact technology.
 
@@
 Artifact size classes indicated by the digits representing data point values.
-The shaded boxes in color indicate TCSA ranges for different weapon types based on Table 1"
+The shaded boxes in color indicate TCSA ranges for different weapon types based on Table 1."
 #| fig-width:   4
@@
 #| label: fig-tcsa-size
-#| fig-cap: "Artifact size and TCSA values by lithic raw material"
+#| fig-cap: "Artifact size and TCSA values by lithic raw material."
 #| fig-height:  3.5

### 2023-09-25 analysis/paper/paper.qmd
minor edit for the submission and add submission materials
Created jpg version of figure for submission.

@@
 
-# ggsave(here::here("analysis/figures/fig-map.png"),
+# ggsave(here::here("analysis/figures/fig-map.jpg"),
 #        width = 4,
 #        height = 4.5,
-#        units = "in")
+#        units = "in",
+#        dpi = 400)
 ```
@@
 
-# ggsave(here::here("analysis/figures/fig-tcsa-all-sp.png"),
+# ggsave(here::here("analysis/figures/fig-tcsa-all-sp.jpg"),
 #        width = 4.45,
 #        height = 4.5,
-#        units = "in")
+#        units = "in",
+#        dpi = 400)
 
@@
 
-# ggsave(here::here("analysis/figures/fig-artifact-size-cluster.png"),
+# ggsave(here::here("analysis/figures/fig-artifact-size-cluster.jpg"),
 #        width = 4,
 #        height = 4,
-#        units = "in")
+#        units = "in",
+#        dpi = 400)
 
@@
 
-# ggsave(here::here("analysis/figures/fig-tcsa-raw-materials.png"),
+# ggsave(here::here("analysis/figures/fig-tcsa-raw-materials.jpg"),
 #        width = 4.3,
@@
 #        units = "in",
-#        bg = "white")
+#        bg = "white",
+#        dpi = 400)
 ```
@@
 
-# ggsave(here::here("analysis/figures/fig-tcsa-size.png"),
+# ggsave(here::here("analysis/figures/fig-tcsa-size.jpg"),
 #        width = 4,
 #        height = 3.3,
-#        units = "in")
+#        units = "in",
+#        dpi = 400)
 
@@
 
-# ggsave(here::here("analysis/figures/fig-tcsa-radio-carbon.png"),
+# ggsave(here::here("analysis/figures/fig-tcsa-radio-carbon.jpg"),
 #        width = 4,
@@
 #        units = "in",
-#        bg = "white")
+#        bg = "white",
+#        dpi = 400)
 
@@
 
-# ggsave(here::here("analysis/figures/fig-tcsa-per-sites.png"),
+# ggsave(here::here("analysis/figures/fig-tcsa-per-sites.jpg"),
 #        width = 4.3,
@@
 
-# ggsave(here::here("analysis/figures/fig-gis-tcsa-sitess.png"),
+# ggsave(here::here("analysis/figures/fig-gis-tcsa-sitess.jpg"),
 #        width = 4,
@@
 
-<!-- The following line inserts a page break  -->
+We are grateful for the insightful feedback provided on an early draft of this paper by Ben Fitzhugh and Peter Lape. Their valuable input greatly enhanced the quality of our work.
 

### 2023-09-26 analysis/paper/paper.qmd
changed dpi of the figures.

@@
 #        units = "in",
-#        dpi = 400)
+#        dpi = 300)
 ```
@@
 #        units = "in",
-#        dpi = 400)
+#        dpi = 300)
 
@@
 #        bg = "white",
-#        dpi = 400)
+#        dpi = 300)
 ```
@@
 #        bg = "white",
-#        dpi = 400)
+#        dpi = 300)
 
@@
 
-# ggsave(here::here("analysis/figures/fig-gis-tcsa-sitess.jpg"),
+# ggsave(here::here("analysis/figures/fig-gis-tcsa-sites.jpg"),
 #        width = 4,

### 2023-11-10 analysis/paper/paper.qmd
Lombard requested tcsa data, which I only had in dataframe. I created excel files for tcsa values. One with all, one without outliers.

@@
   mutate(site_name_top = fct_lump_min(Site_name, 5))
+
+# save the tcsa dataframe as xlsx as Lombard requested
+
+# install.packages('writexl')
+# library(writexl)
+# 
+# write_xlsx(tcsa, here::here("analysis/data/derived_data/tcsa_omit_large_tcsa.xlsx"))
+# 
+# # raw tcsa data including outliers.
+# tcsa_all_data <- mydata %>%
+#   mutate(min_tcsa = 0.5 *Width * Thickness) %>%
+#  # mutate(max_tcsa = 0.5 *Length * Thickness) %>%
+#   # dplyr::filter(min_tcsa <= 250) %>% #omit very large TCSA, could be measurement error
+#   # group rare raw materials < 10
+#   mutate(raw_materials_top = fct_lump_min(Raw_material, 10)) %>%
+#   mutate(site_name_top = fct_lump_min(Site_name, 5))
+# 
+# write_xlsx(tcsa_all_data, here::here("analysis/data/derived_data/tcsa_all.xlsx"))
 
